### 2021-04-05 README.Rmd
Setting up rrtools

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# MethodPaper

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/MattiasSealander/Quantifying-Quartzite/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Sjoelander Mattias, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Sjoelander Mattias, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Sjoelander Mattias, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2021-04-05 method_script.R
initialised script for analysis

@@ -1,0 +1,27 @@
+library(tidyverse)
+library(prospectr)
+
+#Read in raw NIR measurements with meatadata
+NIR <- as.data.frame(read.csv2("analysis/data/raw_data//NIR_raw_data.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#Prepare filtering
+artefact <- c("Point", "Point fragment", "Preform")
+material <- c("Quartzite", "Breccie quartz", "Breccie quartzite")
+
+#Filter data to focus on quartzite points, point fragments and preforms
+Points <-
+  NIR %>%
+    filter(material %in% material,
+           type %in% artefact)
+
+#Select visible range of data
+vis_nir <-
+  Points %>%
+    select(unique_id, X400.0:X700.0)
+
+#mean center measurements
+vis_nir_center <-
+  vis_nir %>%
+    mutate(across(c(2:302), scale, scale = FALSE))
+
+

### 2021-06-08 analysis/paper/paper.Rmd
Big commit

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->
```{r load-packages}
library(tidyverse)
library(rrtools)
library(here)
library(infer)
library(prospectr)
library(FactoMineR)
library(factoextra)

#Read in raw NIR measurements with meatadata
NIR <- as.data.frame(read.csv2(here("analysis/data/raw_data//NIR_raw_data.csv"), sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))

#Prepare filtering
artefact <- c("Point", "Point fragment", "Preform")
material <- c("Quartzite", "Breccie quartz", "Breccie quartzite", "Felsitic porphyric quartz", "Quartz")

#Filter data to focus on quartzite points, point fragments and preforms
Points <-
  NIR %>%
    filter(material %in% material,
           type %in% artefact)
```

```{r vis-nir-scree, fig.cap="Contributing variables to PC1"}
#Select data for visible range of spectrum
vis_nir <-
  Points %>%
    select(unique_id, X400.0:X700.0)

#mean center measurements
vis_nir_center <-
  vis_nir %>%
    mutate(across(c(2:302), scale, scale = FALSE))

#centered  PCA
VNIR.pca <- FactoMineR::PCA(vis_nir_center[,c(2:302)], graph = FALSE, scale.unit = FALSE)

#Screeplot of contributing variables to PC1
#dashed reference line corresponds to the expected value if the contribution was uniform.
fviz_contrib(VNIR.pca,
             choice = "var",
             fill = "black",
             color = "black",
             axes = 1,
             top = 60,
             title = "Contribution of variables to PC1") +
        theme_minimal(base_size = 20) +
        theme(axis.title.x=element_blank(), text = element_text(size=10),
                axis.text.x = element_text(angle=45, hjust=1))
```

Figure \@ref(fig:vis-nir-scree) shows that higher absorbance in the red part of the spectrum contributes more to PC1.

```{r vis-nir-pca, fig.cap="PCA of NIR measurements in the 400 - 700 nm part of the spectrum"}
#PCA of centered data
fviz_pca_ind(VNIR.pca, axes = c(1, 2),
                geom = c("point"),
                geom.ind = c("point"), # show points only (but not "text")
                label = "var",
                palette = c("#BC3C29FF", "#0072B5FF", "#EE4C97FF", "#E18727FF"),
                col.ind = Points$hue, #color by sample hue
                fill.ind = Points$hue, #fill point by sample hue
                addEllipses = FALSE,
                repel = TRUE,
                obs.scale = 1, var.scale = 1,
                legend.title = "Hue",
                title = "PCA - Quartzite points") +
      theme_minimal(base_size = 20) +
      scale_shape_manual(values=c(21,22,23,24)) +
      labs(x = "PC1 (98%)",
           y = "PC2 (1.9%)")
```

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2021-06-08 Barplot.R

@@ -1,0 +1,37 @@
+library(tidyverse)
+library(ggsci)
+
+#Read in metadata file for points
+points <- as.data.frame(read.csv2("analysis/data/raw_data//Samples_GIS_file.csv", sep = ";", dec = ",", header = TRUE))
+
+
+#Prepare filters for data
+artefact <- c("Point", "Point fragment", "Preform")
+material <- c("Quartzite", "Breccie quartz", "Breccie quartzite", "Felsitic porphyric quartz", "Quartz")
+
+
+#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
+Points <-
+  points %>%
+  filter(Material %in% material,
+         Type %in% artefact)
+
+# If you need to flip the order (because you've flipped the orientation)
+# call position_stack() explicitly:
+ggplot(Points, aes(y = Material)) +
+  geom_bar(stat = "count", fill = "black") +
+  geom_text(stat='count', aes(label=..count..), vjust = -0.5) +
+  theme_minimal() +
+  ggtitle("Analyzed Points by Material") +
+  xlab("Count") +
+  theme(plot.title = element_text(size = 22,
+                                  hjust = 0.5,
+                                  face = "bold"),
+        axis.title = element_text(size = 14,
+                                  face = "bold"),
+        axis.text.x = element_text(face = "bold"),
+        axis.title.x = element_blank(),
+        legend.title = element_text(size = 12,
+                                    face = "bold"),
+        legend.position = "bottom") +
+  coord_flip()

### 2021-06-08 Screeplot.R

@@ -1,0 +1,61 @@
+library(robCompositions)
+library(tidyverse)
+library(factoextra)
+library(FactoMineR)
+library(missMDA)
+library(VIM)
+library(ggbiplot)
+library(ggsci)
+library(ggplot2)
+library(gridExtra)
+library(scales)
+library(plyr)
+
+
+#Import eigenvalues from XRF Evince
+XEig <-
+  as.data.frame(read.csv2("analysis/data/raw_data//XRF_Eig_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#Import eigenvalues from NIR Evince
+NEig <-
+  as.data.frame(read.csv2("analysis/data/raw_data//XRF_Eig_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#Import eigenvalues from visNIR Evince
+vNEig <-
+  as.data.frame(read.csv2("analysis/data/raw_data//XRF_Eig_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#Import eigenvalues from Raman Evince
+REig <-
+  as.data.frame(read.csv2("analysis/data/raw_data//XRF_Eig_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#XRF Screeplot of eigenvalues
+ggplot(Eig, aes(x=factor(Vector), y=Eigenvalue, group=1)) +
+  geom_line() +
+  geom_point() +
+  scale_y_continuous(limits = c(0,10), breaks = seq(0,10),"Eigenvalue") +
+  theme_minimal() +
+  xlab("Components")
+
+#NIRScreeplot of eigenvalues
+ggplot(Eig, aes(x=factor(Vector), y=Eigenvalue, group=1)) +
+  geom_line() +
+  geom_point() +
+  scale_y_continuous(limits = c(0,10), breaks = seq(0,10),"Eigenvalue") +
+  theme_minimal() +
+  xlab("Components")
+
+#visNIR Screeplot of eigenvalues
+ggplot(Eig, aes(x=factor(Vector), y=Eigenvalue, group=1)) +
+  geom_line() +
+  geom_point() +
+  scale_y_continuous(limits = c(0,10), breaks = seq(0,10),"Eigenvalue") +
+  theme_minimal() +
+  xlab("Components")
+
+#Raman Screeplot of eigenvalues
[…]

### 2021-06-08 method_script.R

@@ -1,38 +1,148 @@
+library(dplyr)
 library(tidyverse)
 library(prospectr)
 library(FactoMineR)
 library(factoextra)
+library(ChemoSpec)
+library(MASS)
+library(reshape2)
+library(cowplot)
+library(ggplot2)
+library(ggsci)
 
 #Read in raw NIR measurements with meatadata
-NIR <- as.data.frame(read.csv2("analysis/data/raw_data//NIR_raw_data.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+NIR <- as.data.frame(read.csv2("analysis/data/derived_data//nir_mean_20210507.csv", sep = ";", dec = ",", header = TRUE, encoding = "UTF-8"))
 
-#Prepare filtering
+#Read in raw NIR measurements for 1000-2500 nm range
+csv <- read.csv2("analysis/data/raw_data//NIR_only_raw_data_1000_2500.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8")
+
+#make first column row names
+csv2<- csv %>%
+  remove_rownames %>%
+  column_to_rownames(var="unique_id") %>%
+  as.data.frame()
+
+#SNV correction for each row(frequency)
+snv <- standardNormalVariate(csv2)
+
+#Write csv file for SNV treated data
+write.csv2(Points, "a.csv", fileEncoding = "UTF-8")
+
+#Prepare filters for data
+#Didnt work as intended, missed some scrapers etc.
 artefact <- c("Point", "Point fragment", "Preform")
 material <- c("Quartzite", "Breccie quartz", "Breccie quartzite", "Felsitic porphyric quartz", "Quartz")
 
-#Filter data to focus on quartzite points, point fragments and preforms
+#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
 Points <-
   NIR %>%
-    filter(material %in% material,
-           type %in% artefact)
+    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+           type == "Point" | type == "Point fragment" | type == "Preform")
+
+
+#If average is supposed to be done on repeated measurements
+#NIR_averaged <- aggregate(NIR[, 26:2176], list(NIR$Sample_ID), mean)
+
+#Make spectra object between 400-2500 nm
+spec_400 <- matrix2SpectraObject(gr.crit = c("Dark", "Light", "Translucent", "White"),
[…]

### 2021-06-08 xrf script.R

@@ -1,0 +1,632 @@
+library(robCompositions)
+library(tidyverse)
+library(factoextra)
+library(FactoMineR)
+library(missMDA)
+library(VIM)
+library(ggbiplot)
+library(ggsci)
+library(ggplot2)
+library(gridExtra)
+library(scales)
+library(plyr)
+
+#Import xrf data
+XRF.csv <-
+read.csv2("analysis/data/raw_data//XRF//XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE)
+
+#Import eigenvalues from Evince
+Eig <-
+    as.data.frame(read.csv2("analysis/data/raw_data//XRF_Eig_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#Import loadings from Evince
+P <-
+  as.data.frame(read.csv2("analysis/data/raw_data//XRF_P_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+
+#Import sample list, set empty fields to NA
+samples <-
+  as.data.frame(read.csv2("analysis/data/raw_data//bifacial_points.csv", sep = ";", dec = ".", header = TRUE, na.strings=c("", "NA")))
+
+#Set reading no as row number
+XRF <-
+  csv2<- XRF.csv %>%
+  remove_rownames %>%
+  column_to_rownames(var="reading_no") %>%
+  as.data.frame()
+
+#Prepare filters for data
+artefact <- c("Point", "Point fragment", "Preform")
+material <- c("Quartzite", "Breccie quartz", "Breccie quartzite", "Felsitic porphyric quartz", "Quartz")
+
+#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
+Points <-
+  XRF.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
+Points2 <-
+  samples %>%
[…]

### 2021-08-19 analysis/paper/paper.Rmd
Paper framework has been sorted out. Added reference management using references.bib.

@@
 ---
-title: "Title Goes Here"
+title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach"
 author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
-      institute: [UofO]
+  - Mattias Sjölander:
+      email: mattias.sjolander@umu.se
+      institute: [UmU]
       correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
-      correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - UmU: Umeå University
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       - --lua-filter=../templates/pagebreak.lua
+mainfont: Times New Roman
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+csl: "../templates/journal-of-archaeological-and-anthropological-science.csl" # Insert path for the bib-style
 abstract: |
@@
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  Spectroscopy; Quartzite; Characterization; Lithics
 highlights: |
@@
 
+```{r load-packages-and-data}
+library(tidyverse)
+library(ggplot2)
+library(gridExtra)
+library(ggsci)
+
+#Import xrf data
+XRF.csv <-
+read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data//XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE)
+
+#Import sample list, set empty fields to NA
+samples <-
+  as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data//bifacial_points.csv", sep = ";", dec = ".", header = TRUE, na.strings=c("", "NA")))
+
+#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
+Points <-
+  XRF.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Import loadings from Evince
+loadings <-
+ as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data/XRF/XRF_P_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+```
 <!-- The actual document text starts here: -->
@@
 
-Here is a citation [@Marwick2017]
+In European archaeology provenience studies and petrographic characterization has for some time favoured flint and obsidian over other geological materials [@prieto_etal20 p.32].
+However, in regions where local flint and obsidian sources are non-existent other materials have taken on the dominant role.
+In Northern Fennoscandia these include quartz, quartzite and slate  [@baudou78; @rankama_etal06; @olausson_etal12], and while both quartz and quartzite technology have been the target of a number of studies over the years there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
 
-# Background
+One reason for the lack of methodological development could be the geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
+The degree to which geological materials have been mixed and transported means that considerable effort is needed in order to reconstruct the events that have formed the modern day landscape.
+Both quartz and quartzite could potentially have been sourced by local prehistoric groups from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations [@rankama_etal06].
+Due to this ambigous nature of availability of raw materials it is difficult to model sourcing strategies for groups that were based within this region.
+[@prieto_etal19 p.15; @prieto_etal20 p.32] notes that ‘quartzite’ represents a somewhat ambiguous term in archaeology, where it is often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
+This issue has been discussed in geology as well [@howard05], and the definition hence varies somewhat between authors.
+Visual characterization of quartzite on a megascopic level is typically unreliable, and microscopic analysis is recommended for identification of subtypes based on metamorphic grade, but Howard  also emphasises the need for field-based methodologies.
+
+In recent years there has been increased development of petrographic characterization of quartzite using destructive methods [e.g.
+@blomme_etal12; @cnudde_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20].
+These have had success in distinguishing between different groups of quartzite types, allowing for more meaningful studies into quartzite-based resource management strategies.
+In archaeology there are times, however, when destructive techniques simply are not applicable.
+Artefact collections (museum, private, etc.) represent a valuable resource, as well as one of the easier ways of gaining access to large datasets with some geographical distribution.
+While destructive methods may prohibit access to such material there are still non-destructive methods that can generate valuable information.
+The use of spectroscopy in petrographic characterization is not new, but it is often applied in conjunction with thin-section analysis [e.g.
+@blomme_etal12; @prieto_etal20].
+A major advantage of spectroscopy, however, is that it is both non-destructive and feature fast measurements, making it an attractive alternative for field-based screening purposes.
+While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of quantifying the elemental content of a material, there are other instrumentations that may provide a deeper insight into the structural properties.
+Near Infrared spectroscopy (NIRS) goes almost as far back as the XRF @hunt77, and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
+The same can be said for Raman spectroscopy, which also has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts (REF).
+
+In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
+They all provide different information on the structure and composition of the material and will be evaluated in regards to the different features they record, the observed material groupings, and potential for analysis into joint relations.
+
+# Material
+```{r select-point-data}
+#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
+Points2 <-
+  samples %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Reorder factor levels
+Points2$type <-
+  factor(Points2$type, levels=c("Preform", "Point fragment", "Point"))
+```
+
+```{r sample-count, results='hide'}
+#Count number of unique rows in sample ID column for the total number of analysed artefacts
+tot_samples <- 
+  samples %>% 
+    filter(type != "Reference")
+
+tot_uniq <- 
+  unique(tot_samples$sample_id)
+
+#Count number of unique rows in sample ID column for the number of points/preforms included in analysis
+sample_uniq <- 
+  unique(Points2$sample_id)
+```
+
+The current study is based on spectroscopic surface measurements on quartzite points found within the County of Västerbotten, Sweden.
+A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten and Skellefteå museums were analysed using three different spectroscopic instrumentations (Appendix XX).
+The colour of the material has a mainly black (N1-N3), grey or white (N9) hue, as well as opaque and almost translucent material with a beige tint.
+A number of these points have, according to prior classifications, been worked using bifacial technology.
+While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12].
+Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in use of quartzite compared to previous periods [@baudou78].
+
+A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g.
+@andersson15].
+This has likely had a contributing factor to the distribution pattern of the artefacts observed in \@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele, Vilhelmina and Tärnaby.
+This can be seen more clearly in the bar graph in \@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no.
+of samples.
+
+Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable.
+Based on a query of all points labelled as ‘quartz’, ‘brecciated quartz’ or ‘quartzite’ a list was generated from their database.
+Sampling of this pool of artefacts was not random, and based on geographical distribution, in an effort to include as many artefacts as possible from areas with few finds available, such as along the Vindel river.
+None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included.
+This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz/quartzite material.
+
+```{r distribution-map, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates.
+Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.", fig.width=15, fig.height=15, fig.fullwidth=TRUE}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/Distribution.jpeg')
+```
+
+```{r distribution-by-parish, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across parishes in Västerbotten, Sweden."), fig.width=12, fig.height=10, fig.fullwidth=TRUE}
+#Stacked bar chart with counts of samples per parish and type
+ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
+  geom_bar() +
+  theme_minimal() +
+  labs(y = "Count",
+       fill = "Type") +
+  theme(
+    legend.position = "bottom",
+    legend.title = element_text(size = 16, face = "bold"),
+    legend.text = element_text(face = "bold"),
+    axis.title.y=element_blank(),
+    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
+    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
+  scale_fill_jco() +
+  coord_flip() +
+  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
+
+```
+
+While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these.
+It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature.
+In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz, and brecciated quartzite ^[Original Swedish terms are ‘brecciekvarts’/’brecciekvartsit’ , the latter possibly a spelling mistake].
+Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
+The reason for including ‘quartz’ in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.
+
+Another reason for keeping a ‘quartz’ category is the group referred to as ‘brecciated quartz’.
+In a petrographic study of the artefacts recovered along the Ume river valley Åhman (1967:8) briefly touches upon the definition of ‘brecciated quartz’, also referring to it as a ‘mylonite quartz’.
+Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure.
+While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz.
+‘Brecciated quartz’ is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
+@holm91].
 
 # Methods
+All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF.
+The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens.
+Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult.
+The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
+This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind.
+The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.
+Pre-processing of the spectra is different for each instrumentation and based on relevant literature.
+After pre-processing the data was modelled using Principal Component Analysis (PCA), which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020).
+PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].
+
+## Near Infrared
+NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
+Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups.
+The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water (REF).
+NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe.
+The ASD features a detection range of 350 – 2500 nm, with spectral sampling (resp.
+spectral resolution) of 1.4 nm (resp.
+3 nm) in the visible and near infrared range and 1.1 nm (resp.
+10 nm) in the short‐wave infrared range.
+Each object was measured against a white reference background.
+Two parts of the NIR spectra was analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region. In both analyses the data was mean-centred before PCA modelling.
+
+## Raman
+Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01].
+However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the ‘Raman scattering’ effect.
+By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
+The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12].
+The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency.
+As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].
+The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
+The spectral coverage of the instrumentation is 175 – 2 500 cm-1 with a resolution of 9.5 cm-1.
+Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
+All Raman spectra were baseline corrected using modified polynomial fitting in R (R Core Team, 2017) using the ChemoSpec package [@hanson16].
+The spectra were then mean-centred and normalized using Standard Normal Variate (Shaver, 2001: 284-285) in the Prediktera Evince software (www.prediktera.com).
+
+## X-Ray Fluourescence
+XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a Thermo ScientificTM portable test stand.
+The reference calibration used for element quantification was ‘mining’.
+Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light).
+Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions.
+For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
+The XRF data was pre-processed using mean-centering, univariate scaling, and log-transformation in the Prediktera Evince software before PCA modelling.
 
@@
 
-<!-- Here's some example analysis code: -->
-```{r load-packages}
-library(tidyverse)
-library(rrtools)
-library(here)
-library(infer)
-library(prospectr)
-library(FactoMineR)
-library(factoextra)
+## VISNIR
 
-#Read in raw NIR measurements with meatadata
-NIR <- as.data.frame(read.csv2(here("analysis/data/raw_data//NIR_raw_data.csv"), sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+The main explanatory power for the visual region can be found in the first two components, featuring eigenvalues above 1 with a cumulative explained variance of 99.9 %.
+Plotting these in the score plot there is a clear trend along the first component with darker materials being grouped together to the right of the plot and light/white material to the left (fig.
+\@ref(fig:vis-nir-pca)).
+Average spectra of these groups  demonstrate a higher absorbance for the dark group, with both of the light groups featuring higher absorbance in the violet region (fig.
+\@ref(fig:vis-nir-spectra)).
+Despite this trend there are exceptions, and while some of these may be material that exist within the muddled greyscale, there are measurements on dark surfaces (N1-N2) that end up closer to the light group (fig.X).
+There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA.
+Artefact 403 and 502 both have a sample taken on one side in the ‘dark’ group and the other in the ‘greyish’ middle.
+This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering.
+The ‘greyish’ middle group is largely made up of ‘quartz coloured’ opaque material with a beige/grey tint (fig.X).
 
-#Prepare filtering
-artefact <- c("Point", "Point fragment", "Preform")
-material <- c("Quartzite", "Breccie quartz", "Breccie quartzite", "Felsitic porphyric quartz", "Quartz")
-
-#Filter data to focus on quartzite points, point fragments and preforms
-Points <-
-  NIR %>%
-    filter(material %in% material,
-           type %in% artefact)
+```{r vis-nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm part of the spectrum.
+Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/VNIR_PCA_p1_p2_center_20210818.jpg')
 ```
 
-```{r vis-nir-scree, fig.cap="Contributing variables to PC1"}
-#Select data for visible range of spectrum
-vis_nir <-
-  Points %>%
-    select(unique_id, X400.0:X700.0)
+```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region"}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/vis_nir_average_spectra_20210702.png')
+```
+## NIR
+No distinct peaks occurred in the 600 – 1 000 nm region, and thus the analysis focused on the 1 000 – 2 500 nm region to lessen the impact of noise.
+The PCA demonstrates a similar pattern to the one observed in the visual region (fig.
+\@ref(fig:nir-pca)).
+Again the first two components have an eigenvalue above 1 with cumulative explained variance of ~99.9 %.
+While the darker materials remain closely grouped in the right side of the score plot, the ‘light’ cluster seems to split into two separate groups.
+The upper left group houses the main bulk of the ‘brecciated quartz’, as well as the almost translucent and opaque material with a beige tint.
+The bottom left group instead features the majority of the white ‘quartz-coloured’ material.
+In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm.
+In some of the material peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO3 [@clark_etal90].
+The average spectra for the three groups shows that the ‘dark’ group has the highest absorbance with less distinct peaks, although they are noticeable.
+The absorbance in the ‘brecciated’/translucent group is slightly higher in the lower band regions than compared to the ‘white’ group, but with decreased absorbance in the higher regions.
+While the 2 450 nm peak can be difficult to discern in most cases, the average spectra for both the ‘brecciated’/translucent group and ‘white’ group feature the other four peaks.
 
-#mean center measurements
-vis_nir_center <-
-  vis_nir %>%
-    mutate(across(c(2:302), scale, scale = FALSE))
-
-#centered  PCA
-VNIR.pca <- FactoMineR::PCA(vis_nir_center[,c(2:302)], graph = FALSE, scale.unit = FALSE)
-
-#Screeplot of contributing variables to PC1
-#dashed reference line corresponds to the expected value if the contribution was uniform.
-fviz_contrib(VNIR.pca,
-             choice = "var",
-             fill = "black",
-             color = "black",
-             axes = 1,
-             top = 60,
-             title = "Contribution of variables to PC1") +
-        theme_minimal(base_size = 20) +
-        theme(axis.title.x=element_blank(), text = element_text(size=10),
-                axis.text.x = element_text(angle=45, hjust=1))
+```{r nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of light absorbance in 1 000 - 2 500 nm region"}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/NIR_PCA_p1_p2_center_20210818.jpg')
 ```
 
-Figure \@ref(fig:vis-nir-scree) shows that higher absorbance in the red part of the spectrum contributes more to PC1.
+## Raman
+There are two distinctive features present in the Raman data.
+The first can be identified as the ‘quartz’ peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset.
+The second feature shows up as two peaks occurring at around 1 300 cm-1 and 1 600 cm-1, and dominate the spectra in terms of intensity for the darker samples.
+These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
+The D band occurs as a result of disordered carbon structures whilst the G band is caused by stretching in the carbon bonds, which is common to graphite-based materials [@vollebregt_etal12; @lee_etal21].
+The PCA model (fig.X) shows a clear trend along the first component (75.9 %) related to the above two features.
+The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side.
+It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks.
+As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
+The 2nd component (7.82 %) mainly seems to differentiate the measurements that feature a lot of noise and objects that were difficult to sample.
+The same is true for the 3rd component, although it does not seem to be as susceptible to noise as the 2nd component.
+The 3rd component also separates out samples with irregular readings that could possibly be related to sampling issues on surfaces that were difficult to sample.
 
-```{r vis-nir-pca, fig.cap="PCA of NIR measurements in the 400 - 700 nm part of the spectrum"}
-#PCA of centered data
-fviz_pca_ind(VNIR.pca, axes = c(1, 2),
-                geom = c("point"),
-                geom.ind = c("point"), # show points only (but not "text")
-                label = "var",
-                palette = c("#BC3C29FF", "#0072B5FF", "#EE4C97FF", "#E18727FF"),
-                col.ind = Points$hue, #color by sample hue
-                fill.ind = Points$hue, #fill point by sample hue
-                addEllipses = FALSE,
-                repel = TRUE,
-                obs.scale = 1, var.scale = 1,
-                legend.title = "Hue",
-                title = "PCA - Quartzite points") +
-      theme_minimal(base_size = 20) +
-      scale_shape_manual(values=c(21,22,23,24)) +
-      labs(x = "PC1 (98%)",
-           y = "PC2 (1.9%)")
+## XRF
+The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal), which represents the remaining unquantifiable elements in a sample.
+The elements represented in this unquantifiable group are all those with an atomic number lower than 12Mg (REF).
+The statistics of the balance for this dataset can be seen in table 1, and the results for the calculated estimates in figure X.
+Univariate analysis shows that SiO2 makes up the main bulk of the composition of the material, and two major groups can be distinguished based on this content \@ref(fig:xrf-violin-plot).
+The majority of the readings resulted in a SiO2 content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO2 content closer to 90 %.
+The “low” SiO2 group also feature a Bal between 40 – 70 %.
+
+```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
+n represents no.
+of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements.
+Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
+
+#Make long version of Points df to prep for violin plots
+Points_long <-
+  Points %>%
+  select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
+  gather(key = "Element", value = "Value")
+
+#Filter long df to focus on Mg, and generate a violin plot
+Mg <-
+  Points_long %>%
+  filter(Element == "X12Mg") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="salmon1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X12Mg))))) +
+  theme_bw() +
+  ylim(0,5) +
+  xlab("Mg") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Al, and generate a violin plot
+Al <-
+  Points_long %>%
+  filter(Element == "X13Al") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="lightgoldenrod1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X13Al))))) +
+  theme_bw() +
+  ylim(0,10) +
+  xlab("Al") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Si, and generate a violin plot
+Si <-
+  Points_long %>%
+    filter(Element == "X14Si") %>%
+      ggplot(aes(x=Element, y=Value)) +
+      geom_violin(fill="lightgoldenrod1") +
+      ggtitle(paste("n = ", length(which(!is.na(Points$X14Si))))) +
+      theme_bw() +
+      ylim(10,100) +
+      xlab("Si") +
+      theme(
+        axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+        plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+        axis.title.y=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank())
+
+#Filter long df to focus on P, and generate a violin plot
+P <-
+  Points_long %>%
+  filter(Element == "X15P") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="lightgoldenrod1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X15P))))) +
+  theme_bw() +
+  ylim(0,2) +
+  xlab("P") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on S, and generate a violin plot
+S <-
+  Points_long %>%
+  filter(Element == "X16S") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="lightgoldenrod1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X16S))))) +
+  theme_bw() +
+  ylim(0,2.5) +
+  xlab("S") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on K, and generate a violin plot
+K <-
+  Points_long %>%
+  filter(Element == "X19K") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="salmon1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X19K))))) +
+  theme_bw() +
+  ylim(0,10) +
+  xlab("K") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Ca, and generate a violin plot
+Ca <-
+  Points_long %>%
+  filter(Element == "X20Ca") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="salmon1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X20Ca))))) +
+  theme_bw() +
+  ylim(0,1) +
+  xlab("Ca") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on V, and generate a violin plot
+V <-
+  Points_long %>%
+  filter(Element == "X23V") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="skyblue1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X23V))))) +
+  theme_bw() +
+  ylim(0,0.05) +
+  xlab("V") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Mn, and generate a violin plot
+Mn <-
+  Points_long %>%
+  filter(Element == "X25Mn") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="skyblue1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X25Mn))))) +
+  theme_bw() +
+  ylim(0,0.75) +
+  xlab("Mn") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Fe, and generate a violin plot
+Fe <-
+  Points_long %>%
+  filter(Element == "X26Fe") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="skyblue1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X26Fe))))) +
+  theme_bw() +
+  ylim(0,2) +
+  xlab("Fe") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Zn, and generate a violin plot
+Zn <-
+  Points_long %>%
+  filter(Element == "X30Zn") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="skyblue1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X30Zn))))) +
+  theme_bw() +
+  ylim(0,0.05) +
+  xlab("Zn") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Sr, and generate a violin plot
+Sr <-
+  Points_long %>%
+  filter(Element == "X38Sr") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="salmon1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X38Sr))))) +
+  theme_bw() +
+  ylim(0,0.02) +
+  xlab("Sr") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Y, and generate a violin plot
+Y <-
+  Points_long %>%
+  filter(Element == "X39Y") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="skyblue1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X39Y))))) +
+  theme_bw() +
+  ylim(0,0.01) +
+  xlab("Y") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Zr, and generate a violin plot
+Zr <-
+  Points_long %>%
+  filter(Element == "X40Zr") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="skyblue1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X40Zr))))) +
+  theme_bw() +
+  ylim(0,0.1) +
+  xlab("Zr") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Ba, and generate a violin plot
+Ba <-
+  Points_long %>%
+  filter(Element == "X56Ba") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="salmon1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X56Ba))))) +
+  theme_bw() +
+  ylim(0,0.5) +
+  xlab("Ba") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#Filter long df to focus on Ti, and generate a violin plot
+Ti <-
+  Points_long %>%
+  filter(Element == "X81Ti") %>%
+  ggplot(aes(x=Element, y=Value)) +
+  geom_violin(fill="lightgoldenrod1") +
+  ggtitle(paste("n = ", length(which(!is.na(Points$X81Ti))))) +
+  theme_bw() +
+  ylim(0,1.5) +
+  xlab("Ti") +
+  theme(
+    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
+    axis.title.y=element_blank(),
+    axis.text.x=element_blank(),
+    axis.ticks.x=element_blank())
+
+#16 figures arranged in 4 rows and 4 columns
+grid.arrange(Mg,Al,Si,P,S,K,Ca,V,Mn,Fe,Zn,Sr,Y,Zr,Ba,Ti,ncol=4)
 ```
 
-```{r demo-inline-code}
-x <- round(pi, 2)
+The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%.
+The loadings of the three components can be seen in figure X.
+In the score plot of P1 and P2, two large clusters emerges (fig.X).
+The loadings for 13Al, 18Sr, 19K, 20Ca, 40Zr, 81Ti group together in the top right quadrant whilst 12Mg, 14Si, 15P, 16S, 23V, and 56Ba group in the bottom right.
+As seen in figure X, 26Fe is the only element positioned on the left side of the model.
+Interestingly, the left cluster contains the samples with lower 14Si values and the right cluster the higher values.
+These clusters still remain if you were to exclude 14Si from the model, however, suggesting there are other elements of relevance to the clustering.
+The “low 14Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top.
+This may indicate a possible relation to the 26Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite.
+13Al follows a similar trend with higher values recorded at the top of the two clusters.
+The ‘brecciated quartz’ samples largely group together at the bottom of the “low 14Si” cluster, with a couple ending up towards the bottom of the “high 14Si” cluster.
+The white (N9) material is somewhat evenly distributed within the leftmost cluster.
+The group of measurements that are somewhat separated from the “low 14Si” cluster at the top are all but one from the rock painting site of ‘Finnforsberget’, which is located next to the Skellefte river.
+The ‘Finnforsberget’ artefacts were sampled twice with object 491 having one measurement closer to the “low 14Si” group and the other the “high 14Si” group with 13Al estimated at 25 %.
+This artefact was sampled twice on each side, resulting in similar results both times.
+One almost translucent sample also ends up among the darker material.
+Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles.
+Both of these factors may have contributed to its placement in the model.
+
+```{r xrf-loadings, fig.width=8, fig.height=8, fig.fullwidth=TRUE, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
+#Prepare plots for 3-plot bargraph figure of loadings
+#PC1 loadings
+P1 <-
+  ggplot(loadings, aes(x = fct_rev(Element), y = P1)) +
+    geom_bar(
+      stat = "identity", position = "identity",
+      color = "black", fill = "darkgrey") +
+  theme_minimal() +
+  theme(plot.margin = margin(10,10,10,10),
+        axis.title.y=element_blank(),
+        axis.text.y = element_text(size=14, face="bold", colour = "black"),
+        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
+  coord_flip()
+#PC2 loadings
+P2 <-
+  ggplot(loadings, aes(x = fct_rev(Element), y = P2)) +
+  geom_bar(
+    stat = "identity", position = "identity",
+    color = "black", fill = "darkgrey") +
+  theme_minimal() +
+  theme(plot.margin = margin(10,10,10,10),
+        axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
+  coord_flip()
+#PC3 loadings
+P3 <-
+  ggplot(loadings, aes(x = fct_rev(Element), y = P3)) +
+  geom_bar(
+    stat = "identity", position = "identity",
+    color = "black", fill = "darkgrey") +
+  theme_minimal() +
+  theme(plot.margin = margin(10,10,10,10),
+        axis.title.y = element_blank(),
+        axis.text.y = element_blank(),
+        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
+  coord_flip()
+
+#Composite figure of above bar graphs
+grid.arrange(P1, P2, P3, ncol=3)
 ```
 
-Here is an example of inline code `r x` in the middle of a sentence.
+```{r xrf-score-plot, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/XRF_PCA_p1_p2.png')
+```
 
 # Discussion
+The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, none of them are capable of providing a detailed fingerprint on their own based solely on surface measurements.
+Taken together they do provide a more in-depth understanding of the characteristic and structure of the material.
+The most easily recognizable megascopic feature of the material is its colour, and the results of all three instrumentations demonstrate that there is some
+
+Interestingly the ‘brecciated quartz’ remains relatively grouped together in all of the models demonstrating some consistency in the original classification.
+There are exceptions, however, and it would be advisable to follow up with a more directed analysis focused around this group of material……
 

### 2021-08-20 analysis/paper/paper.Rmd
Added references, adjusted layout and code.

@@
 Due to this ambigous nature of availability of raw materials it is difficult to model sourcing strategies for groups that were based within this region.
-[@prieto_etal19 p.15; @prieto_etal20 p.32] notes that ‘quartzite’ represents a somewhat ambiguous term in archaeology, where it is often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
+Prieto et al.
+[-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that ‘quartzite’ represents a somewhat ambiguous term in archaeology, where it is often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
 This issue has been discussed in geology as well [@howard05], and the definition hence varies somewhat between authors.
@@
 While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of quantifying the elemental content of a material, there are other instrumentations that may provide a deeper insight into the structural properties.
-Near Infrared spectroscopy (NIRS) goes almost as far back as the XRF @hunt77, and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
-The same can be said for Raman spectroscopy, which also has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts (REF).
+Near Infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
+The same can be said for Raman spectroscopy, which also has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts [@vandenabeele12].
 
 In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
-They all provide different information on the structure and composition of the material and will be evaluated in regards to the different features they record, the observed material groupings, and potential for analysis into joint relations.
+They all provide different information on the structure and composition of the material and will thus be evaluated in regards to the different features they record in the quartz/quartzite.
+This will be achieved via Principal Component Analysis (PCA) where the observed groupings of the material will be related to the spectral features.
+Lastly, the potential for combining the results of the analyses in order to study joint relations will be discussed.
 
@@
 ```{r sample-count, results='hide'}
-#Count number of unique rows in sample ID column for the total number of analysed artefacts
+#Filter the sample data to exclude reference samples
 tot_samples <- 
@@
 
+#Filter sample data to focus on site Lycksele 301
+lycksele <-
+  samples %>% 
+    filter(site_id == "Lycksele 301")
+
+#Then filter to focus on quartz/quartzite material that are points and preforms 
+lycksele_301 <-
+  lycksele %>% 
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | 
+             material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Create vector of unique samples from site Lycksele 301
+lycksele_301 <-
+unique(lycksele_301$sample_id)
+
+#Create a vector containing all unique sample ID, in order to count number of total samples (incl. types other than points/preforms or materials) 
 tot_uniq <- 
@@
 
-#Count number of unique rows in sample ID column for the number of points/preforms included in analysis
+#Create a vector in order count number of unique rows in sample ID column for the number of points/preforms included in analysis
 sample_uniq <- 
@@
 While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12].
-Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in use of quartzite compared to previous periods [@baudou78].
+Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite compared to previous periods [@baudou78].
 
@@
 @andersson15].
-This has likely had a contributing factor to the distribution pattern of the artefacts observed in \@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele, Vilhelmina and Tärnaby.
-This can be seen more clearly in the bar graph in \@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no.
+This has likely had a contributing factor to the distribution pattern of the artefacts observed in fig.
+\@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele, Vilhelmina and Tärnaby.
+This can be seen more clearly in fig.
+\@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no.
 of samples.
@@
 Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable.
-Based on a query of all points labelled as ‘quartz’, ‘brecciated quartz’ or ‘quartzite’ a list was generated from their database.
+Based on a query of all points labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database.
 Sampling of this pool of artefacts was not random, and based on geographical distribution, in an effort to include as many artefacts as possible from areas with few finds available, such as along the Vindel river.
@@
 ```{r distribution-map, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates.
+The distribution of sampled sites is likely affected by the excavation and survey history of the region.
 Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.", fig.width=15, fig.height=15, fig.fullwidth=TRUE}
@@
 
-```{r distribution-by-parish, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across parishes in Västerbotten, Sweden."), fig.width=12, fig.height=10, fig.fullwidth=TRUE}
+```{r distribution-by-parish, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden.
+Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301"), fig.width=12, fig.height=10, fig.fullwidth=TRUE}
 #Stacked bar chart with counts of samples per parish and type
@@
 It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature.
-In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz, and brecciated quartzite ^[Original Swedish terms are ‘brecciekvarts’/’brecciekvartsit’ , the latter possibly a spelling mistake].
+In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz, and brecciated quartzite ^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake].
 Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
@@
 
-Another reason for keeping a ‘quartz’ category is the group referred to as ‘brecciated quartz’.
-In a petrographic study of the artefacts recovered along the Ume river valley Åhman (1967:8) briefly touches upon the definition of ‘brecciated quartz’, also referring to it as a ‘mylonite quartz’.
+Another reason for keeping a *quartz* category is the group referred to as *brecciated quartz*.
+In a petrographic study of the artefacts recovered along the Ume river valley Åhman (1967:8) briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*.
 Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure.
 While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz.
-‘Brecciated quartz’ is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
+*Brecciated quartz* is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
 @holm91].
@@
 Pre-processing of the spectra is different for each instrumentation and based on relevant literature.
-After pre-processing the data was modelled using Principal Component Analysis (PCA), which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020).
+After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020).
 PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].
@@
 Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01].
-However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the ‘Raman scattering’ effect.
+However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect.
 By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
@@
 The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
-The spectral coverage of the instrumentation is 175 – 2 500 cm-1 with a resolution of 9.5 cm-1.
+The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1.
 Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
 All Raman spectra were baseline corrected using modified polynomial fitting in R (R Core Team, 2017) using the ChemoSpec package [@hanson16].
-The spectra were then mean-centred and normalized using Standard Normal Variate (Shaver, 2001: 284-285) in the Prediktera Evince software (www.prediktera.com).
+The spectra were then mean-centred and normalized using Standard Normal Variate (Shaver, 2001: 284-285) in the Prediktera Evince software.
 
@@
 XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a Thermo ScientificTM portable test stand.
-The reference calibration used for element quantification was ‘mining’.
+The reference calibration used for element quantification was *mining*.
 Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light).
@@
 
-## VISNIR
+<br>
+
+## NIR - visual range
 
@@
 
+<br>
+
 ```{r vis-nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm part of the spectrum.
@@
 
+<br>
+
 ```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region"}
@@
 ```
-## NIR
+
+<br>
+
+## NIR - near infrared range
 No distinct peaks occurred in the 600 – 1 000 nm region, and thus the analysis focused on the 1 000 – 2 500 nm region to lessen the impact of noise.
@@
 Again the first two components have an eigenvalue above 1 with cumulative explained variance of ~99.9 %.
-While the darker materials remain closely grouped in the right side of the score plot, the ‘light’ cluster seems to split into two separate groups.
-The upper left group houses the main bulk of the ‘brecciated quartz’, as well as the almost translucent and opaque material with a beige tint.
-The bottom left group instead features the majority of the white ‘quartz-coloured’ material.
+While the darker materials remain closely grouped in the right side of the score plot, the *light* cluster seems to split into two separate groups.
+The upper left group houses the main bulk of the *brecciated quartz*, as well as the almost translucent and opaque material with a beige tint.
+The bottom left group instead features the majority of the white *quartz-coloured* material.
 In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm.
-In some of the material peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO3 [@clark_etal90].
-The average spectra for the three groups shows that the ‘dark’ group has the highest absorbance with less distinct peaks, although they are noticeable.
-The absorbance in the ‘brecciated’/translucent group is slightly higher in the lower band regions than compared to the ‘white’ group, but with decreased absorbance in the higher regions.
-While the 2 450 nm peak can be difficult to discern in most cases, the average spectra for both the ‘brecciated’/translucent group and ‘white’ group feature the other four peaks.
+In some of the measurements peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO3 [@clark_etal90].
+The average spectra for the three groups show that the *dark* group has the highest absorbance with less distinct peaks, although they are noticeable.
+The absorbance in the *brecciated*/*translucent* group is slightly higher in the lower band regions than compared to the *white* group, but with decreased absorbance in the higher regions.
+While a fifth peak at 2 450 nm can be difficult to discern in most cases, the average spectra for both the *brecciated*/*translucent* group and *white* group feature the above mentioned four peaks.
+
+<br>
 
@@
 
+<br>
+
 ## Raman
 There are two distinctive features present in the Raman data.
-The first can be identified as the ‘quartz’ peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset.
+The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset.
 The second feature shows up as two peaks occurring at around 1 300 cm-1 and 1 600 cm-1, and dominate the spectra in terms of intensity for the darker samples.
@@
 The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal), which represents the remaining unquantifiable elements in a sample.
-The elements represented in this unquantifiable group are all those with an atomic number lower than 12Mg (REF).
-The statistics of the balance for this dataset can be seen in table 1, and the results for the calculated estimates in figure X.
-Univariate analysis shows that SiO2 makes up the main bulk of the composition of the material, and two major groups can be distinguished based on this content \@ref(fig:xrf-violin-plot).
-The majority of the readings resulted in a SiO2 content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO2 content closer to 90 %.
-The “low” SiO2 group also feature a Bal between 40 – 70 %.
+The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF).
+Univariate analysis shows that SiO^2^ makes up the main bulk of the composition of the material, and two major groups can be distinguished based on this content (fig.
+\@ref(fig:xrf-violin-plot)).
+The majority of the readings resulted in a SiO^2^ content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO^2^ content closer to 90 %.
+The “low” SiO^2^ group also feature a Bal between 40 – 70 %.
+
+The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%.
+The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings).
+In the score plot of P1 and P2, two large clusters emerges (fig.
+\@ref(fig:xrf-score-plot-p1-p2)).
+The loadings for ^13^Al, ^18^Sr, ^19^K, ^20^Ca, ^40^Zr, ^81^Ti group together in the top right quadrant whilst ^12^Mg, ^14^Si, ^15^P, ^16^S, ^23^V, and ^56^Ba group in the bottom right.
+As seen in figure \@ref(fig:xrf-score-plot-p1-p2), ^26^Fe is the only element positioned on the left side of the model.
+Interestingly, the left cluster contains the samples with lower ^14^Si values and the right cluster the higher values.
+These clusters still remain if you were to exclude ^14^Si from the model, however, suggesting there are other elements of relevance to the clustering.
+The “low ^14^Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top.
+This may indicate a possible relation to the ^26^Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite.
+^13^Al follows a similar trend with higher values recorded at the top of the two clusters.
+The *brecciated quartz* samples largely group together at the bottom of the “low ^14^Si” cluster, with a couple ending up towards the bottom of the “high ^14^Si” cluster.
+The white (N9) material is somewhat evenly distributed within the leftmost cluster.
+The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of ‘Finnforsberget’, which is located next to the Skellefte river.
+The ‘Finnforsberget’ artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %.
+This artefact was sampled twice on each side, resulting in similar results both times.
+One almost translucent sample also ends up among the darker material.
+Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles.
+Both of these factors may have contributed to its placement in the model.
+
+<br>
 
@@
 
-The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%.
-The loadings of the three components can be seen in figure X.
-In the score plot of P1 and P2, two large clusters emerges (fig.X).
-The loadings for 13Al, 18Sr, 19K, 20Ca, 40Zr, 81Ti group together in the top right quadrant whilst 12Mg, 14Si, 15P, 16S, 23V, and 56Ba group in the bottom right.
-As seen in figure X, 26Fe is the only element positioned on the left side of the model.
-Interestingly, the left cluster contains the samples with lower 14Si values and the right cluster the higher values.
-These clusters still remain if you were to exclude 14Si from the model, however, suggesting there are other elements of relevance to the clustering.
-The “low 14Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top.
-This may indicate a possible relation to the 26Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite.
-13Al follows a similar trend with higher values recorded at the top of the two clusters.
-The ‘brecciated quartz’ samples largely group together at the bottom of the “low 14Si” cluster, with a couple ending up towards the bottom of the “high 14Si” cluster.
-The white (N9) material is somewhat evenly distributed within the leftmost cluster.
-The group of measurements that are somewhat separated from the “low 14Si” cluster at the top are all but one from the rock painting site of ‘Finnforsberget’, which is located next to the Skellefte river.
-The ‘Finnforsberget’ artefacts were sampled twice with object 491 having one measurement closer to the “low 14Si” group and the other the “high 14Si” group with 13Al estimated at 25 %.
-This artefact was sampled twice on each side, resulting in similar results both times.
-One almost translucent sample also ends up among the darker material.
-Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles.
-Both of these factors may have contributed to its placement in the model.
+<br>
 
@@
 
-```{r xrf-score-plot, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
+<br>
+
+```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
 knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/XRF_PCA_p1_p2.png')

### 2021-08-23 analysis/paper/paper.Rmd
Added references and smaller adjustments in text

@@
 Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
-All Raman spectra were baseline corrected using modified polynomial fitting in R (R Core Team, 2017) using the ChemoSpec package [@hanson16].
-The spectra were then mean-centred and normalized using Standard Normal Variate (Shaver, 2001: 284-285) in the Prediktera Evince software.
+All Raman spectra were baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16].
+The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] in the Prediktera Evince software.
 
@@
 In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm.
-In some of the measurements peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO3 [@clark_etal90].
+In some of the measurements peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO^3^ [@clark_etal90].
 The average spectra for the three groups show that the *dark* group has the highest absorbance with less distinct peaks, although they are noticeable.
@@
 The 3rd component also separates out samples with irregular readings that could possibly be related to sampling issues on surfaces that were difficult to sample.
+
+<br>
+
+```{r raman-average, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* around 1 300 cm-1 (D band) and 1 600 cm-1 (G band)"}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/raman_average_spectra_dark_20210823.png')
+```
+
+<br>
 
@@
 ```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
-n represents no.
+**n** represents no.
 of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements.

### 2021-08-30 analysis/paper/paper.Rmd
Updated PCA figures, raman section, and references

@@
 
+\linebreak
+
 # Material
@@
 
+\linebreak
+ \linebreak
+
 The current study is based on spectroscopic surface measurements on quartzite points found within the County of Västerbotten, Sweden.
@@
 
+\linebreak
+
 ```{r distribution-map, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates.
@@
 ```
+
+\linebreak
 
@@
 
+\linebreak
+
 # Methods
+
+\linebreak
+ \linebreak
+
 All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF.
@@
 PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].
+
+\linebreak
 
@@
 Two parts of the NIR spectra was analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region. In both analyses the data was mean-centred before PCA modelling.
+
+\linebreak
 
@@
 
+\linebreak
+
 ## X-Ray Fluourescence
@@
 
+\linebreak
+
 # Results
 
-<br>
+\linebreak
+ \linebreak
 
@@
 
-<br>
+\linebreak
 
-```{r vis-nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm part of the spectrum.
+```{r vis-nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum.
 Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/VNIR_PCA_p1_p2_center_20210818.jpg')
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_1_visnir_averaged_pca_p1_p2.jpg')
 ```
 
-<br>
+\linebreak
 
-```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region"}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/vis_nir_average_spectra_20210702.png')
+```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region.
+Dark material has a higher overall absorbance, with material closer to a white hue feature lower absorbance.
+The *translucent*/*opaque* group also feature higher absorbance in the ultraviolet region."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_average_visnir_spectra.jpg')
 ```
 
-<br>
+\linebreak
 
@@
 
-<br>
+\linebreak
 
-```{r nir-pca, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of light absorbance in 1 000 - 2 500 nm region"}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/NIR_PCA_p1_p2_center_20210818.jpg')
+```{r nir-spectra-comparison, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="Comparison of average spectra for dark, white, and translucent materials.
+Brecciated quartz is included as it is typically translucent with dark particle inclusion, as a comparison with the translucent group that mainly is made up of seemingly pure quarts."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/nir_spectra_comparison.jpg')
 ```
 
-<br>
+\linebreak
+
+```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum.
+Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
+A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
+```
+
+\linebreak
 
@@
 The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset.
-The second feature shows up as two peaks occurring at around 1 300 cm-1 and 1 600 cm-1, and dominate the spectra in terms of intensity for the darker samples.
+The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples.
 These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
-The D band occurs as a result of disordered carbon structures whilst the G band is caused by stretching in the carbon bonds, which is common to graphite-based materials [@vollebregt_etal12; @lee_etal21].
-The PCA model (fig.X) shows a clear trend along the first component (75.9 %) related to the above two features.
+The D band occurs as a result of disordered carbon structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
+Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1.
+It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280].
+This could thus explain the discrepancy between the observed peak in the dataset and the literature.
+
+A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca.
+695 cm-1 and 795 cm-1.
+Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on shale from XXXX.
+There is a fairly large band around 500 cm-1 which can be indicative of the presence of amorphous carbon (aC) [@hodkiewicz10], and it is likely the features observed above are another indication of carbon present within the samples.
+No other features were pr
+
+The resulting PCA model (fig.
+\@ref(fig:nir-pca)) shows a clear trend along the first component (80.3 %) related to the above two features.
 The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side.
@@
 As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-The 2nd component (7.82 %) mainly seems to differentiate the measurements that feature a lot of noise and objects that were difficult to sample.
-The same is true for the 3rd component, although it does not seem to be as susceptible to noise as the 2nd component.
-The 3rd component also separates out samples with irregular readings that could possibly be related to sampling issues on surfaces that were difficult to sample.
+The 2nd component (6.82 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample.
+In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.
 
-<br>
+\linebreak
 
-```{r raman-average, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* around 1 300 cm-1 (D band) and 1 600 cm-1 (G band)"}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/raman_average_spectra_dark_20210823.png')
+```{r raman-average, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
 ```
 
-<br>
+\linebreak
+
+```{r raman-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components of the raman analysis.
+The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1.
+Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
+```
+
+\linebreak
 
@@
 The white (N9) material is somewhat evenly distributed within the leftmost cluster.
-The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of ‘Finnforsberget’, which is located next to the Skellefte river.
-The ‘Finnforsberget’ artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %.
+The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river.
+The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %.
 This artefact was sampled twice on each side, resulting in similar results both times.
@@
 
-<br>
+\linebreak
 
@@
 
-<br>
+\linebreak
 
 ```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/XRF_PCA_p1_p2.png')
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
 ```
 
+\linebreak
+
 # Discussion
+
+\linebreak
+ \linebreak
+
 The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, none of them are capable of providing a detailed fingerprint on their own based solely on surface measurements.
 Taken together they do provide a more in-depth understanding of the characteristic and structure of the material.
-The most easily recognizable megascopic feature of the material is its colour, and the results of all three instrumentations demonstrate that there is some
+The most easily recognizable megascopic feature of the material is its colour, and the results of all three instrumentations demonstrate that there is some truth to this.
+However, a large group of the material, which features different hues of grey, does not conform as easily.
+In each of the PCA models there is a *greyish* middle that makes classification difficult.
 
@@
 
+\linebreak
+
 # Conclusion
 
+\linebreak
+ \linebreak
+
 # Acknowledgements
+
+\linebreak
+ \linebreak
 

### 2021-08-31 analysis/paper/paper.Rmd
Rewrote the code for violin plot to use the more efficient facet wrap function

@@
 
-```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
+```{r xrf-violin-plot, fig.width=15, fig.height=10, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
 **n** represents no.
@@
 Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
-
-#Make long version of Points df to prep for violin plots
+#Prepare data in long format for violin plots
 Points_long <-
@@
 
-#Filter long df to focus on Mg, and generate a violin plot
-Mg <-
+#Remove initial X introduced by read.csv to column names
+Points_long$Element <- substring(Points_long$Element, 2)
+
+# Add column with elemental groups for visualisation
+Points_long <-
   Points_long %>%
-  filter(Element == "X12Mg") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="salmon1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X12Mg))))) +
+    mutate(Group = case_when(
+      endsWith(Element, "Mg") ~ "Alkaline Earth",
+      endsWith(Element, "Al") ~ "Basic metal",
+      endsWith(Element, "Si") ~ "Semimetal",
+      endsWith(Element, "P") ~ "Nonmetal",
+      endsWith(Element, "S") ~ "Nonmetal",
+      endsWith(Element, "K") ~ "Alkali Metal",
+      endsWith(Element, "Ca") ~ "Alkaline Earth",
+      endsWith(Element, "V") ~ "Transition Metal",
+      endsWith(Element, "Mn") ~ "Transition Metal",
+      endsWith(Element, "Fe") ~ "Transition Metal",
+      endsWith(Element, "Zn") ~ "Transition Metal",
+      endsWith(Element, "Sr") ~ "Alkaline Earth",
+      endsWith(Element, "Y") ~ "Transition Metal",
+      endsWith(Element, "Zr") ~ "Transition Metal",
+      endsWith(Element, "Ba") ~ "Alkaline Earth",
+      endsWith(Element, "Ti") ~ "Transition Metal",
+    ))
+
+#Prepare colors
+colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")
+
+#Lookup for element labeller
+Elements <- c(
+  '12Mg' = "Mg",
+  '13Al' = "Al",
+  '14Si' = "Si",
+  '15P' = "P",
+  '16S'= "S",
+  '19K' = "K",
+  '20Ca' ="Ca",
+  '23V' = "V",
+  '25Mn' = "Mn",
+  '26Fe' = "Fe",
+  '30Zn' = "Zn",
+  '38Sr' = "Sr",
+  '39Y' = "Y",
+  '40Zr' = "Zr",
+  '56Ba' = "Ba",
+  '81Ti' = "Ti"
+  )
+
+#Violin plot of elemental content, trimmed, with fill based on elemental groups
+Points_long %>%
+  drop_na(Value) %>%
+    ggplot(aes(x=Element, y=Value, fill = Group)) +
+    geom_violin(trim = TRUE) +
+    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
+    scale_fill_manual(values=colors) +
     theme_bw() +
-  ylim(0,5) +
-  xlab("Mg") +
     theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
+      legend.title = element_text(size = 12, face = "bold", colour = "black"),
+      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
+      axis.title.x = element_blank(),
       plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
@@
       axis.ticks.x=element_blank())
-
-#Filter long df to focus on Al, and generate a violin plot
-Al <-
-  Points_long %>%
-  filter(Element == "X13Al") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="lightgoldenrod1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X13Al))))) +
-  theme_bw() +
-  ylim(0,10) +
-  xlab("Al") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Si, and generate a violin plot
-Si <-
-  Points_long %>%
-    filter(Element == "X14Si") %>%
-      ggplot(aes(x=Element, y=Value)) +
-      geom_violin(fill="lightgoldenrod1") +
-      ggtitle(paste("n = ", length(which(!is.na(Points$X14Si))))) +
-      theme_bw() +
-      ylim(10,100) +
-      xlab("Si") +
-      theme(
-        axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-        plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-        axis.title.y=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank())
-
-#Filter long df to focus on P, and generate a violin plot
-P <-
-  Points_long %>%
-  filter(Element == "X15P") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="lightgoldenrod1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X15P))))) +
-  theme_bw() +
-  ylim(0,2) +
-  xlab("P") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on S, and generate a violin plot
-S <-
-  Points_long %>%
-  filter(Element == "X16S") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="lightgoldenrod1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X16S))))) +
-  theme_bw() +
-  ylim(0,2.5) +
-  xlab("S") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on K, and generate a violin plot
-K <-
-  Points_long %>%
-  filter(Element == "X19K") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="salmon1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X19K))))) +
-  theme_bw() +
-  ylim(0,10) +
-  xlab("K") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Ca, and generate a violin plot
-Ca <-
-  Points_long %>%
-  filter(Element == "X20Ca") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="salmon1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X20Ca))))) +
-  theme_bw() +
-  ylim(0,1) +
-  xlab("Ca") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on V, and generate a violin plot
-V <-
-  Points_long %>%
-  filter(Element == "X23V") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="skyblue1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X23V))))) +
-  theme_bw() +
-  ylim(0,0.05) +
-  xlab("V") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Mn, and generate a violin plot
-Mn <-
-  Points_long %>%
-  filter(Element == "X25Mn") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="skyblue1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X25Mn))))) +
-  theme_bw() +
-  ylim(0,0.75) +
-  xlab("Mn") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Fe, and generate a violin plot
-Fe <-
-  Points_long %>%
-  filter(Element == "X26Fe") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="skyblue1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X26Fe))))) +
-  theme_bw() +
-  ylim(0,2) +
-  xlab("Fe") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Zn, and generate a violin plot
-Zn <-
-  Points_long %>%
-  filter(Element == "X30Zn") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="skyblue1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X30Zn))))) +
-  theme_bw() +
-  ylim(0,0.05) +
-  xlab("Zn") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Sr, and generate a violin plot
-Sr <-
-  Points_long %>%
-  filter(Element == "X38Sr") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="salmon1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X38Sr))))) +
-  theme_bw() +
-  ylim(0,0.02) +
-  xlab("Sr") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Y, and generate a violin plot
-Y <-
-  Points_long %>%
-  filter(Element == "X39Y") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="skyblue1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X39Y))))) +
-  theme_bw() +
-  ylim(0,0.01) +
-  xlab("Y") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Zr, and generate a violin plot
-Zr <-
-  Points_long %>%
-  filter(Element == "X40Zr") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="skyblue1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X40Zr))))) +
-  theme_bw() +
-  ylim(0,0.1) +
-  xlab("Zr") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Ba, and generate a violin plot
-Ba <-
-  Points_long %>%
-  filter(Element == "X56Ba") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="salmon1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X56Ba))))) +
-  theme_bw() +
-  ylim(0,0.5) +
-  xlab("Ba") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#Filter long df to focus on Ti, and generate a violin plot
-Ti <-
-  Points_long %>%
-  filter(Element == "X81Ti") %>%
-  ggplot(aes(x=Element, y=Value)) +
-  geom_violin(fill="lightgoldenrod1") +
-  ggtitle(paste("n = ", length(which(!is.na(Points$X81Ti))))) +
-  theme_bw() +
-  ylim(0,1.5) +
-  xlab("Ti") +
-  theme(
-    axis.title.x = element_text(size = 10, face = "bold", colour = "black"),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
-
-#16 figures arranged in 4 rows and 4 columns
-grid.arrange(Mg,Al,Si,P,S,K,Ca,V,Mn,Fe,Zn,Sr,Y,Zr,Ba,Ti,ncol=4)
 ```

### 2021-08-31 analysis/paper/paper.Rmd
Updated raman text and pca figures

@@
 The resulting PCA model (fig.
-\@ref(fig:nir-pca)) shows a clear trend along the first component (80.3 %) related to the above two features.
+\@ref(fig:nir-pca1)) shows a clear trend along the first component (81.5 %) related to the above two features.
 The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side.
@@
 As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-The 2nd component (6.82 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample.
+The 2nd component (6.29 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample.
 In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.
@@
 
-```{r raman-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components of the raman analysis.
+```{r raman-pca1, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components of the raman analysis.
 The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1.
 Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_6_raman_average_pca_p1_p2.jpg')
+```
+
+\linebreak
+
+```{r raman-pca2, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 3rd components of the raman analysis.
+The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1.
+Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_7_raman_average_pca_p1_p3.jpg')
 ```

### 2021-09-01 analysis/paper/paper.Rmd
Updated figures and explanation for raman results

@@
 The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.
+
 Pre-processing of the spectra is different for each instrumentation and based on relevant literature.
@@
 Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups.
-The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water (REF).
+The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water [@hunt77 p.508].
+
 NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe.
@@
 10 nm) in the short‐wave infrared range.
-Each object was measured against a white reference background.
-Two parts of the NIR spectra was analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region. In both analyses the data was mean-centred before PCA modelling.
+Each object was measured against a white reference background and the collected measurements were averaged for each artefact.
+Two parts of the NIR spectra were analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region.
+In both analyses the data was mean-centred before PCA modelling.
 
@@
 As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].
+
 The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
@@
 Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
-All Raman spectra were baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16].
-The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] in the Prediktera Evince software.
+
+All spectra were averaged for each artefact and the resulting spectra were then baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16].
+The spectra were mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA in the Prediktera Evince software.
 
@@
 
-```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark and light groups in the 400 - 750 nm region.
-Dark material has a higher overall absorbance, with material closer to a white hue feature lower absorbance.
-The *translucent*/*opaque* group also feature higher absorbance in the ultraviolet region."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_average_visnir_spectra.jpg')
+```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region.
+Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance.
+Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
 ```
@@
 While a fifth peak at 2 450 nm can be difficult to discern in most cases, the average spectra for both the *brecciated*/*translucent* group and *white* group feature the above mentioned four peaks.
-
-\linebreak
-
-```{r nir-spectra-comparison, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="Comparison of average spectra for dark, white, and translucent materials.
-Brecciated quartz is included as it is typically translucent with dark particle inclusion, as a comparison with the translucent group that mainly is made up of seemingly pure quarts."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/nir_spectra_comparison.jpg')
-```
 
@@
 695 cm-1 and 795 cm-1.
-Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on shale from XXXX.
-There is a fairly large band around 500 cm-1 which can be indicative of the presence of amorphous carbon (aC) [@hodkiewicz10], and it is likely the features observed above are another indication of carbon present within the samples.
-No other features were pr
+Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on different types of shale, as well as samples of diabase, apatite, orthoclase, and blue quartz.
+Studies of clay minerals [@kloprogge17]
+have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group.
+Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
 

### 2021-09-06 analysis/paper/paper.Rmd
Corrected reference key and added code for performing pca plots in R

@@
 library(ggsci)
+library(prospectr)
+library(patchwork)
+library(factoextra)
+library(ggbiplot)
 
@@
 
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//NIR_master_averaged_20210827.csv", sep = ";",
+            dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+
 #Import sample list, set empty fields to NA
@@
 
-#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
-Points <-
+#Filter xrf, nir and raman data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
   XRF.csv %>%
@@
 
+Points.nir <-
+  nir.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+Points.raman <-
+  raman.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Fill the NA fields in the munsell_hue column and mark them as translucent samples
+Points.nir[8][is.na(Points.nir[8])] <- "Translucent"
+Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
+Points.xrf[8][is.na(Points.xrf[8])] <- "Translucent"
+
+#prepare color/fill and symbols for score plots
+colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
+shapes <- c(24,25,23,22,21)
+
 #Import loadings from Evince
 loadings <-
- as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data/XRF/XRF_P_values.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+ as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data/XRF/XRF_P_values_long.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
 ```
@@
 Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301"), fig.width=12, fig.height=10, fig.fullwidth=TRUE}
+
 #Stacked bar chart with counts of samples per parish and type
@@
 
-```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum.
+<!--```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the #spectrum.
+Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
+A trend #of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
+#knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
+``` -->
+```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum.
 Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
 A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
+
+#perform PCA with SNV normalization and mean-center
+nir.pca <-
+  prcomp(Points.nir[,c(678:2178)], center = TRUE, scale = FALSE)
+
+#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
+basic_plot1 <-
+  fviz_pca_ind(nir.pca, axes = c(1,2), label="none")
+
+#prepare labels for PCs
+pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
+pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
+pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
+pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+
+#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
+ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
+             aes(x=x, y=y, 
+                 shape = material, 
+                 fill = munsell_hue)) +
+  geom_point(size=3) +
+  theme_bw() +
+  ggtitle("Near infrared 1 000 - 2 500 nm") +
+  labs(x = pc1lab,
+       y = pc2lab) +
+  scale_shape_manual(name = "Material",
+                     values=shapes) +
+  scale_fill_manual(name = "Munsell hue",
+                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
+                    values=colors) +
+  guides(shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol=2, 
+                              title.position="top", 
+                              title.hjust = 0.5),
+         fill = guide_legend(override.aes = list(shape = 22,
+                                                 fill = colors,
+                                                 color = "black",
+                                                 size=3),
+                             title.position="top", title.hjust = 0.5))+
+  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
+        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+        legend.title = element_text(size = 12, face = "bold", colour = "black"),
+        legend.background = element_rect(linetype = "solid", color = "black"),
+        legend.position = "bottom")
 ```
@@
 
-```{r raman-average, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
+<!--```{r raman-average, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
 knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
+```-->
+```{r raman-average, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
+
+#perform PCA with SNV normalization and mean-center
+raman.pca <-
+  prcomp(standardNormalVariate(X = Points.raman[,c(29:320)]), center = TRUE, scale = FALSE)
+
+#prepare different basic score plots with fviz_pca_ind
+basic_plot1 <-
+  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")
+
+basic_plot2 <-
+  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")
+
+#prepare labels for PCs
+pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
+pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
+pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
+pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
+pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+pc3lab <- paste0("PC3 (",as.character(pc3var),"%)")
+
+#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
+p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
+       aes(x=x, y=y, 
+           shape = material, 
+           fill = munsell_hue)) +
+  geom_point(size=3) +
+  theme_bw() +
+  ggtitle("Raman") +
+  labs(x = pc1lab,
+       y = pc2lab) +
+  scale_shape_manual(name = "Material",
+                     values=shapes) +
+  scale_fill_manual(name = "Munsell hue",
+                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
+                    values=colors) +
+  guides(shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol=2, 
+                              title.position="top", 
+                              title.hjust = 0.5),
+         fill = guide_legend(override.aes = list(shape = 22,
+                                                 fill = colors,
+                                                 color = "black",
+                                                 size=3),
+                             title.position="top", title.hjust = 0.5))+
+  theme(legend.position = "none")
+
+#bind the basic fviz plot for PC 1 and 3, and use as basis for a more customizeable plot in ggpplot
+p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
+       aes(x=x, y=y, 
+           shape = material, 
+           fill = munsell_hue)) +
+  geom_point(size=3) +
+  theme_bw() +
+  labs(x = pc1lab,
+       y = pc3lab) +
+  scale_shape_manual(name = "Material",
+                     values=shapes) +
+  scale_fill_manual(name = "Munsell hue",
+                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
+                    values=colors) +
+  guides(shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol=2, 
+                              title.position="top", 
+                              title.hjust = 0.5),
+         fill = guide_legend(override.aes = list(shape = 22,
+                                                 fill = colors,
+                                                 color = "black",
+                                                 size=3),
+                             title.position="top", title.hjust = 0.5))+
+  theme(legend.position = "none")
+
+combined <-
+  p1 + p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
+                  axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+                  axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+                  legend.title = element_text(size = 12, face = "bold", colour = "black"),
+                  legend.background = element_rect(linetype = "solid", color = "black"),
+                  legend.position = "bottom")
+
+combined +
+  plot_layout(nrow = 2, guides = "collect")
 ```
@@
 
-```{r raman-pca1, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components of the raman analysis.
-The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1.
-Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_6_raman_average_pca_p1_p2.jpg')
-```
-
-\linebreak
-
-```{r raman-pca2, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 3rd components of the raman analysis.
-The materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G band peaks around 1300 cm-1 and 1600 cm-1.
-Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_7_raman_average_pca_p1_p3.jpg')
+```{r raman-pca1, fig.fullheight=TRUE, fig.fullwidth=TRUE, fig.cap="PCA score plots for the raman analysis.
+Along the 1st component (P1) materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G bands around 1300 cm-1 and 1600 cm-1.
+Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak A) Plot of 1st and 2nd components; noisier spectra drift outwards along P2 B) Plot of 1st and 3rd components; noisier spectra are separated out from other samples, but spectra with peaks around 700-800 cm-1 are drawn towards the bottom right."}
+knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_6_raman_average_pca_p1_p2_p3.jpg')
 ```
@@
 
-```{r xrf-violin-plot, fig.width=15, fig.height=10, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
+```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
 **n** represents no.
@@
 Points_long <-
-  Points %>%
+  Points.xrf %>%
   select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
@@
 #Lookup for element labeller
+#Ändra ordning så att de mer avvikande (Si, Mn etc) hamnar uppe till vänster
 Elements <- c(
@@
 
-```{r xrf-loadings, fig.width=8, fig.height=8, fig.fullwidth=TRUE, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
-#Prepare plots for 3-plot bargraph figure of loadings
-#PC1 loadings
-P1 <-
-  ggplot(loadings, aes(x = fct_rev(Element), y = P1)) +
+```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
+#bar graph of the loadings (P1-P3) for the XRF data  
+loadings %>%
+  ggplot(aes(x=fct_rev(Element), y=Value)) +
     geom_bar(
       stat = "identity", position = "identity",
-      color = "black", fill = "darkgrey") +
-  theme_minimal() +
-  theme(plot.margin = margin(10,10,10,10),
-        axis.title.y=element_blank(),
-        axis.text.y = element_text(size=14, face="bold", colour = "black"),
-        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
-  coord_flip()
-#PC2 loadings
-P2 <-
-  ggplot(loadings, aes(x = fct_rev(Element), y = P2)) +
-  geom_bar(
-    stat = "identity", position = "identity",
-    color = "black", fill = "darkgrey") +
-  theme_minimal() +
-  theme(plot.margin = margin(10,10,10,10),
-        axis.title.y=element_blank(),
-        axis.text.y=element_blank(),
-        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
-  coord_flip()
-#PC3 loadings
-P3 <-
-  ggplot(loadings, aes(x = fct_rev(Element), y = P3)) +
-  geom_bar(
-    stat = "identity", position = "identity",
-    color = "black", fill = "darkgrey") +
-  theme_minimal() +
+      color = "black", fill = "darkgrey",
+      width = 0.5) +
+  facet_wrap( ~ P, scales = "free") +
+  theme_bw() +
   theme(plot.margin = margin(10,10,10,10),
         axis.title.y = element_blank(),
-        axis.text.y = element_blank(),
-        axis.title.x = element_text(size = 14, face = "bold", colour = "black")) +
+        axis.text.y = element_text(size=10, face="bold", colour = "black"),
+        axis.title.x = element_blank(),
+        strip.text.x = element_text(size=14, face="bold", colour="black")) +
   coord_flip()
 
-#Composite figure of above bar graphs
-grid.arrange(P1, P2, P3, ncol=3)
 ```

### 2021-09-06 nir.R
Temporary scripts for testing isolated code

@@ -1,0 +1,65 @@
+library(tidyverse)
+library(ggplot2)
+library(ggbiplot)
+library(prospectr)
+library(patchwork)
+library(factoextra)
+
+#Import raman data
+nir.csv <-
+  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//NIR_master_averaged_20210827.csv", sep = ";",
+            dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+
+#Filter to focus on quartz/quartzite material that are points and preforms
+Points.nir <-
+  nir.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Mark the samples that lack a munsell hue as transclucent
+Points.nir[8][is.na(Points[8])] <- "Translucent"
+
+#perform PCA with SNV normalization and mean-center
+nir.pca <-
+  prcomp(Points.nir[,c(678:2178)], center = TRUE, scale = FALSE)
+
+#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
+basic_plot1 <-
+  fviz_pca_ind(nir.pca, axes = c(1,2), label="none")
+
+#prepare labels for PCs
+pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
+pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
+pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
+pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+
+
+#prepare color/fill and symbols
+colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
+shapes <- c(24,25,23,22,21)
+
+#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
+ggplot(cbind(basic_plot1$data, Points[, c(8,10)]),
[…]

### 2021-09-06 raman.R

@@ -1,0 +1,94 @@
+library(tidyverse)
+library(ggplot2)
+library(ggbiplot)
+library(prospectr)
+library(patchwork)
+library(factoextra)
+
+#Import raman data
+raman.csv <-
+  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//raman_baseline_corrected_20210826.csv", sep = ";",
+            dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+
+#Filter to focus on quartz/quartzite material that are points and preforms
+Points <-
+  raman.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Mark the samples that lack a munsell hue as transclucent
+Points[8][is.na(Points[8])] <- "Translucent"
+
+#perform PCA with SNV normalization and mean-center
+raman.pca <-
+  prcomp(standardNormalVariate(X = Points[,c(29:320)]), center = TRUE, scale = FALSE)
+
+#prepare different basic score plots with fviz_pca_ind
+basic_plot1 <-
+  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")
+
+basic_plot2 <-
+  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")
+
+#prepare labels for PCs
+pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
+pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
+pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
+pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
+pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+pc3lab <- paste0("PC3 (",as.character(pc3var),"%)")
+
+#prepare color/fill and symbols
+colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
[…]

### 2021-09-06 xrf pca.R

@@ -1,0 +1,64 @@
+library(tidyverse)
+library(ggplot2)
+library(missMDA)
+library(ggbiplot)
+library(factoextra)
+
+
+
+#Import xrf data
+XRF.csv <-
+  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data//XRF/XRF_quantitative.csv", sep = ";",
+            dec = ".", header = TRUE)
+
+Points <-
+  XRF.csv %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#select elements to include in pca
+xrf <-
+  Points %>%
+    select(reading_no, sample_id, munsell_hue, material, X12Mg, X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti)
+
+#log transform data
+xrf.log <- log(xrf[,5:20]+1)
+
+#impute missing vales with scale
+xrf.imp <- imputePCA(xrf.log, ncp = 2, scale = TRUE, method = "Regularized", maxiter = 1000)
+#perform pca with mean centering
+xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=FALSE)
+
+#prepare labels for PCs
+pc1var <- round(summary(xrf.pca)$importance[2,1]*100, digits=1)
+pc2var <- round(summary(xrf.pca)$importance[2,2]*100, digits=1)
+pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
+pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+
+#prepare color and symbols
+colors <- c("#E31CFF", "#0080FF", "#FFFF00", "#D5DF9E", "#718EFF", "#FFFFC1", "#C6DF39", "#FFFFC1", "#FFB468", "#FFFF38", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
+shapes <- c(24,25,23,22,21)
+
+#score plot, need to specify to use ggbiplot package
+ggbiplot::ggbiplot(xrf.pca,
+         obs.scale = 1,
[…]

### 2021-09-07 analysis/paper/paper.Rmd
Updated code with relative paths

@@
 library(tidyverse)
+library(gridExtra)
 library(ggplot2)
@@
 XRF.csv <-
-read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data//XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE)
+  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE)
 
@@
 nir.csv <-
-  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//NIR_master_averaged_20210827.csv", sep = ";",
+  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210827.csv", sep = ";",
             dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
@@
 raman.csv <-
-  read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/derived_data//raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
 
@@
 samples <-
-  as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data//bifacial_points.csv", sep = ";", dec = ".", header = TRUE, na.strings=c("", "NA")))
+  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, na.strings=c("", "NA"))
 
@@
 loadings <-
- as.data.frame(read.csv2("C:/Users//masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/data/raw_data/XRF/XRF_P_values_long.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
+ as.data.frame(read.csv2("./analysis/data/raw_data/XRF/XRF_P_values_long.csv", sep = ";", dec = ".", header = TRUE, encoding = "UTF-8"))
 ```
@@
 Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.", fig.width=15, fig.height=15, fig.fullwidth=TRUE}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/Distribution.jpeg')
+knitr::include_graphics('../figures/Distribution.jpeg', error = FALSE)
 ```
@@
 
-```{r vis-nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum.
+<!--```{r vis-nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum.
 Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_1_visnir_averaged_pca_p1_p2.jpg')
+knitr::include_graphics('./analysis/figures/fig_4_1_visnir_averaged_pca_p1_p2.jpg')
+```-->
+```{r vis-nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum.
+Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
+A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
+
+#perform PCA with SNV normalization and mean-center
+visnir.pca <-
+  prcomp(Points.nir[,c(78:428)], center = TRUE, scale = FALSE)
+
+#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
+basic_plot1 <-
+  fviz_pca_ind(visnir.pca, axes = c(1,2), label="none")
+
+#prepare labels for PCs
+pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
+pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
+pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
+pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+
+#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
+ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
+             aes(x=x, y=y, 
+                 shape = material, 
+                 fill = munsell_hue)) +
+  geom_point(size=3) +
+  theme_bw() +
+  ggtitle("Near infrared 400 - 750 nm") +
+  labs(x = pc1lab,
+       y = pc2lab) +
+  scale_shape_manual(name = "Material",
+                     values=shapes) +
+  scale_fill_manual(name = "Munsell hue",
+                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
+                    values=colors) +
+  guides(shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol=2, 
+                              title.position="top", 
+                              title.hjust = 0.5),
+         fill = guide_legend(override.aes = list(shape = 22,
+                                                 fill = colors,
+                                                 color = "black",
+                                                 size=3),
+                             title.position="top", title.hjust = 0.5))+
+  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
+        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+        legend.title = element_text(size = 12, face = "bold", colour = "black"),
+        legend.background = element_rect(linetype = "solid", color = "black"),
+        legend.position = "bottom")
 ```
-
  \linebreak
@@
 Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
+knitr::include_graphics('./analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
 ```
@@
 
-combined <-
+comb <-
   p1 + p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
@@
 
-combined +
+comb +
   plot_layout(nrow = 2, guides = "collect")
@@
 
-```{r raman-pca1, fig.fullheight=TRUE, fig.fullwidth=TRUE, fig.cap="PCA score plots for the raman analysis.
+<!--```{r raman-pca1, fig.fullheight=TRUE, fig.fullwidth=TRUE, fig.cap="PCA score plots for the raman analysis.
 Along the 1st component (P1) materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G bands around 1300 cm-1 and 1600 cm-1.
 Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak A) Plot of 1st and 2nd components; noisier spectra drift outwards along P2 B) Plot of 1st and 3rd components; noisier spectra are separated out from other samples, but spectra with peaks around 700-800 cm-1 are drawn towards the bottom right."}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_6_raman_average_pca_p1_p2_p3.jpg')
-```
+knitr::include_graphics('./analysis/figures/fig_4_6_raman_average_pca_p1_p2_p3.jpg')
+```-->
 
@@
 
-```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
-```
+<!--```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
+knitr::include_graphics('./analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
+```-->
 

### 2021-09-07 analysis/paper/paper.Rmd
Added code for plotting raman spectra

@@
   geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
-
 ```
@@
 
-```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region.
+<!--```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region.
 Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance.
@@
 knitr::include_graphics('./analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
-```
+```-->
 
@@
 ## Raman
-There are two distinctive features present in the Raman data.
+There are two distinctive features present in the Raman data (fig.
+\@ref(fig:raman-spectra)).
 The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset.
@@
 Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on different types of shale, as well as samples of diabase, apatite, orthoclase, and blue quartz.
-Studies of clay minerals [@kloprogge17]
-have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group.
+Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group.
 Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
@@
 The resulting PCA model (fig.
-\@ref(fig:nir-pca1)) shows a clear trend along the first component (81.5 %) related to the above two features.
+\@ref(fig:raman-pca)) shows a clear trend along the first component (81.5 %) related to the above two features.
 The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side.
@@
 
+```{r raman-spectra, fig.width=20, fig.height=8, fig.cap="Spectra of raman measurements sampled on a dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
+#Identify file with raman spectra and import
+file <- '../quantifyingquartzite/analysis/data/derived_data/raman_baseline_corrected_transposed_20210826.csv'
+spectra <- read.csv2(file, sep = ";", dec = ",", header = TRUE)
+
+#Prepare function for plotting spectra
+plot_multi <- function(df, x=1, y=2, cols=y,
+                       xlim=range(df[,x], na.rm = T),
+                       ylim=range(df[,y], na.rm = T),
+                       main="", xlab="", ylab="", ...){
+  # setup plot frame
+  plot(NULL,
+       xlim=xlim,
+       ylim=ylim,
+       main=main, xlab=xlab, ylab=ylab)
+
+  # plot all your y's against your x
+  pb <- sapply(seq_along(y), function(i){
+    points(df[,c(x, y[i])], col=cols[i], ...)
+  })
+}
+
+plot_multi(spectra, y=120, cols = rainbow(ncol(spectra)), type='l', lwd=3,
+           xlab = "Wavenumber (cm-1)", ylab = "Intensity")
+```
+
+\linebreak
+
 <!--```{r raman-average, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
@@
 ```-->
-```{r raman-average, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
+```{r raman-pca, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
 
@@
 ```
-
-\linebreak
 

### 2021-09-09 analysis/paper/paper.Rmd
added code to replicate prediktera PCA figures and loadings with ggplot/ggbiplot

@@
 ```{r load-packages-and-data}
-library(tidyverse)
-library(gridExtra)
-library(ggplot2)
-library(gridExtra)
-library(ggsci)
-library(prospectr)
-library(patchwork)
+library(ChemoSpec)
+library(data.table)
 library(factoextra)
 library(ggbiplot)
+library(ggplot2)
+library(ggsci)
+library(gridExtra)
+library(missMDA)
+library(patchwork)
+library(prospectr)
+library(tidyverse)
 
@@
 XRF.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE)
+  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, na = c("","NA","NULL"))
 
@@
 Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
-Points.xrf[8][is.na(Points.xrf[8])] <- "Translucent"
+Points.xrf[20][is.na(Points.xrf[20])] <- "Translucent"
 
 #prepare color/fill and symbols for score plots
-colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
-shapes <- c(24,25,23,22,21)
+pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
+pca.shapes <- c(24,25,23,22,21)
+pca.munsell_hue <- c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")
 
@@
 #prepare labels for PCs
-pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
-pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
-pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
-pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
+vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
+vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
+vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
 
@@
   ggtitle("Near infrared 400 - 750 nm") +
-  labs(x = pc1lab,
-       y = pc2lab) +
+  labs(x = vis.pc1lab,
+       y = vis.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=shapes) +
+                     values=pca.shapes) +
   scale_fill_manual(name = "Munsell hue",
                     breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=colors) +
+                    values=pca.colors) +
   guides(shape = guide_legend(override.aes = list(fill = "black"), 
@@
          fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = colors,
+                                                 fill = pca.colors,
                                                  color = "black",
                                                  size=3),
-                             title.position="top", title.hjust = 0.5))+
+                             title.position="top", 
+                             title.hjust = 0.5)) +
   theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
@@
 #prepare labels for PCs
-pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
-pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
-pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
-pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
+nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
+nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
+nir.pc1lab <- paste0("PC1 (",as.character(nir.pc1var),"%)")
+nir.pc2lab <- paste0("PC2 (",as.character(nir.pc2var),"%)")
 
@@
   ggtitle("Near infrared 1 000 - 2 500 nm") +
-  labs(x = pc1lab,
-       y = pc2lab) +
+  labs(x = nir.pc1lab,
+       y = nir.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=shapes) +
+                     values=pca.shapes) +
   scale_fill_manual(name = "Munsell hue",
                     breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=colors) +
+                    values=pca.colors) +
   guides(shape = guide_legend(override.aes = list(fill = "black"), 
@@
          fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = colors,
+                                                 fill = pca.colors,
                                                  color = "black",
@@
 
-```{r raman-spectra, fig.width=20, fig.height=8, fig.cap="Spectra of raman measurements sampled on a dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
-#Identify file with raman spectra and import
+```{r raman-spectra, fig.width=20, fig.height=10, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue.
+The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the light and dark groups"}
+#Read and create new file for chemospec
+raman <-matrix2SpectraObject(
+  gr.crit = c("Translucent", "White", "Dark", "Light" ),
+  gr.cols = c("purple", "blue", "black", "yellow"),
+  freq.unit ="Wavenumber (cm-1)",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/raman_baseline_corrected_transposed_20210908.csv",
+  sep = ";",
+  dec = ",",
+  chk = TRUE,
+  out.file = "raman")
+
+surveySpectra(raman, by.gr = TRUE , method = "sem", main = "mean")
+
+```
+<!-- #Identify file with raman spectra and import
 file <- '../quantifyingquartzite/analysis/data/derived_data/raman_baseline_corrected_transposed_20210826.csv'
@@
 plot_multi(spectra, y=120, cols = rainbow(ncol(spectra)), type='l', lwd=3,
-           xlab = "Wavenumber (cm-1)", ylab = "Intensity")
-```
+           xlab = "Wavenumber (cm-1)", ylab = "Intensity") -->
 
@@
 #prepare labels for PCs
-pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
-pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
-pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
-pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
-pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
-pc3lab <- paste0("PC3 (",as.character(pc3var),"%)")
+raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
+raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
+raman.pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
+raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
+raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
+raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")
 
 #bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
-p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
+raman.p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
        aes(x=x, y=y, 
@@
   ggtitle("Raman") +
-  labs(x = pc1lab,
-       y = pc2lab) +
+  labs(x = raman.pc1lab,
+       y = raman.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=shapes) +
+                     values=pca.shapes) +
   scale_fill_manual(name = "Munsell hue",
                     breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=colors) +
+                    values=pca.colors) +
   guides(shape = guide_legend(override.aes = list(fill = "black"), 
@@
 #bind the basic fviz plot for PC 1 and 3, and use as basis for a more customizeable plot in ggpplot
-p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
+raman.p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
        aes(x=x, y=y, 
@@
   theme_bw() +
-  labs(x = pc1lab,
-       y = pc3lab) +
+  labs(x = raman.pc1lab,
+       y = raman.pc3lab) +
   scale_shape_manual(name = "Material",
-                     values=shapes) +
+                     values=pca.shapes) +
   scale_fill_manual(name = "Munsell hue",
                     breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=colors) +
+                    values=pca.colors) +
   guides(shape = guide_legend(override.aes = list(fill = "black"), 
@@
 comb <-
-  p1 + p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
+  raman.p1 + raman.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
                               axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
@@
 #Prepare colors
-colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")
+loading.colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")
 
@@
     facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
-    scale_fill_manual(values=colors) +
+    scale_fill_manual(values=loading.colors) +
     theme_bw() +
@@
 ```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
+#select elements to include in pca, as well as columns to group by
+xrf <-
+  Points.xrf %>%
+    select(reading_no, sample_id, munsell_hue, material, X12Mg, X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti)
+
+#log transform data
+xrf.log <- log(xrf[,5:20])
+
+#impute missing values
+xrf.imp <- imputePCA(xrf.log, ncp = 2, scale = FALSE, method = "Regularized", maxiter = 1000)
+
+#perform pca with mean centering and unite variance scaling
+xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=TRUE)
+
+#Extract loadings from xrf.pca
+loadings <- as.data.frame(xrf.pca$rotation[,1:3])
+#As elements are stored as rownames, set them as first column so it can be used as a variable
+loadings <- setDT(loadings, keep.rownames = "Element")[]
+# remove inital 'X' from element names, which was introduced in initial import to R
+loadings$Element <- substring(loadings$Element, 2)
+#Melt loadings from wide to long format
+loadings.melted <- melt(loadings, id.vars="Element")
+#Set names of columns
+loadings <- setNames(loadings.melted, c("Element", "PC", "Value"))
+
 #bar graph of the loadings (P1-P3) for the XRF data
@@
     width = 0.5) +
-  facet_wrap( ~ P, scales = "free") +
+  facet_wrap( ~ PC, scales = "free") +
   theme_bw() +
@@
   coord_flip()
-
 ```
@@
 ```-->
+```{r xrf-pca1, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 2nd components.
+The smaller group to the left in the plot includes samples with higher Si content, while the larger group includes samples with Si content < 50%."}
+
+#Prepare two score plots
+ggbiplot(xrf.pca,
+         choices=c(1,2),
+         obs.scale = 1,
+         var.scale = 1,
+         group=NULL,
+         varname.size = 4,
+         labels.size=3,
+         ellipse = FALSE,
+         circle = FALSE) +
+theme_minimal() +
+geom_point(aes(shape=Points.xrf$material, 
+               fill = Points.xrf$munsell_hue), 
+           size = 3) +
+coord_cartesian() +
+scale_shape_manual(name = "Material",
+                   values=pca.shapes) +
+scale_fill_manual(name = "Munsell hue",
+                  breaks = pca.munsell_hue,
+                  values=pca.colors) +
+guides(shape = guide_legend(override.aes = list(fill = "black"),
+                            ncol=2,
+                            title.position="top",
+                            title.hjust = 0.5),
+       fill = guide_legend(override.aes = list(shape = 22,
+                                               fill = pca.colors,
+                                               color = "black",
+                                               size=3),
+                           title.position="top", title.hjust = 0.5)) +
+theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
+      axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+      axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+      legend.title = element_text(size = 12, face = "bold", colour = "black"),
+      legend.background = element_rect(linetype = "solid", color = "black"),
+      legend.position = "bottom")
+```
+
+\linebreak
+
+```{r xrf-pca2, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 3rd components.
+The same groupings of low/high Si content remains along PC1, but PC3 seems to separate darker and light samples further in the low Si group."}
+ggbiplot(xrf.pca,
+         choices=c(1,3),
+         obs.scale = 1,
+         var.scale = 1,
+         group=NULL,
+         varname.size = 4,
+         labels.size=3,
+         ellipse = FALSE,
+         circle = FALSE) +
+theme_minimal() +
+geom_point(aes(shape=Points.xrf$material, 
+               fill = Points.xrf$munsell_hue), 
+           size = 3) +
+coord_cartesian() +
+scale_shape_manual(name = "Material",
+                   values=pca.shapes) +
+scale_fill_manual(name = "Munsell hue",
+                  breaks = pca.munsell_hue,
+                  values=pca.colors) +
+guides(shape = guide_legend(override.aes = list(fill = "black"),
+                            ncol=2,
+                            title.position="top",
+                            title.hjust = 0.5),
+       fill = guide_legend(override.aes = list(shape = 22,
+                                               fill = pca.colors,
+                                               color = "black",
+                                               title.position="top", 
+                                               title.hjust = 0.5))) +
+theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
+      axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
+      axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+      legend.title = element_text(size = 12, face = "bold", colour = "black"),
+      legend.background = element_rect(linetype = "solid", color = "black"),
+      legend.position = "bottom")
+```
 

### 2021-09-10 analysis/paper/paper.Rmd
added check.names = FALSE which removes need for substringing data when transposing it. Working on adding spectra plots for raman and nir

@@
 XRF.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, na = c("","NA","NULL"))
+  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
@@
   read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210827.csv", sep = ";",
-            dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+            dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 raman.csv <-
-  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, na = c("","NA","NULL",NULL))
+  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 samples <-
-  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, na.strings=c("", "NA"))
+  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
 
@@
 The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
-The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1.
+The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1.
 Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
@@
 While a fifth peak at 2 450 nm can be difficult to discern in most cases, the average spectra for both the *brecciated*/*translucent* group and *white* group feature the above mentioned four peaks.
+
+\linebreak
+
+```{r nir-spectra, fig.width=20, fig.height=8, fig.cap= "Average NIR spectra by hue.
+Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/translucent samples."}
+
+#Aggregate raw NIR data by inferred hue
+nir.agg <- 
+  aggregate(nir.csv[, 28:2178], list(nir.csv$hue), mean)
+
+#Transpose the data in preparation for working in ChemoSpec 
+# make row names unique & sorted correctly in increasing order
+# by appending numbers in running order
+nir.t <- nir.agg %>%
+  dplyr::mutate(Group.1 = paste(seq(1, n()),
+                         Group.1 ,
+                      sep = "_")) %>%
+  gather(row, value, -Group.1 ) %>% 
+  spread(Group.1 , value)
+
+# strip away the appended numbers from the newly created column names
+names(nir.t) <- sapply(strsplit(names(nir.t), "_"), function(x){x[2]})
+#Change first column name to 'freq'  
+colnames(nir.t)[1] <- "freq"
+
+#Set the first column containing frequencies to numeric
+nir.t[,1] <- sapply(nir.t[,1], as.numeric)
+#Sort dataframe by frequencies asc.
+nir.t <- nir.t[order(nir.t$freq),] 
+#Write to csv file, since ChemoSpec doesn't work with dataframes
+write.csv2(nir.t,"./analysis/data/derived_data/nir_averaged_by_hue.csv", row.names = FALSE)
+
+#Read exported csv and create new file for chemospec
+nir <-matrix2SpectraObject(
+  gr.crit = c("Translucent", "White", "Dark", "Light" ),
+  gr.cols = c("purple", "blue", "black", "red"),
+  freq.unit ="Wavelength (nm)",
+  int.unit ="Absorbance",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/NIR_averaged_by_hue.csv",
+  sep = ";",
+  dec = ",",
+  chk = TRUE,
+  out.file = "nir")
+
+#Plot spectra averaged by inferred hue
+Dark <- plotSpectra(nir,
+            which = 1,
+            yrange = c(1.05,1.2),
+            lab.pos = 0.0,
+            main = "Average spectra - Dark", font.lab = 2)
+#Plot spectra averaged by inferred hue
+Light <- plotSpectra(nir,
+            which = 2,
+            yrange = c(0.7,1),
+            lab.pos = 0.0,
+            main = "Average spectra - Dark", font.lab = 2)
+#Plot spectra averaged by inferred hue
+Translucent <- plotSpectra(nir,
+            which = 3,
+            yrange = c(1.05,1.2),
+            lab.pos = 0.0,
+            main = "Average spectra - Dark", font.lab = 2)
+#Plot spectra averaged by inferred hue
+White <- plotSpectra(nir,
+            which = 4,
+            yrange = c(1.05,1.2),
+            lab.pos = 0.0,
+            main = "Average spectra - Dark", font.lab = 2)
+grid.arrange()
+```
 
@@
   Points.xrf %>%
-  select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
+  select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
   gather(key = "Element", value = "Value")
-
-#Remove initial X introduced by read.csv to column names
-Points_long$Element <- substring(Points_long$Element, 2)
 
@@
   Points.xrf %>%
-    select(reading_no, sample_id, munsell_hue, material, X12Mg, X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti)
+    select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)
 
@@
 loadings <- setDT(loadings, keep.rownames = "Element")[]
-# remove inital 'X' from element names, which was introduced in initial import to R
-loadings$Element <- substring(loadings$Element, 2)
 #Melt loadings from wide to long format

### 2021-09-14 analysis/paper/paper.Rmd
New code for spectra in place. Updating text.

@@
 library(tidyverse)
+library(spectrolab)
 
@@
 nir.csv <-
-  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210827.csv", sep = ";",
+  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210914.csv", sep = ";",
             dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
@@
 The colour of the material has a mainly black (N1-N3), grey or white (N9) hue, as well as opaque and almost translucent material with a beige tint.
+The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *translucent*, *white* ~ N9).
 A number of these points have, according to prior classifications, been worked using bifacial technology.
@@
 
-```{r distribution-map, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates.
+```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates.
 The distribution of sampled sites is likely affected by the excavation and survey history of the region.
-Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet.", fig.width=15, fig.height=15, fig.fullwidth=TRUE}
+Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
 knitr::include_graphics('../figures/Distribution.jpeg', error = FALSE)
@@
 
-```{r distribution-by-parish, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden.
-Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301"), fig.width=12, fig.height=10, fig.fullwidth=TRUE}
+```{r distribution-by-parish, fig.width=12, fig.height=10, fig.fullwidth=TRUE, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden.
+Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301")}
 
@@
  \linebreak
-
 All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF.
@@
 This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind.
-The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that likely were made from the same material, however, for the smaller sites there still remains some uncertainties.
+The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties.
 
@@
 After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020).
-PCA modelling was performed using the Prediktera Evince software (www.prediktera.com), which uses the NIPALS algorithm [@bertils19 pp.9-10].
+PCA modelling was performed using the base R *prcomp()* method [@Rteam17].
 
@@
 
-All spectra were averaged for each artefact and the resulting spectra were then baseline corrected using modified polynomial fitting in R [@Rteam17] using the ChemoSpec package [@hanson16].
-The spectra were mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA in the Prediktera Evince software.
+After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data.
+This was achieved using modified polynomial fitting in R [@Rteam17] with the ChemoSpec package [@hanson16].
+The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA.
 
@@
 XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a Thermo ScientificTM portable test stand.
-The reference calibration used for element quantification was *mining*.
+The reference calibration used for element quantification was set to *mining* mode.
 Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light).
@@
 For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
-The XRF data was pre-processed using mean-centering, univariate scaling, and log-transformation in the Prediktera Evince software before PCA modelling.
+The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling.
+As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA.
+Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* package [@josse_husson16] developed for datasets with incomplete data.
+Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was chosen.
 
@@
 
-The main explanatory power for the visual region can be found in the first two components, featuring eigenvalues above 1 with a cumulative explained variance of 99.9 %.
-Plotting these in the score plot there is a clear trend along the first component with darker materials being grouped together to the right of the plot and light/white material to the left (fig.
-\@ref(fig:vis-nir-pca)).
-Average spectra of these groups  demonstrate a higher absorbance for the dark group, with both of the light groups featuring higher absorbance in the violet region (fig.
-\@ref(fig:vis-nir-spectra)).
-Despite this trend there are exceptions, and while some of these may be material that exist within the muddled greyscale, there are measurements on dark surfaces (N1-N2) that end up closer to the light group (fig.X).
+The main explanatory power for the visual region can be found in the first component with an explained variance of 98.4 %.
+In the PCA (fig.
+\@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/translucent/white material to the left.
+Examining the average spectra of these groups (fig.
+\@ref(fig:vis-nir-spectra)), as expected, the darker material feature higher absorbance of light than the other groups.
+No other distinctive features can be seen in the visual band, that could potentially be used to distinguish the groups.
+It is, however, apparent that while the more extreme hues of *dark* and *white* material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.
+
+A number of artefacts also breaks the observed trend in the PCA, with dark/black (N1-N2) samples ending up closer to the light and translucent group in the PCA.
 There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA.
-Artefact 403 and 502 both have a sample taken on one side in the ‘dark’ group and the other in the ‘greyish’ middle.
+In examination of the non-averaged data artefact 403 and 502 both have a sample taken on one side in the *dark* group and the other in the *greyish* middle.
 This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering.
-The ‘greyish’ middle group is largely made up of ‘quartz coloured’ opaque material with a beige/grey tint (fig.X).
+The *greyish* middle group in the PCA is otherwise largely made up of *quartz coloured* opaque material with a beige/grey tint.
 
@@
 ```-->
-```{r vis-nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum.
+
+```{r vis-nir-pca, fig.width=20, fig.height=8, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum.
 Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
-A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
+The strong trend along PC1 also makes it difficult to separate out the muddled grayscale in the middle."}
 
@@
 ```
- \linebreak
+
+\linebreak
+
+```{r vis-nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue. Darker samples feature higher overall absorbance with white material absorbing the least amount of light."}
+
+#Filter NIR data to focus on the material with a dark hue 
+#and select the NIR range 1 000 - 2 500 nm
+vis.dark <- Points.nir %>% 
+  filter(hue == "Dark") %>% 
+  select(c(1:27),c(78:428))
+
+spec.vd <- 
+      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:27)
+
+vis.light <- Points.nir %>% 
+  filter(hue == "Light") %>% 
+  select(c(1:27),c(78:428))
+
+spec.vl <- 
+  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:27)
+
+vis.translucent <- Points.nir %>% 
+  filter(hue == "Translucent") %>% 
+  select(c(1:27),c(78:428))
+
+spec.vt <- 
+  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:27)
+
+vis.white <- Points.nir %>% 
+  filter(hue == "White") %>% 
+  select(c(1:27),c(78:428))
+
+spec.vw <- 
+  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:27)
+
+par(mfrow=c(4,1))
+
+plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark")
+
+plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
+
+plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Translucent")
+
+plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White")
+```
 
@@
 ## NIR - near infrared range
-No distinct peaks occurred in the 600 – 1 000 nm region, and thus the analysis focused on the 1 000 – 2 500 nm region to lessen the impact of noise.
-The PCA demonstrates a similar pattern to the one observed in the visual region (fig.
+No distinct peaks occurred in the 600 – 1 000 nm band, and thus the analysis focused on the 1 000 – 2 500 nm band to lessen the impact of noise.
+In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm and can be seen in the average spectra (fig.
+\@ref(fig:nir-spectra)).
+Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter could possibly be related to CO^3^ [@clark_etal90].
+While the dark group features a higher overall absorbance the peaks are less distinct.
+The same can be said for the translucent and white group, where the water peak at 1 900 nm is the most pronounced.
+The absorbance in the translucent group is slightly higher in the lower band regions than compared to the white group, but with decreased absorbance in the higher regions.
+While a fifth peak at 2 450 nm can be difficult to discern in most cases....
+
+The PCA demonstrates a similar pattern to the one observed in the visual band (fig.
 \@ref(fig:nir-pca)).
-Again the first two components have an eigenvalue above 1 with cumulative explained variance of ~99.9 %.
+Again the first two components explain most of the variance in the model, ~ 99.9 %.
 While the darker materials remain closely grouped in the right side of the score plot, the *light* cluster seems to split into two separate groups.
@@
 The bottom left group instead features the majority of the white *quartz-coloured* material.
-In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm.
-In some of the measurements peaks also occur at 2 200 nm and 2 350 nm, where the former can be related to combination vibrations of Al-OH and the latter possibly CO^3^ [@clark_etal90].
-The average spectra for the three groups show that the *dark* group has the highest absorbance with less distinct peaks, although they are noticeable.
-The absorbance in the *brecciated*/*translucent* group is slightly higher in the lower band regions than compared to the *white* group, but with decreased absorbance in the higher regions.
-While a fifth peak at 2 450 nm can be difficult to discern in most cases, the average spectra for both the *brecciated*/*translucent* group and *white* group feature the above mentioned four peaks.
-
-\linebreak
-
-```{r nir-spectra, fig.width=20, fig.height=8, fig.cap= "Average NIR spectra by hue.
-Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/translucent samples."}
-
-#Aggregate raw NIR data by inferred hue
-nir.agg <- 
-  aggregate(nir.csv[, 28:2178], list(nir.csv$hue), mean)
-
-#Transpose the data in preparation for working in ChemoSpec 
-# make row names unique & sorted correctly in increasing order
-# by appending numbers in running order
-nir.t <- nir.agg %>%
-  dplyr::mutate(Group.1 = paste(seq(1, n()),
-                         Group.1 ,
-                      sep = "_")) %>%
-  gather(row, value, -Group.1 ) %>% 
-  spread(Group.1 , value)
-
-# strip away the appended numbers from the newly created column names
-names(nir.t) <- sapply(strsplit(names(nir.t), "_"), function(x){x[2]})
-#Change first column name to 'freq'  
-colnames(nir.t)[1] <- "freq"
-
-#Set the first column containing frequencies to numeric
-nir.t[,1] <- sapply(nir.t[,1], as.numeric)
-#Sort dataframe by frequencies asc.
-nir.t <- nir.t[order(nir.t$freq),] 
-#Write to csv file, since ChemoSpec doesn't work with dataframes
-write.csv2(nir.t,"./analysis/data/derived_data/nir_averaged_by_hue.csv", row.names = FALSE)
-
-#Read exported csv and create new file for chemospec
-nir <-matrix2SpectraObject(
-  gr.crit = c("Translucent", "White", "Dark", "Light" ),
-  gr.cols = c("purple", "blue", "black", "red"),
-  freq.unit ="Wavelength (nm)",
-  int.unit ="Absorbance",
-  descrip ="Bifacial points measurements",
-  in.file = "./analysis/data/derived_data/NIR_averaged_by_hue.csv",
-  sep = ";",
-  dec = ",",
-  chk = TRUE,
-  out.file = "nir")
-
-#Plot spectra averaged by inferred hue
-Dark <- plotSpectra(nir,
-            which = 1,
-            yrange = c(1.05,1.2),
-            lab.pos = 0.0,
-            main = "Average spectra - Dark", font.lab = 2)
-#Plot spectra averaged by inferred hue
-Light <- plotSpectra(nir,
-            which = 2,
-            yrange = c(0.7,1),
-            lab.pos = 0.0,
-            main = "Average spectra - Dark", font.lab = 2)
-#Plot spectra averaged by inferred hue
-Translucent <- plotSpectra(nir,
-            which = 3,
-            yrange = c(1.05,1.2),
-            lab.pos = 0.0,
-            main = "Average spectra - Dark", font.lab = 2)
-#Plot spectra averaged by inferred hue
-White <- plotSpectra(nir,
-            which = 4,
-            yrange = c(1.05,1.2),
-            lab.pos = 0.0,
-            main = "Average spectra - Dark", font.lab = 2)
-grid.arrange()
-```
 
@@
 ``` -->
+
+```{r nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue.
+Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/translucent samples."}
+
+#Filter NIR data to focus on the material with a dark hue 
+#and select the NIR range 1 000 - 2 500 nm
+dark <- Points.nir %>% 
+  filter(hue == "Dark") %>% 
+  select(c(1:27),c(678:2178))
+
+#create a spectra object for the dark material, and then repeat for light, white and translucent groups
+spec.d <- 
+  as_spectra(dark, name_idx = 1, meta_idxs = 2:27)
+
+light <- Points.nir %>% 
+  filter(hue == "Light") %>% 
+  select(c(1:27),c(678:2178))
+
+spec.l <- 
+  as_spectra(light, name_idx = 1, meta_idxs = 2:27)
+
+translucent <- Points.nir %>% 
+  filter(hue == "Translucent") %>% 
+  select(c(1:27),c(678:2178))
+
+spec.t <- 
+  as_spectra(translucent, name_idx = 1, meta_idxs = 2:27)
+
+white <- Points.nir %>% 
+  filter(hue == "White") %>% 
+  select(c(1:27),c(678:2178))
+
+spec.w <- 
+  as_spectra(white, name_idx = 1, meta_idxs = 2:27)
+
+#Prepare a layout with 1 column and four rows
+par(mfrow=c(4,1))
+#Plot the mean spectrum for each spectra group above
+plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark")
+
+plot(mean(spec.l), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
+
+plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Translucent")
+
+plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White")
+
+```
+
+\linebreak
+
 ```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum.
@@
 
-```{r raman-spectra, fig.width=20, fig.height=10, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue.
-The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the light and dark groups"}
+```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue.
+The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the dark group"}
 #Read and create new file for chemospec
-raman <-matrix2SpectraObject(
-  gr.crit = c("Translucent", "White", "Dark", "Light" ),
-  gr.cols = c("purple", "blue", "black", "yellow"),
-  freq.unit ="Wavenumber (cm-1)",
-  int.unit ="Intensity",
-  descrip ="Bifacial points measurements",
-  in.file = "./analysis/data/derived_data/raman_baseline_corrected_transposed_20210908.csv",
-  sep = ";",
-  dec = ",",
-  chk = TRUE,
-  out.file = "raman")
-
-surveySpectra(raman, by.gr = TRUE , method = "sem", main = "mean")
-
-```
-<!-- #Identify file with raman spectra and import
-file <- '../quantifyingquartzite/analysis/data/derived_data/raman_baseline_corrected_transposed_20210826.csv'
-spectra <- read.csv2(file, sep = ";", dec = ",", header = TRUE)
-
-#Prepare function for plotting spectra
-plot_multi <- function(df, x=1, y=2, cols=y,
-                       xlim=range(df[,x], na.rm = T),
-                       ylim=range(df[,y], na.rm = T),
-                       main="", xlab="", ylab="", ...){
-  # setup plot frame
-  plot(NULL,
-       xlim=xlim,
-       ylim=ylim,
-       main=main, xlab=xlab, ylab=ylab)
-
-  # plot all your y's against your x
-  pb <- sapply(seq_along(y), function(i){
-    points(df[,c(x, y[i])], col=cols[i], ...)
-  })
-}
-
-plot_multi(spectra, y=120, cols = rainbow(ncol(spectra)), type='l', lwd=3,
-           xlab = "Wavenumber (cm-1)", ylab = "Intensity") -->
+#raman <-matrix2SpectraObject(
+  #gr.crit = c("Translucent", "White", "Dark", "Light" ),
+  #gr.cols = c("purple", "blue", "black", "yellow"),
+  #freq.unit ="Wavenumber (cm-1)",
+  #int.unit ="Intensity",
+  #descrip ="Bifacial points measurements",
+  #in.file = "./analysis/data/derived_data/raman_baseline_corrected_transposed_20210908.csv",
+  #sep = ";",
+  #dec = ",",
+ # chk = TRUE,
+ # out.file = "raman")
+
+#surveySpectra(raman, by.gr = TRUE , method = "sem", main = "mean")
+
+r.dark <- Points.raman %>% 
+  filter(hue == "Dark")
+
+spec.rd <- 
+  as_spectra(r.dark, name_idx = 1, meta_idxs = 2:28)
+
+r.light <- Points.raman %>% 
+  filter(hue == "Light")
+
+spec.rl <- 
+  as_spectra(r.light, name_idx = 1, meta_idxs = 2:28)
+
+r.translucent <- Points.raman %>% 
+  filter(hue == "Translucent")
+
+spec.rt <- 
+  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:28)
+
+r.white <- Points.raman %>% 
+  filter(hue == "White")
+
+spec.rw <- 
+  as_spectra(r.white, name_idx = 1, meta_idxs = 2:28)
+
+par(mfrow=c(4,1))
+
+plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - Dark")
+
+plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - Light")
+
+plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - Translucent")
+
+plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - White")
+```
 
@@
 ```-->
-```{r raman-pca, fig.width=20, fig.height=15, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
+
+```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra.
+The main explanatory power can be found in the 1^st^ component, which groups the data based on the quartz and D/G bands.
+As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA.
+The 3^rd^ component does, however, also separate out the few samples that feature peaks at 700-800 cm-1 that may result from Al, Si and OH vibrations."}
 

### 2021-09-27 analysis/paper/paper.Rmd
Removed old project file name, added river classification to data files and updated code

@@
   filter(hue == "Dark") %>% 
-  select(c(1:27),c(78:428))
+  select(c(1:28),c(79:429))
 
 spec.vd <- 
-      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:27)
+      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "Light") %>% 
-  select(c(1:27),c(78:428))
+  select(c(1:28),c(79:429))
 
 spec.vl <- 
-  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "Translucent") %>% 
-  select(c(1:27),c(78:428))
+  select(c(1:28),c(79:429))
 
 spec.vt <- 
-  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "White") %>% 
-  select(c(1:27),c(78:428))
+  select(c(1:28),c(79:429))
 
 spec.vw <- 
-  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "Dark") %>% 
-  select(c(1:27),c(678:2178))
+  select(c(1:28),c(679:2179))
 
@@
 spec.d <- 
-  as_spectra(dark, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(dark, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "Light") %>% 
-  select(c(1:27),c(678:2178))
+  select(c(1:28),c(679:2179))
 
 spec.l <- 
-  as_spectra(light, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(light, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "Translucent") %>% 
-  select(c(1:27),c(678:2178))
+  select(c(1:28),c(679:2179))
 
 spec.t <- 
-  as_spectra(translucent, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(translucent, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "White") %>% 
-  select(c(1:27),c(678:2178))
+  select(c(1:28),c(679:2179))
 
 spec.w <- 
-  as_spectra(white, name_idx = 1, meta_idxs = 2:27)
+  as_spectra(white, name_idx = 1, meta_idxs = 2:28)
 
@@
 #Plot the mean spectrum for each spectra group above
-plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark")
-
-plot(mean(spec.l), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
-
-plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Translucent")
-
-plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White")
+plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
+     xlab = "Wavelength (nm)", main = "Average - Dark")
+
+plot(mean(spec.l), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
+     xlab = "Wavelength (nm)", main = "Average - Light")
+
+plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
+     xlab = "Wavelength (nm)", main = "Average - Translucent")
+
+plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
+     xlab = "Wavelength (nm)", main = "Average - White")
 
@@
 nir.pca <-
-  prcomp(Points.nir[,c(678:2178)], center = TRUE, scale = FALSE)
+  prcomp(Points.nir[,c(679:2179)], center = TRUE, scale = FALSE)
 
@@
 nir.pc2lab <- paste0("PC2 (",as.character(nir.pc2var),"%)")
+
+#NIR LOADING LINES
+#par(mfrow=c(3,1))
+#sapply(1:3, function(comp){
+#  plot(1000:2500, nir.pca$rotation[,comp], type='l', lwd=2,
+#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1000,2500,6))})
 
@@
 695 cm-1 and 795 cm-1.
-Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks turned up in measurements collected on different types of shale, as well as samples of diabase, apatite, orthoclase, and blue quartz.
+Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale.
 Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group.
@@
 The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the dark group"}
-#Read and create new file for chemospec
-#raman <-matrix2SpectraObject(
-  #gr.crit = c("Translucent", "White", "Dark", "Light" ),
-  #gr.cols = c("purple", "blue", "black", "yellow"),
-  #freq.unit ="Wavenumber (cm-1)",
-  #int.unit ="Intensity",
-  #descrip ="Bifacial points measurements",
-  #in.file = "./analysis/data/derived_data/raman_baseline_corrected_transposed_20210908.csv",
-  #sep = ";",
-  #dec = ",",
- # chk = TRUE,
- # out.file = "raman")
-
-#surveySpectra(raman, by.gr = TRUE , method = "sem", main = "mean")
 
@@
 spec.rd <- 
-  as_spectra(r.dark, name_idx = 1, meta_idxs = 2:28)
+  as_spectra(r.dark, name_idx = 1, meta_idxs = 2:29)
 
@@
 spec.rl <- 
-  as_spectra(r.light, name_idx = 1, meta_idxs = 2:28)
+  as_spectra(r.light, name_idx = 1, meta_idxs = 2:29)
 
@@
 spec.rt <- 
-  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:28)
+  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:29)
 
@@
 spec.rw <- 
-  as_spectra(r.white, name_idx = 1, meta_idxs = 2:28)
+  as_spectra(r.white, name_idx = 1, meta_idxs = 2:29)
 
@@
 
-plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - Dark")
-
-plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - Light")
-
-plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - Translucent")
-
-plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavenumber (cm-1)", main = "Average - White")
+plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
+     xlab = "Wavenumber (cm-1)", main = "Average - Dark")
+
+plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
+     xlab = "Wavenumber (cm-1)", main = "Average - Light")
+
+plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
+     xlab = "Wavenumber (cm-1)", main = "Average - Translucent")
+
+plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
+     xlab = "Wavenumber (cm-1)", main = "Average - White")
 ```
@@
 raman.pca <-
-  prcomp(standardNormalVariate(X = Points.raman[,c(29:320)]), center = TRUE, scale = FALSE)
+  prcomp(standardNormalVariate(X = Points.raman[,c(30:321)]), center = TRUE, scale = FALSE)
 
@@
 raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")
+
+#RAMAN LOADING LINES
+#par(mfrow=c(2,2))
+#sapply(1:4, function(comp){
+#  plot(row.names(raman.pca$rotation), raman.pca$rotation[,comp], type='l', lwd=2, 
+#       main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(92.88,1798.3,6))})
 

### 2021-10-05 analysis/paper/paper.Rmd
Updated text

@@
 No distinct peaks occurred in the 600 – 1 000 nm band, and thus the analysis focused on the 1 000 – 2 500 nm band to lessen the impact of noise.
-In the case of NIRS most features relate to the presence of molecular water in the material [@hunt77; @sciuto_etal19], these mainly occur at around 1 400 nm and 1 900 nm and can be seen in the average spectra (fig.
+In the case of NIRS similar features can be observed for all four colour groups, with darker materials exhibiting higher absorbance and less defined peaks.
+The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig.
 \@ref(fig:nir-spectra)).
-Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter could possibly be related to CO^3^ [@clark_etal90].
-While the dark group features a higher overall absorbance the peaks are less distinct.
-The same can be said for the translucent and white group, where the water peak at 1 900 nm is the most pronounced.
-The absorbance in the translucent group is slightly higher in the lower band regions than compared to the white group, but with decreased absorbance in the higher regions.
-While a fifth peak at 2 450 nm can be difficult to discern in most cases....
-
-The PCA demonstrates a similar pattern to the one observed in the visual band (fig.
+Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falls within one of the stronger CO^3^ bands [@clark_etal90].
+Inspection of individual spectra for the *light* group, however, also reveals a 5th peak for some samples at 2 450 nm.
+Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
+It is therefore possibly that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other cases Al-OH.
+
+The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig.
 \@ref(fig:nir-pca)).
-Again the first two components explain most of the variance in the model, ~ 99.9 %.
-While the darker materials remain closely grouped in the right side of the score plot, the *light* cluster seems to split into two separate groups.
-The upper left group houses the main bulk of the *brecciated quartz*, as well as the almost translucent and opaque material with a beige tint.
-The bottom left group instead features the majority of the white *quartz-coloured* material.
+Darker quartzite cluster to the right with white and translucent samples grouping to the opposite side.
+While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left.
+A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points.
+These three samples come from different sites and exhibit no distinguishing peaks in their spectra, making them an interesting set of outliers.
 
@@
 The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples.
-These peak positions fall within what is referred to as the D and G band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
-The D band occurs as a result of disordered carbon structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
+These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
+The D band occurs as a result of disordered C structures whilst the G band is indicative of sp^2^ C networks and caused by stretching in the C bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
 Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1.
@@
 Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group.
-Considering the diagenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
+Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
 
 The resulting PCA model (fig.
-\@ref(fig:raman-pca)) shows a clear trend along the first component (81.5 %) related to the above two features.
-The measurements with mainly quartz peaks gravitate towards the positive side of the 1st component, and those where the graphite peaks are dominant towards the opposite side.
+\@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D/G peaks.
+PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after preprocessing.
+Interestingly, PC 4 does catch the Al-O-Si peaks around the  700 cm-1 - 800 cm-1 band, which helps in separating out those samples where this feature is present.
+The measurements with mainly quartz peaks gravitates towards the positive side of the PC 1, and those where the graphite peaks are dominant towards the opposite side.
 It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks.
 As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-The 2nd component (6.29 %) mainly seems to differentiate spectra with a lot of noise, where the objects were difficult to sample.
-In contrast, along the 3rd component those samples that feature more pronounced peaks around 700 - 800 cm-1 groups up separately from the *quartz group*.
+When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the *quartz* group.
 
@@
 basic_plot2 <-
-  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")
+  fviz_pca_ind(raman.pca, axes = c(1,4), label="none")
 
@@
 raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
-raman.pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
+raman.pc4var <- round(summary(raman.pca)$importance[2,4]*100, digits=1)
 raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
 raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
-raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")
+raman.pc4lab <- paste0("PC4 (",as.character(raman.pc4var),"%)")
 
@@
 
-The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 86.5%.
+The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 84.7%.
 The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings).
-In the score plot of P1 and P2, two large clusters emerges (fig.
+In the score plot of PC 1 and PC 2, two large clusters can be seen (fig.
 \@ref(fig:xrf-score-plot-p1-p2)).
-The loadings for ^13^Al, ^18^Sr, ^19^K, ^20^Ca, ^40^Zr, ^81^Ti group together in the top right quadrant whilst ^12^Mg, ^14^Si, ^15^P, ^16^S, ^23^V, and ^56^Ba group in the bottom right.
-As seen in figure \@ref(fig:xrf-score-plot-p1-p2), ^26^Fe is the only element positioned on the left side of the model.
-Interestingly, the left cluster contains the samples with lower ^14^Si values and the right cluster the higher values.
-These clusters still remain if you were to exclude ^14^Si from the model, however, suggesting there are other elements of relevance to the clustering.
-The “low ^14^Si” cluster shows a trend along P2 of quartzite with a darker hue grouping towards the top.
-This may indicate a possible relation to the ^26^Fe content of the material, as a darker hue may be the result of inclusions of hematite, or possibly biotite.
-^13^Al follows a similar trend with higher values recorded at the top of the two clusters.
-The *brecciated quartz* samples largely group together at the bottom of the “low ^14^Si” cluster, with a couple ending up towards the bottom of the “high ^14^Si” cluster.
-The white (N9) material is somewhat evenly distributed within the leftmost cluster.
-The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the top are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river.
+Interestingly, the smaller cluster on the left side in the model includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content.
+It is not ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that drives this.
+The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom.
+This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite ([@howard05] p.711).
+The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up towards closer to the “high ^14^Si” cluster.
+The white (N9) material is somewhat evenly distributed within the right cluster.
+
+The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river.
 The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %.
 This artefact was sampled twice on each side, resulting in similar results both times.
-One almost translucent sample also ends up among the darker material.
-Upon closer inspection this sample is extremely small, 3 by 1 mm, and also feature inclusions of dark particles.
+
+One almost translucent sample also ends up in the "high ^14^Si" cluster among the darker material.
+Upon closer inspection this sample is small, ~ 3 by 1 mm, and also feature inclusions of dark particles.
 Both of these factors may have contributed to its placement in the model.

### 2021-10-06 analysis/paper/paper.Rmd
Added text

@@
 Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
-A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
+A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen.
+Three black samples at the top, from different sites, all feature no distinguishing peaks in their spectra thus separating them from the other samples in the dataset."}
 
@@
 nir.pca <-
-  prcomp(Points.nir[,c(679:2179)], center = TRUE, scale = FALSE)
+  prcomp(Points.nir[,c(680:2179)], center = TRUE, scale = FALSE)
 
@@
 #NIR LOADING LINES
-#par(mfrow=c(3,1))
-#sapply(1:3, function(comp){
-#  plot(1000:2500, nir.pca$rotation[,comp], type='l', lwd=2,
-#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1000,2500,6))})
+par(mfrow=c(3,1))
+sapply(1:3, function(comp){
+  plot(1001:2500, nir.pca$rotation[,comp], type='l', lwd=2,
+      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1001,2500,6))})
 
@@
 \@ref(fig:raman-spectra)).
-The first can be identified as the *quartz* peak occurring at 460 cm-1 [@gillet_etal90], which is present for all samples in the dataset.
+The first are three peaks in the 100 cm-1 - 500 cm-1 band that are typical for SiO~2~ and quartz-based materials, with the strongest occuring at 460 cm-1 [@gillet_etal90].
+These peaks are present for all samples in the dataset.
 The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples.
@@
 The D band occurs as a result of disordered C structures whilst the G band is indicative of sp^2^ C networks and caused by stretching in the C bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
-Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being 1 330 - 1 350 cm-1.
+Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm-1 region.
 It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280].
@@
 \@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D/G peaks.
-PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after preprocessing.
-Interestingly, PC 4 does catch the Al-O-Si peaks around the  700 cm-1 - 800 cm-1 band, which helps in separating out those samples where this feature is present.
-The measurements with mainly quartz peaks gravitates towards the positive side of the PC 1, and those where the graphite peaks are dominant towards the opposite side.
+PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
+Interestingly, the loading for PC 4 does catch the Al-O-Si peaks around the  700 cm-1 - 800 cm-1 band, which helps in separating out those samples where this feature is present.
+The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side.
 It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks.
@@
 
-par(mfrow=c(4,1))
+par(mfrow=c(2,2))
 
@@
   labs(x = raman.pc1lab,
-       y = raman.pc3lab) +
+       y = raman.pc4lab) +
   scale_shape_manual(name = "Material",
@@
 
-The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, none of them are capable of providing a detailed fingerprint on their own based solely on surface measurements.
-Taken together they do provide a more in-depth understanding of the characteristic and structure of the material.
-The most easily recognizable megascopic feature of the material is its colour, and the results of all three instrumentations demonstrate that there is some truth to this.
-However, a large group of the material, which features different hues of grey, does not conform as easily.
-In each of the PCA models there is a *greyish* middle that makes classification difficult.
+ The structures seen in the NIRS PCA are
+
+However, a large group of the material that feature different hues of grey, does not conform as easily.
+In each of the PCA models there is a *greyish* middle of samples that fall in between the extremes of the observable trends that make classification difficult.
 
@@
 # Conclusion
+
+The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint on their own based solely on surface measurements.
+Taken together, however, they do provide a more in-depth understanding of the characteristic and structure of the material.
+
+The most easily recognizable megascopic feature of the material is its colour, and this has been one of the main means of describing quartzite in Northern Sweden (e.g.
+[@forsberg85], []).
+The overall structure in the spectroscopic results for each of the instrumentations, and especially the NIRS, groups samples of similar color (given as munsell hue) together, however, the presence of samples that break these trends demonstrate the presence of structural and compositional features that do not necessarily align with the perceived color of a material.
+
+In the case of the raman data the PCA is somewhat lacking in exploratory and descriptive power.
+Apart from the 700 cm-1 - 800 cm-1 band peaks the major features in the data are explained by PC 1.
+Intensity in raman spectroscopy is not a direct quantitative indicator, thus making it difficult to group materials along the PC 1, apart from presence/absence of the *graphite peaks*.
+
+The metamorphic background of quartzite makes it a complex rock for characterization, but that is not to say that there aren't potential markers that may be used for fingerprinting.
+The presence of sp^2^-C and AlOH peaks are worthwhile to follow up.
 

### 2021-10-07 analysis/paper/paper.Rmd
Added to discussion

@@
   raman.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
-         type == "Point" | type == "Point fragment" | type == "Preform")
+    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
 
@@
 
- The structures seen in the NIRS PCA are
-
-However, a large group of the material that feature different hues of grey, does not conform as easily.
-In each of the PCA models there is a *greyish* middle of samples that fall in between the extremes of the observable trends that make classification difficult.
+The most easily recognizable megascopic feature of petrographic material is its colour, a feature that has been one of the main means of grouping quartzite in North Swedish archaeological research (e.g.
+[@forsberg85] p.60-64, [@kaeck09] p.
+53-55).
+While it is true that some inference can be made based on colour, such as ^26^Fe-bearing minerals contributing to a darker colour, this becomes increasingly difficult with heterogeneous material.
+Should a quartzite where the matrix is mainly light gray in colour, but streaked with black mineral veins, be considered a *light* or *dark* quartzite?
+The results above echoes this complexity.
+While the overall structure in the spectroscopic results for each of the instrumentations, and especially the NIRS, groups samples of similar colour (given as *munsell hue*) together, there are samples that break these trends and thus demonstrate the presence of structural and compositional features that do not necessarily align with the perceived colour of a material.
+
+In the case of the raman data the PCA is somewhat lacking in exploratory and descriptive power.
+Apart from the few samples that feature the 700 cm-1 - 800 cm-1 band peaks the major features in the data are all explained by the first component.
+As the intensity in raman spectroscopy is not a direct quantitative indicator this makes it difficult to group materials based on the intensity of the D/G bands, making more of a binary separation of presence/absence of the *graphite peaks*.
+That is not to say that it is not possible to attempt quantification using raman spectroscopy, but rather that this would demand a different experimental setup.
+That said, the presence of the *graphite peaks* may represent an important clue in determining the potential geological origin for the material.
+The presence of the C structures suggest the petrogenesis included sediments enriched with organic material, providing a good filter for continued research.
 
@@
 
-The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint on their own based solely on surface measurements.
-Taken together, however, they do provide a more in-depth understanding of the characteristic and structure of the material.
-
-The most easily recognizable megascopic feature of the material is its colour, and this has been one of the main means of describing quartzite in Northern Sweden (e.g.
-[@forsberg85], []).
-The overall structure in the spectroscopic results for each of the instrumentations, and especially the NIRS, groups samples of similar color (given as munsell hue) together, however, the presence of samples that break these trends demonstrate the presence of structural and compositional features that do not necessarily align with the perceived color of a material.
-
-In the case of the raman data the PCA is somewhat lacking in exploratory and descriptive power.
-Apart from the 700 cm-1 - 800 cm-1 band peaks the major features in the data are explained by PC 1.
-Intensity in raman spectroscopy is not a direct quantitative indicator, thus making it difficult to group materials along the PC 1, apart from presence/absence of the *graphite peaks*.
-
-The metamorphic background of quartzite makes it a complex rock for characterization, but that is not to say that there aren't potential markers that may be used for fingerprinting.
-The presence of sp^2^-C and AlOH peaks are worthwhile to follow up.
+The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint based solely on surface measurements.
+Nevertheless they do provide a more in-depth understanding of the composition and structure of the material, and combining these results may be the key to more robust non-destructive characterization of petrographic material.
 

### 2021-10-15 analysis/paper/paper.Rmd
More references and updated introduction

@@
   fig.path = "../figures/",
-  dpi = 300
+  dpi = 600
 )
@@
 
-#Import xrf data
+#Import xrf data, set empty fields to NA
 XRF.csv <-
@@
 
-#Import loadings from Evince
-loadings <-
- as.data.frame(read.csv2("./analysis/data/raw_data/XRF/XRF_P_values_long.csv", sep = ";", dec = ".", header = TRUE,  encoding = "UTF-8"))
 ```
@@
 
-In European archaeology provenience studies and petrographic characterization has for some time favoured flint and obsidian over other geological materials [@prieto_etal20 p.32].
-However, in regions where local flint and obsidian sources are non-existent other materials have taken on the dominant role.
-In Northern Fennoscandia these include quartz, quartzite and slate  [@baudou78; @rankama_etal06; @olausson_etal12], and while both quartz and quartzite technology have been the target of a number of studies over the years there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-
-One reason for the lack of methodological development could be the geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
-The degree to which geological materials have been mixed and transported means that considerable effort is needed in order to reconstruct the events that have formed the modern day landscape.
-Both quartz and quartzite could potentially have been sourced by local prehistoric groups from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations [@rankama_etal06].
-Due to this ambigous nature of availability of raw materials it is difficult to model sourcing strategies for groups that were based within this region.
+In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials like quartz, quartzite and slate have been the primary target for lithic production^[This is not to say that flint was not available or used in this region] [@baudou78; @rankama_etal06; @olausson_etal12].
+Non-flint lithics and raw material studies in  European archaeology have seen a steady increase during the last decades, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; -@prieto_etal20; -@prieto_etal21; @ramacciotti_etal19].
+With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region.
+While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
+
+One reason for the lack of methodological development could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
+The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape.
+While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06].
+Due to this ambigous nature of availability of raw materials lprovenience studies are increasingly complex and fraught with uncertainty.
+
+The defintition for what constitutes a *quartzite* is not so stratightforward.
 Prieto et al.
-[-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that ‘quartzite’ represents a somewhat ambiguous term in archaeology, where it is often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
-This issue has been discussed in geology as well [@howard05], and the definition hence varies somewhat between authors.
-Visual characterization of quartzite on a megascopic level is typically unreliable, and microscopic analysis is recommended for identification of subtypes based on metamorphic grade, but Howard  also emphasises the need for field-based methodologies.
-
-In recent years there has been increased development of petrographic characterization of quartzite using destructive methods [e.g.
+[-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
+The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and the definition hence varies somewhat between authors.
+Visual characterization of quartzite on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies.
+
+In recent years there has been increased development of petrographic characterization of quartzite using a combination of destructive methods and spectroscopy [e.g.
 @blomme_etal12; @cnudde_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20].
@@
 While destructive methods may prohibit access to such material there are still non-destructive methods that can generate valuable information.
-The use of spectroscopy in petrographic characterization is not new, but it is often applied in conjunction with thin-section analysis [e.g.
-@blomme_etal12; @prieto_etal20].
-A major advantage of spectroscopy, however, is that it is both non-destructive and feature fast measurements, making it an attractive alternative for field-based screening purposes.
-While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of quantifying the elemental content of a material, there are other instrumentations that may provide a deeper insight into the structural properties.
+Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive alternative for field-based screening purposes.
+While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data of the elemental content of a material, there are other instrumentations that may provide a deeper insight into the structural properties.
 Near Infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
-The same can be said for Raman spectroscopy, which also has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts [@vandenabeele12].
+The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts [@vandenabeele12].
 
@@
 ```{r select-point-data}
-#Filter data to focus on quartzite points, point fragments and preforms made from quartz/quartzite materials
+#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
 Points2 <-
@@
 It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature.
-In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz, and brecciated quartzite ^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake].
+In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake], and brecciated quartzite.
 Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
@@
 par(mfrow=c(4,1))
-#Plot the mean spectrum for each spectra group above
+#Plot the mean spectrum for each spectra group above with text annotations for the peaks in the first dark spectra
 plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
      xlab = "Wavelength (nm)", main = "Average - Dark")
+text(x=1405, y=1.098, labels="OH", font=2, cex = 1.5)
+text(x=1935, y=1.123, labels="OH", font=2, cex = 1.5)
+text(x=2215, y=1.130, labels="AlOH", font=2, cex = 1.5)
+text(x=2350, y=1.130, labels="CO3", font=2, cex = 1.5)
 
@@
 #NIR LOADING LINES
-par(mfrow=c(3,1))
-sapply(1:3, function(comp){
-  plot(1001:2500, nir.pca$rotation[,comp], type='l', lwd=2,
-      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1001,2500,6))})
+#par(mfrow=c(3,1))
+#sapply(1:3, function(comp){
+#  plot(1001:2500, nir.pca$rotation[,comp], type='l', lwd=2,
+#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1001,2500,6))})
 
@@
 
+#Prepare a layout with 2 columns and 2 rows
 par(mfrow=c(2,2))
 
+#Plot the mean spectrum for each spectra group above with text annotations for the peaks in the first dark spectra
 plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
      xlab = "Wavenumber (cm-1)", main = "Average - Dark")
+text(x=120, y=2900, labels="SiO", font=2, cex = 1.5)
+text(x=200, y=4300, labels="SiO", font=2, cex = 1.5)
+text(x=460, y=9250, labels="SiO", font=2, cex = 1.5)
+text(x=1285, y=6300, labels="D", font=2, cex = 2.0)
+text(x=1600, y=4000, labels="G", font=2, cex = 2.0)
 
@@
 The most easily recognizable megascopic feature of petrographic material is its colour, a feature that has been one of the main means of grouping quartzite in North Swedish archaeological research (e.g.
-[@forsberg85] p.60-64, [@kaeck09] p.
+@forsberg85 p.60-64, @kaeck09 p.
 53-55).
@@
 
-Interestingly the ‘brecciated quartz’ remains relatively grouped together in all of the models demonstrating some consistency in the original classification.
-There are exceptions, however, and it would be advisable to follow up with a more directed analysis focused around this group of material……
+Interestingly the *brecciated quartz* remains relatively grouped together in all of the models demonstrating some consistency with the original classification.
+As the material does group with other similarly semi-translucent material it does not seem to distinguish itself enough in the models in order to consider it its own group.
+The original classification by @ahman67 refers to the microscopic fragmented configuration of the quartz crystals as being the main distinguishing feature, and while it is possible to discern the presence of SiO structure through non-destructive spectroscopy, there have been no other discernible features related to this group.
+
+Keeping in mind the
 

### 2021-10-19 analysis/paper/paper.Rmd
Rewritten text, added some visualisation to figures

@@
       fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      reference_docx: "../templates/template.docx" 
       pandoc_args:
@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-and-anthropological-science.csl" # Insert path for the bib-style
+csl: "../templates/journal-of-archaeological-and-anthropological-science.csl"
 abstract: |
@@
 ---
-
-<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
 
@@
 
+<!-- The following code chunk defines prepares most of the data that will be needed and reads in the necessary libraries -->
+
 ```{r load-packages-and-data}
@@
 library(ggsci)
+library(grid)
+library(gridBase)
 library(gridExtra)
@@
   XRF.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
          type == "Point" | type == "Point fragment" | type == "Preform")
@@
   nir.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
          type == "Point" | type == "Point fragment" | type == "Preform")
@@
   raman.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
 
 #Fill the NA fields in the munsell_hue column and mark them as translucent samples
-Points.nir[8][is.na(Points.nir[8])] <- "Translucent"
-Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
-Points.xrf[20][is.na(Points.xrf[20])] <- "Translucent"
+Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
+Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"
 
@@
 pca.shapes <- c(24,25,23,22,21)
-pca.munsell_hue <- c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")
-
-```
-<!-- The actual document text starts here: -->
+pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")
+
+```
 
@@
 
-In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials like quartz, quartzite and slate have been the primary target for lithic production^[This is not to say that flint was not available or used in this region] [@baudou78; @rankama_etal06; @olausson_etal12].
-Non-flint lithics and raw material studies in  European archaeology have seen a steady increase during the last decades, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; -@prieto_etal20; -@prieto_etal21; @ramacciotti_etal19].
+In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] [@baudou78; @rankama_etal06; @olausson_etal12].
+The last decades have seen an increase in interest for non-flint lithics and raw material studies in  European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
 With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region.
-While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-
-One reason for the lack of methodological development could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
+While quartz and quartzite technology have been part of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
+One reason for the lack of methodological development in North Swedish research could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
 The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape.
-While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06].
-Due to this ambigous nature of availability of raw materials lprovenience studies are increasingly complex and fraught with uncertainty.
-
-The defintition for what constitutes a *quartzite* is not so stratightforward.
+While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
+Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.
+
+Attempting to define what constitutes a *quartzite* is not so stratightforward.
 Prieto et al.
 [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
-The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and the definition hence varies somewhat between authors.
-Visual characterization of quartzite on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies.
-
-In recent years there has been increased development of petrographic characterization of quartzite using a combination of destructive methods and spectroscopy [e.g.
-@blomme_etal12; @cnudde_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20].
-These have had success in distinguishing between different groups of quartzite types, allowing for more meaningful studies into quartzite-based resource management strategies.
-In archaeology there are times, however, when destructive techniques simply are not applicable.
+The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and it hence varies somewhat between authors.
+Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies.
+There are also situations where destructive techniques simply are not applicable.
 Artefact collections (museum, private, etc.) represent a valuable resource, as well as one of the easier ways of gaining access to large datasets with some geographical distribution.
@@
 Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive alternative for field-based screening purposes.
-While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data of the elemental content of a material, there are other instrumentations that may provide a deeper insight into the structural properties.
-Near Infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
-The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and the analysis of artwork and manuscripts [@vandenabeele12].
+While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
+The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
+The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].
 
 In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
-They all provide different information on the structure and composition of the material and will thus be evaluated in regards to the different features they record in the quartz/quartzite.
-This will be achieved via Principal Component Analysis (PCA) where the observed groupings of the material will be related to the spectral features.
-Lastly, the potential for combining the results of the analyses in order to study joint relations will be discussed.
-
+The goal is to evaluate their capabilities for fingerprinting quartz and quartzite in the North Swedish region.
+This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to the spectral features for each instrumentation.
  \linebreak
@@
   samples %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | material  == "Felsitic porphyric quartz",
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
          type == "Point" | type == "Point fragment" | type == "Preform")
@@
   lycksele %>% 
-    filter(material == "Quartz" | material == "Quartzite" | material == "Breccie quartz" | material == "Breccie quartzite" | 
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | 
              material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
@@
 
-The current study is based on spectroscopic surface measurements on quartzite points found within the County of Västerbotten, Sweden.
-A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten and Skellefteå museums were analysed using three different spectroscopic instrumentations (Appendix XX).
-The colour of the material has a mainly black (N1-N3), grey or white (N9) hue, as well as opaque and almost translucent material with a beige tint.
-The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *translucent*, *white* ~ N9).
+The current study is based on spectroscopic surface measurements on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden.
+A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy (Appendix XX).
+The material features a mainly black (N1-N3), grey or white (N9) hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
+The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *white* ~ N9, *Colourless* ~ lack of colour, opaque to translucent) for visualization purposes.
 A number of these points have, according to prior classifications, been worked using bifacial technology.
@@
 This has likely had a contributing factor to the distribution pattern of the artefacts observed in fig.
-\@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele, Vilhelmina and Tärnaby.
+\@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele^[In this case the majority of the samples from Lycksele are from the rock painting site of Lycksele 301], Vilhelmina and Tärnaby.
 This can be seen more clearly in fig.
@@
 
-Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable.
+Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online.
 Based on a query of all points labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database.
@@
 Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
-The reason for including ‘quartz’ in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.
+The reason for including *quartz* in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.
 
@@
  \linebreak
-All measurements were performed at the Environmental Archaeology Lab at Umeå University using three different instrumentations; NIRS, Raman scattering, and XRF.
+All measurements were performed at the Environmental Archaeology Lab at Umeå University.
 The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens.
@@
 The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties.
+For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
 
 Pre-processing of the spectra is different for each instrumentation and based on relevant literature.
-After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset (Geladi and Linderholm, 2020).
-PCA modelling was performed using the base R *prcomp()* method [@Rteam17].
+After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20].
+PCA modelling was performed using the base R *prcomp* function [@Rteam17; @rcoreteam21].
 
@@
 NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe.
-The ASD features a detection range of 350 – 2500 nm, with spectral sampling (resp.
+The ASD features a detection range of 350 nm – 2500 nm, with spectral sampling (resp.
 spectral resolution) of 1.4 nm (resp.
@@
 Each object was measured against a white reference background and the collected measurements were averaged for each artefact.
-Two parts of the NIR spectra were analysed separately, the visual region (400 – 750 nm) and the 1 000 – 2 500 nm region.
-In both analyses the data was mean-centred before PCA modelling.
+Two parts of the As there were no notable features in the visible range the analysis was restricted to the 1 000 nm – 2 500 nm region in order to lessen the impact of noise.
+The data was mean-centred before PCA modelling.
 
@@
 The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
-The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1.
+The spectral coverage of the instrumentation is 175 cm-1 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1.
 Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
@@
 ## X-Ray Fluourescence
-XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a Thermo ScientificTM portable test stand.
+XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand.
 The reference calibration used for element quantification was set to *mining* mode.
@@
 Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions.
-For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
 The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling.
 As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA.
-Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* package [@josse_husson16] developed for datasets with incomplete data.
-Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was chosen.
+Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
+Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was used.
 
@@
 \linebreak
-
-<!--```{r vis-nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 400 nm – 750 nm region of the spectrum.
-Quartzite with a darker surface group to the right in the model, while white objects group in the bottom left quadrant."}
-knitr::include_graphics('./analysis/figures/fig_4_1_visnir_averaged_pca_p1_p2.jpg')
-```-->
 
@@
   scale_fill_manual(name = "Munsell hue",
-                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
+                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                     values=pca.colors) +
@@
 vis.translucent <- Points.nir %>% 
-  filter(hue == "Translucent") %>% 
+  filter(hue == "Colourless") %>% 
   select(c(1:28),c(79:429))
@@
 
-plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark")
+plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
 
@@
 
-plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Translucent")
-
-plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White")
-```
-
-<!--```{r vis-nir-spectra, fig.width=15, fig.height=5, fig.fullwidth=TRUE, fig.cap="Average spectra for dark (N1), translucent (MV) and white (N1) groups in the 400 - 750 nm region.
-Dark material has a higher overall absorbance, whereas white and translucent material feature lower absorbance.
-Brecciated quartz, which is typically described as a translucent material, falls somewhere inbetween the above groups."}
-knitr::include_graphics('./analysis/figures/fig_4_2_breccie_averaged_visnir_spectra.jpg')
-```-->
+plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")
+
+plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
+```
 
@@
 ## NIR - near infrared range
-No distinct peaks occurred in the 600 – 1 000 nm band, and thus the analysis focused on the 1 000 – 2 500 nm band to lessen the impact of noise.
-In the case of NIRS similar features can be observed for all four colour groups, with darker materials exhibiting higher absorbance and less defined peaks.
+In the case of NIRS similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks.
 The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig.
@@
 Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
-It is therefore possibly that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other cases Al-OH.
+It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH.
 
@@
 A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points.
-These three samples come from different sites and exhibit no distinguishing peaks in their spectra, making them an interesting set of outliers.
-
-\linebreak
-
-<!--```{r nir-pca, fig.width=18, fig.height=5, fig.fullwidth=TRUE, fig.cap="PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the #spectrum.
-Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
-A trend #of translucent material and opaque material with a beige tint grouping in the top left can also be seen."}
-#knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_3_nir_averaged_pca_p1_p2.jpg')
-``` -->
+These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
+
+\linebreak
 
@@
 translucent <- Points.nir %>% 
-  filter(hue == "Translucent") %>% 
+  filter(hue == "Colourless") %>% 
   select(c(1:28),c(679:2179))
@@
 par(mfrow=c(4,1))
-#Plot the mean spectrum for each spectra group above with text annotations for the peaks in the first dark spectra
+
+#plot the mean spectra for each group
 plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - Dark")
-text(x=1405, y=1.098, labels="OH", font=2, cex = 1.5)
-text(x=1935, y=1.123, labels="OH", font=2, cex = 1.5)
-text(x=2215, y=1.130, labels="AlOH", font=2, cex = 1.5)
-text(x=2350, y=1.130, labels="CO3", font=2, cex = 1.5)
+     xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
 
@@
 plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - Translucent")
+     xlab = "Wavelength (nm)", main = "Average - Colourless")
 
 plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - White")
-
+     xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
+
+pushViewport(vps1)
+X <- convertX(unit(1413,"native"), "npc")
+X2 <- convertX(unit(1935,"native"), "npc")
+X3 <- convertX(unit(2215,"native"), "npc")
+X4 <- convertX(unit(2350,"native"), "npc")
+#For some reason grid.text for peak labels does not always align with the x value for the dashed line, so need to add a specific value for the text
+x1 <- convertX(unit(1436,"native"), "npc")
+x2 <- convertX(unit(1936,"native"), "npc")
+x3 <- convertX(unit(2212,"native"), "npc")
+x4 <- convertX(unit(2335,"native"), "npc")
+popViewport(3)
+
+grid.move.to(x = X, y = unit(1, "npc"), vp = vps1)
+grid.line.to(x = X, y = unit(0, "npc"), vp = vps4, 
+             gp = gpar(col = "gray", lty = 2, lwd = 2))
+grid.text("OH", x = x1, y = unit(0.87, "npc"), just="center", gp=gpar(font=2))
+
+grid.move.to(x = X2, y = unit(1, "npc"), vp = vps1)
+grid.line.to(x = X2, y = unit(0, "npc"), vp = vps4, 
+             gp = gpar(col = "gray", lty = 2, lwd = 2))
+grid.text("OH", x = x2, y = unit(0.93, "npc"), just="center", gp=gpar(font=2))
+
+grid.move.to(x = X3, y = unit(1, "npc"), vp = vps1)
+grid.line.to(x = X3, y = unit(0, "npc"), vp = vps4, 
+             gp = gpar(col = "gray", lty = 2, lwd = 2))
+grid.text("AlOH", x = x3, y = unit(0.95, "npc"), just="center", gp=gpar(font=2))
+
+grid.move.to(x = X4, y = unit(1, "npc"), vp = vps1)
+grid.line.to(x = X4, y = unit(0, "npc"), vp = vps4, 
+             gp = gpar(col = "gray", lty = 2, lwd = 2))
+grid.text("CO3", x = x4, y = unit(0.95, "npc"), just="center", gp=gpar(font=2))
 ```
@@
 \@ref(fig:raman-spectra)).
-The first are three peaks in the 100 cm-1 - 500 cm-1 band that are typical for SiO~2~ and quartz-based materials, with the strongest occuring at 460 cm-1 [@gillet_etal90].
+The first are three peaks in the 100 cm-1 - 500 cm-1 band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm-1 [@gillet_etal90].
 These peaks are present for all samples in the dataset.
@@
 These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
-The D band occurs as a result of disordered C structures whilst the G band is indicative of sp^2^ C networks and caused by stretching in the C bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
+The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
 Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm-1 region.
@@
 r.translucent <- Points.raman %>% 
-  filter(hue == "Translucent")
+  filter(hue == "Colourless")
 
@@
 plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm-1)", main = "Average - Translucent")
+     xlab = "Wavenumber (cm-1)", main = "Average - Colourless")
 
@@
 \linebreak
-
-<!--```{r raman-average, fig.height=15, fig.fullwidth=TRUE, fig.cap="Average spectra of raman measurements sampled on grey-to-dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
-knitr::include_graphics('C:/Users/masj0062/Documents/DOKTORANDSPROJEKT/VERSION CONTROL/MethodPaper/analysis/figures/fig_4_4_raman_spectra_dark.jpg')
-```-->
 
@@
 ```
-
-<!--```{r raman-pca1, fig.fullheight=TRUE, fig.fullwidth=TRUE, fig.cap="PCA score plots for the raman analysis.
-Along the 1st component (P1) materials group based on two main features, the quartz-peak at 460 cm-1 and the D/G bands around 1300 cm-1 and 1600 cm-1.
-Darker materials feature more pronounced D/G band peaks while the white and more translucent materials feature only the quartz peak A) Plot of 1st and 2nd components; noisier spectra drift outwards along P2 B) Plot of 1st and 3rd components; noisier spectra are separated out from other samples, but spectra with peaks around 700-800 cm-1 are drawn towards the bottom right."}
-knitr::include_graphics('./analysis/figures/fig_4_6_raman_average_pca_p1_p2_p3.jpg')
-```-->
 
@@
 
-#perform pca with mean centering and unite variance scaling
+#perform pca with mean centering and unit variance scaling
 xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=TRUE)
@@
 
-<!--```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
-knitr::include_graphics('./analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
-```-->
 ```{r xrf-pca1, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 2nd components.

### 2021-11-01 analysis/paper/paper.Rmd
Developed discussion, changed "translucent" to "colourless", and other slight updates to text

@@
   fig.path = "../figures/",
-  dpi = 600
+  dpi = 300
 )
@@
 
-#Fill the NA fields in the munsell_hue column and mark them as translucent samples
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
 Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
@@
 It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature.
-In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit* , the latter possibly a spelling mistake], and brecciated quartzite.
+In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz and brecciated quartzite^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit*].
 Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
@@
 The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
-The spectral coverage of the instrumentation is 175 cm-1 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1.
+The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^.
 Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
@@
 Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
-Two methods for imputation can be implemented, iterative or regularized, since the former suffers when there are a large number of missing values the latter was used.
+Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.
 
@@
 In the PCA (fig.
-\@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/translucent/white material to the left.
-Examining the average spectra of these groups (fig.
-\@ref(fig:vis-nir-spectra)), as expected, the darker material feature higher absorbance of light than the other groups.
+\@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/colourless/white material to the left.
+Examining the average spectra of these groups, as expected, the darker material feature higher absorbance of light than the other groups.
 No other distinctive features can be seen in the visual band, that could potentially be used to distinguish the groups.
 It is, however, apparent that while the more extreme hues of *dark* and *white* material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.
-
-A number of artefacts also breaks the observed trend in the PCA, with dark/black (N1-N2) samples ending up closer to the light and translucent group in the PCA.
+There are also a number of artefacts with a dark hue that gravitate closer to the lighter material in the PCA.
+While the small size of these artefacts could be an explanation for the placement of these within the model, there are plenty of similarly sized artefacts that do not follow the same trend.
+
 There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA.
@@
 This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering.
-The *greyish* middle group in the PCA is otherwise largely made up of *quartz coloured* opaque material with a beige/grey tint.
-
-\linebreak
-
-```{r vis-nir-pca, fig.width=20, fig.height=8, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum.
+The *greyish* middle group in the PCA is otherwise largely made up of opaque material with a beige/grey tint.
+
+\linebreak
+
+```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum.
 Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
@@
   fviz_pca_ind(visnir.pca, axes = c(1,2), label="none")
+basic_plot2 <-
+  fviz_pca_ind(visnir.pca, axes = c(1,3), label="none")
 
@@
 vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
+vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=1)
 vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
 vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
+vis.pc3lab <- paste0("PC3 (",as.character(vis.pc3var),"%)")
+
+#VISNIR LOADING LINES
+#par(mfrow=c(3,1))
+#sapply(1:3, function(comp){
+#  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
+#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,6))})
 
 #bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
+visnir.p1 <-
   ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
@@
                              title.hjust = 0.5)) +
-  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
+  theme(legend.position = "none")
+
+visnir.p2 <-
+  ggplot(cbind(basic_plot2$data, Points.nir[, c(8,10)]),
+             aes(x=x, y=y, 
+                 shape = material, 
+                 fill = munsell_hue)) +
+  geom_point(size=3) +
+  theme_bw() +
+  ggtitle("Near infrared 400 - 750 nm") +
+  labs(x = vis.pc1lab,
+       y = vis.pc3lab) +
+  scale_shape_manual(name = "Material",
+                     values=pca.shapes) +
+  scale_fill_manual(name = "Munsell hue",
+                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
+                    values=pca.colors) +
+  guides(shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol=2, 
+                              title.position="top", 
+                              title.hjust = 0.5),
+         fill = guide_legend(override.aes = list(shape = 22,
+                                                 fill = pca.colors,
+                                                 color = "black",
+                                                 size=3),
+                             title.position="top", 
+                             title.hjust = 0.5)) +
+  theme(legend.position = "none")
+
+comb1 <-
+  visnir.p1 + visnir.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
                               axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
@@
                               legend.position = "bottom")
+
+comb1 +
+  plot_layout(nrow = 2, guides = "collect")
 ```
@@
 
-vis.translucent <- Points.nir %>% 
+vis.colourless <- Points.nir %>% 
   filter(hue == "Colourless") %>% 
@@
 spec.vt <- 
-  as_spectra(vis.translucent, name_idx = 1, meta_idxs = 2:28)
+  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)
 
@@
 \@ref(fig:nir-pca)).
-Darker quartzite cluster to the right with white and translucent samples grouping to the opposite side.
+Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side.
 While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left.
@@
 ```{r nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue.
-Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/translucent samples."}
+Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/colourless samples."}
 
@@
   filter(hue == "Dark") %>% 
-  select(c(1:28),c(679:2179))
-
-#create a spectra object for the dark material, and then repeat for light, white and translucent groups
+  select(c(1:28),c(680:2179))
+
+#create a spectra object for the dark material, and then repeat for light, white and colourless groups
 spec.d <- 
@@
   filter(hue == "Light") %>% 
-  select(c(1:28),c(679:2179))
+  select(c(1:28),c(680:2179))
 
@@
 
-translucent <- Points.nir %>% 
+colourless <- Points.nir %>% 
   filter(hue == "Colourless") %>% 
-  select(c(1:28),c(679:2179))
+  select(c(1:28),c(680:2179))
 
 spec.t <- 
-  as_spectra(translucent, name_idx = 1, meta_idxs = 2:28)
+  as_spectra(colourless, name_idx = 1, meta_idxs = 2:28)
 
@@
   filter(hue == "White") %>% 
-  select(c(1:28),c(679:2179))
+  select(c(1:28),c(680:2179))
 
@@
 \@ref(fig:raman-spectra)).
-The first are three peaks in the 100 cm-1 - 500 cm-1 band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm-1 [@gillet_etal90].
+The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90].
 These peaks are present for all samples in the dataset.
-The second feature shows up as two peaks occurring at around 1 287 cm-1 and 1 598 cm-1, and dominate the spectra in terms of intensity for the darker samples.
+The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples.
 These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
 The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
-Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm-1 region.
+Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
 It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280].
@@
 A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca.
-695 cm-1 and 795 cm-1.
+695 cm^-1^ and 795 cm^-1^.
 Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale.
-Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm-1 band which has been related to Al-O-Si vibrations for the Kaolin group.
+Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band which has been related to Al-O-Si vibrations for the Kaolin group.
 Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
@@
 PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
-Interestingly, the loading for PC 4 does catch the Al-O-Si peaks around the  700 cm-1 - 800 cm-1 band, which helps in separating out those samples where this feature is present.
+Interestingly, the loading for PC 4 does catch the Al-O-Si peaks around the  700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
 The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side.
-It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more translucent objects mainly feature quartz peaks.
+It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks.
 As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
@@
 ```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue.
-The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the dark group"}
+The *quartz peak* at 460 cm^-1^ is present in all four groups, but the *graphite peaks* at 1 287  cm^-1^ (D band) and 1 598 cm^-1^ (G band) mainly appear in the dark group"}
 
@@
 
-r.translucent <- Points.raman %>% 
+r.colourless <- Points.raman %>% 
   filter(hue == "Colourless")
@@
 spec.rt <- 
-  as_spectra(r.translucent, name_idx = 1, meta_idxs = 2:29)
+  as_spectra(r.colourless, name_idx = 1, meta_idxs = 2:29)
 
@@
 plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm-1)", main = "Average - Dark")
+     xlab = "Wavenumber (cm^-1^)", main = "Average - Dark")
 text(x=120, y=2900, labels="SiO", font=2, cex = 1.5)
@@
 plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm-1)", main = "Average - Light")
+     xlab = "Wavenumber (cm^-1^)", main = "Average - Light")
 
 plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm-1)", main = "Average - Colourless")
+     xlab = "Wavenumber (cm^-1^)", main = "Average - Colourless")
 
 plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm-1)", main = "Average - White")
+     xlab = "Wavenumber (cm^-1^)", main = "Average - White")
 ```
@@
 As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA.
-The 3^rd^ component does, however, also separate out the few samples that feature peaks at 700-800 cm-1 that may result from Al, Si and OH vibrations."}
+The 3^rd^ component does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ that may result from Al, Si and OH vibrations."}
 
@@
 
-The three first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 84.7%.
-The loadings of the three components can be seen in figure \@ref(fig:xrf-loadings).
-In the score plot of PC 1 and PC 2, two large clusters can be seen (fig.
+The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 81.3%.
+The loadings can be be seen in figure \@ref(fig:xrf-loadings).
+In the score plot of PC 1 and PC 2 two large clusters can be seen (fig.
 \@ref(fig:xrf-score-plot-p1-p2)).
 Interestingly, the smaller cluster on the left side in the model includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content.
-It is not ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that drives this.
+It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr.
 The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom.
-This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite ([@howard05] p.711).
-The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up towards closer to the “high ^14^Si” cluster.
-The white (N9) material is somewhat evenly distributed within the right cluster.
+This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711].
+The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster.
+The white (N9) material is fairly evenly distributed within the right cluster.
 
@@
 
-One almost translucent sample also ends up in the "high ^14^Si" cluster among the darker material.
+One of the *colourless* samples also ends up in the "high ^14^Si" cluster among the darker material.
 Upon closer inspection this sample is small, ~ 3 by 1 mm, and also feature inclusions of dark particles.
@@
 Points_long <-
-  Points.xrf %>%
+  drop_na(Points.xrf %>%
     select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
-  gather(key = "Element", value = "Value")
+    gather(key = "Element", value = "Value"))
 
@@
 
+#Calculate and add value equal to 95% of the value range for each element, this is in preparation for 
+#text annotations in final plot where a text will be positioned at this values height in each plot
+Points_long <- 
+  Points_long %>% 
+  group_by(Element) %>% 
+    mutate(ypos = min(Value)+0.95*diff(range(Value)))
+
 #Prepare colors
@@
 
-#Violin plot of elemental content, trimmed, with fill based on elemental groups
+#Prepare data frame with elements for annotating facets in final plot
+na_values <- 
+  data.frame("Element" = unique(Points_long$Element))
+
+#Calculate and add number of missing values for each element             
+na_values <- 
+  na_values %>% 
+    mutate(na_values = c(
+      paste("n = ", length(which(!is.na(Points.xrf$`12Mg`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`13Al`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`14Si`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`15P`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`16S`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`19K`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`20Ca`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`23V`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`25Mn`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`26Fe`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`30Zn`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`38Sr`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`39Y`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`40Zr`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`56Ba`)))),
+      paste("n = ", length(which(!is.na(Points.xrf$`81Ti`)))))
+    )
+
+#Merge na_values and Points_long data frames in order to get the y position that the text annotation should be displayed
+#in for each element facet in final plot
+na_values <- 
+  merge(na_values, unique(Points_long[,c("Element", "ypos")]), by = "Element", all.x = TRUE)
+
+#Violin plot of elemental content, trimmed, with fill based on elemental groups and number of values under LOD as text annotation
 Points_long %>%
-  drop_na(Value) %>%
     ggplot(aes(x=Element, y=Value, fill = Group)) +
@@
     theme_bw() +
+    geom_text(data=na_values, aes(x=0.2, y=ypos, label=na_values, fontface = "bold"), nudge_x = 0.4, inherit.aes = FALSE) +
     theme(
@@
 
-```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
+```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each loading (element)"}
 #select elements to include in pca, as well as columns to group by
@@
 #Extract loadings from xrf.pca
-loadings <- as.data.frame(xrf.pca$rotation[,1:3])
+loadings <- as.data.frame(xrf.pca$rotation[,1:2])
 #As elements are stored as rownames, set them as first column so it can be used as a variable
@@
 
-#bar graph of the loadings (P1-P3) for the XRF data
+#bar graph of the loadings (P1-P2) for the XRF data
 loadings %>%
@@
 
-#Prepare two score plots
+#XRF score plot
 ggbiplot(xrf.pca,
@@
 \linebreak
-
+<!--
 ```{r xrf-pca2, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 3rd components.
@@
 ```
-
-\linebreak
+-->
 
@@
 
-The most easily recognizable megascopic feature of petrographic material is its colour, a feature that has been one of the main means of grouping quartzite in North Swedish archaeological research (e.g.
-@forsberg85 p.60-64, @kaeck09 p.
-53-55).
+The most easily recognizable megascopic feature of petrographic material is its colour, a feature that has been one of the main means of grouping quartzite in North Swedish archaeological research [e.g.
+@forsberg85 p.60-64; @kaeck09 p.
+53-55].
 While it is true that some inference can be made based on colour, such as ^26^Fe-bearing minerals contributing to a darker colour, this becomes increasingly difficult with heterogeneous material.
@@
 The results above echoes this complexity.
-While the overall structure in the spectroscopic results for each of the instrumentations, and especially the NIRS, groups samples of similar colour (given as *munsell hue*) together, there are samples that break these trends and thus demonstrate the presence of structural and compositional features that do not necessarily align with the perceived colour of a material.
-
-In the case of the raman data the PCA is somewhat lacking in exploratory and descriptive power.
-Apart from the few samples that feature the 700 cm-1 - 800 cm-1 band peaks the major features in the data are all explained by the first component.
-As the intensity in raman spectroscopy is not a direct quantitative indicator this makes it difficult to group materials based on the intensity of the D/G bands, making more of a binary separation of presence/absence of the *graphite peaks*.
-That is not to say that it is not possible to attempt quantification using raman spectroscopy, but rather that this would demand a different experimental setup.
-That said, the presence of the *graphite peaks* may represent an important clue in determining the potential geological origin for the material.
-The presence of the C structures suggest the petrogenesis included sediments enriched with organic material, providing a good filter for continued research.
+While the overall structure in the spectroscopic results for each of the instrumentations, and especially the NIRS, groups samples of similar colour (given as *Munsell hue*) together, there are samples that break these trends and thus demonstrate the presence of structural and compositional features that do not necessarily align with the perceived colour of a material.
+
+Rather than assigning color based on how it is perceived visually it can be approached in a more systematic manner [e.g. @fairchild13].
+
+In the case of the Raman data the PCA is somewhat lacking in exploratory and descriptive power.
+Apart from the few samples that feature the 700 cm^-1^ - 800 cm^-1^ band peaks the major features in the data are all explained by the first component.
+As the intensity in Raman spectroscopy is not a direct quantitative indicator this makes it difficult to group materials based on the intensity of the D/G bands, making more of a binary separation of presence/absence of the *graphite peaks*.
+That is not to say that it is not possible to attempt quantification using Raman spectroscopy, but rather that this would necessitate a different experimental setup.
+That said, the presence of the *graphite peaks* may present an important clue in determining the potential geological origin for the material.
+Mentions of *graphitic quartzite* can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@motet25], but otherwise does not seem to be a frequently used term in the modern day.
+
+The Raman data is also a good example of the potential value of PCA components with a low degree of explained variance.
+While the typical statistical approach is to disregard components with an eigenvalue below 1, in the case of spectral data there may still be meaningful features explained by those very same components.
+The eigenvalue of the 4^th^ component in the Raman PCA is below 1, but amidst the noise there were still meaningful features that may be used for characterization.
 
@@
 As the material does group with other similarly semi-translucent material it does not seem to distinguish itself enough in the models in order to consider it its own group.
-The original classification by @ahman67 refers to the microscopic fragmented configuration of the quartz crystals as being the main distinguishing feature, and while it is possible to discern the presence of SiO structure through non-destructive spectroscopy, there have been no other discernible features related to this group.
-
-Keeping in mind the
+The original classification by @ahman67 refers to the microscopic fragmented configuration of the quartz crystals as being the main distinguishing feature, and while it is possible to discern the presence of SiO structure through non-destructive spectroscopy, there have been no other discernible features related to this group at this stage.
+The question remains whether the material classified as *brecciated quartzite*
 
@@
 The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint based solely on surface measurements.
-Nevertheless they do provide a more in-depth understanding of the composition and structure of the material, and combining these results may be the key to more robust non-destructive characterization of petrographic material.
+Nevertheless they do provide a more in-depth understanding of the composition and structure of the material as well as the complexity which exists in quartz and quartzite materials.
+The study represents a good example of an exploratory approach in the characterization of petrographic material, where there is an absence of reference material.
+With a large enough dataset exploratory multivariate statistics can be used in order to discern groupings based on the spectral features, taking advantage of the short integration
+
+and combining these results may be the key to more robust non-destructive characterization of petrographic material.
 
@@
 # Acknowledgements
+
+I would like to thank my supervisors, Philip I. Buckland and Johan Linderholm, for valuable comments on previous versions of this paper.
 

### 2021-11-09 analysis/paper/paper.Rmd
Overhauled text, code changes and more

@@
 
-Attempting to define what constitutes a *quartzite* is not so stratightforward.
 Prieto et al.
 [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
-The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and it hence varies somewhat between authors.
-Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies.
-There are also situations where destructive techniques simply are not applicable.
-Artefact collections (museum, private, etc.) represent a valuable resource, as well as one of the easier ways of gaining access to large datasets with some geographical distribution.
-While destructive methods may prohibit access to such material there are still non-destructive methods that can generate valuable information.
-Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive alternative for field-based screening purposes.
-While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
+It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05].
+Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
+In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable.
+Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.
+While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information.
+Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive option for field-based screening purposes.
+While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
 The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
@@
 
-In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
-The goal is to evaluate their capabilities for fingerprinting quartz and quartzite in the North Swedish region.
-This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to the spectral features for each instrumentation.
+In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
+The aim is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting.
+This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features.
  \linebreak
@@
 
-The current study is based on spectroscopic surface measurements on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden.
-A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy (Appendix XX).
-The material features a mainly black (N1-N3), grey or white (N9) hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
-The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *white* ~ N9, *Colourless* ~ lack of colour, opaque to translucent) for visualization purposes.
+The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden.
+A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy.
+The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
+The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).
 A number of these points have, according to prior classifications, been worked using bifacial technology.
@@
 Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online.
-Based on a query of all points labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database.
-Sampling of this pool of artefacts was not random, and based on geographical distribution, in an effort to include as many artefacts as possible from areas with few finds available, such as along the Vindel river.
+Based on a query of all points and preforms labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database.
+Sampling of this list of artefacts was not random.
+Sites located in areas with few finds were included first in order to increase the geographical distribution, with sampling of areas with higher site density making up the remaining bulk of the dataset.
 None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included.
-This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz/quartzite material.
-
-\linebreak
-
-```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates.
+This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz or quartzite material.
+
+\linebreak
+
+```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of sampled sites within Västerbotten County, Sweden.
 The distribution of sampled sites is likely affected by the excavation and survey history of the region.
+White: Sampled sites; Black: Parish centroid for measured samples lacking site coordinates.
 Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
@@
 Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
-The reason for including *quartz* in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.
-
-Another reason for keeping a *quartz* category is the group referred to as *brecciated quartz*.
-In a petrographic study of the artefacts recovered along the Ume river valley Åhman (1967:8) briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*.
+The reason for including quartz in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.
+
+Another reason for keeping a quartz category is the group referred to as *brecciated quartz*.
+In a petrographic study of the artefacts recovered along the Ume river valley Åhman [-@ahman67 p.8] briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*.
 Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure.
@@
 *Brecciated quartz* is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
-@holm91].
+@holm91 p.24].
 
@@
  \linebreak
-All measurements were performed at the Environmental Archaeology Lab at Umeå University.
+
+All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/forskning/infrastruktur/mal/].
 The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens.
-Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult.
+Each artefact was measured twice, on different sides, with additional measurements in cases where sampling was difficult.
 The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
@@
 After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20].
-PCA modelling was performed using the base R *prcomp* function [@Rteam17; @rcoreteam21].
+PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].
 
@@
 NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe.
-The ASD features a detection range of 350 nm – 2500 nm, with spectral sampling (resp.
+The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp.
 spectral resolution) of 1.4 nm (resp.
@@
 Each object was measured against a white reference background and the collected measurements were averaged for each artefact.
-Two parts of the As there were no notable features in the visible range the analysis was restricted to the 1 000 nm – 2 500 nm region in order to lessen the impact of noise.
+The visible spectrum (~ 400 nm - 750 nm) and the NIR range (~ 1 000 nm – 2 500 nm) were analysed separately, with the former focused on colour analysis and the latter the chemical structure of the material.
 The data was mean-centred before PCA modelling.
@@
 Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01].
-However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect.
+However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect.
 By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
@@
 The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^.
-Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.
+Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå^[www.slu.se/en/departments/forest-biomaterials-technology/].
 
@@
 ## X-Ray Fluourescence
+XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method for petrographic analysis in archaeology [@hall60; @shackley11a].
+Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e.
+*fluourescence*).
+As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can tested statistically [@tykot03].
+
 XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand.
@@
 Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions.
+
 The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling.
-As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA.
+As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA.
 Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
@@
 ## NIR - visual range
-
-The main explanatory power for the visual region can be found in the first component with an explained variance of 98.4 %.
-In the PCA (fig.
-\@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/colourless/white material to the left.
-Examining the average spectra of these groups, as expected, the darker material feature higher absorbance of light than the other groups.
-No other distinctive features can be seen in the visual band, that could potentially be used to distinguish the groups.
-It is, however, apparent that while the more extreme hues of *dark* and *white* material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.
-There are also a number of artefacts with a dark hue that gravitate closer to the lighter material in the PCA.
-While the small size of these artefacts could be an explanation for the placement of these within the model, there are plenty of similarly sized artefacts that do not follow the same trend.
-
-There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA.
-In examination of the non-averaged data artefact 403 and 502 both have a sample taken on one side in the *dark* group and the other in the *greyish* middle.
-This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering.
-The *greyish* middle group in the PCA is otherwise largely made up of opaque material with a beige/grey tint.
-
-\linebreak
-
-```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum.
-Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
-The strong trend along PC1 also makes it difficult to separate out the muddled grayscale in the middle."}
+Examining the average spectra of the different colour groups (dark, light, white and colourless), as expected, the darker material feature higher absorbance of light than the other groups.
+No other distinctive features can be seen in the visual band, that could potentially be used to distinguish them.
+It is, however, apparent that while the more extreme hues of dark and white material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.
+
+The cumulative explained variance for the first three components in the PCA is ~ 99.9%, with the accompanying loadings in fig.
+\@ref(fig:vis-nir-loadings).
+Most of the contribution to the first Principal Component (PC) can be found within the red part of the spectrum (~ 600 nm - 750 nm).
+This can be seen in the score plot where darker and lighter materials are separated along PC 1, as darker samples would feature a higher absorbance within the red part of the spectrum (fig.
+\@ref(fig:vis-nir-pca)).
+In contrast, most of the contribution to PC 2 occurs within the blue-violet part of the spectrum, which does not seem to alter the structure of the dataset in any significant way.
+PC 3 feature high contributions within both the blue-violet and red parts of the spectrum, which also seems to draw the light and dark samples that feature a blue-gray tint towards the top of the score plot.
+
+While the overall structure of the model seems to group colourless and white materials towards the bottom left, with dark material to the right, there are a number of samples that break this pattern.
+This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light.
+It is also difficult to separate out the greyscale of colours that occur inbetween the "black and white" groups, where both light gray and semi-translucent samples end up.
+All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.
+
+\linebreak
+
+```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data.
+Top: PC 1 and PC 2, bottom: PC 1 and PC 3 PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure.
+PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}
 
@@
 #prepare labels for PCs
-vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
-vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
-vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=1)
+vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=2)
+vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=2)
+vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=2)
 vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
@@
 
-```{r vis-nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue. Darker samples feature higher overall absorbance with white material absorbing the least amount of light."}
+```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3.
+The two first components mainly describes features with a linear increase in absorbance across the visual spectrum, but the 3rd component demonstrates more variation, with lower absorbance within the blue-yellow region.
+"}
 
@@
 #and select the NIR range 1 000 - 2 500 nm
-vis.dark <- Points.nir %>% 
-  filter(hue == "Dark") %>% 
-  select(c(1:28),c(79:429))
-
-spec.vd <- 
-      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)
-
-vis.light <- Points.nir %>% 
-  filter(hue == "Light") %>% 
-  select(c(1:28),c(79:429))
-
-spec.vl <- 
-  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)
-
-vis.colourless <- Points.nir %>% 
-  filter(hue == "Colourless") %>% 
-  select(c(1:28),c(79:429))
-
-spec.vt <- 
-  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)
-
-vis.white <- Points.nir %>% 
-  filter(hue == "White") %>% 
-  select(c(1:28),c(79:429))
-
-spec.vw <- 
-  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)
-
-par(mfrow=c(4,1))
-
-plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
-
-plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
-
-plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")
-
-plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
+#vis.dark <- Points.nir %>% 
+#  filter(hue == "Dark") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vd <- 
+#      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)
+
+#vis.light <- Points.nir %>% 
+#  filter(hue == "Light") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vl <- 
+#  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)
+
+#vis.colourless <- Points.nir %>% 
+#  filter(hue == "Colourless") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vt <- 
+#  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)
+
+#vis.white <- Points.nir %>% 
+#  filter(hue == "White") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vw <- 
+#  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)
+
+#VISNIR LOADING LINES
+#For some reason invisible() has to be used to prevent console output showing up in knitted document
+par(mfrow=c(3,1))
+invisible(sapply(1:3, function(comp){
+  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
+      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,5))}))
+
+#par(mfrow=c(4,1))
+
+#plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
+
+#plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
+
+#plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")
+
+#plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
 ```
@@
 ## NIR - near infrared range
-In the case of NIRS similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks.
-The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig.
+In the case of the 1 000 nm - 2 500 nm band of the NIR data, similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks.
+The presence of molecular water can be inferred based on the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig.
 \@ref(fig:nir-spectra)).
-Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falls within one of the stronger CO^3^ bands [@clark_etal90].
-Inspection of individual spectra for the *light* group, however, also reveals a 5th peak for some samples at 2 450 nm.
-Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
-It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH.
+Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90].
+Inspection of individual spectra for the light group, however, also reveals a 5th peak for some samples at 2 450 nm.
+Comparisons to spectra collected on muscovite [@clark_etal90 p.12 675] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
+It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.
 
@@
 Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side.
-While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left.
-A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points.
+While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the brecciated quartz, seem to cluster towards the top-left.
+At the top of the score plot 3 black samples separate themselves clearly from the rest of the data points.
 These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
@@
 ```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum.
-Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant.
+Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left.
 A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen.
@@
 #prepare labels for PCs
-nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
-nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
+nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=2)
+nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=2)
 nir.pc1lab <- paste0("PC1 (",as.character(nir.pc1var),"%)")
@@
 The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples.
-These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99].
-The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21].
+These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99].
+The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21].
 Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
@@
 
-A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ca.
-695 cm^-1^ and 795 cm^-1^.
+A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ~ 695 cm^-1^ and 795 cm^-1^.
 Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale.
-Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band which has been related to Al-O-Si vibrations for the Kaolin group.
+Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band.
 Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
@@
 The resulting PCA model (fig.
-\@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D/G peaks.
+\@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D - G peaks.
 PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
-Interestingly, the loading for PC 4 does catch the Al-O-Si peaks around the  700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
+Interestingly, the loading for PC 4 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
 The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side.
@@
 As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the *quartz* group.
-
-\linebreak
-
-```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue.
+When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
+
+\linebreak
+
+```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Average Raman spectra grouped by their hue.
 The *quartz peak* at 460 cm^-1^ is present in all four groups, but the *graphite peaks* at 1 287  cm^-1^ (D band) and 1 598 cm^-1^ (G band) mainly appear in the dark group"}
@@
      xlab = "Wavenumber (cm^-1^)", main = "Average - Dark")
-text(x=120, y=2900, labels="SiO", font=2, cex = 1.5)
-text(x=200, y=4300, labels="SiO", font=2, cex = 1.5)
-text(x=460, y=9250, labels="SiO", font=2, cex = 1.5)
+text(x=120, y=2900, labels="Quartz", font=2, cex = 1.5)
+text(x=200, y=4300, labels="Quartz", font=2, cex = 1.5)
+text(x=460, y=9250, labels="Quartz", font=2, cex = 1.5)
 text(x=1285, y=6300, labels="D", font=2, cex = 2.0)
@@
 ```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra.
-The main explanatory power can be found in the 1^st^ component, which groups the data based on the quartz and D/G bands.
+The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands.
 As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA.
-The 3^rd^ component does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ that may result from Al, Si and OH vibrations."}
+PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the bottom left, that may result from Al, Si and OH vibrations."}
 
@@
 ## XRF
-The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal), which represents the remaining unquantifiable elements in a sample.
+Four samples had dimensions that were < 10 mm in size of their smallest dimension, and were therefore excluded from the analysis as there may be complications when comparing these to larger samples [@shackley11b pp.9-10].
+The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal) that represents the remaining unquantifiable elements in a sample.
 The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF).
@@
 \@ref(fig:xrf-violin-plot)).
-The majority of the readings resulted in a SiO^2^ content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO^2^ content closer to 90 %.
-The “low” SiO^2^ group also feature a Bal between 40 – 70 %.
+The majority of the readings resulted in a SiO^2^ content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO^2^ content closer to 90%.
+The “low” SiO^2^ group also feature a Bal between 40 – 70%.
 
@@
 In the score plot of PC 1 and PC 2 two large clusters can be seen (fig.
-\@ref(fig:xrf-score-plot-p1-p2)).
-Interestingly, the smaller cluster on the left side in the model includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content.
-It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr.
+\@ref(fig:xrf-score-plot)).
+Interestingly, the smaller cluster on the left side in the score plot includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content.
+It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that are the main contributors.
 The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom.
 This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711].
-The *brecciated quartz* samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster.
+The brecciated quartz samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster.
 The white (N9) material is fairly evenly distributed within the right cluster.
@@
 
-One of the *colourless* samples also ends up in the "high ^14^Si" cluster among the darker material.
-Upon closer inspection this sample is small, ~ 3 by 1 mm, and also feature inclusions of dark particles.
-Both of these factors may have contributed to its placement in the model.
-
-\linebreak
-
-```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF.
+\linebreak
+
+```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of percentage element content in quartzite samples measured using pXRF.
 **n** represents no.
-of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements.
+of measurements above Limit of Detection (LOD) for that element out of a total of 474 measurements.
 Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
-#Prepare data in long format for violin plots
+#Prepare data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
 Points_long <-
   drop_na(Points.xrf %>%
+    filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
     select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
@@
 
-```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each loading (element)"}
+```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each variable (element)"}
+#filter on sample size (only include samples with smallest dimension >= 10mm), 
 #select elements to include in pca, as well as columns to group by
@@
   Points.xrf %>%
+  filter(max_length_mm >= 10 & max_width_mm >= 10) %>% 
   select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)
@@
 
-```{r xrf-pca1, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 2nd components.
+```{r xrf-score-plot, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of PC 1 and PC 2.
 The smaller group to the left in the plot includes samples with higher Si content, while the larger group includes samples with Si content < 50%."}
@@
 theme_minimal() +
-geom_point(aes(shape=Points.xrf$material, 
-               fill = Points.xrf$munsell_hue), 
+geom_point(aes(shape=xrf$material, 
+               fill = xrf$munsell_hue), 
            size = 3) +
@@
 53-55].
-While it is true that some inference can be made based on colour, such as ^26^Fe-bearing minerals contributing to a darker colour, this becomes increasingly difficult with heterogeneous material.
+While it is true that some inference can be made based on colour, such as ^26^Fe-bearing minerals contributing to a darker colour, this is significantly more difficult with heterogeneous material.
 Should a quartzite where the matrix is mainly light gray in colour, but streaked with black mineral veins, be considered a *light* or *dark* quartzite?
-The results above echoes this complexity.
-While the overall structure in the spectroscopic results for each of the instrumentations, and especially the NIRS, groups samples of similar colour (given as *Munsell hue*) together, there are samples that break these trends and thus demonstrate the presence of structural and compositional features that do not necessarily align with the perceived colour of a material.
-
-Rather than assigning color based on how it is perceived visually it can be approached in a more systematic manner [e.g. @fairchild13].
+The results above reflects this complexity.
+While the above models tend to group samples of a similar hue together, especially in the case of NIRS, there are samples that break these trends and thus demonstrate the presence of structural and compositional features that do not necessarily align with the perceived colour of a material.
+Rather than assigning color based on how it is perceived visually it can be approached in a more systematic manner [e.g.
+@fairchild13].
+Restricting the NIRS model to the visual range (~ 400 nm - 750 nm) opens up for some inference of color based on the absorbance of light within certain bands.
+However, as the data was collected using a contact probe the issue of how heterogeneous samples should be treated remains.
+This issue can be alleviated somewhat with a different sampling approach, for instance using a hyperspectral camera where different pixels can be selected and analysed separately [@sciuto19].
 
@@
 Apart from the few samples that feature the 700 cm^-1^ - 800 cm^-1^ band peaks the major features in the data are all explained by the first component.
-As the intensity in Raman spectroscopy is not a direct quantitative indicator this makes it difficult to group materials based on the intensity of the D/G bands, making more of a binary separation of presence/absence of the *graphite peaks*.
+As the intensity in Raman spectroscopy is not a direct quantitative indicator this makes it difficult to group materials based on the intensity of the D and G bands, making more of a binary separation of samples with or without *graphite peaks*.
 That is not to say that it is not possible to attempt quantification using Raman spectroscopy, but rather that this would necessitate a different experimental setup.
-That said, the presence of the *graphite peaks* may present an important clue in determining the potential geological origin for the material.
-Mentions of *graphitic quartzite* can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@motet25], but otherwise does not seem to be a frequently used term in the modern day.
-
-The Raman data is also a good example of the potential value of PCA components with a low degree of explained variance.
-While the typical statistical approach is to disregard components with an eigenvalue below 1, in the case of spectral data there may still be meaningful features explained by those very same components.
-The eigenvalue of the 4^th^ component in the Raman PCA is below 1, but amidst the noise there were still meaningful features that may be used for characterization.
-
-Interestingly the *brecciated quartz* remains relatively grouped together in all of the models demonstrating some consistency with the original classification.
-As the material does group with other similarly semi-translucent material it does not seem to distinguish itself enough in the models in order to consider it its own group.
-The original classification by @ahman67 refers to the microscopic fragmented configuration of the quartz crystals as being the main distinguishing feature, and while it is possible to discern the presence of SiO structure through non-destructive spectroscopy, there have been no other discernible features related to this group at this stage.
-The question remains whether the material classified as *brecciated quartzite*
+That said, the presence of the *graphite peaks* may present an important piece of information in the identification of the potential geological origin of the material.
+Interestingly, mentions of *graphitic quartzite* can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
+
+The material described as *brecciated quartz* remains relatively grouped together in all of the models, demonstrating some consistency with the original classification.
+As the samples group with other similarly semi-translucent material, however, it does not seem to distinguish itself enough in the models in order to consider it its own group.
+This is perhaps not so strange as the original classification by @ahman67 refers to the microscopic fragmented configuration of the quartz crystals as being the main distinguishing feature, and while it is possible to discern the presence of SiO structure through non-destructive spectroscopy, there have been no other discernible features related to this group at this stage.
+It is questionable whether all of the samples in this group truly exhibit brecciation, given the context under which they were collected and classified, which further complicates analysis of this material.
+Nevertheless, the results of the analysis suggests that there is some in-group variation, regardless of whether the material exhibits breccia.
 
@@
 
-The results demonstrate clearly that all three instrumentations are capable of characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint based solely on surface measurements.
-Nevertheless they do provide a more in-depth understanding of the composition and structure of the material as well as the complexity which exists in quartz and quartzite materials.
-The study represents a good example of an exploratory approach in the characterization of petrographic material, where there is an absence of reference material.
-With a large enough dataset exploratory multivariate statistics can be used in order to discern groupings based on the spectral features, taking advantage of the short integration
-
-and combining these results may be the key to more robust non-destructive characterization of petrographic material.
+The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification.
+Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials.
+The study represents a good example of a multivariate exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
+There is a need for methodologies that can be applied in regions with a complex geological setting and vague distribution of potential geological sources.
+Taking full advantage of the short integration time of these instrumentations in order to build large comprehensive datasets it is possible to discern spectral fingerprints using these methodologies.
 
@@
 # References
+
+\linebreak
 

### 2021-12-06 analysis/paper/paper.Rmd
Updated xrf violin plot

@@
 library(ggplot2)
+library(ggh4x)
 library(ggsci)
@@
     mutate(Group = case_when(
-      endsWith(Element, "Mg") ~ "Alkaline Earth",
-      endsWith(Element, "Al") ~ "Basic metal",
-      endsWith(Element, "Si") ~ "Semimetal",
-      endsWith(Element, "P") ~ "Nonmetal",
-      endsWith(Element, "S") ~ "Nonmetal",
-      endsWith(Element, "K") ~ "Alkali Metal",
-      endsWith(Element, "Ca") ~ "Alkaline Earth",
-      endsWith(Element, "V") ~ "Transition Metal",
-      endsWith(Element, "Mn") ~ "Transition Metal",
-      endsWith(Element, "Fe") ~ "Transition Metal",
-      endsWith(Element, "Zn") ~ "Transition Metal",
-      endsWith(Element, "Sr") ~ "Alkaline Earth",
-      endsWith(Element, "Y") ~ "Transition Metal",
-      endsWith(Element, "Zr") ~ "Transition Metal",
-      endsWith(Element, "Ba") ~ "Alkaline Earth",
-      endsWith(Element, "Ti") ~ "Transition Metal",
+      endsWith(Element, "Mg") ~ "Lithophile",
+      endsWith(Element, "Al") ~ "Lithophile",
+      endsWith(Element, "Si") ~ "Lithophile",
+      endsWith(Element, "P") ~ "Lithophile",
+      endsWith(Element, "S") ~ "Chalcophile",
+      endsWith(Element, "K") ~ "Lithophile",
+      endsWith(Element, "Ca") ~ "Lithophile",
+      endsWith(Element, "V") ~ "Lithophile",
+      endsWith(Element, "Mn") ~ "Siderophile",
+      endsWith(Element, "Fe") ~ "Siderophile",
+      endsWith(Element, "Zn") ~ "Chalcophile",
+      endsWith(Element, "Sr") ~ "Lithophile",
+      endsWith(Element, "Y") ~ "Lithophile",
+      endsWith(Element, "Zr") ~ "Lithophile",
+      endsWith(Element, "Ba") ~ "Lithophile",
+      endsWith(Element, "Ti") ~ "Chalcophile",
     ))
 
-#Calculate and add value equal to 95% of the value range for each element, this is in preparation for 
+#Calculate and add value equal to 90% of the value range for each element, this is in preparation for 
 #text annotations in final plot where a text will be positioned at this values height in each plot
@@
     group_by(Element) %>% 
-    mutate(ypos = min(Value)+0.95*diff(range(Value)))
+    dplyr::mutate(ypos = quantile(Value, 0.95))
 
 #Prepare colors
-loading.colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")
+loading.colors <- c("lightgoldenrod1", "darkorange", "lightpink1")
 
@@
 
+#Prepare ylimits for each group in order to exclude extreme outliers in the violin plot graphic
+ylims <- Points_long %>%
+  group_by(Element) %>% 
+  dplyr::summarise(y1 = 0, y2 = quantile(Value, 0.95))
+
+#Prepare a list with formula for setting new ylimits for each group in ggh4x package "facetted_pos_scales() which makes it possible to work with facet_wrap()
+scales <- c(
+  Element == '12Mg' ~ scale_y_continuous(limits = c(as.vector(t(ylims[1,2:3])))),
+  Element == '13Al' ~ scale_y_continuous(limits = c(as.vector(t(ylims[2,2:3])))),
+  Element == '14Si' ~ scale_y_continuous(limits = c(as.vector(t(ylims[3,2:3])))),
+  Element == '15P' ~ scale_y_continuous(limits = c(as.vector(t(ylims[4,2:3])))),
+  Element == '16S' ~ scale_y_continuous(limits = c(as.vector(t(ylims[5,2:3])))),
+  Element == '19K' ~ scale_y_continuous(limits = c(as.vector(t(ylims[6,2:3])))),
+  Element == '20Ca' ~ scale_y_continuous(limits = c(as.vector(t(ylims[7,2:3])))),
+  Element == '23V' ~ scale_y_continuous(limits = c(as.vector(t(ylims[8,2:3])))),
+  Element == '25Mn' ~ scale_y_continuous(limits = c(as.vector(t(ylims[9,2:3])))),
+  Element == '26Fe' ~ scale_y_continuous(limits = c(as.vector(t(ylims[10,2:3])))),
+  Element == '30Zn' ~ scale_y_continuous(limits = c(as.vector(t(ylims[11,2:3])))),
+  Element == '38Sr' ~ scale_y_continuous(limits = c(as.vector(t(ylims[12,2:3])))),
+  Element == '39Y' ~ scale_y_continuous(limits = c(as.vector(t(ylims[13,2:3])))),
+  Element == '40Zr' ~ scale_y_continuous(limits = c(as.vector(t(ylims[14,2:3])))),
+  Element == '56Ba' ~ scale_y_continuous(limits = c(as.vector(t(ylims[15,2:3])))),
+  Element == '81Ti' ~ scale_y_continuous(limits = c(as.vector(t(ylims[16,2:3]))))
+)
+
 #Violin plot of elemental content, trimmed, with fill based on elemental groups and number of values under LOD as text annotation
@@
     scale_fill_manual(values=loading.colors) +
+    facetted_pos_scales(y = scales) +
     theme_bw() +

### 2021-12-20 analysis/paper/paper.Rmd
Updated material section and placeholder visualization code if including references in models

@@
 nir.csv <-
-  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210914.csv", sep = ";",
-            dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210914.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
   raman.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz", 
+           type == "Point" | type == "Point fragment" | type == "Preform")
 
@@
 pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")
+
+#If including references
+#pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#DF5555", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#BFBFBF", "#DFDFDF", "#FFFFFF")
+#pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "2.5YR", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9")
 
@@
 
+# Study region
+
+\linebreak
+ \linebreak
+
+```{r geo-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geological map of Västerbotten County, Sweden, showing both the tectonic units of the mountains and the widely distributed moraine.
+Made with Lantmäteriet (CC0), Natural Earth, and Sveriges Geologiska Undersökning (SGU)"}
+knitr::include_graphics('../figures/geo_map2.png', error = FALSE)
+```
+
+\linebreak
+
+The effects of the last glacial on the landscape is visible in the widely distributed moraine formations and glaciofluvial deposits (fig.
+\@ref(fig:distribution-map)).
+In Västerbotten the Köli Nappe complex exhibits greensschist - amphibolite facies metamorphism, overlying the Seve Nappe complex where High Pressure (HP) and Ultra High Pressure (UHP) metamoprhism has been detected in a number of localities.
+
+In 1925 the Geological Society arranged an exursion to Södra Storfjället in Västerbotten.
+In the description of the regional stratigraphy it is noted that the lower parts are made up of, among others, "*...dark schists and graphitic quartzite...*" [@geologiska25 p.147].
+Apart from these quartzite with little impurities and Alum shale are also noted.
+
 # Material
@@
 A number of these points have, according to prior classifications, been worked using bifacial technology.
-While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12].
-Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite compared to previous periods [@baudou78].
+While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts[@baudou92 p.99; @forsberg85 p.5; @forsberg89].
+Taken together the research suggests a period of use that falls between ca.
+3 800 - 2 000 BP [@kaeck09 p.60].
+Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].
 

### 2021-12-22 analysis/paper/paper.Rmd
added Aims section

@@
 
+##Aims
+
 In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
-The aim is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting.
+The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting.
+The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use.
+As flint is a non-local material to Northern Sweden it follows that any find of this material within the region most likely have been brought there through anthropogenic activity.
+In contrast, determining the degree to which the distribution pattern of quartz and quartzite have been influenced by cultural and/or environmental factors however, is far more difficult.
+While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for comparison.
+The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes.
+Although models for settlement and mobility can be used to test ideas of material distribution as influenced by cultural processes [e.g.
+@forsberg85], it is also necessary to attempt to estimate the degree of environmental impact on this pattern.
+
 This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features.
- \linebreak
-
-# Study region
+
+\linebreak
+
+<!--# Study region
 
@@
 Apart from these quartzite with little impurities and Alum shale are also noted.
-
+-->
 # Material

### 2022-01-20 analysis/paper/paper.Rmd
Updated code and text

@@
 XRF.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
@@
 nir.csv <-
-  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210914.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/NIR_master_averaged_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 raman.csv <-
-  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_baseline_corrected_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 samples <-
-  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
 
@@
 
-In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] [@baudou78; @rankama_etal06; @olausson_etal12].
-The last decades have seen an increase in interest for non-flint lithics and raw material studies in  European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
+The last decades have seen an increase in the interest for non-flint lithics and raw material studies in European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
+In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] (e.g.
+quartz, quartzite, slate) [@baudou78; @rankama_etal06; @olausson_etal12].
 With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region.
@@
 
-##Aims
-
-In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF).
-The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting.
+\linebreak
+
+## Aims
+
+In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF).
+The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region based on surface measurements.
 The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use.
 As flint is a non-local material to Northern Sweden it follows that any find of this material within the region most likely have been brought there through anthropogenic activity.
-In contrast, determining the degree to which the distribution pattern of quartz and quartzite have been influenced by cultural and/or environmental factors however, is far more difficult.
-While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for comparison.
+In contrast, determining the degree to which cultural and/or environmental factors have influenced the distribution pattern of quartz and quartzite artefacts, however, is far more difficult.
+While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for spectral comparison.
 The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes.
@@
 
-This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features.
+In order to engage in any meaningful analysis, however, it is first necessary to distinguish between different quartz and quartzite groups.
+Principal Component Analysis (PCA) has been used to model the spectral data in order to discern material groupings related to different spectral features and their contributions to the Principal Components (PC).
+The results form the basis of a discussion of the viability of using non-destructive spectroscopy in petrographic characterization of quartz and quartzite material, and thus future efforts of developing models for raw material sourcing in Northern Sweden.
 
@@
 
-#Reorder factor levels
+#Reorder type factor levels
 Points2$type <-
   factor(Points2$type, levels=c("Preform", "Point fragment", "Point"))
+
+#Reorder parish factor levels
+Points2$parish <- 
+  factor(Points2$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
 ```
@@
 A number of these points have, according to prior classifications, been worked using bifacial technology.
-While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts[@baudou92 p.99; @forsberg85 p.5; @forsberg89].
-Taken together the research suggests a period of use that falls between ca.
+While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89].
+Previous research suggests a period of use that falls between ca.
 3 800 - 2 000 BP [@kaeck09 p.60].
@@
 
+\linebreak
+
 While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these.
@@
 PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].
-
-\linebreak
 
@@
 As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA.
-Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
+Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
 Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.
 
-\linebreak
+\pagebreak
 
@@
 
-```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data.
-Top: PC 1 and PC 2, bottom: PC 1 and PC 3 PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure.
-PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}
+```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3.
+The two first components mainly describes features with a linear increase in absorbance across the visual spectrum, but the 3rd component demonstrates more variation, with lower absorbance within the blue-yellow region.
+"}
 
@@
   prcomp(Points.nir[,c(78:428)], center = TRUE, scale = FALSE)
+
+#Filter NIR data to focus on the material with a dark hue 
+#and select the NIR range 1 000 - 2 500 nm
+#vis.dark <- Points.nir %>% 
+#  filter(hue == "Dark") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vd <- 
+#      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)
+
+#vis.light <- Points.nir %>% 
+#  filter(hue == "Light") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vl <- 
+#  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)
+
+#vis.colourless <- Points.nir %>% 
+#  filter(hue == "Colourless") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vt <- 
+#  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)
+
+#vis.white <- Points.nir %>% 
+#  filter(hue == "White") %>% 
+#  select(c(1:28),c(79:429))
+
+#spec.vw <- 
+#  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)
+
+#VISNIR LOADING LINES
+#For some reason invisible() has to be used to prevent console output showing up in knitted document
+par(mfrow=c(3,1))
+invisible(sapply(1:3, function(comp){
+  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
+      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,5))}))
+
+#par(mfrow=c(4,1))
+
+#plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
+
+#plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
+
+#plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")
+
+#plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
+```
+
+\linebreak
+
+```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data.
+Top: PC 1 and PC 2, bottom: PC 1 and PC 3.
+PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure.
+PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}
 
@@
   plot_layout(nrow = 2, guides = "collect")
-```
-
-\linebreak
-
-```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3.
-The two first components mainly describes features with a linear increase in absorbance across the visual spectrum, but the 3rd component demonstrates more variation, with lower absorbance within the blue-yellow region.
-"}
-
-#Filter NIR data to focus on the material with a dark hue 
-#and select the NIR range 1 000 - 2 500 nm
-#vis.dark <- Points.nir %>% 
-#  filter(hue == "Dark") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vd <- 
-#      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)
-
-#vis.light <- Points.nir %>% 
-#  filter(hue == "Light") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vl <- 
-#  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)
-
-#vis.colourless <- Points.nir %>% 
-#  filter(hue == "Colourless") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vt <- 
-#  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)
-
-#vis.white <- Points.nir %>% 
-#  filter(hue == "White") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vw <- 
-#  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)
-
-#VISNIR LOADING LINES
-#For some reason invisible() has to be used to prevent console output showing up in knitted document
-par(mfrow=c(3,1))
-invisible(sapply(1:3, function(comp){
-  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
-      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,5))}))
-
-#par(mfrow=c(4,1))
-
-#plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
-
-#plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
-
-#plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")
-
-#plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
 ```
@@
 Interestingly, the loading for PC 4 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
-The measurements with mainly quartz peaks gravitates towards the negative side of the PC 1, and those where the graphite peaks are dominant towards the opposite side.
+The measurements with mainly quartz peaks gravitates towards the negative side of PC 1, and those where the graphite peaks are dominant towards the opposite side.
 It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks.

### 2022-01-27 analysis/paper/paper2.Rmd
Version with SNV performed before detrending for raman

--- 2022-01-20 analysis/paper/paper.Rmd
@@
 
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
+
 #Import xrf data, set empty fields to NA
@@
 raman.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_baseline_corrected_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 
-Points.raman <-
-  raman.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-           type == "Point" | type == "Point fragment" | type == "Preform")
-
 #Fill the NA fields in the munsell_hue column and mark them as colourless samples
 Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
-Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+
 Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"
@@
 The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden.
-A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy.
+A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman scattering and XRF.
 The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
@@
 The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens.
-Each artefact was measured twice, on different sides, with additional measurements in cases where sampling was difficult.
+Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult.
 The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
@@
 
+```{r raman-prep-data}
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#perform Standard Normal Variate on raman data
+raman.averaged[,2:466] <- 
+  standardNormalVariate(raman.averaged[,2:466])
+
+#merge data with metadata
+#raman_merged <- 
+#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+    select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <-matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_spec_object")
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
+#spectra are stored in "Spectra" object, not a data frame
+#need to extract samples and frequencies from that object 
+raman.tmp <- 
+  cbind(raman.baseline$freq, t(raman.baseline$data))
+#transpose data to get frequencies as columns
+raman.tmp <- 
+  as.data.frame(t(raman.tmp))
+#Add sample_id column and remove the initial "X" character introduced to sample names by R
+raman.tmp$sample_id <- 
+  c("sample_id", sub("^X*", "", raman.baseline$names))
+
+#Make first row headers and remove the first row with the names
+names(raman.tmp) <-
+  raman.tmp[1,]
+raman.tmp <- 
+  raman.tmp[-1,]
+
+#merge raman de-trended data with metadata
+raman.baseline.merged <- 
+  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))
+
+#select and filter raman data to focus on relevant material and artefact types
+Points.raman <-
+  raman.baseline.merged %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
+Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+```
+
 ```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Average Raman spectra grouped by their hue.
@@
 PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the bottom left, that may result from Al, Si and OH vibrations."}
-
-#perform PCA with SNV normalization and mean-center
+#perform PCA with SNV normalization and mean-center on the 
 raman.pca <-
-  prcomp(standardNormalVariate(X = Points.raman[,c(30:321)]), center = TRUE, scale = FALSE)
+  prcomp(Points.raman[,c(30:321)], center = TRUE, scale = FALSE)
 
@@
 basic_plot2 <-
-  fviz_pca_ind(raman.pca, axes = c(1,4), label="none")
+  fviz_pca_ind(raman.pca, axes = c(2,3), label="none")
 
@@
 raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
-raman.pc4lab <- paste0("PC4 (",as.character(raman.pc4var),"%)")
+raman.pc4lab <- paste0("PC3 (",as.character(raman.pc4var),"%)")
 
@@
   plot_layout(nrow = 2, guides = "collect")
+
 ```

### 2022-01-27 analysis/paper/paper.Rmd
Code updated so that metadata is merged with data files. Pretreatment done in document, instead of outside of it.

@@
 
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
+
 #Import xrf data, set empty fields to NA
-XRF.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+xrf.csv <-
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/xrf_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
@@
 nir.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/NIR_master_averaged_20220119.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
   read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
-
-#Filter xrf, nir and raman data to focus on points and preforms made from quartz/quartzite material 
-Points.xrf <-
-  XRF.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-           type == "Point" | type == "Point fragment" | type == "Preform")
-
-Points.nir <-
-  nir.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-           type == "Point" | type == "Point fragment" | type == "Preform")
-
-Points.raman <-
-  raman.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-           type == "Point" | type == "Point fragment" | type == "Preform")
-
-#Fill the NA fields in the munsell_hue column and mark them as colourless samples
-Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
-Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
-Points.xrf[20][is.na(Points.xrf[20])] <- "Colourless"
 
@@
 #pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "2.5YR", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9")
-
 ```
@@
 #Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
-Points2 <-
-  samples %>%
+sample.list <-
+  metadata.csv %>%
     filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
@@
 #Reorder type factor levels
-Points2$type <-
-  factor(Points2$type, levels=c("Preform", "Point fragment", "Point"))
+sample.list$type <-
+  factor(sample.list$type, levels=c("Preform", "Point fragment", "Point"))
 
 #Reorder parish factor levels
-Points2$parish <- 
-  factor(Points2$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
+sample.list$parish <- 
+  factor(sample.list$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
 ```
@@
 tot_samples <- 
-  samples %>% 
+  metadata.csv %>% 
     filter(type != "Reference")
@@
 lycksele <-
-  samples %>% 
+  metadata.csv %>% 
     filter(site_id == "Lycksele 301")
@@
 sample_uniq <- 
-  unique(Points2$sample_id)
+  unique(sample.list$sample_id)
 ```
@@
 #Stacked bar chart with counts of samples per parish and type
-ggplot(Points2, aes(x= fct_rev(parish), fill = type)) +
+ggplot(sample.list, aes(x= fct_rev(parish), fill = type)) +
   geom_bar() +
@@
 
+```{r nir-prep}
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#select and filter NIR data to focus on relevant material and artefact types
+Points.nir <-
+  nir.merged %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
+```
+
 ```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3.
@@
 
-#perform PCA with SNV normalization and mean-center
+#perform PCA with mean-centering
 visnir.pca <-
-  prcomp(Points.nir[,c(78:428)], center = TRUE, scale = FALSE)
+  prcomp(Points.nir[,c(80:430)], center = TRUE, scale = FALSE)
 
@@
   filter(hue == "Dark") %>% 
-  select(c(1:28),c(680:2179))
+  select(c(1:28),c(680:2180))
 
@@
   filter(hue == "Light") %>% 
-  select(c(1:28),c(680:2179))
+  select(c(1:28),c(680:2180))
 
@@
   filter(hue == "Colourless") %>% 
-  select(c(1:28),c(680:2179))
+  select(c(1:28),c(680:2180))
 
@@
   filter(hue == "White") %>% 
-  select(c(1:28),c(680:2179))
+  select(c(1:28),c(680:2180))
 
@@
 nir.pca <-
-  prcomp(Points.nir[,c(680:2179)], center = TRUE, scale = FALSE)
+  prcomp(Points.nir[,c(680:2180)], center = TRUE, scale = FALSE)
 
@@
 
+```{r raman-prep-data}
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#perform Standard Normal Variate on raman data
+#raman.averaged[,2:466] <- 
+#  standardNormalVariate(raman.averaged[,2:466])
+
+#merge data with metadata
+#raman_merged <- 
+#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+    select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <-matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_spec_object")
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
+#spectra are stored in "Spectra" object, not a data frame
+#need to extract samples and frequencies from that object 
+raman.tmp <- 
+  cbind(raman.baseline$freq, t(raman.baseline$data))
+#transpose data to get frequencies as columns
+raman.tmp <- 
+  as.data.frame(t(raman.tmp))
+#Add sample_id column and remove the initial "X" character introduced to sample names by R
+raman.tmp$sample_id <- 
+  c("sample_id", sub("^X*", "", raman.baseline$names))
+
+#Make first row headers and remove the first row with the names
+names(raman.tmp) <-
+  raman.tmp[1,]
+raman.tmp <- 
+  raman.tmp[-1,]
+
+#merge raman de-trended data with metadata
+raman.baseline.merged <- 
+  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))
+
+#select and filter raman data to focus on relevant material and artefact types
+Points.raman <-
+  raman.baseline.merged %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
+Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+```
+
 ```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Average Raman spectra grouped by their hue.
@@
 basic_plot1 <-
-  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")
+  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")
 
@@
   plot_layout(nrow = 2, guides = "collect")
+#ADD CIRCLES AROUND 700-800 cm-1 points , try https://datavizpyr.com/how-to-add-circles-around-specific-data-points-in-r/
+
 ```
@@
 \linebreak
+
+```{r xrf-data-prep}
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
+           type == "Point" | type == "Point fragment" | type == "Preform")
+
+#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
+Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
+```
 

### 2022-01-27 analysis/paper/paper.Rmd
updated some text

@@
 This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light.
-It is also difficult to separate out the greyscale of colours that occur inbetween the "black and white" groups, where both light gray and semi-translucent samples end up.
+It is also difficult to separate out the greyscale of colours that occur inbetween the "black" and "white" groups, where both light gray and semi-translucent samples end up.
 All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.
@@
 \@ref(fig:nir-spectra)).
-Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90].
+Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90; @ramanaidou15].
 Inspection of individual spectra for the light group, however, also reveals a 5th peak for some samples at 2 450 nm.
-Comparisons to spectra collected on muscovite [@clark_etal90 p.12 675] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
+Comparisons to spectra collected on muscovite [@clark_etal90 p.12; @ramanaidou15 p.220] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
 It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.
@@
 ## Raman
-There are two distinctive features present in the Raman data (fig.
+There are two distinguishing features present in the Raman data (fig.
 \@ref(fig:raman-spectra)).
@@
 Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band.
-Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset.
+Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset, although measurements on iron ore minerals have also generated a response in the 600 - 700 cm^-1^ band region [@ramanaidou15].
 
@@
 
-```{r raman-prep-data}
+```{r raman-prep-data, results="hide", fig.keep="none"}
 #aggregate observations by group(sample) and calculate average of wavelength measurements

### 2022-03-09 analysis/paper/paper.Rmd
Text update and code altered for raman pca to highlight 700-800 wavenumber samples

@@
 ---
-title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach"
+title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach"
 author:
@@
 abstract: |
-  Text of abstract
+  Northern Fennoscandia represents a geologically complex region with both glacial and postglacial processes having an impact on the archaeology.
+The widely distributed moraine makes provenience studies difficult, and there is a severe lack of methodological development in the characterization of local geological materials used for tool making.
+The question is how to approach provenience studies in a region where there is a lack of reference samples as well as documentation of known geological sources.
+Quartz and quartzite are two central material types utilized among hunter-gatherers in Northern Sweden, and thus important to the understanding of raw material procurement and material flow within the region.
+A dataset of 445 quartz and quartzite points were analyzed non-destructively with three separate spectroscopic instrumentations, including Near-Infrared, Raman scattering and X-Ray Fluorescence.
+The aim was to test whether non-destructive application of these spectroscopic techniques can distinguish between different quartz and quartzite types.
+Multivariate modelling of the spectral data was performed using Principal Component Analysis (PCA) in which the features observed in the spectra contribute to the clustering within the model.
+While the results demonstrate that there are a number of structural and compositional differences within the dataset it is difficult to identify any distinct material classes based on any singular technique.
+The features identified do, however, provide good grounds for survey efforts and the identification of potential geological origin of some of the material.
+Future efforts will endeavor to model the joint relations between the separate blocks in order to test whether the combined information can aid in discerning different material classes.
 keywords: |
@@
 metadata.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
+  read.csv2("./analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
 
@@
 xrf.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/XRF/xrf_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2("./analysis/data/raw_data/XRF/xrf_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
@@
 nir.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 raman.csv <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
@@
 samples <-
-  read.csv2("C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
+  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
 
@@
 While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information.
-Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive option for field-based screening purposes.
+Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes.
 While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
@@
 In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF).
-The aim of the study is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region based on surface measurements.
+The aim is to evaluate their viability for non-destructive fingerprinting of quartz and quartzite material in the North Swedish region.
 The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use.
-As flint is a non-local material to Northern Sweden it follows that any find of this material within the region most likely have been brought there through anthropogenic activity.
-In contrast, determining the degree to which cultural and/or environmental factors have influenced the distribution pattern of quartz and quartzite artefacts, however, is far more difficult.
+In contrast to non-local material such as flint, determining the degree to which cultural and/or environmental factors have influenced the distribution pattern of quartz and quartzite artefacts is difficult.
 While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for spectral comparison.
@@
 
-In order to engage in any meaningful analysis, however, it is first necessary to distinguish between different quartz and quartzite groups.
-Principal Component Analysis (PCA) has been used to model the spectral data in order to discern material groupings related to different spectral features and their contributions to the Principal Components (PC).
-The results form the basis of a discussion of the viability of using non-destructive spectroscopy in petrographic characterization of quartz and quartzite material, and thus future efforts of developing models for raw material sourcing in Northern Sweden.
-
-\linebreak
-
-<!--# Study region
-
-\linebreak
- \linebreak
-
-```{r geo-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geological map of Västerbotten County, Sweden, showing both the tectonic units of the mountains and the widely distributed moraine.
-Made with Lantmäteriet (CC0), Natural Earth, and Sveriges Geologiska Undersökning (SGU)"}
-knitr::include_graphics('../figures/geo_map2.png', error = FALSE)
-```
-
-\linebreak
-
-The effects of the last glacial on the landscape is visible in the widely distributed moraine formations and glaciofluvial deposits (fig.
-\@ref(fig:distribution-map)).
-In Västerbotten the Köli Nappe complex exhibits greensschist - amphibolite facies metamorphism, overlying the Seve Nappe complex where High Pressure (HP) and Ultra High Pressure (UHP) metamoprhism has been detected in a number of localities.
-
-In 1925 the Geological Society arranged an exursion to Södra Storfjället in Västerbotten.
-In the description of the regional stratigraphy it is noted that the lower parts are made up of, among others, "*...dark schists and graphitic quartzite...*" [@geologiska25 p.147].
-Apart from these quartzite with little impurities and Alum shale are also noted.
--->
+In order to engage in any meaningful analysis of the above settlement pattern, however, it is first necessary to distinguish between different quartz and quartzite groups.
+In doing so the distribution of different material groups can be spatially analyzed, and the above models for settlement and mobility tested.
+In this study Principal Component Analysis (PCA) is used in order to group the materials according to their spectral features and their contributions to the Principal Components (PC) [@geladi_linderholm20].
+Any clustering or trends observed in the PCA should relate to the chemical characteristic of the material, and thus potentially similar geological origin.
+The results form the basis of a discussion of the viability of using non-destructive spectroscopy in petrographic characterization of quartz and quartzite material, and thus future efforts in the development of models for raw material sourcing in Northern Sweden.
+
+\linebreak
+
 # Material
@@
 
-While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these.
+While some lithics were petrographically analysed in the *Early Norrland* project [@biornstad68; @ahman67] it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these.
 It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature.
@@
 Another reason for keeping a quartz category is the group referred to as *brecciated quartz*.
-In a petrographic study of the artefacts recovered along the Ume river valley Åhman [-@ahman67 p.8] briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*.
+In the above mentioned petrographic study of the artefacts recovered along the Ume river valley Åhman [-@ahman67 p.8] briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*.
 Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure.
@@
 
-All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/forskning/infrastruktur/mal/].
+All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/].
 The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens.
@@
 In contrast, most of the contribution to PC 2 occurs within the blue-violet part of the spectrum, which does not seem to alter the structure of the dataset in any significant way.
-PC 3 feature high contributions within both the blue-violet and red parts of the spectrum, which also seems to draw the light and dark samples that feature a blue-gray tint towards the top of the score plot.
+PC 3 feature high contributions within the blue-green range, which also seems to draw the light and dark samples that feature a blue-gray tint towards the top of the score plot.
 
@@
 This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light.
-It is also difficult to separate out the greyscale of colours that occur inbetween the "black" and "white" groups, where both light gray and semi-translucent samples end up.
+It is also difficult to separate out the greyscale of colours that occur in between the "black" and "white" groups, where both light gray and semi-translucent samples end up.
 All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.
@@
 It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.
-
-The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig.
-\@ref(fig:nir-pca)).
-Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side.
-While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the brecciated quartz, seem to cluster towards the top-left.
-At the top of the score plot 3 black samples separate themselves clearly from the rest of the data points.
-These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
 
@@
 
+The PCA shows a strong trend along PC 1 (`r nir.pc1var` %) that seems to correlate with the colour of the material(fig.
+\@ref(fig:nir-pca)).
+Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side.
+While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the brecciated quartz, seem to cluster towards the top-left.
+At the top of the score plot 3 black samples separate themselves clearly from the rest of the data points.
+These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
+
+\linebreak
+
+```{r raman-prep-data, results="hide", fig.keep="none"}
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#perform Standard Normal Variate on raman data
+#raman.averaged[,2:466] <- 
+#  standardNormalVariate(raman.averaged[,2:466])
+
+#merge data with metadata
+#raman_merged <- 
+#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+    select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <-matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "./analysis/data/derived_data/raman_spec_object")
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
+#spectra are stored in "Spectra" object, not a data frame
+#need to extract samples and frequencies from that object 
+raman.tmp <- 
+  cbind(raman.baseline$freq, t(raman.baseline$data))
+#transpose data to get frequencies as columns
+raman.tmp <- 
+  as.data.frame(t(raman.tmp))
+#Add sample_id column and remove the initial "X" character introduced to sample names by R
+raman.tmp$sample_id <- 
+  c("sample_id", sub("^X*", "", raman.baseline$names))
+
+#Make first row headers and remove the first row with the names
+names(raman.tmp) <-
+  raman.tmp[1,]
+raman.tmp <- 
+  raman.tmp[-1,]
+
+#merge raman de-trended data with metadata
+raman.baseline.merged <- 
+  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))
+
+#select and filter raman data to focus on relevant material and artefact types
+Points.raman <-
+  raman.baseline.merged %>%
+  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
+         type == "Point" | type == "Point fragment" | type == "Preform")
+
+#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
+Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+```
+
+\linebreak
+
 ## Raman
@@
 A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ~ 695 cm^-1^ and 795 cm^-1^.
-Comparing these to reference samples from the mineral and rock collections kept at the Department of Ecology and Environmental Science at Umeå University, the same peaks are present in measurements collected on, among others, albite, apatite, diabase and shale.
 Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band.
@@
 
-The resulting PCA model (fig.
-\@ref(fig:raman-pca)) shows a clear trend along PC 1 (81.5 %) related to the quartz and D - G peaks.
-PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
-Interestingly, the loading for PC 4 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
-The measurements with mainly quartz peaks gravitates towards the negative side of PC 1, and those where the graphite peaks are dominant towards the opposite side.
-It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks.
-As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-When PC 4 is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
-
-\linebreak
-
-```{r raman-prep-data, results="hide", fig.keep="none"}
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
-
-#perform Standard Normal Variate on raman data
-#raman.averaged[,2:466] <- 
-#  standardNormalVariate(raman.averaged[,2:466])
-
-#merge data with metadata
-#raman_merged <- 
-#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))
-
-#select sample_id and data columns and transpose in prep for baseline de-trend
-raman.transposed <- 
-  raman.averaged %>% 
-    select(sample_id, `92.88`:`2503.59`)
-
-raman.transposed <- 
-  t(raman.transposed)
-
-#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
-#ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
-
-#prepare vector with sample_id for reading raman data into spectra object
-sample_id <- 
-  raman.averaged$sample_id
-
-#read transposed raman data into spectra object
-raman <-matrix2SpectraObject(
-  gr.crit = sample_id,
-  gr.cols = c("auto"),
-  freq.unit ="Wavelength cm-1",
-  int.unit ="Intensity",
-  descrip ="Bifacial points measurements",
-  in.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_transposed.csv",
-  sep = ";",
-  dec = ".",
-  chk = TRUE,
-  out.file = "C:/Users/sjola/Documents/VERSION CONTROL/quantifying-quartzite/analysis/data/derived_data/raman_spec_object")
-
-#call on relevant baseline method (modified polynomial fitting) from baseline package and 
-#return corrected spectra to spectra object
-raman.baseline <- 
-  baselineSpectra(raman,
-                  int = FALSE,
-                  method = "modpolyfit",
-                  retC = TRUE)
-
-#spectra are stored in "Spectra" object, not a data frame
-#need to extract samples and frequencies from that object 
-raman.tmp <- 
-  cbind(raman.baseline$freq, t(raman.baseline$data))
-#transpose data to get frequencies as columns
-raman.tmp <- 
-  as.data.frame(t(raman.tmp))
-#Add sample_id column and remove the initial "X" character introduced to sample names by R
-raman.tmp$sample_id <- 
-  c("sample_id", sub("^X*", "", raman.baseline$names))
-
-#Make first row headers and remove the first row with the names
-names(raman.tmp) <-
-  raman.tmp[1,]
-raman.tmp <- 
-  raman.tmp[-1,]
-
-#merge raman de-trended data with metadata
-raman.baseline.merged <- 
-  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))
-
-#select and filter raman data to focus on relevant material and artefact types
-Points.raman <-
-  raman.baseline.merged %>%
-  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-         type == "Point" | type == "Point fragment" | type == "Preform")
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
-```
+\linebreak
 
@@
 plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm^-1^)", main = "Average - Dark")
+     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - Dark")
 text(x=120, y=2900, labels="Quartz", font=2, cex = 1.5)
@@
 plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm^-1^)", main = "Average - Light")
+     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - Light")
 
 plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm^-1^)", main = "Average - Colourless")
+     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - Colourless")
 
 plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = "Wavenumber (cm^-1^)", main = "Average - White")
+     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - White")
 ```
@@
 As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA.
-PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the bottom left, that may result from Al, Si and OH vibrations."}
+PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the top left (marked with red circle), that may result from Al, Si and OH vibrations."}
 
@@
 basic_plot1 <-
+  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")
+
+basic_plot2 <-
   fviz_pca_ind(raman.pca, axes = c(1,3), label="none")
-
-basic_plot2 <-
-  fviz_pca_ind(raman.pca, axes = c(1,4), label="none")
 
@@
 raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
-raman.pc4var <- round(summary(raman.pca)$importance[2,4]*100, digits=1)
+raman.pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
 raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
 raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
-raman.pc4lab <- paste0("PC4 (",as.character(raman.pc4var),"%)")
+raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")
 
@@
                                                  size=3),
-                             title.position="top", title.hjust = 0.5))+
+                             title.position="top", title.hjust = 0.5)) +
   theme(legend.position = "none")
 
-#bind the basic fviz plot for PC 1 and 3, and use as basis for a more customizeable plot in ggpplot
-raman.p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
-       aes(x=x, y=y, 
+#bind the basic fviz plot for PC 1 and 3, with metadata for visualization, and use as basis for a more customizeable plot in ggpplot
+#In order to be able to filter out certain points that have the 700-800 wavenumber peaks and give them a red circle in a second geom_point statement this 
+#has to be done outside of ggplot statement
+r.data2 <- 
+  cbind(basic_plot2$data, Points.raman[, c(1,8,10)])
+
+#Set up the score plot and 
+raman.p2 <- 
+  r.data2 %>%  
+  ggplot(aes(x=x, y=y, 
            shape = material, 
@@
   labs(x = raman.pc1lab,
-       y = raman.pc4lab) +
+       y = raman.pc3lab) +
   scale_shape_manual(name = "Material",
@@
                                                  size=3),
-                             title.position="top", title.hjust = 0.5))+
+                             title.position="top", title.hjust = 0.5)) +
+  #Filter out points with 700 - 800 wavenumber peaks and give them a red circle and slightly increase their size for better identification
+  geom_point(data = r.data2 %>% 
+               filter(sample_id == 6 | sample_id == 34 | sample_id == 42 | sample_id == 127 | sample_id == 445 | sample_id == 482 | sample_id == 491 | sample_id == 494 | sample_id == 495 | sample_id == 501 |sample_id == 502 | sample_id == 503 | sample_id == 504),
+             pch=21,
+             size=4, 
+             colour ="red") +
   theme(legend.position = "none")
@@
   plot_layout(nrow = 2, guides = "collect")
-#ADD CIRCLES AROUND 700-800 cm-1 points , try https://datavizpyr.com/how-to-add-circles-around-specific-data-points-in-r/
-
-```
+
+```
+
+\linebreak
+
+The resulting PCA model (fig.
+\@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks.
+PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
+Interestingly, the loading for PC 3 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
+The measurements with mainly quartz peaks gravitates towards the negative side of PC 1, and those where the graphite peaks are dominant towards the opposite side.
+It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks.
+As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
+When PC 3 (`r raman.pc3var` %) is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
 
@@
 
-The most easily recognizable megascopic feature of petrographic material is its colour, a feature that has been one of the main means of grouping quartzite in North Swedish archaeological research [e.g.
-@forsberg85 p.60-64; @kaeck09 p.
-53-55].
-While it is true that some inference can be made based on colour, such as ^26^Fe-bearing minerals contributing to a darker colour, this is significantly more difficult with heterogeneous material.
-Should a quartzite where the matrix is mainly light gray in colour, but streaked with black mineral veins, be considered a *light* or *dark* quartzite?
-The results above reflects this complexity.
-While the above models tend to group samples of a similar hue together, especially in the case of NIRS, there are samples that break these trends and thus demonstrate the presence of structural and compositional features that do not necessarily align with the perceived colour of a material.
-Rather than assigning color based on how it is perceived visually it can be approached in a more systematic manner [e.g.
-@fairchild13].
-Restricting the NIRS model to the visual range (~ 400 nm - 750 nm) opens up for some inference of color based on the absorbance of light within certain bands.
-However, as the data was collected using a contact probe the issue of how heterogeneous samples should be treated remains.
-This issue can be alleviated somewhat with a different sampling approach, for instance using a hyperspectral camera where different pixels can be selected and analysed separately [@sciuto19].
-
-In the case of the Raman data the PCA is somewhat lacking in exploratory and descriptive power.
-Apart from the few samples that feature the 700 cm^-1^ - 800 cm^-1^ band peaks the major features in the data are all explained by the first component.
-As the intensity in Raman spectroscopy is not a direct quantitative indicator this makes it difficult to group materials based on the intensity of the D and G bands, making more of a binary separation of samples with or without *graphite peaks*.
-That is not to say that it is not possible to attempt quantification using Raman spectroscopy, but rather that this would necessitate a different experimental setup.
-That said, the presence of the *graphite peaks* may present an important piece of information in the identification of the potential geological origin of the material.
-Interestingly, mentions of *graphitic quartzite* can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
-
-The material described as *brecciated quartz* remains relatively grouped together in all of the models, demonstrating some consistency with the original classification.
-As the samples group with other similarly semi-translucent material, however, it does not seem to distinguish itself enough in the models in order to consider it its own group.
-This is perhaps not so strange as the original classification by @ahman67 refers to the microscopic fragmented configuration of the quartz crystals as being the main distinguishing feature, and while it is possible to discern the presence of SiO structure through non-destructive spectroscopy, there have been no other discernible features related to this group at this stage.
-It is questionable whether all of the samples in this group truly exhibit brecciation, given the context under which they were collected and classified, which further complicates analysis of this material.
-Nevertheless, the results of the analysis suggests that there is some in-group variation, regardless of whether the material exhibits breccia.
+The most easily recognizable megascopic feature of petrographic material is its color, which has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g.
+@forsberg85 pp.60–64; @kaeck09 pp.53–55].
+While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite.
+Indeed, destructive techniques are typically necessary in order to document the various metamorphic features characteristic of quartzite [@howard05; @prieto_etal19].
+Non-destructive application of the above spectroscopic techniques does, however, bring to light some interesting characteristics of the material.
+Although there are trends that align with the perceived color of the objects in the PCA there are also exceptions to this rule, as well as a “muddled” grey zone of material that does not conform to the light and dark extremes.
+It is also worth mentioning that the color classification does not take the heterogeneous nature of the material into consideration.
+
+Although no distinct enough groupings were observed as to make classification possible, a number of structural and compositional characteristics were identified.
+Despite the Raman PCA lacking descriptive power, due to the few features observed in the spectra, the presence of the “graphite peaks” may present an important piece of information in the identification of the potential geological origin of the material.
+Interestingly, mentions of graphitic quartzite can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
+
+Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies are necessary in order to infer resource management strategies from this evidence.
+The characterization of quartz and quartzite is fundamental, however, if questions of resource procurement, trade and exchange and human mobility are to be developed.
+Being able to distinguish between different types of quartz and quartzite would enable various spatial analyses and modelling of distribution patterns, regardless of whether the geological source is known or not.
+Moving forwards, the answer may lie in the field of multi-block analysis, where the potential for joint relations between separate blocks of data can be tested.
 
@@
 The study represents a good example of a multivariate exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
-There is a need for methodologies that can be applied in regions with a complex geological setting and vague distribution of potential geological sources.
-Taking full advantage of the short integration time of these instrumentations in order to build large comprehensive datasets it is possible to discern spectral fingerprints using these methodologies.
+There is a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
+Taking full advantage of the short integration time of these instrumentations it is possible to build large comprehensive datasets that then can be explored and mined using multivariate models.
 

### 2022-03-10 analysis/paper/paper.Rmd
Updated text and overhauled certain figures

@@
 PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].
+ \linebreak
 
@@
 
+#Set the legend for the combined figure of both score plots
 comb1 <-
@@
 
+#Call on the finished figure and layout in 2 rows
 comb1 +
@@
 The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands.
-As these are the main features in the data, the other components mainly describe the remaining noise, producing the arch-shape in the PCA.
-PC 4 does, however, also separate out the few samples that feature peaks at 700-800 cm^-1^ in the top left (marked with red circle), that may result from Al, Si and OH vibrations."}
+As these are the main features in the data, the other components mainly describe the remaining noise.
+PC 3 does, however, separate out the few samples that feature peaks at 700-800 cm^-1^ in the top left (marked with red circle), that may result from Al, Si and OH vibrations."}
 
@@
 
-#Set up the score plot and 
+#Set up the score plot
 raman.p2 <- 
@@
                              title.position="top", title.hjust = 0.5)) +
-  #Filter out points with 700 - 800 wavenumber peaks and give them a red circle and slightly increase their size for better identification
+  #Filter out points with 700 - 800 wavenumber peaks and give them a red circle for easier identification
   geom_point(data = r.data2 %>% 
@@
              pch=21,
-             size=4, 
-             colour ="red") +
+             size=7, 
+             colour ="red",
+             fill = NA,
+             stroke = 2) +
   theme(legend.position = "none")
 
+#Set the legend for the combined figure of both score plots
 comb <-
@@
 
+#Call on the finished figure and layout in 2 rows
 comb +
   plot_layout(nrow = 2, guides = "collect")
-
 ```
@@
 
+```{r xrf-data-prep}
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
+           type == "Point" | type == "Point fragment" | type == "Preform")
+
+#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
+Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
+
+#Filter out Si values above 50%
+high_si <- 
+  Points.xrf %>% 
+    filter(`14Si` > 50)
+
+#Prepare a calculation of percentage measurements above 50% to call on in text
+high_si <- 
+  round(length(high_si$`14Si`) / length(Points.xrf$`14Si`) * 100)
+
+#Prepare XRF data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
+Points_long <-
+  drop_na(Points.xrf %>%
+    filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
+    select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
+    gather(key = "Element", value = "Value"))
+
+#filter XRF data on sample size (only include samples with smallest dimension >= 10mm), 
+#select elements to include in PCA, as well as columns to group by
+xrf <-
+  Points.xrf %>%
+  filter(max_length_mm >= 10 & max_width_mm >= 10) %>% 
+  select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)
+
+#log transform XRF data
+xrf.log <- log(xrf[,5:20])
+
+#impute missing values
+xrf.imp <- imputePCA(xrf.log, ncp = 2, scale = FALSE, method = "Regularized", maxiter = 1000)
+
+#perform pca with mean centering and unit variance scaling
+xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=TRUE)
+
+#Prepare cumulative explained variance of two first XRF PCs
+xrf_variance <- 
+  round(summary(xrf.pca)$importance[2,1]*100, digits=1) + round(summary(xrf.pca)$importance[2,2]*100, digits=1)
+
+```
+
 ## XRF
@@
 The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF).
-Univariate analysis shows that SiO^2^ makes up the main bulk of the composition of the material, and two major groups can be distinguished based on this content (fig.
+Univariate analysis shows that ^14^Si makes up the main bulk of the composition of the material, with a bimodal distribution (fig.
 \@ref(fig:xrf-violin-plot)).
-The majority of the readings resulted in a SiO^2^ content of less than 50%, and about ¼ of the readings (~ 27%) featured a SiO^2^ content closer to 90%.
-The “low” SiO^2^ group also feature a Bal between 40 – 70%.
-
-The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ 81.3%.
+The majority of the readings resulted in a ^14^Si content of less than 50%, and about ¼ of the readings (~ `r high_si`%) featured a ^14^Si content closer to 90%.
+The “low” ^14^Si group also feature a Bal between 40 – 70%.
+
+The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ `r high_si`%.
 The loadings can be be seen in figure \@ref(fig:xrf-loadings).
@@
 \@ref(fig:xrf-score-plot)).
-Interestingly, the smaller cluster on the left side in the score plot includes the samples with a SiO^2^ content higher than 50%, whilst the larger cluster includes those with lower SiO^2^ content.
+The smaller cluster on the left side in the score plot includes the samples with a ^14^Si content higher than 50%, whilst the larger cluster includes those with lower ^14^Si content.
 It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that are the main contributors.
@@
 The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river.
-The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25 %.
+The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25%.
 This artefact was sampled twice on each side, resulting in similar results both times.
@@
 \linebreak
-
-```{r xrf-data-prep}
-
-#merge XRF data with metadata
-xrf.merged <- 
-  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
-
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
-Points.xrf <-
-  xrf.merged %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-           type == "Point" | type == "Point fragment" | type == "Preform")
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
-```
 
@@
 Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
-#Prepare data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
-Points_long <-
-  drop_na(Points.xrf %>%
-    filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
-    select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
-    gather(key = "Element", value = "Value"))
 
@@
 ```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each variable (element)"}
-#filter on sample size (only include samples with smallest dimension >= 10mm), 
-#select elements to include in pca, as well as columns to group by
-xrf <-
-  Points.xrf %>%
-  filter(max_length_mm >= 10 & max_width_mm >= 10) %>% 
-  select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)
-
-#log transform data
-xrf.log <- log(xrf[,5:20])
-
-#impute missing values
-xrf.imp <- imputePCA(xrf.log, ncp = 2, scale = FALSE, method = "Regularized", maxiter = 1000)
-
-#perform pca with mean centering and unit variance scaling
-xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=TRUE)
 
@@
 
-Although no distinct enough groupings were observed as to make classification possible, a number of structural and compositional characteristics were identified.
+Although no distinct enough groupings were observed as to make classification possible, a number of structural and chemical characteristics were identified.
 Despite the Raman PCA lacking descriptive power, due to the few features observed in the spectra, the presence of the “graphite peaks” may present an important piece of information in the identification of the potential geological origin of the material.
 Interestingly, mentions of graphitic quartzite can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
-
-Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies are necessary in order to infer resource management strategies from this evidence.
-The characterization of quartz and quartzite is fundamental, however, if questions of resource procurement, trade and exchange and human mobility are to be developed.
-Being able to distinguish between different types of quartz and quartzite would enable various spatial analyses and modelling of distribution patterns, regardless of whether the geological source is known or not.
-Moving forwards, the answer may lie in the field of multi-block analysis, where the potential for joint relations between separate blocks of data can be tested.
+Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies would be necessary in order to infer resource management strategies from this evidence.
+
+The exploratory approach presented in this paper holds potential for enabling provenance studies within geologically complex regions.
+Screening a large enough dataset makes it possible to look at in-group variations and isolate spectroscopic distinct fingerprints [e.g.
+@sciuto_etal19].
+These can then be used as the basis for later classification of the material into structurally different groups.
+The characterization of quartz and quartzite is fundamental if questions of resource procurement, trade and exchange and human mobility are to be developed within Northern Sweden.
+Being able to distinguish between different types of quartz and quartzite would enable the spatial analysis and modelling of distribution patterns, regardless of whether the geological source is known or not.
+In the current case study modelling the results of single instrumentation may have been limited for the purpose of classification, but modelling the joint relations between the separate data blocks may provide further aid in isolating distinct types of quartz and quartzite.
 
@@
 There is a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
-Taking full advantage of the short integration time of these instrumentations it is possible to build large comprehensive datasets that then can be explored and mined using multivariate models.
+Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined using multivariate models.
 

### 2022-05-13 analysis/paper/001-nir-spectra-summary.R
New subscripts for better code problemshooting

@@ -1,0 +1,150 @@
+# Loading package
+library(data.table)
+library(ggplot2)
+library(gridBase)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-05-13 analysis/paper/002-nir-dark-filtered-summary.R

@@ -1,0 +1,153 @@
+# Loading package
+library(gridBase)
+library(spectrolab)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+Points.nir[,1] <- as.numeric(Points.nir[,1]) 
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-05-13 analysis/paper/003-nir-savitzky-transform.R

@@ -1,0 +1,126 @@
+#Load packages
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-05-13 analysis/paper/004-nir-pca.R

@@ -1,0 +1,92 @@
+#Load packages
+suppressPackageStartupMessages(library(plotly))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+        read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+        read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+        aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+        as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+        nir.merged %>%
+        filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+                       site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+                       site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+                       site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-05-13 analysis/paper/005-raman-spectra-summary.R

@@ -1,0 +1,200 @@
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(spectrolab))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+  select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = ".")
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <- suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "./analysis/data/derived_data/raman_spec_object"))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
[…]

### 2022-05-13 analysis/paper/006-raman-peak-identification.R

@@ -1,0 +1,160 @@
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(spectrolab))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+  select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <- suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "./analysis/data/derived_data/raman_spec_object"))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
[…]

### 2022-05-13 analysis/paper/007-raman-pca.R

@@ -1,0 +1,146 @@
+#Load packages
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(plotly))
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#transpose in prep for baseline de-trend
+raman.transposed <- 
+  t(raman.averaged)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <-suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "./analysis/data/derived_data/raman_spec_object"))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
+#prevent baseline from plotting a figure
+dev.off()
+
[…]

### 2022-05-13 analysis/paper/008-xrf-violin-plot.R

@@ -1,0 +1,172 @@
+library(ggplot2)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
[…]

### 2022-05-13 analysis/paper/009-xrf-pca.R

@@ -1,0 +1,127 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-05-13 analysis/paper/010-O2PLS.R

@@ -1,0 +1,133 @@
+#Load packages
+suppressPackageStartupMessages(library(OmicsPLS))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[,4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#merge NIR data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
[…]

### 2022-05-13 analysis/paper/paper.Rmd
Restructured paper with subscripts

@@
   warning = FALSE,
+  error = TRUE,
   message = FALSE,
@@
   fig.path = "../figures/",
-  dpi = 300
+  dpi = 300,
+  encoding = "UTF-8"
 )
@@
 
-<!-- The following code chunk defines prepares most of the data that will be needed and reads in the necessary libraries -->
-
-```{r load-packages-and-data}
-library(ChemoSpec)
-library(data.table)
-library(factoextra)
-library(ggbiplot)
-library(ggplot2)
-library(ggh4x)
-library(ggsci)
-library(grid)
-library(gridBase)
-library(gridExtra)
-library(missMDA)
-library(patchwork)
-library(prospectr)
-library(tidyverse)
-library(spectrolab)
-
-#Import descriptive metadata
-metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220119.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
-
-#Import xrf data, set empty fields to NA
-xrf.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/xrf_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
-
-#Import nir data, set empty fields to NA
-nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
-
-#Import raman data, set empty fields to NA
-raman.csv <-
-  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220127.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
-
-#Import sample list, set empty fields to NA
-samples <-
-  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))
-
-#prepare color/fill and symbols for score plots
-pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
-pca.shapes <- c(24,25,23,22,21)
-pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")
-
-#If including references
-#pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#DF5555", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#BFBFBF", "#DFDFDF", "#FFFFFF")
-#pca.munsell_hue <- c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "2.5YR", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9")
+<!-- The following function can be used to suppress message outputs -->
+```{r}	
+quietly <- function(x) {	
+  sink(tempfile())	
+  on.exit(sink())	
+  invisible(force(suppressMessages(x)))	
+}	
 ```
@@
 
-The last decades have seen an increase in the interest for non-flint lithics and raw material studies in European archaeology, providing much needed development in a previously neglected field [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
-In regions where local flint and obsidian sources are non-existent, such as Northern Fennoscandia, other materials have been the primary target for lithic production^[This is not to say that flint was not available or used in these regions] (e.g.
-quartz, quartzite, slate) [@baudou78; @rankama_etal06; @olausson_etal12].
-With the lack of any natural geological source of flint in Northern Sweden both quartz and quartzite are instrumental for the inference of settlement and mobility within the region.
-While quartz and quartzite technology have been part of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-One reason for the lack of methodological development in North Swedish research could be the complex geological background of the region and the abundance of moraine formations [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
-The degree to which geological materials have been mixed and transported demands considerable effort in order to reconstruct the events that have formed the modern day landscape.
-While tentaive interpretations of lithic material sourcing tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12], both quartz and quartzite could potentially have been sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
+Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF).
+The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
+Although the petrographic characterization of the material mainly establishes a point of geological origin for the object it is still an important first step in reconstructing the *chaîne opératoire*.
+Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these.
+The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
+Prieto et al.
+[-@prieto_etal19 p.
+15] highlights the current bias that exists in the understanding of raw material management in regions where the local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions].
+In northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region.
+While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
+Although flint and slate have featured in discussions of trade and exchange [@halen94 pp.
+115-122; @lundberg97 pp.
+161-171] within the regon, another reason for the lack of methodological development in North Swedish research could be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
+The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape.
+The tentative interpretations of lithic material sourcing within northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+This does not account for the possibility that the material sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
 Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.
@@
 # Material
-```{r select-point-data}
-#Filter data to focus on points, point fragments and preforms made from quartz/quartzite materials
-sample.list <-
-  metadata.csv %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-         type == "Point" | type == "Point fragment" | type == "Preform")
+```{r sample-count, results='hide'}
+suppressPackageStartupMessages(library(tidyverse))
 
-#Reorder type factor levels
-sample.list$type <-
-  factor(sample.list$type, levels=c("Preform", "Point fragment", "Point"))
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220407.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
 
-#Reorder parish factor levels
-sample.list$parish <- 
-  factor(sample.list$parish, levels = c("Bjurholm","Bureå","Burträsk", "Bygdeå", "Degerfors", "Dorotea", "Lycksele", "Norsjö", "Skellefteå", "Sorsele", "Stensele", "Tärna", "Umeå", "Vännäs", "Vilhelmina", "Åsele"))
-```
-
-```{r sample-count, results='hide'}
 #Filter the sample data to exclude reference samples
@@
   metadata.csv %>% 
-    filter(type != "Reference")
-
-#Filter sample data to focus on site Lycksele 301
-lycksele <-
-  metadata.csv %>% 
-    filter(site_id == "Lycksele 301")
-
-#Then filter to focus on quartz/quartzite material that are points and preforms 
-lycksele_301 <-
-  lycksele %>% 
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | 
-             material  == "Felsitic porphyric quartz", type == "Point" | type == "Point fragment" | type == "Preform")
-
-#Create vector of unique samples from site Lycksele 301
-lycksele_301 <-
-unique(lycksele_301$sample_id)
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
+           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
+           site_id == "Åsele 182" |  site_id == "Åsele 188" |site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
+         type == "Point" | type == "Point fragment" | type == "Preform", 
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")
 
@@
 
-#Create a vector in order count number of unique rows in sample ID column for the number of points/preforms included in analysis
-sample_uniq <- 
-  unique(sample.list$sample_id)
+bquartz <- unique(tot_samples %>% 
+  filter(material == "Brecciated quartz"))$sample_id
+
+quartz <- unique(tot_samples %>% 
+  filter(material == "Quartz"))$sample_id
+
+quartzite <- unique(tot_samples %>% 
+  filter(material == "Quartzite"))$sample_id
 ```
@@
 
-The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden.
-A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman scattering and XRF.
-The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
+The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites along the river of Ångermanälven, Västerbotten County, Sweden.
+The study area ranges from the lake of Kultsjön in the southwest of the County to the town of Åsele in the southeast.
+A dataset of `r length(tot_uniq)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
+The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
 The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).
@@
 @andersson15].
-This has likely had a contributing factor to the distribution pattern of the artefacts observed in fig.
-\@ref(fig:distribution-map) as the known sites with finds of points and preforms are concentrated around the parishes of Dorotea, Lycksele^[In this case the majority of the samples from Lycksele are from the rock painting site of Lycksele 301], Vilhelmina and Tärnaby.
-This can be seen more clearly in fig.
-\@ref(fig:distribution-by-parish) with the four parishes accounting for almost ¾ of the total no.
-of samples.
-
-Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online.
-Based on a query of all points and preforms labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database.
-Sampling of this list of artefacts was not random.
-Sites located in areas with few finds were included first in order to increase the geographical distribution, with sampling of areas with higher site density making up the remaining bulk of the dataset.
-None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included.
-This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz or quartzite material.
+Vilhelmina parish in particular has been the target for numerous excavations and survey projects during this period.
 
@@
 
-```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of sampled sites within Västerbotten County, Sweden.
-The distribution of sampled sites is likely affected by the excavation and survey history of the region.
-White: Sampled sites; Black: Parish centroid for measured samples lacking site coordinates.
+```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.
 Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
-knitr::include_graphics('../figures/Distribution.jpeg', error = FALSE)
+knitr::include_graphics('../figures/site_distribution.jpeg', error = FALSE)
 ```
@@
 \linebreak
-
-```{r distribution-by-parish, fig.width=12, fig.height=10, fig.fullwidth=TRUE, fig.cap= paste("Distribution of the", length(sample_uniq), "points and preforms across the parishes in Västerbotten, Sweden.
-Note that while Lycksele feature a large number of samples", length(lycksele_301), "of these are from the site of Lycksele 301")}
-
-#Stacked bar chart with counts of samples per parish and type
-ggplot(sample.list, aes(x= fct_rev(parish), fill = type)) +
-  geom_bar() +
-  theme_minimal() +
-  labs(y = "Count",
-       fill = "Type") +
-  theme(
-    legend.position = "bottom",
-    legend.title = element_text(size = 16, face = "bold"),
-    legend.text = element_text(face = "bold"),
-    axis.title.y=element_blank(),
-    axis.text.y = element_text(size = 10, face = "bold", colour = "black"),
-    axis.title.x = element_text(size = 12, face = "bold", colour = "black")) +
-  scale_fill_jco() +
-  coord_flip() +
-  geom_text(stat ='count', aes(group = parish, label = ..count..), hjust = -0.5)
-```
+The dataset includes three main types of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe.
+"*brecciekvarts*").
+Those that had no prior classification available were sorted into either the quartz or quartzite category.This distribution of samples into these categories are as follows: `r length(bquartz)` brecciated quartz, `r length(quartz)` quartz, and `r length(quartzite)` quartzite samples.
+The inclusion of quartz and brecciated quartz in the study serves two purposes.
+The first being to test the potential for separating the material types based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
+In his study of the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] analysed artefacts from sites located along the Ume river valley.
+He briefly touches upon the definition of *brecciated quartz* (also referred to as *mylonite quartz*) and describes the material as having formed from a quartz vein which has fractured and then recrystallized via pressure.
+While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz.
+*Brecciated quartz* in particular is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
+@holm91 p.24].
+As the material frequently has been characterized based on megascopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.
 
@@
 
-While some lithics were petrographically analysed in the *Early Norrland* project [@biornstad68; @ahman67] it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these.
-It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature.
-In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz and brecciated quartzite^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit*].
-Those that had no prior material classification registered were sorted into either the quartz or quartzite category.
-The reason for including quartz in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.
-
-Another reason for keeping a quartz category is the group referred to as *brecciated quartz*.
-In the above mentioned petrographic study of the artefacts recovered along the Ume river valley Åhman [-@ahman67 p.8] briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*.
-Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure.
-While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz.
-*Brecciated quartz* is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
-@holm91 p.24].
-
-\linebreak
+```{r sample-hue-pictures, fig.cap= "An assortment of points demonstrating the variation in color and heterogeneity of the material."}
+knitr::include_graphics('../figures/sample_hue.jpg', error = FALSE)
+```
 
@@
 After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20].
-PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].
+PCA has proven useful for exploring complex spectroscopic datasets [@sciuto19], and identifying clusters and trends that relate to the spectral fingerprints of the material.
+These groupings can then be tested with supervised methods, and in the construction of classification models.
+Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12].
+The compositional data add up to a constant (100%), which is described as the *closed sum* effect.
+Correlation analysis is directly affected by this *closure* effect and therefore not suitable.
+With suitable preprocessing the PCA model can reveal the structure and relations between the variables (elements) more accurately.
+PCA modelling in this case was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].
  \linebreak
@@
 The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling.
-As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA.
+As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA.
+A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement.
 Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
@@
 \linebreak
- \linebreak
 
-## NIR - visual range
-Examining the average spectra of the different colour groups (dark, light, white and colourless), as expected, the darker material feature higher absorbance of light than the other groups.
-No other distinctive features can be seen in the visual band, that could potentially be used to distinguish them.
-It is, however, apparent that while the more extreme hues of dark and white material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.
+## NIR - near infrared range
+```{r nir-sample-count, results='hide'}
+#Filter out the dark quartzite samples that feature peaks of interest in their spectra 
+points.d <-
+  metadata.csv %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
+           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
+           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
+           site_id == "Åsele 99", 
+         type == "Point" | type == "Point fragment" | type == "Preform", 
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+    filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
+                           "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
+                           "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
+  filter(hue == "Dark")
 
-The cumulative explained variance for the first three components in the PCA is ~ 99.9%, with the accompanying loadings in fig.
-\@ref(fig:vis-nir-loadings).
-Most of the contribution to the first Principal Component (PC) can be found within the red part of the spectrum (~ 600 nm - 750 nm).
-This can be seen in the score plot where darker and lighter materials are separated along PC 1, as darker samples would feature a higher absorbance within the red part of the spectrum (fig.
-\@ref(fig:vis-nir-pca)).
-In contrast, most of the contribution to PC 2 occurs within the blue-violet part of the spectrum, which does not seem to alter the structure of the dataset in any significant way.
-PC 3 feature high contributions within the blue-green range, which also seems to draw the light and dark samples that feature a blue-gray tint towards the top of the score plot.
+#pass to vector so that length() can be used in text to quote nr. of dark samples
+spec.d <- 
+  points.d$sample_id
+```
 
-While the overall structure of the model seems to group colourless and white materials towards the bottom left, with dark material to the right, there are a number of samples that break this pattern.
-This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light.
-It is also difficult to separate out the greyscale of colours that occur in between the "black" and "white" groups, where both light gray and semi-translucent samples end up.
-All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.
+It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the abscence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77].
+Similarly it can be difficult to work with dark materials as most of the light will be absorbed, thus obscuring any potential spectral features.
+This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group.
+The dark group generally features less distinct peaks than the material with a lighter hue.
+This is not to say that all spectra collected on dark quartzite features less information, but any model including them will affect the clustering of the dataset as a whole.
+Filtering out those spectra that feature no peaks or mainly noise leaves `r length(spec.d)` samples in the dark group.
+Excluding the noisier spectra, and those with no discernable peaks, there is still a noticeable difference between the dark and light group.
+By applying 2^nd^ derivative Savitszky-Golay transformation on the dark spectra broad spectral features can be filtered out [@chryssikos_gates pp.78-81].
+Comparing the centered spectra to the derivative it becomes clear that there are significant peaks that may be of relevance to the characterization of the material (fig.
+\@ref(fig:nir-savitzky-transform)).
 
@@
 
-```{r nir-prep}
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-nir.averaged <- 
-  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
-
-#merge NIR data with metadata
-nir.merged <- 
-  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
-
-#select and filter NIR data to focus on relevant material and artefact types
-Points.nir <-
-  nir.merged %>%
-  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-         type == "Point" | type == "Point fragment" | type == "Preform")
-
-#Fill the NA fields in the munsell_hue column and mark them as colourless samples
-Points.nir[8][is.na(Points.nir[8])] <- "Colourless"
-```
-
-```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3.
-The two first components mainly describes features with a linear increase in absorbance across the visual spectrum, but the 3rd component demonstrates more variation, with lower absorbance within the blue-yellow region.
-"}
-
-#perform PCA with mean-centering
-visnir.pca <-
-  prcomp(Points.nir[,c(80:430)], center = TRUE, scale = FALSE)
-
-#Filter NIR data to focus on the material with a dark hue 
-#and select the NIR range 1 000 - 2 500 nm
-#vis.dark <- Points.nir %>% 
-#  filter(hue == "Dark") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vd <- 
-#      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)
-
-#vis.light <- Points.nir %>% 
-#  filter(hue == "Light") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vl <- 
-#  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)
-
-#vis.colourless <- Points.nir %>% 
-#  filter(hue == "Colourless") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vt <- 
-#  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)
-
-#vis.white <- Points.nir %>% 
-#  filter(hue == "White") %>% 
-#  select(c(1:28),c(79:429))
-
-#spec.vw <- 
-#  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)
-
-#VISNIR LOADING LINES
-#For some reason invisible() has to be used to prevent console output showing up in knitted document
-par(mfrow=c(3,1))
-invisible(sapply(1:3, function(comp){
-  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
-      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,5))}))
-
-#par(mfrow=c(4,1))
-
-#plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
-
-#plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")
-
-#plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")
-
-#plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
+```{r nir-spectra-summary, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
+quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
+knitr::include_graphics('../figures/001-nir-spectra-summary.png', error = FALSE)
 ```
@@
  
-```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data.
-Top: PC 1 and PC 2, bottom: PC 1 and PC 3.
-PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure.
-PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}
-
-#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
-basic_plot1 <-
-  fviz_pca_ind(visnir.pca, axes = c(1,2), label="none")
-basic_plot2 <-
-  fviz_pca_ind(visnir.pca, axes = c(1,3), label="none")
-
-#prepare labels for PCs
-vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=2)
-vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=2)
-vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=2)
-vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
-vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
-vis.pc3lab <- paste0("PC3 (",as.character(vis.pc3var),"%)")
-
-#VISNIR LOADING LINES
-#par(mfrow=c(3,1))
-#sapply(1:3, function(comp){
-#  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
-#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,6))})
-
-#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
-visnir.p1 <-
-  ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
-             aes(x=x, y=y, 
-                 shape = material, 
-                 fill = munsell_hue)) +
-  geom_point(size=3) +
-  theme_bw() +
-  ggtitle("Near infrared 400 - 750 nm") +
-  labs(x = vis.pc1lab,
-       y = vis.pc2lab) +
-  scale_shape_manual(name = "Material",
-                     values=pca.shapes) +
-  scale_fill_manual(name = "Munsell hue",
-                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol=2, 
-                              title.position="top", 
-                              title.hjust = 0.5),
-         fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = pca.colors,
-                                                 color = "black",
-                                                 size=3),
-                             title.position="top", 
-                             title.hjust = 0.5)) +
-  theme(legend.position = "none")
-
-visnir.p2 <-
-  ggplot(cbind(basic_plot2$data, Points.nir[, c(8,10)]),
-             aes(x=x, y=y, 
-                 shape = material, 
-                 fill = munsell_hue)) +
-  geom_point(size=3) +
-  theme_bw() +
-  ggtitle("Near infrared 400 - 750 nm") +
-  labs(x = vis.pc1lab,
-       y = vis.pc3lab) +
-  scale_shape_manual(name = "Material",
-                     values=pca.shapes) +
-  scale_fill_manual(name = "Munsell hue",
-                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol=2, 
-                              title.position="top", 
-                              title.hjust = 0.5),
-         fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = pca.colors,
-                                                 color = "black",
-                                                 size=3),
-                             title.position="top", 
-                             title.hjust = 0.5)) +
-  theme(legend.position = "none")
-
-#Set the legend for the combined figure of both score plots
-comb1 <-
-  visnir.p1 + visnir.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
-                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
-                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
-                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
-                              legend.background = element_rect(linetype = "solid", color = "black"),
-                              legend.position = "bottom")
-
-#Call on the finished figure and layout in 2 rows
-comb1 +
-  plot_layout(nrow = 2, guides = "collect")
+```{r nir-spectra-dark, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-dark-filtered-summary.R"), echo = FALSE}
+quietly(source("./analysis/paper/002-nir-dark-filtered-summary.R", encoding = "UTF-8"))
+knitr::include_graphics('../figures/002-nir-dark-filtered-summary.png', error = FALSE)
 ```
@@
 
-## NIR - near infrared range
-In the case of the 1 000 nm - 2 500 nm band of the NIR data, similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks.
-The presence of molecular water can be inferred based on the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig.
-\@ref(fig:nir-spectra)).
-Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90; @ramanaidou15].
-Inspection of individual spectra for the light group, however, also reveals a 5th peak for some samples at 2 450 nm.
-Comparisons to spectra collected on muscovite [@clark_etal90 p.12; @ramanaidou15 p.220] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm.
-It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.
+The most common features in quartz-based material are two peaks located at 1 400 nm and 1 900 nm (fig.
+XX), indicating the presence of molecular water [@hunt77; @sciuto_etal19].
+Apart from these the main features can be found in  the > 2 000 nm range.
+Hydroxyl (OH) groups can produce a response in a number of different fields in the same material, and largely depend on the element they are attached to [@hunt77 pp.
+509-510].
+In a large group of the samples these peaks occur at 2 200 nm and 2 350 nm, with the former often being the more intense of the two.
+A general rule has been that aluminium (Al) contributes to a more intense peak at 2 200 nm, and magnesium (Mg) at 2 300 nm [@hunt77 p.509].
+While this often may be the case there are also exceptions to this rule, and it should thus not be considered a definitive answer [@clark_etal90 p.12664].
+These absorption bands can be found in minerals like muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207].
+The 2 350 nm band is also one of the stronger carbonate (CO) bands, but considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching.
 
@@
 
-```{r nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue.
-Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/colourless samples."}
-
-#Filter NIR data to focus on the material with a dark hue 
-#and select the NIR range 1 000 - 2 500 nm
-dark <- Points.nir %>% 
-  filter(hue == "Dark") %>% 
-  select(c(1:28),c(680:2180))
-
-#create a spectra object for the dark material, and then repeat for light, white and colourless groups
-spec.d <- 
-  as_spectra(dark, name_idx = 1, meta_idxs = 2:28)
-
-light <- Points.nir %>% 
-  filter(hue == "Light") %>% 
-  select(c(1:28),c(680:2180))
-
-spec.l <- 
-  as_spectra(light, name_idx = 1, meta_idxs = 2:28)
-
-colourless <- Points.nir %>% 
-  filter(hue == "Colourless") %>% 
-  select(c(1:28),c(680:2180))
-
-spec.t <- 
-  as_spectra(colourless, name_idx = 1, meta_idxs = 2:28)
-
-white <- Points.nir %>% 
-  filter(hue == "White") %>% 
-  select(c(1:28),c(680:2180))
-
-spec.w <- 
-  as_spectra(white, name_idx = 1, meta_idxs = 2:28)
-
-#Prepare a layout with 1 column and four rows
-par(mfrow=c(4,1))
-
-#plot the mean spectra for each group
-plot(mean(spec.d), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())
-
-plot(mean(spec.l), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - Light")
-
-plot(mean(spec.t), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - Colourless")
-
-plot(mean(spec.w), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xaxp=c(1000,2500,6),
-     xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
-
-pushViewport(vps1)
-X <- convertX(unit(1413,"native"), "npc")
-X2 <- convertX(unit(1935,"native"), "npc")
-X3 <- convertX(unit(2215,"native"), "npc")
-X4 <- convertX(unit(2350,"native"), "npc")
-#For some reason grid.text for peak labels does not always align with the x value for the dashed line, so need to add a specific value for the text
-x1 <- convertX(unit(1436,"native"), "npc")
-x2 <- convertX(unit(1936,"native"), "npc")
-x3 <- convertX(unit(2212,"native"), "npc")
-x4 <- convertX(unit(2335,"native"), "npc")
-popViewport(3)
-
-grid.move.to(x = X, y = unit(1, "npc"), vp = vps1)
-grid.line.to(x = X, y = unit(0, "npc"), vp = vps4, 
-             gp = gpar(col = "gray", lty = 2, lwd = 2))
-grid.text("OH", x = x1, y = unit(0.87, "npc"), just="center", gp=gpar(font=2))
-
-grid.move.to(x = X2, y = unit(1, "npc"), vp = vps1)
-grid.line.to(x = X2, y = unit(0, "npc"), vp = vps4, 
-             gp = gpar(col = "gray", lty = 2, lwd = 2))
-grid.text("OH", x = x2, y = unit(0.93, "npc"), just="center", gp=gpar(font=2))
-
-grid.move.to(x = X3, y = unit(1, "npc"), vp = vps1)
-grid.line.to(x = X3, y = unit(0, "npc"), vp = vps4, 
-             gp = gpar(col = "gray", lty = 2, lwd = 2))
-grid.text("AlOH", x = x3, y = unit(0.95, "npc"), just="center", gp=gpar(font=2))
-
-grid.move.to(x = X4, y = unit(1, "npc"), vp = vps1)
-grid.line.to(x = X4, y = unit(0, "npc"), vp = vps4, 
-             gp = gpar(col = "gray", lty = 2, lwd = 2))
-grid.text("CO3", x = x4, y = unit(0.95, "npc"), just="center", gp=gpar(font=2))
+```{r nir-savitzky-transform, fig.cap= "nir savitzky comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform.R"), echo = FALSE}
+quietly(source("./analysis/paper/003-nir-savitzky-transform.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/003-nir-savitzky-transform.png", error = FALSE)
 ```
@@
 
-```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum.
-Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left.
-A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen.
-Three black samples at the top, from different sites, all feature no distinguishing peaks in their spectra thus separating them from the other samples in the dataset."}
-
-#perform PCA with SNV normalization and mean-center
-nir.pca <-
-  prcomp(Points.nir[,c(680:2180)], center = TRUE, scale = FALSE)
-
-#prepare a basic score plot with fviz_pca_ind using PC1 and PC2
-basic_plot1 <-
-  fviz_pca_ind(nir.pca, axes = c(1,2), label="none")
-
-#prepare labels for PCs
-nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=2)
-nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=2)
-nir.pc1lab <- paste0("PC1 (",as.character(nir.pc1var),"%)")
-nir.pc2lab <- paste0("PC2 (",as.character(nir.pc2var),"%)")
-
-#NIR LOADING LINES
-#par(mfrow=c(3,1))
-#sapply(1:3, function(comp){
-#  plot(1001:2500, nir.pca$rotation[,comp], type='l', lwd=2,
-#      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(1001,2500,6))})
-
-#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
-ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
-             aes(x=x, y=y, 
-                 shape = material, 
-                 fill = munsell_hue)) +
-  geom_point(size=3) +
-  theme_bw() +
-  ggtitle("Near infrared 1 000 - 2 500 nm") +
-  labs(x = nir.pc1lab,
-       y = nir.pc2lab) +
-  scale_shape_manual(name = "Material",
-                     values=pca.shapes) +
-  scale_fill_manual(name = "Munsell hue",
-                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol=2, 
-                              title.position="top", 
-                              title.hjust = 0.5),
-         fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = pca.colors,
-                                                 color = "black",
-                                                 size=3),
-                             title.position="top", title.hjust = 0.5))+
-  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
-        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
-        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
-        legend.title = element_text(size = 12, face = "bold", colour = "black"),
-        legend.background = element_rect(linetype = "solid", color = "black"),
-        legend.position = "bottom")
+```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/004-nir-pca.png", error = FALSE)
 ```
@@
 
-The PCA shows a strong trend along PC 1 (`r nir.pc1var` %) that seems to correlate with the colour of the material(fig.
+The PCA shows a strong trend along PC 1 ( %) that seems to correlate with the colour of the material(fig.
 \@ref(fig:nir-pca)).
@@
 These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
-
-\linebreak
-
-```{r raman-prep-data, results="hide", fig.keep="none"}
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-raman.averaged <- 
-  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
-
-#perform Standard Normal Variate on raman data
-#raman.averaged[,2:466] <- 
-#  standardNormalVariate(raman.averaged[,2:466])
-
-#merge data with metadata
-#raman_merged <- 
-#  as.data.frame(merge(metadata, raman_averaged, by='sample_id'))
-
-#select sample_id and data columns and transpose in prep for baseline de-trend
-raman.transposed <- 
-  raman.averaged %>% 
-    select(sample_id, `92.88`:`2503.59`)
-
-raman.transposed <- 
-  t(raman.transposed)
-
-#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
-#ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
-
-#prepare vector with sample_id for reading raman data into spectra object
-sample_id <- 
-  raman.averaged$sample_id
-
-#read transposed raman data into spectra object
-raman <-matrix2SpectraObject(
-  gr.crit = sample_id,
-  gr.cols = c("auto"),
-  freq.unit ="Wavelength cm-1",
-  int.unit ="Intensity",
-  descrip ="Bifacial points measurements",
-  in.file = "./analysis/data/derived_data/raman_transposed.csv",
-  sep = ";",
-  dec = ".",
-  chk = TRUE,
-  out.file = "./analysis/data/derived_data/raman_spec_object")
-
-#call on relevant baseline method (modified polynomial fitting) from baseline package and 
-#return corrected spectra to spectra object
-raman.baseline <- 
-  baselineSpectra(raman,
-                  int = FALSE,
-                  method = "modpolyfit",
-                  retC = TRUE)
-
-#spectra are stored in "Spectra" object, not a data frame
-#need to extract samples and frequencies from that object 
-raman.tmp <- 
-  cbind(raman.baseline$freq, t(raman.baseline$data))
-#transpose data to get frequencies as columns
-raman.tmp <- 
-  as.data.frame(t(raman.tmp))
-#Add sample_id column and remove the initial "X" character introduced to sample names by R
-raman.tmp$sample_id <- 
-  c("sample_id", sub("^X*", "", raman.baseline$names))
-
-#Make first row headers and remove the first row with the names
-names(raman.tmp) <-
-  raman.tmp[1,]
-raman.tmp <- 
-  raman.tmp[-1,]
-
-#merge raman de-trended data with metadata
-raman.baseline.merged <- 
-  as.data.frame(merge(metadata.csv, raman.tmp, by='sample_id'))
-
-#select and filter raman data to focus on relevant material and artefact types
-Points.raman <-
-  raman.baseline.merged %>%
-  filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-         type == "Point" | type == "Point fragment" | type == "Preform")
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
-```
 
@@
 
-```{r raman-spectra, fig.width=20, fig.height=15, fullwidth = TRUE, fig.cap="Average Raman spectra grouped by their hue.
-The *quartz peak* at 460 cm^-1^ is present in all four groups, but the *graphite peaks* at 1 287  cm^-1^ (D band) and 1 598 cm^-1^ (G band) mainly appear in the dark group"}
-
-r.dark <- Points.raman %>% 
-  filter(hue == "Dark")
-
-spec.rd <- 
-  as_spectra(r.dark, name_idx = 1, meta_idxs = 2:29)
-
-r.light <- Points.raman %>% 
-  filter(hue == "Light")
-
-spec.rl <- 
-  as_spectra(r.light, name_idx = 1, meta_idxs = 2:29)
-
-r.colourless <- Points.raman %>% 
-  filter(hue == "Colourless")
-
-spec.rt <- 
-  as_spectra(r.colourless, name_idx = 1, meta_idxs = 2:29)
-
-r.white <- Points.raman %>% 
-  filter(hue == "White")
-
-spec.rw <- 
-  as_spectra(r.white, name_idx = 1, meta_idxs = 2:29)
-
-#Prepare a layout with 2 columns and 2 rows
-par(mfrow=c(2,2))
-
-#Plot the mean spectrum for each spectra group above with text annotations for the peaks in the first dark spectra
-plot(mean(spec.rd), lwd = 1.2, col = "black", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - Dark")
-text(x=120, y=2900, labels="Quartz", font=2, cex = 1.5)
-text(x=200, y=4300, labels="Quartz", font=2, cex = 1.5)
-text(x=460, y=9250, labels="Quartz", font=2, cex = 1.5)
-text(x=1285, y=6300, labels="D", font=2, cex = 2.0)
-text(x=1600, y=4000, labels="G", font=2, cex = 2.0)
-
-plot(mean(spec.rl), lwd = 1.2, col = "red", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - Light")
-
-plot(mean(spec.rt), lwd = 1.2, col = "purple", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - Colourless")
-
-plot(mean(spec.rw), lwd = 1.2, col = "blue", ylab = "Intensity", cex.lab = 1.5, cex.main = 1.5,
-     xlab = parse(text=paste("Wavenumber (cm^-1)")), main = "Average - White")
+```{r raman-spectra-summary, fig.cap= "raman spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/005-raman-spectra-summary.R"), echo = FALSE}
+quietly(source("./analysis/paper/005-raman-spectra-summary.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/005-raman-spectra-summary.png", error = FALSE)
 ```
@@
 
-```{r raman-pca, fig.width=20, fig.height=15, fig.cap="PCA score plots for the raman spectra.
-The main explanatory power can be found in PC 1, which groups the data based on the quartz and D-G bands.
-As these are the main features in the data, the other components mainly describe the remaining noise.
-PC 3 does, however, separate out the few samples that feature peaks at 700-800 cm^-1^ in the top left (marked with red circle), that may result from Al, Si and OH vibrations."}
-
-#perform PCA with SNV normalization and mean-center
-raman.pca <-
-  prcomp(standardNormalVariate(X = Points.raman[,c(30:321)]), center = TRUE, scale = FALSE)
-
-#prepare different basic score plots with fviz_pca_ind
-basic_plot1 <-
-  fviz_pca_ind(raman.pca, axes = c(1,2), label="none")
-
-basic_plot2 <-
-  fviz_pca_ind(raman.pca, axes = c(1,3), label="none")
-
-#prepare labels for PCs
-raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
-raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
-raman.pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
-raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
-raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
-raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")
-
-#RAMAN LOADING LINES
-#par(mfrow=c(2,2))
-#sapply(1:4, function(comp){
-#  plot(row.names(raman.pca$rotation), raman.pca$rotation[,comp], type='l', lwd=2, 
-#       main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(92.88,1798.3,6))})
-
-#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
-raman.p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
-      aes(x=x, y=y, 
-           shape = material, 
-           fill = munsell_hue)) +
-  geom_point(size=3) +
-  theme_bw() +
-  ggtitle("Raman") +
-  labs(x = raman.pc1lab,
-       y = raman.pc2lab) +
-  scale_shape_manual(name = "Material",
-                     values=pca.shapes) +
-  scale_fill_manual(name = "Munsell hue",
-                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol=2, 
-                              title.position="top", 
-                              title.hjust = 0.5),
-         fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = pca.colors,
-                                                 color = "black",
-                                                 size=3),
-                             title.position="top", title.hjust = 0.5)) +
-  theme(legend.position = "none")
-
-#bind the basic fviz plot for PC 1 and 3, with metadata for visualization, and use as basis for a more customizeable plot in ggpplot
-#In order to be able to filter out certain points that have the 700-800 wavenumber peaks and give them a red circle in a second geom_point statement this 
-#has to be done outside of ggplot statement
-r.data2 <- 
-  cbind(basic_plot2$data, Points.raman[, c(1,8,10)])
-
-#Set up the score plot
-raman.p2 <- 
-  r.data2 %>%  
-  ggplot(aes(x=x, y=y, 
-           shape = material, 
-           fill = munsell_hue)) +
-  geom_point(size=3) +
-  theme_bw() +
-  labs(x = raman.pc1lab,
-       y = raman.pc3lab) +
-  scale_shape_manual(name = "Material",
-                     values=pca.shapes) +
-  scale_fill_manual(name = "Munsell hue",
-                    breaks = c("Colourless", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
-                    values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol=2, 
-                              title.position="top", 
-                              title.hjust = 0.5),
-         fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = pca.colors,
-                                                 color = "black",
-                                                 size=3),
-                             title.position="top", title.hjust = 0.5)) +
-  #Filter out points with 700 - 800 wavenumber peaks and give them a red circle for easier identification
-  geom_point(data = r.data2 %>% 
-               filter(sample_id == 6 | sample_id == 34 | sample_id == 42 | sample_id == 127 | sample_id == 445 | sample_id == 482 | sample_id == 491 | sample_id == 494 | sample_id == 495 | sample_id == 501 |sample_id == 502 | sample_id == 503 | sample_id == 504),
-             pch=21,
-             size=7, 
-             colour ="red",
-             fill = NA,
-             stroke = 2) +
-  theme(legend.position = "none")
-
-#Set the legend for the combined figure of both score plots
-comb <-
-  raman.p1 + raman.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
-                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
-                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
-                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
-                              legend.background = element_rect(linetype = "solid", color = "black"),
-                              legend.position = "bottom")
-
-#Call on the finished figure and layout in 2 rows
-comb +
-  plot_layout(nrow = 2, guides = "collect")
+```{r raman-peak-ident, fig.cap= "raman peak identification", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/006-raman-peak-identification.R"), echo = FALSE}
+quietly(source("./analysis/paper/006-raman-peak-identification.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/006-raman-peak-identification.png", error = FALSE)
 ```
@@
 The resulting PCA model (fig.
-\@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks.
+\@ref(fig:raman-pca)) shows a clear trend along PC 1 ( %) related to the quartz and D - G peaks.
 PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
@@
 As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-When PC 3 (`r raman.pc3var` %) is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
+When PC 3 ( %) is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
 
@@
 
-```{r xrf-data-prep}
-
-#merge XRF data with metadata
-xrf.merged <- 
-  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
-
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
-Points.xrf <-
-  xrf.merged %>%
-    filter(material == "Quartz" | material == "Quartzite" | material == "Brecciated quartz" | material == "Brecciated quartzite" | material  == "Felsitic porphyric quartz",
-           type == "Point" | type == "Point fragment" | type == "Preform")
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
-
-#Filter out Si values above 50%
-high_si <- 
-  Points.xrf %>% 
-    filter(`14Si` > 50)
-
-#Prepare a calculation of percentage measurements above 50% to call on in text
-high_si <- 
-  round(length(high_si$`14Si`) / length(Points.xrf$`14Si`) * 100)
-
-#Prepare XRF data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
-Points_long <-
-  drop_na(Points.xrf %>%
-    filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
-    select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
-    gather(key = "Element", value = "Value"))
-
-#filter XRF data on sample size (only include samples with smallest dimension >= 10mm), 
-#select elements to include in PCA, as well as columns to group by
-xrf <-
-  Points.xrf %>%
-  filter(max_length_mm >= 10 & max_width_mm >= 10) %>% 
-  select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)
-
-#log transform XRF data
-xrf.log <- log(xrf[,5:20])
-
-#impute missing values
-xrf.imp <- imputePCA(xrf.log, ncp = 2, scale = FALSE, method = "Regularized", maxiter = 1000)
-
-#perform pca with mean centering and unit variance scaling
-xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=TRUE)
-
-#Prepare cumulative explained variance of two first XRF PCs
-xrf_variance <- 
-  round(summary(xrf.pca)$importance[2,1]*100, digits=1) + round(summary(xrf.pca)$importance[2,2]*100, digits=1)
-
+```{r raman-pca, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/007-raman-pca.png", error = FALSE)
 ```
@@
 \@ref(fig:xrf-violin-plot)).
-The majority of the readings resulted in a ^14^Si content of less than 50%, and about ¼ of the readings (~ `r high_si`%) featured a ^14^Si content closer to 90%.
+The majority of the readings resulted in a ^14^Si content of less than 50%, and about ¼ of the readings (~ %) featured a ^14^Si content closer to 90%.
 The “low” ^14^Si group also feature a Bal between 40 – 70%.
 
-The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ `r high_si`%.
+The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ %.
 The loadings can be be seen in figure \@ref(fig:xrf-loadings).
@@
 
-```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of percentage element content in quartzite samples measured using pXRF.
-**n** represents no.
-of measurements above Limit of Detection (LOD) for that element out of a total of 474 measurements.
-Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
-
-# Add column with elemental groups for visualisation
-Points_long <-
-  Points_long %>%
-    mutate(Group = case_when(
-      endsWith(Element, "Mg") ~ "Lithophile",
-      endsWith(Element, "Al") ~ "Lithophile",
-      endsWith(Element, "Si") ~ "Lithophile",
-      endsWith(Element, "P") ~ "Lithophile",
-      endsWith(Element, "S") ~ "Chalcophile",
-      endsWith(Element, "K") ~ "Lithophile",
-      endsWith(Element, "Ca") ~ "Lithophile",
-      endsWith(Element, "V") ~ "Lithophile",
-      endsWith(Element, "Mn") ~ "Siderophile",
-      endsWith(Element, "Fe") ~ "Siderophile",
-      endsWith(Element, "Zn") ~ "Chalcophile",
-      endsWith(Element, "Sr") ~ "Lithophile",
-      endsWith(Element, "Y") ~ "Lithophile",
-      endsWith(Element, "Zr") ~ "Lithophile",
-      endsWith(Element, "Ba") ~ "Lithophile",
-      endsWith(Element, "Ti") ~ "Chalcophile",
-    ))
-
-#Calculate and add value equal to 90% of the value range for each element, this is in preparation for 
-#text annotations in final plot where a text will be positioned at this values height in each plot
-Points_long <- 
-  Points_long %>% 
-    group_by(Element) %>% 
-    dplyr::mutate(ypos = quantile(Value, 0.95))
-
-#Prepare colors
-loading.colors <- c("lightgoldenrod1", "darkorange", "lightpink1")
-
-#Lookup for element labeller
-#Ändra ordning så att de mer avvikande (Si, Mn etc) hamnar uppe till vänster
-Elements <- c(
-  '12Mg' = "Mg",
-  '13Al' = "Al",
-  '14Si' = "Si",
-  '15P' = "P",
-  '16S'= "S",
-  '19K' = "K",
-  '20Ca' ="Ca",
-  '23V' = "V",
-  '25Mn' = "Mn",
-  '26Fe' = "Fe",
-  '30Zn' = "Zn",
-  '38Sr' = "Sr",
-  '39Y' = "Y",
-  '40Zr' = "Zr",
-  '56Ba' = "Ba",
-  '81Ti' = "Ti"
-  )
-
-#Prepare data frame with elements for annotating facets in final plot
-na_values <- 
-  data.frame("Element" = unique(Points_long$Element))
-
-#Calculate and add number of missing values for each element             
-na_values <- 
-  na_values %>% 
-    mutate(na_values = c(
-      paste("n = ", length(which(!is.na(Points.xrf$`12Mg`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`13Al`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`14Si`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`15P`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`16S`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`19K`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`20Ca`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`23V`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`25Mn`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`26Fe`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`30Zn`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`38Sr`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`39Y`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`40Zr`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`56Ba`)))),
-      paste("n = ", length(which(!is.na(Points.xrf$`81Ti`)))))
-    )
-
-#Merge na_values and Points_long data frames in order to get the y position that the text annotation should be displayed
-#in for each element facet in final plot
-na_values <- 
-  merge(na_values, unique(Points_long[,c("Element", "ypos")]), by = "Element", all.x = TRUE)
-
-#Prepare ylimits for each group in order to exclude extreme outliers in the violin plot graphic
-ylims <- Points_long %>%
-  group_by(Element) %>% 
-  dplyr::summarise(y1 = 0, y2 = quantile(Value, 0.95))
-
-#Prepare a list with formula for setting new ylimits for each group in ggh4x package "facetted_pos_scales() which makes it possible to work with facet_wrap()
-scales <- c(
-  Element == '12Mg' ~ scale_y_continuous(limits = c(as.vector(t(ylims[1,2:3])))),
-  Element == '13Al' ~ scale_y_continuous(limits = c(as.vector(t(ylims[2,2:3])))),
-  Element == '14Si' ~ scale_y_continuous(limits = c(as.vector(t(ylims[3,2:3])))),
-  Element == '15P' ~ scale_y_continuous(limits = c(as.vector(t(ylims[4,2:3])))),
-  Element == '16S' ~ scale_y_continuous(limits = c(as.vector(t(ylims[5,2:3])))),
-  Element == '19K' ~ scale_y_continuous(limits = c(as.vector(t(ylims[6,2:3])))),
-  Element == '20Ca' ~ scale_y_continuous(limits = c(as.vector(t(ylims[7,2:3])))),
-  Element == '23V' ~ scale_y_continuous(limits = c(as.vector(t(ylims[8,2:3])))),
-  Element == '25Mn' ~ scale_y_continuous(limits = c(as.vector(t(ylims[9,2:3])))),
-  Element == '26Fe' ~ scale_y_continuous(limits = c(as.vector(t(ylims[10,2:3])))),
-  Element == '30Zn' ~ scale_y_continuous(limits = c(as.vector(t(ylims[11,2:3])))),
-  Element == '38Sr' ~ scale_y_continuous(limits = c(as.vector(t(ylims[12,2:3])))),
-  Element == '39Y' ~ scale_y_continuous(limits = c(as.vector(t(ylims[13,2:3])))),
-  Element == '40Zr' ~ scale_y_continuous(limits = c(as.vector(t(ylims[14,2:3])))),
-  Element == '56Ba' ~ scale_y_continuous(limits = c(as.vector(t(ylims[15,2:3])))),
-  Element == '81Ti' ~ scale_y_continuous(limits = c(as.vector(t(ylims[16,2:3]))))
-)
-
-#Violin plot of elemental content, trimmed, with fill based on elemental groups and number of values under LOD as text annotation
-Points_long %>%
-    ggplot(aes(x=Element, y=Value, fill = Group)) +
-    geom_violin(trim = TRUE) +
-    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
-    scale_fill_manual(values=loading.colors) +
-    facetted_pos_scales(y = scales) +
-    theme_bw() +
-    geom_text(data=na_values, aes(x=0.2, y=ypos, label=na_values, fontface = "bold"), nudge_x = 0.4, inherit.aes = FALSE) +
-    theme(
-      legend.title = element_text(size = 12, face = "bold", colour = "black"),
-      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
-      axis.title.x = element_blank(),
-      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-      axis.title.y=element_blank(),
-      axis.text.x=element_blank(),
-      axis.ticks.x=element_blank())
+```{r xrf-violin-plot, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/008-xrf-violin-plot.png", error = FALSE)
 ```
@@
 
-```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="Loading vectors (P1-P2) with scores for each variable (element)"}
-
-#Extract loadings from xrf.pca
-loadings <- as.data.frame(xrf.pca$rotation[,1:2])
-#As elements are stored as rownames, set them as first column so it can be used as a variable
-loadings <- setDT(loadings, keep.rownames = "Element")[]
-#Melt loadings from wide to long format
-loadings.melted <- melt(loadings, id.vars="Element")
-#Set names of columns
-loadings <- setNames(loadings.melted, c("Element", "PC", "Value"))
-
-#bar graph of the loadings (P1-P2) for the XRF data
-loadings %>%
-  ggplot(aes(x=fct_rev(Element), y=Value)) +
-  geom_bar(
-    stat = "identity", position = "identity",
-    color = "black", fill = "darkgrey",
-    width = 0.5) +
-  facet_wrap( ~ PC, scales = "free") +
-  theme_bw() +
-  theme(plot.margin = margin(10,10,10,10),
-        axis.title.y = element_blank(),
-        axis.text.y = element_text(size=10, face="bold", colour = "black"),
-        axis.title.x = element_blank(),
-        strip.text.x = element_text(size=14, face="bold", colour="black")) +
-  coord_flip()
+```{r xrf-score-plot, fig.cap= "xrf score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca.png", error = FALSE)
 ```
@@
 
-```{r xrf-score-plot, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of PC 1 and PC 2.
-The smaller group to the left in the plot includes samples with higher Si content, while the larger group includes samples with Si content < 50%."}
+# O2PLS
+ \linebreak
+ \linebreak
 
-#XRF score plot
-ggbiplot(xrf.pca,
-         choices=c(1,2),
-         obs.scale = 1,
-         var.scale = 1,
-         group=NULL,
-         varname.size = 4,
-         labels.size=3,
-         ellipse = FALSE,
-         circle = FALSE) +
-theme_minimal() +
-geom_point(aes(shape=xrf$material, 
-               fill = xrf$munsell_hue), 
-           size = 3) +
-coord_cartesian() +
-scale_shape_manual(name = "Material",
-                   values=pca.shapes) +
-scale_fill_manual(name = "Munsell hue",
-                  breaks = pca.munsell_hue,
-                  values=pca.colors) +
-guides(shape = guide_legend(override.aes = list(fill = "black"),
-                            ncol=2,
-                            title.position="top",
-                            title.hjust = 0.5),
-       fill = guide_legend(override.aes = list(shape = 22,
-                                               fill = pca.colors,
-                                               color = "black",
-                                               size=3),
-                           title.position="top", title.hjust = 0.5)) +
-theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
-      axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
-      axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
-      legend.title = element_text(size = 12, face = "bold", colour = "black"),
-      legend.background = element_rect(linetype = "solid", color = "black"),
-      legend.position = "bottom")
+```{r o2pls, fig.cap= "o2pls score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/010-O2PLS.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/010-O2PLS.png", error = FALSE)
 ```
-
-\linebreak
-<!--
-```{r xrf-pca2, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 3rd components.
-The same groupings of low/high Si content remains along PC1, but PC3 seems to separate darker and light samples further in the low Si group."}
-ggbiplot(xrf.pca,
-         choices=c(1,3),
-         obs.scale = 1,
-         var.scale = 1,
-         group=NULL,
-         varname.size = 4,
-         labels.size=3,
-         ellipse = FALSE,
-         circle = FALSE) +
-theme_minimal() +
-geom_point(aes(shape=xrf$material, 
-               fill = xrf$munsell_hue), 
-           size = 3) +
-coord_cartesian() +
-scale_shape_manual(name = "Material",
-                   values=pca.shapes) +
-scale_fill_manual(name = "Munsell hue",
-                  breaks = pca.munsell_hue,
-                  values=pca.colors) +
-guides(shape = guide_legend(override.aes = list(fill = "black"),
-                            ncol=2,
-                            title.position="top",
-                            title.hjust = 0.5),
-       fill = guide_legend(override.aes = list(shape = 22,
-                                               fill = pca.colors,
-                                               color = "black",
-                                               title.position="top", 
-                                               title.hjust = 0.5))) +
-theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
-      axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
-      axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
-      legend.title = element_text(size = 12, face = "bold", colour = "black"),
-      legend.background = element_rect(linetype = "solid", color = "black"),
-      legend.position = "bottom")
-```
--->
 

### 2022-05-23 analysis/paper/010-X-O2PLS.R
Updated metadata

@@ -1,0 +1,101 @@
+#Load packages
+suppressPackageStartupMessages(library(OmicsPLS))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220407.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
+
+#Import nir data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[,4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#merge NIR data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
[…]

### 2022-05-23 analysis/paper/paper.Rmd
Updated paper knitting

@@
 Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
-knitr::include_graphics('../figures/site_distribution.jpeg', error = FALSE)
+knitr::include_graphics('../figures/site_distribution.png', error = FALSE)
 ```

### 2022-06-21 analysis/paper/001-nir-spectra-summary.R
New and updated subscripts

@@ -94,7 +94,7 @@ p.c <-
 #plot the dark spectra
 p.d <-
   ggplot(Points.d, aes(x = as.numeric(Wavelength))) + 
-  geom_line(aes(y = Absorbance, colour = "", group = sample_id), size = 1, stat = "identity") +
+  geom_line(aes(y = Absorbance, colour = "", group = sample_id), size = 0.5, stat = "identity") +
   xlab("Wavelength (nm)") +
   ylab("Absorbance") +
   scale_color_manual(name = "Dark",
@@ -144,7 +144,7 @@ ggsave("001-nir-spectra-summary.png",
        fig,
        device = "png",
        here::here("analysis/figures/"),
-       width=20, 
+       width=25, 
        height=20,
        units = "cm",
        dpi = 300)

### 2022-06-21 analysis/paper/002-nir-peak-summary.R

@@ -1,0 +1,163 @@
+# Loading package
+library(gridBase)
+library(spectrolab)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+Points.nir[,1] <- as.numeric(Points.nir[,1]) 
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+                  site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+                  site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+                  site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-06-21 analysis/paper/003-nir-savitzky-transform.R

@@ -90,10 +90,10 @@ p.d258 <-
         legend.title = element_text(size = 12, face = "bold", colour = "black"))
 
 #Perform Savitzky-Golay 2nd derivative on sample 258
-sg.d <- as.data.frame(gapDer(X = Points.d258[,2:1501], m = 2, w = 11, s = 5))
+sg.d <- as.data.table(gapDer(X = Points.d258[,2:1501], m = 2, w = 11, s = 5))
 
 #Transpose the data for ggplot
-sg.d.long <- reshape2::melt(sg.d, variable.name = "Wavelength", value.name = "Absorbance")
+sg.d.long <-  suppressWarnings(melt(setDT(sg.d), variable.name = "Wavelength", value.name = "Absorbance", variable.factor = FALSE))
 
 #plot the transformed dark spectra
 p.sg.d <-

### 2022-06-21 analysis/paper/009-xrf-pca.R

@@ -61,7 +61,7 @@ xrf.pc2lab <- paste0("PC2 (",as.character(xrf.pc2var),"%)")
 xrf.pc3lab <- paste0("PC3 (",as.character(xrf.pc3var),"%)")
 
 #prepare color/fill and symbols for score plots
-pca.colors <- c("#D62728FF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
+pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
 pca.hue <- c("Colourless", "Dark", "Light", "White")
 
 #Show eigenvalue scree plot
@@ -95,25 +95,43 @@ fig <-
                 fill.var = "red", 
                 repel = TRUE,
                 pointshape = 21, 
-                pointsize = 3) +
+                pointsize = 4) +
   theme_bw() +
   labs(x = xrf.pc1lab,
        y = xrf.pc2lab) +
-  geom_point(aes(fill = xrf$hue), shape=21, size = 3) +
-  scale_fill_manual(name = "Hue", 
-                    values = pca.colors) +
+  geom_point(aes(fill = xrf$hue, 
+                 shape = factor(xrf$material)), 
+             size = 3) +
+  scale_shape_manual(name = "Material", 
+                     values=c(24,22,21),
+                     guide = guide_legend(override.aes = list(fill = "black"),
+                                          title.position="top",
+                                          title.hjust = 0.5,
+                                          order = 1)) +
+  scale_color_manual(values = pca.colors,
+                    guide = "none") +
+  scale_fill_manual(name = "Hue",
+                    values = pca.colors,
+                    guide = guide_legend(override.aes = list(shape = 21,
+                                                             fill = pca.colors,
+                                                             color = "black",
+                                                             size=3),
+                                         title.position="top",
+                                         title.hjust = 0.5,
+                                         order = 2)) +
   ggnewscale::new_scale_fill() +
   stat_chull(aes(fill = xrf$cluster),
              alpha = 0.3, 
              geom = "polygon") + 
   scale_fill_manual(name = "Cluster", 
-                    values = c( "#0072B2", "#D55E00", "#CC79A7", "#009E73")) +
+                    values = c( "#0072B2", "#D55E00", "#CC79A7", "#009E73"),
+                    guide_legend(order = 1)) +
   theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
[…]

### 2022-06-21 analysis/paper/010-O2PLS-model-fit.R

@@ -1,4 +1,6 @@
 #Load packages
+suppressPackageStartupMessages(library(grid))
+suppressPackageStartupMessages(library(gridExtra))
 suppressPackageStartupMessages(library(OmicsPLS))
 library(parallel)
 suppressPackageStartupMessages(library(rgr))
@@ -85,87 +87,49 @@ Points.nir[,30:2180] <-
 xrf.merged <- 
   as.data.frame(merge(Points.xrf, nir.averaged, by='sample_id'))
 
-#fit the o2pls model, in this case MSE of 1.95734 was suggested with n=2, nx=3, and ny=4
+#perform cross-validation to estimate number of components to fit to the o2pls model
+cval <- 
+  crossval_o2m_adjR2(Points.nir[,681:2180], Points.xrf[,2:10], 1:4, 0:4, 0:4,
+                     nr_folds = 5, nr_cores = detectCores())
+
+#fit the o2pls model, in this case the second row provided by cval was used as at least 2 joint components is desirable 
 fit0 = o2m(Points.nir[,681:2180], Points.xrf[,2:10], 2, 3, 4)
 
-#plot the joint scores of o2pls model
-p.x <- 
-  #score plot of joint X scores 
-  qplot(x=fit0$Tt[,1],
-        y=fit0$Tt[,2],
-        label = Points.nir$sample_id) +
-  theme_bw() +
-  labs(x = "Joint X - Component 1",
-       y = "Joint X - Component 2") +
-  geom_point(aes(shape = Points.nir$material, 
-                 fill = Points.nir$hue), 
-             size = 4) +
-  geom_text(hjust=0,
-            vjust=-1,
-            size=3) +
-  scale_color_manual(values = c("purple", "black","red","blue")) +
-  scale_shape_manual(name = "Material", 
-                     values=c(24,22,21)) +
-  scale_fill_manual(name = "Hue", 
-                    values = c("purple", "black","red","blue")) +
-  guides(color = "none",
-    shape = guide_legend(override.aes = list(fill = "black"),
-                              title.position="top",
-                              title.hjust = 0.5,
-                         order = 2),
-    fill = guide_legend(override.aes = list(shape = 21,
-                                                  fill = c("purple", "black","red","blue"),
-                                                  color = "black",
-                                                  size=3),
-                        title.position="top",
[…]

### 2022-06-21 analysis/paper/paper.Rmd
Updated draft

@@
 
-# Introduction
-
-Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF).
-The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
-Although the petrographic characterization of the material mainly establishes a point of geological origin for the object it is still an important first step in reconstructing the *chaîne opératoire*.
-Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these.
-The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
-Prieto et al.
-[-@prieto_etal19 p.
-15] highlights the current bias that exists in the understanding of raw material management in regions where the local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions].
-In northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region.
-While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-Although flint and slate have featured in discussions of trade and exchange [@halen94 pp.
-115-122; @lundberg97 pp.
-161-171] within the regon, another reason for the lack of methodological development in North Swedish research could be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
-The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape.
-The tentative interpretations of lithic material sourcing within northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
-This does not account for the possibility that the material sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
-Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.
-
-Prieto et al.
-[-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
-It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05].
-Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
-In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable.
-Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.
-While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information.
-Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes.
-While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
-The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19].
-The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].
-
-\linebreak
-
-## Aims
-
-In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF).
-The aim is to evaluate their viability for non-destructive fingerprinting of quartz and quartzite material in the North Swedish region.
-The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use.
-In contrast to non-local material such as flint, determining the degree to which cultural and/or environmental factors have influenced the distribution pattern of quartz and quartzite artefacts is difficult.
-While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for spectral comparison.
-The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes.
-Although models for settlement and mobility can be used to test ideas of material distribution as influenced by cultural processes [e.g.
-@forsberg85], it is also necessary to attempt to estimate the degree of environmental impact on this pattern.
-
-In order to engage in any meaningful analysis of the above settlement pattern, however, it is first necessary to distinguish between different quartz and quartzite groups.
-In doing so the distribution of different material groups can be spatially analyzed, and the above models for settlement and mobility tested.
-In this study Principal Component Analysis (PCA) is used in order to group the materials according to their spectral features and their contributions to the Principal Components (PC) [@geladi_linderholm20].
-Any clustering or trends observed in the PCA should relate to the chemical characteristic of the material, and thus potentially similar geological origin.
-The results form the basis of a discussion of the viability of using non-destructive spectroscopy in petrographic characterization of quartz and quartzite material, and thus future efforts in the development of models for raw material sourcing in Northern Sweden.
-
-\linebreak
-
-# Material
 ```{r sample-count, results='hide'}
@@
 
-#Create a vector containing all unique sample ID, in order to count number of total samples (incl. types other than points/preforms or materials) 
-tot_uniq <- 
+#Create a vector containing all unique sample ID
+tot_uniq_samples <- 
   unique(tot_samples$sample_id)
 
+#Create a vector containing all unique site ID
+tot_uniq_sites <-
+  unique(tot_samples$site_id)
+
+#Create a vector containing all unique sample ID made in breacciated quartz
 bquartz <- unique(tot_samples %>% 
@@
 
+#Create a vector containing all unique sample ID made in quartz
 quartz <- unique(tot_samples %>% 
@@
 
+#Create a vector containing all unique sample ID made in quartzite
 quartzite <- unique(tot_samples %>% 
@@
 
+# Introduction
+
+Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF).
+The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
+Although the petrographic characterization of the material mainly establishes a point of geological origin for the object it is still an important first step in reconstructing the *chaîne opératoire*.
+Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these.
+The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
+Prieto et al.
+[-@prieto_etal19 p.
+15] highlights the current bias that exists in the understanding of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions].
+In northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region.
+While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
+While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp.
+115-122; @lundberg97 pp.
+161-171], there has been little development of on the topic of quartz and quartzite.
+A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
+The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape.
+The tentative interpretations of lithic material sourcing within northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
+Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.
+
+Prieto et al.
+[-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
+It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05].
+Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
+In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable.
+Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.
+While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information.
+Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes.
+While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
+The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
+The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].
+
+In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF).
+The study includes `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig.
+\@ref(fig:nir-spectra-summary)).
+The aim is to chemically characterize the quartzite points recovered from the sites using the above mentioned suite of instrumenations, and assess the potential of non-destructive techniques for producing reliable fingerprints of different types of quartzite.
+
+<!--
+The study is a part of a larger project wherein models of settlement and mobility are being evaluated and developed in regards to raw material procurement and use.
+In contrast to non-local material such as flint, determining the degree to which cultural and/or environmental factors have influenced the distribution pattern of quartz and quartzite artefacts is difficult.
+While there is a good understanding of the geological makeup of the Scandes [@stephens_bergmanweihed20], the quartzite itself remains largely uncharacterized, with few reference samples available for spectral comparison.
+The glacial impact on the landscape further compounds the issue as geological material has been transported and deposited by glacial movements, as well as subsequent postglacial processes.
+Although models for settlement and mobility can be used to test ideas of material distribution as influenced by cultural processes [e.g.
+@forsberg85], it is also necessary to attempt to estimate the degree of environmental impact on this pattern.
+
+In order to engage in any meaningful analysis of the above settlement pattern, however, it is first necessary to distinguish between different quartz and quartzite groups.
+In doing so the distribution of different material groups can be spatially analyzed, and the above models for settlement and mobility tested.
+In this study Principal Component Analysis (PCA) is used in order to group the materials according to their spectral features and their contributions to the Principal Components (PC) [@geladi_linderholm20].
+Any clustering or trends observed in the PCA should relate to the chemical characteristic of the material, and thus potentially similar geological origin.
+The results form the basis of a discussion of the viability of using non-destructive spectroscopy in petrographic characterization of quartz and quartzite material, and thus future efforts in the development of models for raw material sourcing in Northern Sweden.
+-->
+
+\linebreak
+
+# Material
+
 \linebreak
@@
 
-The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites along the river of Ångermanälven, Västerbotten County, Sweden.
-The study area ranges from the lake of Kultsjön in the southwest of the County to the town of Åsele in the southeast.
-A dataset of `r length(tot_uniq)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
-The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent.
+The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden.
+The sites are distributed along the Ångermanälven river valley, from the lake Kultsjön in the north, down towards the town of Åsele in the south.
+A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
+The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint and some being almost fully translucent.
 The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).
+
 A number of these points have, according to prior classifications, been worked using bifacial technology.
-While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89].
-Previous research suggests a period of use that falls between ca.
+While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca.
 3 800 - 2 000 BP [@kaeck09 p.60].
@@
 
+<!--
 A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g.
@@
 Vilhelmina parish in particular has been the target for numerous excavations and survey projects during this period.
+-->
+ \linebreak
+
+```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine, with The tectonic  Made with ©SGU, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
+knitr::include_graphics('../figures/site_distribution_soils_annotated.png', error = FALSE)
+```
 
@@
 
-```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.
-Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
-knitr::include_graphics('../figures/site_distribution.png', error = FALSE)
-```
-
-\linebreak
-The dataset includes three main types of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe.
+The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe.
 "*brecciekvarts*").
-Those that had no prior classification available were sorted into either the quartz or quartzite category.This distribution of samples into these categories are as follows: `r length(bquartz)` brecciated quartz, `r length(quartz)` quartz, and `r length(quartzite)` quartzite samples.
+Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection.
+This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples.
 The inclusion of quartz and brecciated quartz in the study serves two purposes.
-The first being to test the potential for separating the material types based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
-In his study of the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] analysed artefacts from sites located along the Ume river valley.
-He briefly touches upon the definition of *brecciated quartz* (also referred to as *mylonite quartz*) and describes the material as having formed from a quartz vein which has fractured and then recrystallized via pressure.
-While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz.
-*Brecciated quartz* in particular is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
+The first being to test the potential for separating quartz from quartzitebased on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
+When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
+Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis.
+*Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
 @holm91 p.24].
@@
 
-Pre-processing of the spectra is different for each instrumentation and based on relevant literature.
-After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20].
-PCA has proven useful for exploring complex spectroscopic datasets [@sciuto19], and identifying clusters and trends that relate to the spectral fingerprints of the material.
-These groupings can then be tested with supervised methods, and in the construction of classification models.
-Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12].
-The compositional data add up to a constant (100%), which is described as the *closed sum* effect.
-Correlation analysis is directly affected by this *closure* effect and therefore not suitable.
-With suitable preprocessing the PCA model can reveal the structure and relations between the variables (elements) more accurately.
-PCA modelling in this case was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].
+Pre-processing of the data is different for each instrumentation and based on relevant literature.
+
 \linebreak
 
-## Near Infrared
+## Instrumentaion and pre-processing
+
+\linebreak
+ \linebreak
+
+### Near Infrared
 NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
@@
 
-## Raman
+### Raman
 Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01].
@@
 After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data.
-This was achieved using modified polynomial fitting in R [@Rteam17] with the ChemoSpec package [@hanson16].
+This was achieved using modified polynomial fitting in R [@Rteam21] with the ChemoSpec package [@hanson16].
 The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA.
@@
 
-## X-Ray Fluourescence
+### X-Ray Fluourescence
 XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method for petrographic analysis in archaeology [@hall60; @shackley11a].
@@
 Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.
+
+Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12].
+The compositional data add up to a constant (100%), which is described as the *closed sum* effect.
+Correlation analysis is directly affected by this *closure* effect and therefore not suitable.
+With suitable preprocessing the PCA model can reveal the structure and relations between the variables (elements) more accurately.
+
+\linebreak
+
+## PCA
+
+The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20].
+PCA has proven useful for exploring complex spectroscopic datasets [@sciuto18], and identifying clusters and trends that relate to the spectral fingerprints of the material.
+These groupings can then be followed up with supervised methods, and in the construction of classification models.
+
+PCA analysis was performed using the base R *prcomp()* function [@Rteam21; @rcoreteam21].
+
+\linebreak
+
+## O2PLS
+
+O2PLS is a multivariate regression method used to integrate data from different sources into a single model [@trygg02].
+It builds on the Orthogonal Partial Least Squares (OPLS) method, which was developed to correct for systematic noise in a single dataset and identify latent structures [@bylesjo_etal07].
+The O2PLS method takes the systematic variation in two datasets and divides it into two parts, joint variation between the two blocks of data, and the orthogonal variation that is block specific.
+This makes it possible to test the potential overlap between separate instrumentation, in this case NIRS and XRF.
+
+O2PLS was implemented in R [@rcoreteam21] using the *OmicsPLS* package [@bouhaddani_etal18].
+The data was pre-processed according to the above descriptions, and then cross-validation was used in order to determine the number of components to fit to the model.
 
@@
 
-It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the abscence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77].
-Similarly it can be difficult to work with dark materials as most of the light will be absorbed, thus obscuring any potential spectral features.
+It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77].
+Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, thus obscuring any potential features in the spectra.
 This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group.
 The dark group generally features less distinct peaks than the material with a lighter hue.
-This is not to say that all spectra collected on dark quartzite features less information, but any model including them will affect the clustering of the dataset as a whole.
-Filtering out those spectra that feature no peaks or mainly noise leaves `r length(spec.d)` samples in the dark group.
-Excluding the noisier spectra, and those with no discernable peaks, there is still a noticeable difference between the dark and light group.
-By applying 2^nd^ derivative Savitszky-Golay transformation on the dark spectra broad spectral features can be filtered out [@chryssikos_gates pp.78-81].
-Comparing the centered spectra to the derivative it becomes clear that there are significant peaks that may be of relevance to the characterization of the material (fig.
+This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.
+Transforming the data collected on the dark samples using derivatives can help in the evaluation of the spectra.
+Savitzky-Golay filtering (SG) is a smoothing technique that has been widely applied in spectroscopy (although there has been recent suggestions for adaptions, see [@schmid_etal22]).
+SG filters out broad spectral features, which is beneficial in NIRS, but may also amplify any existing noise in the spectra [@chryssikos_gates17 pp.78-81; @mark_workman21 pp.
+351-397].
+Comparing one of the centered spectra from the dark group to its derivative counterpart it becomes clear that there indeed are significant peaks that may be of relevance to the characterization of the material, despite the high absorbance (fig.
 \@ref(fig:nir-savitzky-transform)).
-
-\linebreak
+Filtering out those spectra that feature no discernible peaks or mainly noise leaves `r length(spec.d)` samples in the dark group.
+There is still, however, a significant difference between the dark and light group, as can be seen by comparing fig.
+\@ref(fig:nir-spectra-summary) and fig.
+\@ref(fig:nir-savitzky-transform):A.
 
@@
 \linebreak
- 
-```{r nir-spectra-dark, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-dark-filtered-summary.R"), echo = FALSE}
-quietly(source("./analysis/paper/002-nir-dark-filtered-summary.R", encoding = "UTF-8"))
-knitr::include_graphics('../figures/002-nir-dark-filtered-summary.png', error = FALSE)
-```
-
-\linebreak
-
-The most common features in quartz-based material are two peaks located at 1 400 nm and 1 900 nm (fig.
-XX), indicating the presence of molecular water [@hunt77; @sciuto_etal19].
-Apart from these the main features can be found in  the > 2 000 nm range.
-Hydroxyl (OH) groups can produce a response in a number of different fields in the same material, and largely depend on the element they are attached to [@hunt77 pp.
-509-510].
-In a large group of the samples these peaks occur at 2 200 nm and 2 350 nm, with the former often being the more intense of the two.
-A general rule has been that aluminium (Al) contributes to a more intense peak at 2 200 nm, and magnesium (Mg) at 2 300 nm [@hunt77 p.509].
-While this often may be the case there are also exceptions to this rule, and it should thus not be considered a definitive answer [@clark_etal90 p.12664].
-These absorption bands can be found in minerals like muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207].
-The 2 350 nm band is also one of the stronger carbonate (CO) bands, but considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching.
 
@@
 
-```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca.png", error = FALSE)
+The most common features in quartz-based material are two peaks located at 1 400 nm and 1 900 nm (fig.
+\@ref(fig:nir-peak-summary)), indicating the presence of molecular water [@hunt77; @sciuto_etal19].
+Apart from these the main features can be found in  the > 2 000 nm range.
+Hydroxyl (OH) groups can produce a response in a number of different fields in the same material, and largely depend on the element they are attached to [@hunt77 pp.
+509-510].
+In a large group of the samples these peaks occur at 2 200 nm and 2 350 nm, with the former often being the more intense of the two.
+A general rule in NIRS has been that aluminium (Al) contributes to a more intense peak at 2 200 nm, and magnesium (Mg) at 2 300 nm [@hunt77 p.509].
+While this often may be the case there are also exceptions to this rule, and it should thus not be considered a definitive answer [@clark_etal90 p.12664].
+These absorption bands can be found in minerals like muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207].
+The 2 350 nm band is also one of the stronger carbonate (CO) bands, but considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching in this case.
+
+\linebreak
+
+```{r nir-peak-summary, fig.cap= "nir peak summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
+quietly(source("./analysis/paper/002-nir-peak-summary.R", encoding = "UTF-8"))
+knitr::include_graphics('../figures/002-nir-peak-summary.png', error = FALSE)
 ```
@@
 These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
+
+\linebreak
+
+```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/004-nir-pca.png", error = FALSE)
+```
 
@@
 
-```{r o2pls, fig.cap= "o2pls score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS.R"), echo = FALSE}
+The cross-validation was performed using the *crossval_o2m_adjR2()* function from the *OmicsPLS* package, and based on the results it was decided to use 2 joint components, 3 X specific (NIRS block), and 4 Y specific (XRF block).
+The data was then fitted to the O2PLS model, with summary statistics of the joint and orthogonal variation in table \@ref(tab:o2pls-summary).
+
+\linebreak
+
+```{r o2pls-summary, fig.width=8, fig.height=10, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS-model-fit.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/010-O2PLS-model-fit.R", encoding = "UTF-8")))
+#setup a tempd data.frame with the image of the table, and then us kable to give it a table caption
+temp.df = data.frame(image="![](../figures/010-O2PLS-model-fit.png)")
+temp.mat <- as.matrix(temp.df)
+colnames(temp.mat) <- NULL
+knitr::kable(temp.mat, caption="A summary of the O2PLS estimated variation for the X (NIRS) and Y (XRF) data")
+```
+
+```{r o2pls-plot, fig.cap= "o2pls score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS.R"), echo = FALSE}
 quietly(eval(parse("./analysis/paper/010-O2PLS.R", encoding = "UTF-8")))

### 2022-06-30 analysis/paper/001-nir-spectra-summary.R
Updated subscripts

@@ -39,52 +39,53 @@ Points.nir <-
 
 #Filter NIR data to focus on colourless 
 #and select the NIR range 1 000 - 2 500 nm
-Points.c <- Points.nir %>%
+n.colourless <- Points.nir %>%
   dplyr::filter(hue == "Colourless") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.c <- 
-  suppressWarnings(melt(setDT(Points.c), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
+n.colourless.long <- 
+  suppressWarnings(melt(setDT(n.colourless), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with dark hues 
 #and select the NIR range 1 000 - 2 500 nm
-Points.d <- Points.nir %>%
+n.dark <- Points.nir %>%
   dplyr::filter(hue == "Dark") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.d <- 
-  suppressWarnings(melt(setDT(Points.d), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
+n.dark.long <- 
+  suppressWarnings(melt(setDT(n.dark), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with light hues 
 #and select the NIR range 1 000 - 2 500 nm
-Points.l <- Points.nir %>%
+n.light <- Points.nir %>%
   dplyr::filter(hue == "Light") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.l <- 
-  suppressWarnings(melt(setDT(Points.l), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
+n.light.long <- 
+  suppressWarnings(melt(setDT(n.light), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with white hues 
 #and select the NIR range 1 000 - 2 500 nm
-Points.w <- Points.nir %>%
+n.white <- Points.nir %>%
   dplyr::filter(hue == "White") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.w <- 
[…]

### 2022-06-30 analysis/paper/002-nir-dark-filtered-summary.R

@@ -42,52 +42,53 @@ Points.nir <-
 
 #Filter NIR data to focus on colourless 
 #and select the NIR range 1 000 - 2 500 nm
-Points.c <- Points.nir %>%
+n.colourless <- Points.nir %>%
   dplyr::filter(hue == "Colourless") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.c <- 
-  suppressWarnings(melt(setDT(Points.c), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
+n.colourless.long <- 
+  suppressWarnings(melt(setDT(n.colourless), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with dark hues 
 #and select the NIR range 1 000 - 2 500 nm
-Points.d <- Points.nir %>%
+n.dark <- Points.nir %>%
   dplyr::filter(hue == "Dark") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.d <- 
-  suppressWarnings(melt(setDT(Points.d), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
+n.dark.long <- 
+  suppressWarnings(melt(setDT(n.dark), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with light hues 
 #and select the NIR range 1 000 - 2 500 nm
-Points.l <- Points.nir %>%
+n.light <- Points.nir %>%
   dplyr::filter(hue == "Light") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.l <- 
-  suppressWarnings(melt(setDT(Points.l), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
+n.light.long <- 
+  suppressWarnings(melt(setDT(n.light), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with white hues 
 #and select the NIR range 1 000 - 2 500 nm
-Points.w <- Points.nir %>%
+n.white <- Points.nir %>%
   dplyr::filter(hue == "White") %>% 
   dplyr::select(1, c(681:2180))
 
 #Melt into long format
-Points.w <- 
[…]

### 2022-06-30 analysis/paper/003-nir-savitzky-transform.R

@@ -57,6 +57,7 @@ p.d <-
   ylab("Absorbance") +
   scale_color_manual(name = "Dark",
                      values = "black") + 
+  scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 10)) +
   theme_classic() +
   theme(legend.position = c(.1,.95),
         axis.title.x = element_blank(),
@@ -83,6 +84,7 @@ p.d258 <-
                      values = "black") + 
   #coord_cartesian(x=c(1000,2500), clip = "off") +
   #annotate("label", x = 890, y = 1.20, label = "B", label.size=NA, size= 5, fontface = "bold", fill = "white") +
+  scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 10)) +
   theme_classic() +
   theme(legend.position = c(.15,.95),
         axis.title.x = element_blank(),
@@ -103,6 +105,7 @@ p.sg.d <-
   ylab("Absorbance") +
   scale_color_manual(name = "Sample 258",
                      values = "black") + 
+  scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 10)) +
   theme_classic() +
   theme(legend.position = c(.1,.95),
         axis.title.x = element_blank(),

### 2022-06-30 analysis/paper/004-nir-pca.R

@@ -42,6 +42,8 @@ Points.nir <-
 nir.pca <-
         prcomp(Points.nir[,c(681:2180)], center = TRUE, scale = FALSE)
 
+nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=2)
+
 #Prepare axis labels with variance in %
 nir.pc1lab <- as.data.frame(paste0("<b> PC1 (",as.character(round(summary(nir.pca)$importance[2,1]*100, digits=2)),"%) </b>"))
 nir.pc2lab <- as.data.frame(paste0("<b> PC2 (",as.character(round(summary(nir.pca)$importance[2,2]*100, digits=2)),"%) </b>"))
@@ -58,6 +60,8 @@ components$PC3 <- -components$PC3
 
 components = cbind(components, Points.nir$hue)
 
+pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
+
 
 #Plot using plotly with hue as color group
 fig <- 
@@ -69,7 +73,7 @@ fig <-
                   y = ~PC3, 
                   z = ~PC2,
                   color = ~factor(Points.nir$hue),
-                  colors = "Set1",
+                  colors = pca.colors,
                   marker = list(opacity = 0.8,
                                 line = list(
                                         color = 'rgba(0,0,0)',

### 2022-06-30 analysis/paper/005-raman-spectra-summary.R

@@ -129,59 +129,59 @@ r.white.long <-
 
 #plot the colourless spectra
 p.c <-
-  ggplot(r.colourless.long,aes(x = as.numeric(Wavenumber))) + 
-  geom_line(aes(y = Intensity, colour = "", group = sample_id), size = 1, stat = "identity") +
-  xlab("Wavenumber (cm-1)") +
-  ylab("Intensity") +
-  scale_color_manual(name = "Colourless",
-                     values = "purple") + 
-  theme_classic() +
-  theme(legend.position = c(.9,.95),
-        axis.title.x = element_blank(),
-        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
-        legend.title = element_text(size = 12, face = "bold", colour = "black"))
+  ggplot(r.colourless.long,aes(x = as.numeric(as.character(Wavenumber)))) + 
+    geom_line(aes(y = Intensity, colour = "", group = sample_id), size = 1, stat = "identity") +
+    xlab("Wavenumber (cm-1)") +
+    ylab("Intensity") +
+    scale_color_manual(name = "Colourless",
+                      values = "purple") + 
+    theme_classic() +
+    theme(legend.position = c(.9,.95),
+          axis.title.x = element_blank(),
+          axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
+          legend.title = element_text(size = 12, face = "bold", colour = "black"))
 
 #plot the dark spectra
 p.d <-
-  ggplot(r.dark.long, aes(x = as.numeric(Wavenumber))) + 
-  geom_line(aes(y = Intensity, colour = "", group = sample_id), size = 1, stat = "identity") +
-  xlab("Wavenumber (cm-1)") +
-  ylab("Intensity") +
-  scale_color_manual(name = "Dark",
-                     values = "black") + 
-  theme_classic() +
-  theme(legend.position = c(.1,.95),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.title = element_text(size = 12, face = "bold", colour = "black"))
+  ggplot(r.dark.long, aes(x = as.numeric(as.character(Wavenumber)))) + 
+    geom_line(aes(y = Intensity, colour = "", group = sample_id), size = 1, stat = "identity") +
+    xlab("Wavenumber (cm-1)") +
+    ylab("Intensity") +
+    scale_color_manual(name = "Dark",
+                      values = "black") + 
+    theme_classic() +
+    theme(legend.position = c(.1,.95),
[…]

### 2022-06-30 analysis/paper/007-raman-pca.R

@@ -112,6 +112,7 @@ components$PC3 <- -components$PC3
 
 components = cbind(components, Points.raman$hue)
 
+pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
 
 #Plot using plotly with hue as color group
 fig <- 
@@ -123,7 +124,7 @@ fig <-
             y = ~PC3, 
             z = ~PC2,
             color = ~factor(Points.raman$hue),
-            colors = "Set1",
+            colors = pca.colors,
             marker = list(opacity = 0.8,
                           line = list(
                             color = 'rgba(0,0,0)',

### 2022-06-30 analysis/paper/009-xrf-pca.R

@@ -65,7 +65,7 @@ pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
 pca.hue <- c("Colourless", "Dark", "Light", "White")
 
 #Show eigenvalue scree plot
-#fviz_eig(xrf.pca, choice = "eigenvalue")
+fviz_eig(xrf.pca, choice = "eigenvalue")
 
 #keep PCs above eigenvalue 1
 xrf.transform = as.data.frame(-xrf.pca$x[,1:3])

### 2022-06-30 analysis/paper/003-nir-savitzky-transform-alt.R
new subscripts

@@ -1,0 +1,81 @@
+#Load packages
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-06-30 analysis/paper/004-nir-pca-alt.R

@@ -1,0 +1,155 @@
+#Load packages
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(tidyverse))
+
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-06-30 analysis/paper/004-nir-pca-loadings.R

@@ -1,0 +1,108 @@
+#Load packages
+suppressPackageStartupMessages(library(plotly))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#Fill the NA fields in the munsell_hue column and mark them as colourless samples
+nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-06-30 analysis/paper/009-xrf-pca-loadings-scree.R

@@ -1,0 +1,116 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-06-30 analysis/paper/009-xrf-pca-loadings.R

@@ -1,0 +1,90 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-06-30 analysis/paper/009-xrf-pca-scree.R

@@ -1,0 +1,78 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-06-30 analysis/paper/009-xrf-pca.R
updated subscript

@@ -65,7 +65,7 @@ pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
 pca.hue <- c("Colourless", "Dark", "Light", "White")
 
 #Show eigenvalue scree plot
-fviz_eig(xrf.pca, choice = "eigenvalue")
+#fviz_eig(xrf.pca, choice = "eigenvalue")
 
 #keep PCs above eigenvalue 1
 xrf.transform = as.data.frame(-xrf.pca$x[,1:3])

### 2022-06-30 analysis/paper/paper.Rmd
Updated text and modified layout

@@
 ```{r nir-sample-count, results='hide'}
+#Select all dark samples
+points.d.tot <-
+  metadata.csv %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
+           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
+           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
+           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
+           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
+           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
+           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
+           site_id == "Åsele 99", 
+         type == "Point" | type == "Point fragment" | type == "Preform", 
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  filter(hue == "Dark")
+
 #Filter out the dark quartzite samples that feature peaks of interest in their spectra 
-points.d <-
+points.d.filt <-
   metadata.csv %>%
@@
 #pass to vector so that length() can be used in text to quote nr. of dark samples
-spec.d <- 
-  points.d$sample_id
+spec.d.total <- 
+  points.d.tot$sample_id
+
+#pass to vector so that length() can be used in text to quote nr. of dark samples
+spec.d.filtered <- 
+  points.d.filt$sample_id
 ```
@@
 It is well-known that NIRS is less effective on translucent material, such as a pure quartz crystal, as the absence of impurities in the material will result in little to no reflection/absorbance of light [@hunt77].
-Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, thus obscuring any potential features in the spectra.
+Similarly it can be difficult to work with dark materials as most of the light produced by the instrument will be absorbed, with the high absorbance potentially obscuring visible features in the spectra.
 This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group.
@@
 This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.
-Transforming the data collected on the dark samples using derivatives can help in the evaluation of the spectra.
-Savitzky-Golay filtering (SG) is a smoothing technique that has been widely applied in spectroscopy (although there has been recent suggestions for adaptions, see [@schmid_etal22]).
-SG filters out broad spectral features, which is beneficial in NIRS, but may also amplify any existing noise in the spectra [@chryssikos_gates17 pp.78-81; @mark_workman21 pp.
-351-397].
-Comparing one of the centered spectra from the dark group to its derivative counterpart it becomes clear that there indeed are significant peaks that may be of relevance to the characterization of the material, despite the high absorbance (fig.
-\@ref(fig:nir-savitzky-transform)).
-Filtering out those spectra that feature no discernible peaks or mainly noise leaves `r length(spec.d)` samples in the dark group.
-There is still, however, a significant difference between the dark and light group, as can be seen by comparing fig.
+
+The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81].
+2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the cost of potentially amplifying noise [@mark_workman21 pp.
+351-397; @schmid_etal22].
+This effect can be seen in fig.
+\@ref(fig:nir-savitzky-transform), where one of the centered spectra from the dark group is compared to its derivative counterpart.
+The SG transformation amplifies a number of significant peaks in the spectra, as well as a shift in the instrumentation filter at 1800 nm.
+Out of the total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were left in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise leaves.
+Despite this there is still a considerable difference in light absorbance between the dark and light group, as can be seen by comparing fig.
 \@ref(fig:nir-spectra-summary) and fig.
 \@ref(fig:nir-savitzky-transform):A.
+
+\linebreak
 
@@
 
-The most common features in quartz-based material are two peaks located at 1 400 nm and 1 900 nm (fig.
-\@ref(fig:nir-peak-summary)), indicating the presence of molecular water [@hunt77; @sciuto_etal19].
-Apart from these the main features can be found in  the > 2 000 nm range.
-Hydroxyl (OH) groups can produce a response in a number of different fields in the same material, and largely depend on the element they are attached to [@hunt77 pp.
-509-510].
-In a large group of the samples these peaks occur at 2 200 nm and 2 350 nm, with the former often being the more intense of the two.
-A general rule in NIRS has been that aluminium (Al) contributes to a more intense peak at 2 200 nm, and magnesium (Mg) at 2 300 nm [@hunt77 p.509].
-While this often may be the case there are also exceptions to this rule, and it should thus not be considered a definitive answer [@clark_etal90 p.12664].
-These absorption bands can be found in minerals like muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207].
-The 2 350 nm band is also one of the stronger carbonate (CO) bands, but considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching in this case.
+```{r nir-savitzky-transform-alt, fig.cap= "nir savitzky comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
+quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/003-nir-savitzky-transform-alt.png", error = FALSE)
+```
+
+\linebreak
+ 
+Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm.
+These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig.
+\@ref(fig:nir-peak-summary)) [@hunt77 p.
+508-510; @sciuto_etal19].
+OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region.
+As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664].
+While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH.
+In the current spectra the 2 200 nm band is typically the more intense band.
+
+The above absorption bands can be found in minerals such as muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207].
+Although the 2 350 nm band is one of the stronger carbonate (CO) bands these are typically not associated to water features, and considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching in this case.
 
@@
 
-The PCA shows a strong trend along PC 1 ( %) that seems to correlate with the colour of the material(fig.
-\@ref(fig:nir-pca)).
-Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side.
-While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the brecciated quartz, seem to cluster towards the top-left.
-At the top of the score plot 3 black samples separate themselves clearly from the rest of the data points.
-These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.
+```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
+nir.pc1var <- nir.pc1var
+```
 
@@
 
-```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca.png", error = FALSE)
+```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-loadings.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca-loadings.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/004-nir-pca-loadings.png", error = FALSE)
+```
+
+\linebreak
+
+The loading lines for the PCA can be seen in figure \@ref(fig:nir-pca-loadings).
+The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material \@ref(fig:nir-pca-fig).
+Darker material cluster towards the positive side of PC 1, with lighter towards the negative.
+The light group also separate into two general groups along PC 2.
+Material that feature less impurities and some translucence group towards the top left of the plot, with white/light grey samples grouping in the bottom left.
+
+\linebreak
+
+```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-alt.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/004-nir-pca-alt.png", error = FALSE)
 ```
@@
 \linebreak
+<!--
+```{r xrf-pca-scree, fig.cap= "xrf loadings", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-scree.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca-scree.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca-scree.png", error = FALSE)
+```
 
-```{r xrf-score-plot, fig.cap= "xrf score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca.R"), echo = FALSE}
+\linebreak
+ 
+```{r xrf-pca-loadings, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-loadings.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca-loadings.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca-loadings.png", error = FALSE)
+```
+-->
+
+```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-loadings-scree.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca-loadings-scree.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
+```
+
+\linebreak
+
+```{r xrf-pca, fig.cap= "xrf score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca.R"), echo = FALSE}
 quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))

### 2022-08-25 analysis/paper/paper.Rmd
Updated manuscript

@@
 A number of these points have, according to prior classifications, been worked using bifacial technology.
-While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca.
+While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, @mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca.
 3 800 - 2 000 BP [@kaeck09 p.60].
@@
 351-397; @schmid_etal22].
-This effect can be seen in fig.
+An example of this can be seen in fig.
 \@ref(fig:nir-savitzky-transform), where one of the centered spectra from the dark group is compared to its derivative counterpart.
 The SG transformation amplifies a number of significant peaks in the spectra, as well as a shift in the instrumentation filter at 1800 nm.
-Out of the total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were left in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise leaves.
+Out of a total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were kept in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise.
 Despite this there is still a considerable difference in light absorbance between the dark and light group, as can be seen by comparing fig.
@@
 nir.pc1var <- nir.pc1var
+nir.pc2var <- nir.pc2var
+nir.pc3var <- nir.pc3var
 ```
@@
 The loading lines for the PCA can be seen in figure \@ref(fig:nir-pca-loadings).
-The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material \@ref(fig:nir-pca-fig).
-Darker material cluster towards the positive side of PC 1, with lighter towards the negative.
-The light group also separate into two general groups along PC 2.
-Material that feature less impurities and some translucence group towards the top left of the plot, with white/light grey samples grouping in the bottom left.
+The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig.
+\@ref(fig:nir-pca-fig)).
+This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative.
+A slight difference in the clustering of light coloured material can also be seen along PC 2 (`r nir.pc2var` %).
+Material with a white or light grey hue gravitates towards the negative side (bottom left), while material that showcases translucence, and leaning more towards a yellow-beige hue, cluster towards the positive (top left).
+This can be seen more clearly in figure XX, which shows a score plot of PC 1 and PC 2 with examples of the points added to illustrate the differences in appearance of the material.
+As can be seen there are a couple of exemptions to the aforementioned trends, such as the translucent samples 260 (middle right) or 390 (top middle), possibly due to the darker impurities in the material.
+
+While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1).
+Adding PC 3 to a score plot (fig.
+\@ref(fig:nir-pca-fig):B), however,  there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.
 
@@
 There are two distinguishing features present in the Raman data (fig.
-\@ref(fig:raman-spectra)).
-The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90].
-These peaks are present for all samples in the dataset.
+\@ref(fig:raman-spectra-summary) & fig.
+\@ref(fig:raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures.
+The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
 The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples.
@@
 The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21].
-Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
+Whereas the 2^nd^ peak falls within the expected position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
 It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280].
@@
 
-A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ~ 695 cm^-1^ and 795 cm^-1^.
-Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band.
-Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset, although measurements on iron ore minerals have also generated a response in the 600 - 700 cm^-1^ band region [@ramanaidou15].
+Aside from the above more intense features a couple of weaker peaks can also be seen in a number of samples.
+Some of these could potentially be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band.
+Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been found to be one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
 
@@
 
+```{r raman-pca-var, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
+raman.pc1var <- raman.pc1var
+raman.pc2var <- raman.pc2var
+raman.pc3var <- raman.pc3var
+```
+
+\linebreak
+
 The resulting PCA model (fig.
-\@ref(fig:raman-pca)) shows a clear trend along PC 1 ( %) related to the quartz and D - G peaks.
-PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
-Interestingly, the loading for PC 3 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present.
-The measurements with mainly quartz peaks gravitates towards the negative side of PC 1, and those where the graphite peaks are dominant towards the opposite side.
-It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks.
-As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks.
-When PC 3 ( %) is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.
+\@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks.
+Samples where the D/G peaks dominate in terms of intensity are clustered towards the positive side of PC 1, with those that lack D/G peaks clustered to the opposite side.
+As can be seen in both the plot, and the spectra, the darker samples typically feature more intense D/G bands.
+PC 2 (`r raman.pc2var` %) captures similar features as PC 1, due to the main features being a presence or lack of D/G bands, which causes the "arch" visible in the plot.
+The "extremes" of PC 1 gravitate to the negative side of PC 2, with those that feature both distinct quartz and D/G bands clustering towards the positive side.
+PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
 
@@
 
+```{r xrf-si, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))
+
+#extract samples above 80% Si 
+Si_90 <- Points.xrf %>% 
+  filter(`Si` > 90)
+#total number of samples in dataset
+
+#calculate percentage of samples above 80% Si
+#si_perc <- round(((length(unique(Si_80$sample_id))/length(unique(Points.xrf$sample_id)))*100), 0)
+si_90 <- length(unique(Si_90$sample_id))
+```
+
+```{r xrf-pca-var, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca-2.R", encoding = "UTF-8")))
+
+xrf_var <- xrf.pc1var + xrf.pc2var + xrf.pc3var + xrf.pc4var
+```
+
+\linebreak
+
 ## XRF
-Four samples had dimensions that were < 10 mm in size of their smallest dimension, and were therefore excluded from the analysis as there may be complications when comparing these to larger samples [@shackley11b pp.9-10].
-The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal) that represents the remaining unquantifiable elements in a sample.
-The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF).
-Univariate analysis shows that ^14^Si makes up the main bulk of the composition of the material, with a bimodal distribution (fig.
+The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
+The elements represented in this group are all those with an atomic number lower than ~12~Mg (REF).
+
+The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (fig.
 \@ref(fig:xrf-violin-plot)).
-The majority of the readings resulted in a ^14^Si content of less than 50%, and about ¼ of the readings (~ %) featured a ^14^Si content closer to 90%.
-The “low” ^14^Si group also feature a Bal between 40 – 70%.
+Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
+This could both be due to the limits of the instrumentation, as well as the limited sampling.
+Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content of > 90%.
 
-The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ %.
-The loadings can be be seen in figure \@ref(fig:xrf-loadings).
-In the score plot of PC 1 and PC 2 two large clusters can be seen (fig.
-\@ref(fig:xrf-score-plot)).
-The smaller cluster on the left side in the score plot includes the samples with a ^14^Si content higher than 50%, whilst the larger cluster includes those with lower ^14^Si content.
-It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that are the main contributors.
-The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom.
-This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711].
-The brecciated quartz samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster.
-The white (N9) material is fairly evenly distributed within the right cluster.
+~23~V, ~39~Y, and ~30~Zn and ~25~Mn recorded a high number missing values (below LOD), although in the case ~25~Mn the values recorded reached up to 0.3%.
+As the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g.
+P, S).
+Taking these two factors into consideration the PCA was limited to the following elements:  ~12~Mg, ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe, ~38~Sr and ~40~Zr.
 
-The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river.
-The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25%.
-This artefact was sampled twice on each side, resulting in similar results both times.
+The initial PCA model of the full dataset produced 5 outliers, which significantly impacted the interpretation of the data (sample 183, 193, 259, 359 and 400).
+These samples were therefore excluded and the PCA run again, which resulted in 4 PCs with an eigenvalue above 1 (fig.
+\@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%.
+K-means cluster analysis was then performed on the PCA data in order to help identify potential groups exhibiting similar chemofacies.
+Using the elbow graph method 3 clusters were deemed suitable for the analysis.
+
+As can be seen in the score plot of PC 1-2 (fig.
+\@ref(fig:xrf-pca-a)), the three K-means clusters remain separated with next to no overlap.
+Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%.
+It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
+Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
+Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples.
+As in the case of cluster 3, apart from 1 outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%.
+Except for ~22~Ti, ~26~Fe and ~40~Zr, the cluster features comparable values to cluster 3.
+Where it stands out is in the large amount of samples with ~38~Sr readings.
+~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays, and has been the focus of study of carbonate diagenesis/paragenesis [@ahijado_etal05;@banner95].
 
@@
 
-```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-loadings-scree.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca-loadings-scree.R", encoding = "UTF-8")))
+```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2.R"), echo = FALSE}
 knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
@@
 
-```{r xrf-pca, fig.cap= "xrf score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca.png", error = FALSE)
+```{r xrf-pca-a, fig.cap= "xrf score plot", echo = FALSE}
+#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca-a.png", error = FALSE)
 ```
@@
 
-# O2PLS
+```{r xrf-pca-b, fig.cap= "xrf score plot", echo = FALSE}
+#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca-b.png", error = FALSE)
+```
+
+\linebreak
+
+```{r xrf-kmeans-boxplot, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
+#quietly(eval(parse("./analysis/paper/009-xrf-kmeans-boxplot.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-kmeans-boxplot.png", error = FALSE)
+```
+
+\linebreak
+
+<!--# O2PLS
  \linebreak
@@
 ```
+-->
 
@@
 
-The most easily recognizable megascopic feature of petrographic material is its color, which has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g.
+The most easily recognizable megascopic feature of petrographic material is its color, which, together with texture, has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g.
 @forsberg85 pp.60–64; @kaeck09 pp.53–55].
 While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite.
-Indeed, destructive techniques are typically necessary in order to document the various metamorphic features characteristic of quartzite [@howard05; @prieto_etal19].
+Indeed, destructive techniques are typically necessary in order to document the various metamorphic features in quartzite [@howard05; @prieto_etal19].
 Non-destructive application of the above spectroscopic techniques does, however, bring to light some interesting characteristics of the material.
-Although there are trends that align with the perceived color of the objects in the PCA there are also exceptions to this rule, as well as a “muddled” grey zone of material that does not conform to the light and dark extremes.
+Although there are trends that align with the perceived color of the objects there are also frequent exceptions to this rule.
 It is also worth mentioning that the color classification does not take the heterogeneous nature of the material into consideration.
@@
 Although no distinct enough groupings were observed as to make classification possible, a number of structural and chemical characteristics were identified.
-Despite the Raman PCA lacking descriptive power, due to the few features observed in the spectra, the presence of the “graphite peaks” may present an important piece of information in the identification of the potential geological origin of the material.
-Interestingly, mentions of graphitic quartzite can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
-Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies would be necessary in order to infer resource management strategies from this evidence.
+Despite the lack of explanatory power in the Raman PCA the features observed in the spectra are important for future studies.
+The presence of the “graphite peaks” is important to understanding the diagenetic background of the material, and thus its potential origin.
+Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
+Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, but it is nonetheless a key point in understanding the material flow within the region.
+
+Taking the sampling of the XRF data into consideration, some interesting patterns did emerge from the analysis. The first being the bimodal distribution of ~14~Si
 

### 2022-08-25 analysis/paper/009-xrf-pca-2.R
Temporary alternative pca subscript handling all pca related images

@@ -1,0 +1,325 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-08-25 analysis/paper/009-xrf-pca-2NOSR.R
Version of pca subscript without strontium and magnesium

@@ -1,0 +1,323 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-09-05 analysis/paper/011-nir-sample-comparison.R
New subscript comparing data for three samples from different k-means groups

@@ -1,0 +1,167 @@
+# Loading package
+suppressPackageStartupMessages(library(cowplot))
+suppressPackageStartupMessages(library(data.table))
+suppressPackageStartupMessages(library(flextable))
+suppressPackageStartupMessages(library(grid))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+nir.averaged <- 
+  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+
+#merge NIR data with metadata
+nir.merged <- 
+  as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+Points.nir <-
+  nir.merged %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+                  site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
[…]

### 2022-09-05 analysis/paper/009-xrf-pca-noSr.R
Added code plotting k-means clusters on NIR pca

@@ -1,0 +1,320 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(rgr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-09-05 analysis/paper/paper.Rmd
Updated paper with new figures and text

@@
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220407.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"))
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
@@
 In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF).
-The study includes `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig.
+The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig.
 \@ref(fig:nir-spectra-summary)).
-The aim is to chemically characterize the quartzite points recovered from the sites using the above mentioned suite of instrumenations, and assess the potential of non-destructive techniques for producing reliable fingerprints of different types of quartzite.
+The aim is to assess the potential of non-destructive techniques for characterization of quartz and quartzite material without the use of references from known geological sources.
+Due to the geological complexity of both the region, as well as the material, raw material management studies struggles with a significant amount of uncertainties.
+This study represents a first look into the chemical characteristic of the material, and will test whether the field-based instrumentation can be used to identify the variation between the samples in  order to form the basis for a predictive model.
 
@@
 
-```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine, with The tectonic  Made with ©SGU, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
+```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine.
+Made with ©SGU, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
 knitr::include_graphics('../figures/site_distribution_soils_annotated.png', error = FALSE)
@@
 XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand.
-The reference calibration used for element quantification was set to *mining* mode.
-Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light).
-Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions.
+The reference calibration used for element quantification was set to *mining* mode [@knight_etal21 p.2].
+The integration time was set to 120 seconds in order for the instrument to cycle through all of the filters (main, low, high, light).
+Due to time constraints, and the number of objects included in the project, it was only possible to sample each object once, apart from some exceptions.
 
@@
 As can be seen in both the plot, and the spectra, the darker samples typically feature more intense D/G bands.
+Worth noting are two dark samples that completely lack D/G peaks, placing them in the tight cluster that otherwise only contains light/white/colourless material.
+These two objects have, with all likelihood, been mischaracterized as quartzite.
+One of them is black with a fine texture, but also exhibits layering and resembles more a sandstone, possibly a sedimentary orthoquartzite.
+The second object likewise exhibits layering, but has a more rough texture with large quartz inclusions.
 PC 2 (`r raman.pc2var` %) captures similar features as PC 1, due to the main features being a presence or lack of D/G bands, which causes the "arch" visible in the plot.
@@
 
-```{r xrf-pca-var, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca-2.R", encoding = "UTF-8")))
+```{r xrf-pca-var, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2NOSR.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca-2NOSR.R", encoding = "UTF-8")))
 
-xrf_var <- xrf.pc1var + xrf.pc2var + xrf.pc3var + xrf.pc4var
+xrf_var <- xrf.pc1var + xrf.pc2var #+ xrf.pc3var + xrf.pc4var
 ```
@@
 The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
-The elements represented in this group are all those with an atomic number lower than ~12~Mg (REF).
+The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18].
 
@@
 Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
-This could both be due to the limits of the instrumentation, as well as the limited sampling.
-Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content of > 90%.
+This could both be due to the limits of the instrumentation, as well as the limitations in the sampling.
+Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
 
-~23~V, ~39~Y, and ~30~Zn and ~25~Mn recorded a high number missing values (below LOD), although in the case ~25~Mn the values recorded reached up to 0.3%.
-As the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g.
+~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the values reached up to 0.3% and 3% respectively.
+Due to the number of missing values these elements were therefore excluded from the PCA.
+Similarly, as the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g.
 P, S).
-Taking these two factors into consideration the PCA was limited to the following elements:  ~12~Mg, ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe, ~38~Sr and ~40~Zr.
+Element variables that may be influenced by remains from physical management of the objects were thus also excluded.
+The following elements were included in the PCA:  ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe, ~38~Sr and ~40~Zr.
 
-The initial PCA model of the full dataset produced 5 outliers, which significantly impacted the interpretation of the data (sample 183, 193, 259, 359 and 400).
+<!--The initial PCA model of the full dataset produced 5 outliers, which significantly impacted the interpretation of the data (sample 183, 193, 259, 359 and 400).
 These samples were therefore excluded and the PCA run again, which resulted in 4 PCs with an eigenvalue above 1 (fig.
@@
 ~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays, and has been the focus of study of carbonate diagenesis/paragenesis [@ahijado_etal05;@banner95].
+-->
+
+The PCA resulted in 2 PCs with an eigenvalue above 1 (fig.
+\@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%.
+K-means cluster analysis was then performed on the PCA data in order to help identify potential groups exhibiting similar chemofacies.
+Using the elbow graph method 3 clusters were deemed suitable for the analysis.
+As can be seen in the score plot (fig.
+\@ref(fig:xrf-pca-a)), the three K-means clusters remain separated.
+Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%.
+It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
+Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
+The remaining elemental variables are all higher on average than the other clusters, as is to be expected considering the higher ~14~Si value.
+Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples.
+As in the case of cluster 3, apart from one outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%.
+While the clay mineral related elements (Al, Ca, K) are comparable to cluster 3, cluster 1 measured lower in ~22~Ti, ~26~Fe and ~40~Z.
+Where it stands out, however, is in the large amount of samples with ~38~Sr readings, as almost all ~38~Sr records come from samples in cluster 3.
+~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95].
 
@@
 
-```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2.R"), echo = FALSE}
-knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
+```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", echo = FALSE}
+knitr::include_graphics("../figures/009-xrf-pca-load-screeNOSR.png", error = FALSE)
 ```
@@
 #quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-a.png", error = FALSE)
+knitr::include_graphics("../figures/009-xrf-pca-aNOSR.png", error = FALSE)
 ```
@@
 #quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-b.png", error = FALSE)
+knitr::include_graphics("../figures/009-xrf-pca-bNOSR.png", error = FALSE)
 ```
@@
 #quietly(eval(parse("./analysis/paper/009-xrf-kmeans-boxplot.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-kmeans-boxplot.png", error = FALSE)
+knitr::include_graphics("../figures/009-xrf-kmeans-boxplotNOSR.png", error = FALSE)
 ```
@@
 \linebreak
+
+Visualising the XRF k-means clusters onto the PCA for the NIRS data further emphasizes the variation within the general color groups (fig.
+\@ref(fig:nir-xrf-kmeans)).
+While the ~26~Fe/~40~Zr enriched samples largely are made up of those objects within the dark cluster in the NIR PCA.
+There are, however, a fair number of exceptions to this rule.
+A closer look at three of the objects that are part of the dark color group, but with different k-means, highlights the different nuances (fig.
+\@ref(fig:nir-sample-comparison)).
+
+```{r nir-xrf-kmeans, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
+knitr::include_graphics("../figures/010-nir-pca-xrf-kmeans.png", error = FALSE)
+```
+
+\linebreak
+
+```{r nir-sample-comparison, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/011-sample-comparison.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/011-nir-sample-comparison.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/011-nir-sample-comparison.png", error = FALSE)
+```
 
@@
 While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite.
-Indeed, destructive techniques are typically necessary in order to document the various metamorphic features in quartzite [@howard05; @prieto_etal19].
-Non-destructive application of the above spectroscopic techniques does, however, bring to light some interesting characteristics of the material.
+Indeed, destructive techniques have been identified as necessary in order to document the metamorphic features that are necessary for distinguishing between sedimentary and metamorphic quartzite [@howard05; @prieto_etal19].
+Non-destructive application of the above field-based spectroscopic techniques does, however, bring to light some interesting characteristics of the material collected in Vilhelmina and Åsele parish.
+
+Despite the limitations inherent with dark and translucent (colourless) materials in the near-infrared when using the ASD, it does provide a good estimation of differences related to general hue.
+That being said, the results also highlights the complexity inherent with categorizing the color of a heterogeneous material which lacks a uniform surface.
+Although the surface area sampled by the contact probe is large enough to cover any potential variation in color it will ultimately generate an average spectra of these features.
+What differences could be observed between the samples in the spectra, apart from absorbance, were limited, and mainly occurred within the metal-OH range.
+This metal-OH overtone region could possibly relate to the presence of clay minerals (AlOH), such as muscovite, in the material.
+
+The presence of the “graphite peaks” is another interesting aspect of the material revealed by the Raman analysis.
+These graphite-like carbon structures likely relate to the diagenetic environment in which the material formed, and should thus be critical to identifying the area of plausible geological origin.
+Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
+Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp.
+482-ff], it would not be strange to encounter carbon structures in quartzite.
+Naturally, identifying the geological origin for these materials does not necessitate a direct link between that area and the procurement of the material, but it is nonetheless a key point for understanding the material flow within the region.
+
+Sampling...xrf
+
 Although there are trends that align with the perceived color of the objects there are also frequent exceptions to this rule.
@@
 
-Although no distinct enough groupings were observed as to make classification possible, a number of structural and chemical characteristics were identified.
-Despite the lack of explanatory power in the Raman PCA the features observed in the spectra are important for future studies.
-The presence of the “graphite peaks” is important to understanding the diagenetic background of the material, and thus its potential origin.
-Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
-Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, but it is nonetheless a key point in understanding the material flow within the region.
+The heterogenous nature of the material also raises the issue of sampling.
+While NIRS provides a somewhat less subjective approximation of the color, the limited integration surface could result in a biased represenation of the material as a whole.
+This is true for all of the instrumentations used in the current study, especially for Raman and XRF where the spot size is far smaller than compared to the ASD contact probe.
+A significant number of points in the dataset feature veins of mineral inclusions running throughout the material, making it difficult to get a representative reading using probes.
+Such issues could be avoided using spectral cameras where the entire surface might be captured, but it would also defeat the purpose of using field-based methods.
+
+NIR 
+- absorbance
+- small metal-OH differences
+- material with translucence can still have dark impurities -> higher absorbance and placed next to black matrix
+- difficulties in sampling lead to issues with normalization methods? Or just material itself?
+
+Raman
+- graphite indicating
+
+XRF
+- clay minerals (K, Ca, Al)
+- Higher iron (dark samples)
+- High/Low Si
+
+Sampling 
+-
+
+The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, and as was to be expected there seems to be some possible mischaracterization due to this.
+The intention of keeping these original classifications was to further test whether the instrumentation could pick up on these possible discrepancies.
+While no extreme outliers showed up in the analysis there were some instances where objects were located contrary to the observed trend.
+The two examples of points/preforms with a dark hue in the Raman model is a good example of this.
+It highlights the complexity of the material and
 

### 2022-09-12 analysis/paper/009-xrf-pca-2NOSR.R
Revision and replacement of previous XRF PCA script. Deletion of previous temporary script.

@@ -1,323 +1,0 @@
-suppressPackageStartupMessages(library(factoextra))
-suppressPackageStartupMessages(library(ggpubr))
-suppressPackageStartupMessages(library(rgr))
-suppressPackageStartupMessages(library(robCompositions))
-suppressPackageStartupMessages(library(tidyverse))
-
-#Import xrf data, set empty fields to NA
-xrf.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
-
-#Import descriptive metadata
-metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
-
-#merge XRF data with metadata
-xrf.merged <- 
-  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
-
-#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
-Points.xrf <-
-  xrf.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
[…]

### 2022-09-12 analysis/paper/009-xrf-pca.R

@@ -8,9 +8,13 @@ suppressPackageStartupMessages(library(tidyverse))
 xrf.csv <-
   read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #merge XRF data with metadata
 xrf.merged <- 
@@ -27,22 +31,21 @@ Points.xrf <-
            site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
+         #!sample_id %in% c('183', '193', '259', '359', '400'),
          type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") 
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>%
+  replace_na(list(munsell_hue = "Colourless"))
 
[…]

### 2022-09-12 analysis/paper/001-nir-spectra-summary.R
Updated scripts to use new metadata file with UTF-8 encoding

@@ -6,7 +6,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-

### 2022-09-12 analysis/paper/002-nir-dark-filtered-summary.R

@@ -5,7 +5,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-

### 2022-09-12 analysis/paper/004-nir-pca-loadings.R
Updated scripts to use new metadata file with UTF-8 encoding

@@ -4,7 +4,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-

### 2022-09-12 analysis/paper/004-nir-pca.R

@@ -4,7 +4,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-        read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+        read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-

### 2022-09-12 analysis/paper/005-raman-spectra-summary.R
Updated script to use more tidyverse code for readability

@@ -4,7 +4,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import raman data, set empty fields to NA
 raman.csv <-
@@ -51,25 +51,37 @@ raman.baseline <-
                   method = "modpolyfit",
                   retC = TRUE)
 
-#prevent baseline from plotting a figure
+#prevent baselineSpectra from plotting a figure which will show up in the knitted document
 dev.off()
 
-#spectra are stored in "Spectra" object, not a data frame
+#spectra are stored in a "Spectra" object, not a data frame
 #need to extract samples and frequencies from that object 
 raman.tmp <- 
-  cbind(raman.baseline$freq, t(raman.baseline$data))
+  #bind data from spectra object
+  as.data.frame(cbind(raman.baseline$freq, t(raman.baseline$data))) %>% 
+  #transpose data so that frequencies are column headers and observations rows
+  tibble::rownames_to_column() %>% 
+  pivot_longer(-rowname) %>% 
+  pivot_wider(names_from=rowname, values_from=value) %>%
+  #remove unnecessary column with previous column headers
+  select(-name) %>% 
+  #Set frequencies as headers, as they are still first row
+  purrr::set_names(as.character(slice(., 1))) %>%
+  slice(-1) %>% 
+  #add sample id from spectra object
+  add_column(sample_id = sub("^X*", "", raman.baseline$names))
 #transpose data to get frequencies as columns
-raman.tmp <- 
-  as.data.frame(t(raman.tmp))
+#raman.tmp <- 
+#  as.data.frame(t(raman.tmp))
 #Add sample_id column and remove the initial "X" character introduced to sample names by R
-raman.tmp$sample_id <- 
-  c("sample_id", sub("^X*", "", raman.baseline$names))
+#raman.tmp$sample_id <- 
+#  c("sample_id", sub("^X*", "", raman.baseline$names))
 
 #Make first row headers and remove the first row with the names
-names(raman.tmp) <-
-  raman.tmp[1,]
[…]

### 2022-09-12 analysis/paper/006-raman-peak-identification.R
Updated script to use new metadata file with UTF-8 encoding

@@ -4,7 +4,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import raman data, set empty fields to NA
 raman.csv <-

### 2022-09-12 analysis/paper/007-raman-pca.R

@@ -9,7 +9,7 @@ raman.csv <-
   read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata_20220510.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 

### 2022-09-12 analysis/paper/011-nir-sample-comparison.R
Updated script, added raman spectra to figure as well as changed one sample for more appropriate example

@@ -1,4 +1,7 @@
 # Loading package
+library(png)
+library(grid)
+library(jpeg)
 suppressPackageStartupMessages(library(cowplot))
 suppressPackageStartupMessages(library(data.table))
 suppressPackageStartupMessages(library(flextable))
@@ -12,6 +15,10 @@ metadata.csv <-
 #Import nir data, set empty fields to NA
 nir.csv <-
   read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
@@ -29,6 +36,10 @@ nir.merged <-
 xrf.merged <- 
   as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
 
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
 #Filter nir data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
@@ -43,11 +54,85 @@ Points.nir <-
                 type == "Point" | type == "Point fragment" | type == "Preform", 
                 material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") 
 
+#select sample_id and data columns and transpose in prep for baseline de-trend
+raman.transposed <- 
+  raman.averaged %>% 
+  dplyr::select(sample_id, `92.88`:`2503.59`)
+
+raman.transposed <- 
+  t(raman.transposed)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = ".")
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
[…]

### 2022-09-12 analysis/paper/paper.Rmd
Updated text and structure

@@
 The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
-Although the petrographic characterization of the material mainly establishes a point of geological origin for the object it is still an important first step in reconstructing the *chaîne opératoire*.
+Petrographic characterization of the material represents the important first step in the reconstruction of the *chaîne opératoire*, as well as questions regarding raw material selection and management.
 Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these.
 The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
-Prieto et al.
+The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al.
 [-@prieto_etal19 p.
-15] highlights the current bias that exists in the understanding of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions].
-In northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region.
+15].
+In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region.
 While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
@@
 115-122; @lundberg97 pp.
-161-171], there has been little development of on the topic of quartz and quartzite.
+161-171], there has been little development on the topic of quartz and quartzite.
 A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
 The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape.
-The tentative interpretations of lithic material sourcing within northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+The tentative interpretations of lithic material sourcing within Northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
 This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
@@
 While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
-The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
+The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
 The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].
@@
 A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
-The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint and some being almost fully translucent.
+The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
 The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).
@@
 The inclusion of quartz and brecciated quartz in the study serves two purposes.
-The first being to test the potential for separating quartz from quartzitebased on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
+The first being to test the potential for separating quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
+
 When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
@@
 
-Pre-processing of the data is different for each instrumentation and based on relevant literature.
+Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16], missMDA [@josse_husson16], ggplot2 [@wickham16] and plotly [@sievert20].
+Pre-processing and analysis of the data is different for each instrumentation and based on relevant literature.
+Data exploration and analysis was performed using Principal Component Analysis (PCA).
 
@@
 Each object was measured against a white reference background and the collected measurements were averaged for each artefact.
-The visible spectrum (~ 400 nm - 750 nm) and the NIR range (~ 1 000 nm – 2 500 nm) were analysed separately, with the former focused on colour analysis and the latter the chemical structure of the material.
-The data was mean-centred before PCA modelling.
+The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before PCA modelling.
 
@@
 
-The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling.
+The XRF data was transformed using *centred log-ratio transformation*, mean-centered and scaled to unit variance before PCA.
 As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA.
@@
 Correlation analysis is directly affected by this *closure* effect and therefore not suitable.
-With suitable preprocessing the PCA model can reveal the structure and relations between the variables (elements) more accurately.
+The application of *centred log-ratio transformation* was done in order to address the closed sum effect, and in combination PCA it can reveal the structure and relations between the variables (elements) more accurately.
 
@@
 \linebreak
-
+<!--
 ## O2PLS
@@
 The data was pre-processed according to the above descriptions, and then cross-validation was used in order to determine the number of components to fit to the model.
-
+-->
 \pagebreak
@@
 
-```{r nir-savitzky-transform-alt, fig.cap= "nir savitzky comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
-quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/003-nir-savitzky-transform-alt.png", error = FALSE)
-```
-
-\linebreak
-
 Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm.
 These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig.
-\@ref(fig:nir-peak-summary)) [@hunt77 p.
+\@ref(fig:nir-savitzky-transform-alt) & \@ref(fig:nir-peak-summary)) [@hunt77 p.
 508-510; @sciuto_etal19].
@@
  
-```{r nir-peak-summary, fig.cap= "nir peak summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
+```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
+quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/003-nir-savitzky-transform-alt.png", error = FALSE)
+```
+
+\linebreak
+
+```{r nir-peak-summary, fig.cap= "Average NIR spectra for each colour group ", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
 quietly(source("./analysis/paper/002-nir-peak-summary.R", encoding = "UTF-8"))
@@
 
-While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1).
+While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) (fig.
+\@ref(fig:nir-pca-loadings)).
 Adding PC 3 to a score plot (fig.
@@
 The "extremes" of PC 1 gravitate to the negative side of PC 2, with those that feature both distinct quartz and D/G bands clustering towards the positive side.
-PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.
+PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and pre-processing the data.
 
@@
 Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
-This could both be due to the limits of the instrumentation, as well as the limitations in the sampling.
+This could both be due to the limitations of the instrumentation, as well as the sampling.
 Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
 
-~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the values reached up to 0.3% and 3% respectively.
+~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the values reached up to 0.3% and 3% respectively.
 Due to the number of missing values these elements were therefore excluded from the PCA.
@@
 Element variables that may be influenced by remains from physical management of the objects were thus also excluded.
-The following elements were included in the PCA:  ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe, ~38~Sr and ~40~Zr.
+The following elements were included in the PCA:  ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe and ~40~Zr.
+The inclusion of ~38~Sr resulted in a couple of strong outliers and overall impacted the interpretation of the model, and was therefore excluded.
 
@@
 \@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%.
-K-means cluster analysis was then performed on the PCA data in order to help identify potential groups exhibiting similar chemofacies.
+K-means cluster analysis was then performed on the PCA data in order identify potential groups exhibiting similar chemofacies.
 Using the elbow graph method 3 clusters were deemed suitable for the analysis.
@@
 \@ref(fig:xrf-pca-a)), the three K-means clusters remain separated.
-Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%.
-It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
-Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
+Cluster 1 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%.
+It mainly features samples with a dark hue, with the few samples from the light group that are grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
+Cluster 3 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
 The remaining elemental variables are all higher on average than the other clusters, as is to be expected considering the higher ~14~Si value.
-Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples.
-As in the case of cluster 3, apart from one outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%.
-While the clay mineral related elements (Al, Ca, K) are comparable to cluster 3, cluster 1 measured lower in ~22~Ti, ~26~Fe and ~40~Z.
-Where it stands out, however, is in the large amount of samples with ~38~Sr readings, as almost all ~38~Sr records come from samples in cluster 3.
-~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95].
+
+Cluster 2 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples.
+As in the case of cluster 1, apart from a couple of outliers, most of the samples in cluster 2 have a recorded ~14~Si content below 50%.
+While the clay mineral related elements (Al, Ca, K) are comparable to cluster 1, cluster 2 measured lower in ~22~Ti, ~26~Fe and ~40~Z.
+Where it stands out, however, is in the large amount of samples with ~38~Sr readings, as almost all ~38~Sr records come from samples in cluster 2.
+~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 2.
 
@@
 
-```{r xrf-pca-b, fig.cap= "xrf score plot", echo = FALSE}
-#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-bNOSR.png", error = FALSE)
-```
-
-\linebreak
-
 ```{r xrf-kmeans-boxplot, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
@@
 
-Visualising the XRF k-means clusters onto the PCA for the NIRS data further emphasizes the variation within the general color groups (fig.
+Visualising the K-means cluster analysis of the XRF PCA in the NIRS model further emphasizes the variation within the general color groups (fig.
 \@ref(fig:nir-xrf-kmeans)).
-While the ~26~Fe/~40~Zr enriched samples largely are made up of those objects within the dark cluster in the NIR PCA.
-There are, however, a fair number of exceptions to this rule.
-A closer look at three of the objects that are part of the dark color group, but with different k-means, highlights the different nuances (fig.
+Most of the dark material belongs to K-means cluster 1, which on average features higher ~22~Ti, ~26~Fe, and ~40~Zr values.
+This is not surprising as Fe may contribute to a darker hue in the material.
+Cluster 2, which features low Si values and clustered alongside clay mineral indicating elements, is fairly evenly distributed in the NIR PCA, including some of the darker objects.
+The same can be said about cluster 3, which indicates no real patterning in the NIR model.
+
+This variation can be exemplified by looking closer at three samples selected from the dark dark group, that each belong to a different K-means cluster.
+They all share a similar hue with D/G peaks detected in the Raman analysis (fig.
 \@ref(fig:nir-sample-comparison)).
+All three samples have similar absorbance values, but sample 391 exhibits some differences in the spectra compared to the other two samples.
+Out of the three objects sample 391 seems to stand out more clearly from the others, with the XRF analysis being what mainly seems to separate the other two.
 
-```{r nir-xrf-kmeans, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
+```{r nir-xrf-kmeans, fig.cap= "XRF PCA K-means cluster on NIR PCA", echo = FALSE}
 knitr::include_graphics("../figures/010-nir-pca-xrf-kmeans.png", error = FALSE)
@@
 
-```{r nir-sample-comparison, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/011-sample-comparison.R"), echo = FALSE}
+```{r nir-sample-comparison, fig.cap= "sample comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/011-sample-comparison.R"), echo = FALSE}
 quietly(eval(parse("./analysis/paper/011-nir-sample-comparison.R", encoding = "UTF-8")))
@@
 
+The study represents a good overview of the range of material which exists within the frames of what archaeologists refer to as *quartzite*, at least within Northern Sweden.
+Visually the material varies greatly, both in terms of texture as well as color.
+The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, and as was to be expected there seems to be some possible mischaracterization due to this.
+The intention of keeping these original classifications was to further test whether the instrumentation could pick up on these possible discrepancies.
+These instances are not represented as outliers in the different PCA models, however, but rather distinguish themselves by behaving contrary to expectations.
+In a similar manner the *brecciated quartz* exhibited no unique signals in the analysis, and behaved much like the other semi-translucent material in the dataset.
+The two quartz samples, on the other hand, generated somewhat different results.
+While they did not contrast with the dataset as a whole they placed in different cluster within the models, with only one them recording > 90% ~14~Si.
+This, with a visual inspection of the material, suggests that at least one of them have been mislabeled.
+
 The most easily recognizable megascopic feature of petrographic material is its color, which, together with texture, has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g.
@@
 While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite.
-Indeed, destructive techniques have been identified as necessary in order to document the metamorphic features that are necessary for distinguishing between sedimentary and metamorphic quartzite [@howard05; @prieto_etal19].
-Non-destructive application of the above field-based spectroscopic techniques does, however, bring to light some interesting characteristics of the material collected in Vilhelmina and Åsele parish.
+Indeed, destructive techniques have been identified as necessary in order to document the metamorphic features that are necessary for distinguishing between sedimentary and metamorphic quartzite [@prieto_etal19].
+Alhough field-based megascopic methods for petrographic characterization of quartzite have been proposed [e.g.
+@howard05], it would be fair to suggest that these have seen little to no use among archaeologists.
+As such, archaeological classification of quartzite is mainly based on a general understanding of the material.
+Development of field-based spectroscopic methods is therefore desirable, as these not only would increase the accuracy, but also open up for comparative analysis of the material and aid in guiding the survey and excavation efforts.
 
+The three instrumentations (NIRS, Raman, XRF) are capable of characterizing some aspects of quartzite.
+They highlight the complexity of the heterogeneous material, as well as the limitations of restricting oneself to one singular technique.
 Despite the limitations inherent with dark and translucent (colourless) materials in the near-infrared when using the ASD, it does provide a good estimation of differences related to general hue.
@@
 Although the surface area sampled by the contact probe is large enough to cover any potential variation in color it will ultimately generate an average spectra of these features.
-What differences could be observed between the samples in the spectra, apart from absorbance, were limited, and mainly occurred within the metal-OH range.
-This metal-OH overtone region could possibly relate to the presence of clay minerals (AlOH), such as muscovite, in the material.
+Including other techniques, such as imaging techniques [@sciuto_etal19], or Vis-spectrophotometers [@ramacciotti_etal22], would open up for a more systematic way of characterizing color.
+Apart from the absorbance of the material, what differences could be observed between the samples in the NIR spectra were limited, and mainly occurred within the metal-OH range.
 
 The presence of the “graphite peaks” is another interesting aspect of the material revealed by the Raman analysis.
-These graphite-like carbon structures likely relate to the diagenetic environment in which the material formed, and should thus be critical to identifying the area of plausible geological origin.
+These graphite-like carbon structures likely relate to the diagenetic/paragenetic environment in which the material formed, and should thus be critical to identifying the area of plausible geological origin.
 Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
 Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp.
-482-ff], it would not be strange to encounter carbon structures in quartzite.
-Naturally, identifying the geological origin for these materials does not necessitate a direct link between that area and the procurement of the material, but it is nonetheless a key point for understanding the material flow within the region.
+482-ff], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.
+Naturally, identifying the geological origin for these materials does not necessitate a direct link between that area and the procurement strategies employed by people, but it is nonetheless a key point for understanding the material flow within the region.
 
-Sampling...xrf
+Perhaps the most clear overview of the material comes from the XRF analysis, with reservations to the sampling.
+The results hint at a general division between material with a higher and lower amount of impurities related to Fe/Ti/Zr, as well as higher and lower amounts of Si.
+Perhaps the biggest conundrum in this case are those samples with < 50% ~14~Si, which is accompanied with a high BAL.
+At this point it is difficult to say with any certainty whether this is a result of instrumentation, sampling or merely the reality of the situation.
+While 1 sample was taken per object as a rule, several of these points and preforms came from the same site, and likely stem from the same material.
+Nevertheless, more data is needed in order to form a clearer picture of the situation.
+
+The exploratory approach presented in this paper holds potential for enabling provenance studies within geologically complex regions.
+Screening a large enough dataset makes it possible to look at in-group variations with the potential for isolating spectroscopic distinct fingerprints [e.g.
+@sciuto_etal19].
+The characterization of quartz and quartzite is fundamental if questions of resource procurement, trade and exchange and human mobility are to be developed within Northern Sweden.
+Currently, the resolution at which these differences manifest themselves within the models are limiting.
+However, by applying data fusion techniques there may be the potential to further develop these models.
+Being able to distinguish between different types of quartz and quartzite with these techniques would enable the spatial analysis and modelling of distribution patterns, regardless of whether the geological source is known or not.
+
+<!--
 
@@
 
-The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, and as was to be expected there seems to be some possible mischaracterization due to this.
-The intention of keeping these original classifications was to further test whether the instrumentation could pick up on these possible discrepancies.
  While no extreme outliers showed up in the analysis there were some instances where objects were located contrary to the observed trend.
@@
 
+-->
  \linebreak
@@
 Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials.
-The study represents a good example of a multivariate exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
-There is a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
-Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined using multivariate models.
+The study provides an example of an exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
+Not only is there a need for methodological development in the study of quartzite in archaeology, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
+Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined for information, and ultimately used to produce predictive models which can be applied to spatiotemporal questions of raw material management.
 

### 2022-12-07 analysis/paper/007-raman-pca-alt.R
Updated plot to feature correct labelling

@@ -1,0 +1,153 @@
+#Load packages
+suppressPackageStartupMessages(library(ChemoSpec))
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(prospectr))
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import raman data, set empty fields to NA
+raman.csv <-
+  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#aggregate observations by group(sample) and calculate average of wavelength measurements
+raman.averaged <- 
+  aggregate(raman.csv[, 5:469], list(sample_id = raman.csv$sample_id), mean)
+
+#transpose in prep for baseline de-trend
+raman.transposed <- 
+  t(raman.averaged)
+
+#Write csv with transposed raman data for baseline de-trending in ChemoSpec package
+#ChemoSpec requires data to be read from file
+write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+
+#prepare vector with sample_id for reading raman data into spectra object
+sample_id <- 
+  raman.averaged$sample_id
+
+#read transposed raman data into spectra object
+raman <-suppressWarnings(matrix2SpectraObject(
+  gr.crit = sample_id,
+  gr.cols = c("auto"),
+  freq.unit ="Wavelength cm-1",
+  int.unit ="Intensity",
+  descrip ="Bifacial points measurements",
+  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  sep = ";",
+  dec = ".",
+  chk = TRUE,
+  out.file = "./analysis/data/derived_data/raman_spec_object"))
+
+#call on relevant baseline method (modified polynomial fitting) from baseline package and 
+#return corrected spectra to spectra object
+raman.baseline <- 
+  baselineSpectra(raman,
+                  int = FALSE,
+                  method = "modpolyfit",
+                  retC = TRUE)
+
+#prevent baseline from plotting a figure
+dev.off()
+
+#spectra are stored in "Spectra" object, not a data frame
[…]

### 2022-12-07 analysis/paper/paper.Rmd
Updated discussion with reference to graphite-schists
Continued update and rewrite of manuscript and layout

@@
 The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
-Petrographic characterization of the material represents the important first step in the reconstruction of the *chaîne opératoire*, as well as questions regarding raw material selection and management.
-Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these.
+Characterization of the material, and identification of its potential source(s), represents the important first step in the reconstruction of the *chaîne opératoire*, and inferences of strategies for raw material selection and management.
+Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these materials.
 The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
@@
 15].
-In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region.
+In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region as they represent the most common material groups at archaeological sites.
 While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
@@
 115-122; @lundberg97 pp.
-161-171], there has been little development on the topic of quartz and quartzite.
+161-171], in contrast, there has been little development on the topic of quartz and quartzite.
 A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
 The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape.
-The tentative interpretations of lithic material sourcing within Northern Sweden tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+What interpretations of lithic material sourcing within Northern Sweden there are tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
 This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
@@
 Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
-In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable.
+Spectroscopic instrumentations feature fast measurements that can be collected directly on the surface of an object.
+With newer mobile models of these instruments readily available, this makes it an attractive option for field-based screening purposes.
+In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable.
 Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.
 While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information.
-Aside from being non-destructive spectroscopic instrumentations feature fast measurements, making it an attractive option for field-based screening purposes.
-While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties.
-The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
-The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].
+While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other complementary instrumentations that may provide further insight into the structural properties of a material.
+The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
+The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].
 
@@
 \@ref(fig:nir-spectra-summary)).
-The aim is to assess the potential of non-destructive techniques for characterization of quartz and quartzite material without the use of references from known geological sources.
-Due to the geological complexity of both the region, as well as the material, raw material management studies struggles with a significant amount of uncertainties.
-This study represents a first look into the chemical characteristic of the material, and will test whether the field-based instrumentation can be used to identify the variation between the samples in  order to form the basis for a predictive model.
+The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources.
+The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management.
+Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models.
+The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.
 
@@
 The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
-The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent).
+The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
 
@@
 3 800 - 2 000 BP [@kaeck09 p.60].
-Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].
+Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].
 
@@
 The inclusion of quartz and brecciated quartz in the study serves two purposes.
-The first being to test the potential for separating quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
+The first being to test the potential for the instrumentation to separate quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
 
@@
 Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult.
-The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
+For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
+An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
 This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind.
-The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties.
-For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
+The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.
 
-Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16], missMDA [@josse_husson16], ggplot2 [@wickham16] and plotly [@sievert20].
+Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16] and zCompositions [@palarea-albaladejo_martin-fernandez15] for data processing, and ggplot2 [@wickham16] for visualisation.
 Pre-processing and analysis of the data is different for each instrumentation and based on relevant literature.
@@
 ### Raman
-Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01].
+Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01].
 However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect.
 By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
-The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12].
+The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12].
 The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency.
@@
 After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data.
-This was achieved using modified polynomial fitting in R [@Rteam21] with the ChemoSpec package [@hanson16].
-The spectra were then mean-centred and normalized using Standard Normal Variate [@shaver01 pp.284-285] before PCA.
+This was achieved using modified polynomial fitting [@lieber_mahadevan-jansen03] from the ChemoSpec package [@hanson16].
+The spectra were then mean-centred <!--and normalized using Standard Normal Variate [@shaver01 pp.284-285] --> before PCA.
 
@@
 ### X-Ray Fluourescence
-XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method for petrographic analysis in archaeology [@hall60; @shackley11a].
+XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
 Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e.
 *fluourescence*).
-As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can tested statistically [@tykot03].
+As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can be tested statistically [@tykot03].
 
@@
 The reference calibration used for element quantification was set to *mining* mode [@knight_etal21 p.2].
-The integration time was set to 120 seconds in order for the instrument to cycle through all of the filters (main, low, high, light).
+The integration time was set to 130 seconds in order for the instrument to cycle through all of the filters (main, low, high, light).
 Due to time constraints, and the number of objects included in the project, it was only possible to sample each object once, apart from some exceptions.
 
-The XRF data was transformed using *centred log-ratio transformation*, mean-centered and scaled to unit variance before PCA.
-As the resulting dataset also features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA.
+Before engagaging in any analysis it is important to address the *closed sum* effect of compositional data [@pawlowsky-glahn_egozcue06].
+As the variables are given as proportions of a constant sum (e.g.
+percentage, parts per million) this means that if one element was to change in value, others would follow [@grunsky_caritat20].
+This is managed by transforming the data using log-ratio transformation [@aitchison99], and currently there are three main methods for achieving this; additive log-ratio (alr), centered log-ratio (clr) or isometric log-ratio.
+They all have their own nuances and as such the choice of method is dependent on the type of analysis and material [@pawlowsky-glahn_etal15].
+In the PCA the Principal Components are orthonormal, and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
+@reimann_etal12].
+This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.
+
+As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA.
 A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement.
-Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data.
-Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.
-
-Multivariate analysis methods, such as PCA, are also useful for overcoming some of the issues related to so-called "data closure" [@reimann_etal12].
-The compositional data add up to a constant (100%), which is described as the *closed sum* effect.
-Correlation analysis is directly affected by this *closure* effect and therefore not suitable.
-The application of *centred log-ratio transformation* was done in order to address the closed sum effect, and in combination PCA it can reveal the structure and relations between the variables (elements) more accurately.
+Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15].
+*lrEM* stands for "Log-ratio Expectation-Maximisation", which is an algorithm used to impute "left-censored data" (e.g.
+values below LOD, rounded zeros).
+Following log-transformation the data was then mean-centered and scaled to unit variance before PCA.
 
@@
 The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20].
-PCA has proven useful for exploring complex spectroscopic datasets [@sciuto18], and identifying clusters and trends that relate to the spectral fingerprints of the material.
-These groupings can then be followed up with supervised methods, and in the construction of classification models.
+Multivariate approaches, such as PCA,  has proven useful for exploring complex spectroscopic datasets [@sciuto18].
+Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material.
+These groupings can then be followed up with supervised methods, and aid in the construction of predicitive models.
 
@@
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
-    filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
-                           "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
-                           "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
+  filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
+                           "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
+                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56")) %>% 
   filter(hue == "Dark")
@@
 This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group.
-The dark group generally features less distinct peaks than the material with a lighter hue.
+Apart from the dark material featuring a wider range of absorbance, likely owing some to the range of hues within the group, they generally features less distinct peaks than the material with a lighter hue.
 This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.
@@
 An example of this can be seen in fig.
-\@ref(fig:nir-savitzky-transform), where one of the centered spectra from the dark group is compared to its derivative counterpart.
-The SG transformation amplifies a number of significant peaks in the spectra, as well as a shift in the instrumentation filter at 1800 nm.
-Out of a total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were kept in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise.
-Despite this there is still a considerable difference in light absorbance between the dark and light group, as can be seen by comparing fig.
-\@ref(fig:nir-spectra-summary) and fig.
-\@ref(fig:nir-savitzky-transform):A.
+\@ref(fig:nir-savitzky-transform), where the spectra from two different dark objects are compared.
+The non-processed spectra are fairly indistinct, but the derivative transformation of sample 258 amplifies a number of significant peaks.
+The derivative spectra of sample 411, however, does not have as clear and distinct peaks and features mainly noise.
+In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation.
+After reviewing the dark spectra, out of a total of `r length(spec.d.total)` spectra, `r length(spec.d.filtered)` were kept in the dark group.
 
@@
 
-```{r nir-spectra-summary, fig.cap= "nir spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
+```{r nir-spectra-summary, fig.cap= "Summary of the raw spectra for each colour group.
+Note the differeces in absorbance between the groups, with the dark group featuring featuring a larger range of absorbance values as well as less distinct peaks.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
 quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
@@
 
-```{r nir-savitzky-transform, fig.cap= "nir savitzky comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform.R"), echo = FALSE}
-quietly(source("./analysis/paper/003-nir-savitzky-transform.R", encoding = "UTF-8"))
+```{r nir-savitzky-transform, fig.cap= "Comparison of NIR spectra between two samples from the dark group.
+a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform2.R"), echo = FALSE}
+quietly(source("./analysis/paper/003-nir-savitzky-transform2.R", encoding = "UTF-8"))
 knitr::include_graphics("../figures/003-nir-savitzky-transform.png", error = FALSE)
@@
 OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region.
-As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664].
+As a general rule, the 2 200 nm and 2 350 nm positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664].
 While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH.
-In the current spectra the 2 200 nm band is typically the more intense band.
+In the current dataset the 2 200 nm band is typically the more intense band, although there are exceptions.
 
-The above absorption bands can be found in minerals such as muscovite, which would not be strange to find in quartzite [@clark_etal90 p.12; hunt77 p.510; @ramanaidou15 p.207].
-Although the 2 350 nm band is one of the stronger carbonate (CO) bands these are typically not associated to water features, and considering that the 2 350 nm peak always is accompanied by the 2 200 nm peak it would be unlikely to be related to CO stretching in this case.
+The above absorption bands can be found in mica, such as muscovite and biotite, which would not be strange to find in quartzite [@clark_etal90 p.12; @hunt77 p.510; @ramanaidou_etal15 p.207].
+Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2250 nm band for biotite [@duke94].
+As can be seen in the average spectra in fig.
+\@ref(fig:nir-peak-summary) the dark material exhibits an additional peak at ~2 250 nm, which may indicate the presence of biotite and a different metamorphic grade.
+
+Two smaller features appear at around 1 065 nm and 1 300 nm (fig.
+\@ref(fig:nir-savitzky-transform-alt)), and are often vague and difficult to make out in the non-processed spectra.
+These could be caused by ferrous iron (Fe^2+^) [@hunt77 pp.503-504], which has been demonstrated to produce a response in the NIR in a number of minerals.
+In studies of olivine the absorption region has been determined to be broad and centered around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13].
+Weak water features also have been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water, however, these overlap with the Fe^2+^ bands in, for instance, biotite [@adams75 pp.104-106].
+It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this point.
 
@@
 
-```{r nir-peak-summary, fig.cap= "Average NIR spectra for each colour group ", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
+```{r nir-peak-summary, fig.cap= "The main features in the NIR spectra can be related to overtones and vibration combinations of OH groups.
+The 1 400 nm and 1 900 nm peaks are indicative of molecular water, and common in quartz-based materials.
+The 2 200 nm and 2 350 nm bands indicate metal-OH combinations, where a more intense 2 200 nm band has typically been related to the presence of AlOH, and a more intense 2 350 nm band MgOH.
+Although difficult to detect in the non-processed spectra, weaker features also appear at 1 065 nm and 1 300 nm, that may be the result of ferrous iron (Fe^2+^)", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
 quietly(source("./analysis/paper/002-nir-peak-summary.R", encoding = "UTF-8"))
@@
 
-```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
+```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-alt.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
 nir.pc1var <- nir.pc1var
@@
 
-```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-loadings.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca-loadings.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca-loadings.png", error = FALSE)
-```
-
-\linebreak
-
-The loading lines for the PCA can be seen in figure \@ref(fig:nir-pca-loadings).
+The loading lines for the PCA can be seen in Appendix \@ref(fig:nir-pca-loadings).
 The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig.
@@
 This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative.
-A slight difference in the clustering of light coloured material can also be seen along PC 2 (`r nir.pc2var` %).
-Material with a white or light grey hue gravitates towards the negative side (bottom left), while material that showcases translucence, and leaning more towards a yellow-beige hue, cluster towards the positive (top left).
-This can be seen more clearly in figure XX, which shows a score plot of PC 1 and PC 2 with examples of the points added to illustrate the differences in appearance of the material.
-As can be seen there are a couple of exemptions to the aforementioned trends, such as the translucent samples 260 (middle right) or 390 (top middle), possibly due to the darker impurities in the material.
+A slight difference in the clustering of material with a light hue can also be seen along PC 2 (`r nir.pc2var` %).
+Objects with a  white or light grey hue gravitate towards the bottom left of the score plot, while material that showcases some translucence cluster towards the top left.
+There are objects that break this trend, however, such as sample 255 in the bottom right of fig.
+\@ref(fig:nir-pca-fig):A.
 
-While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) (fig.
-\@ref(fig:nir-pca-loadings)).
+While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) (Appendix \@ref(fig:nir-pca-loadings)).
 Adding PC 3 to a score plot (fig.
@@
 ```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-alt.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca-alt.png", error = FALSE)
+#quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/004-nir-pca-alt-with-photos.png", error = FALSE)
 ```
@@
 The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90].
+
 The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples.
@@
 Some of these could potentially be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band.
-Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been found to be one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
+Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been found to be one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou_etal15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.
 
@@
 
-```{r raman-spectra-summary, fig.cap= "raman spectra summary", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/005-raman-spectra-summary.R"), echo = FALSE}
-quietly(source("./analysis/paper/005-raman-spectra-summary.R", encoding = "UTF-8"))
+```{r raman-spectra-summary, fig.cap= "The non-processed Raman spectra visualised by colour group.
+Note that the dark material in general feature more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/005-raman-spectra-summary2.R"), echo = FALSE}
+quietly(source("./analysis/paper/005-raman-spectra-summary2.R", encoding = "UTF-8"))
 knitr::include_graphics("../figures/005-raman-spectra-summary.png", error = FALSE)
@@
 
-```{r raman-peak-ident, fig.cap= "raman peak identification", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/006-raman-peak-identification.R"), echo = FALSE}
+```{r raman-peak-ident, fig.cap= "The two features that show up in the Raman spectra can be related to quartz and carbon structures in the material.
+The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
+The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/006-raman-peak-identification.R"), echo = FALSE}
 quietly(source("./analysis/paper/006-raman-peak-identification.R", encoding = "UTF-8"))
@@
 
-```{r raman-pca-var, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
+```{r raman-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-alt.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/007-raman-pca-alt.R", encoding = "UTF-8")))
 raman.pc1var <- raman.pc1var
@@
 
-The resulting PCA model (fig.
-\@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks.
-Samples where the D/G peaks dominate in terms of intensity are clustered towards the positive side of PC 1, with those that lack D/G peaks clustered to the opposite side.
-As can be seen in both the plot, and the spectra, the darker samples typically feature more intense D/G bands.
-Worth noting are two dark samples that completely lack D/G peaks, placing them in the tight cluster that otherwise only contains light/white/colourless material.
-These two objects have, with all likelihood, been mischaracterized as quartzite.
-One of them is black with a fine texture, but also exhibits layering and resembles more a sandstone, possibly a sedimentary orthoquartzite.
-The second object likewise exhibits layering, but has a more rough texture with large quartz inclusions.
-PC 2 (`r raman.pc2var` %) captures similar features as PC 1, due to the main features being a presence or lack of D/G bands, which causes the "arch" visible in the plot.
-The "extremes" of PC 1 gravitate to the negative side of PC 2, with those that feature both distinct quartz and D/G bands clustering towards the positive side.
-PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and pre-processing the data.
+The loading lines for the PCA can be seen in Appendix \@ref(fig:raman-pca-loadings).
+The PCA model (fig.
+\@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D/G peaks.
+Samples where the D/G peaks dominate in terms of intensity cluster towards the positive side of PC 1, with those that completely lack or have weaker D/G bands on the opposite side.
+Similarly, along PC 2 (`r raman.pc2var` %) the samples that have higher intensity D/G peaks gravitate towards the top of the score plot.
+Worth noting are two dark samples that completely lack D/G peaks, placing them among the light material in the left side of the plot.
+One of these is a brecciated quartz sample with a dark yellow tint, and thus more a sign of the difficulties with assigning a generic hue to the material.
+The second sample, however, stands out from the rest of the material.
+It is a point preform made from a grey material with a rough texture, and possible layering, with larger quartz crystals.
+It does not fit with the "expected" image of a quartzite, and may be a case of mischaraterization.
+The addition of PC 3 (`r raman.pc3var` %) provides no new information, but rather separates out those samples with a higher level of noise.
 
@@
 
-```{r raman-pca, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/007-raman-pca.png", error = FALSE)
+```{r raman-pca, fig.cap= "PCA score plot for Raman data.
+Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributeted towards the right of the plot.
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-alt.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/007-raman-pca-alt.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/raman-pca-with-photos.png", error = FALSE)
 ```
@@
 
-```{r xrf-pca-var, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2NOSR.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca-2NOSR.R", encoding = "UTF-8")))
+```{r xrf-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-alt-impute.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/009-xrf-pca-alt-impute.R", encoding = "UTF-8")))
 
-xrf_var <- xrf.pc1var + xrf.pc2var #+ xrf.pc3var + xrf.pc4var
+xrf_var <- xrf.pc1var + xrf.pc2var
+
+xrf_var_low <- xrf.pc1var.low + xrf.pc2var.low + xrf.pc3var.low + xrf.pc4var.low
 ```
 
-\linebreak
+## XRF
 
-## XRF
 The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample.
 The elements represented in this group are all those with an atomic number lower than ~12~Mg [@thermo18].
-
 The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (fig.
@@
 Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown.
-This could both be due to the limitations of the instrumentation, as well as the sampling.
+While this could be due to the limitations of the instrumentation, it is perhaps more likely a result of the sampling.
+As quartzite is a metamorph rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on measurement.
 Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
-
-~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the values reached up to 0.3% and 3% respectively.
-Due to the number of missing values these elements were therefore excluded from the PCA.
-Similarly, as the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g.
-P, S).
-Element variables that may be influenced by remains from physical management of the objects were thus also excluded.
-The following elements were included in the PCA:  ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe and ~40~Zr.
-The inclusion of ~38~Sr resulted in a couple of strong outliers and overall impacted the interpretation of the model, and was therefore excluded.
+~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
 
@@
 -->
-
-The PCA resulted in 2 PCs with an eigenvalue above 1 (fig.
-\@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%.
-K-means cluster analysis was then performed on the PCA data in order identify potential groups exhibiting similar chemofacies.
+The PCA was run twice, once with the > 90% ~14~Si samples and a second time without them, as their inclusion would certainly influence the model as a whole.
+The loadings can be seen in Appendix \@ref(fig:xrf-pca-loadings-scree).
+The initial PCA with the full dataset included resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition.
 Using the elbow graph method 3 clusters were deemed suitable for the analysis.
 As can be seen in the score plot (fig.
-\@ref(fig:xrf-pca-a)), the three K-means clusters remain separated.
-Cluster 1 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%.
-It mainly features samples with a dark hue, with the few samples from the light group that are grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
-Cluster 3 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
+\@ref(fig:xrf-pca):A), the three K-means clusters remain separated.
+As expected, cluster 1 is made up of those samples that feature a high ~14~Si content, and what would traiditionally be considered a quartzite.
+There is no pattern in terms of hue within this group, and interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1.
+
+Cluster 2 is marked by a higher ~22~Ti, ~26~Fe and ~40~Zr content on average.
+It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
+Cluster 3 is mainly marked by its, on average, lower elemental content than the other two clusters.
+That said, it also features a more limited range of values between the samples.
+The material tends to feature a lighter hue with all of the brecciated quartz samples ending up in this cluster.
+
+In the second PCA model, with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%.
+The elbow graph method still suggests 3 clusters to be adequate.
+
+separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
 The remaining elemental variables are all higher on average than the other clusters, as is to be expected considering the higher ~14~Si value.
@@
 
-```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", echo = FALSE}
-knitr::include_graphics("../figures/009-xrf-pca-load-screeNOSR.png", error = FALSE)
+\linebreak
+
+```{r xrf-pca, fig.cap= "xrf score plot", echo = FALSE}
+#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/009-xrf-pca.png", error = FALSE)
 ```
@@
 
-```{r xrf-pca-a, fig.cap= "xrf score plot", echo = FALSE}
-#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-aNOSR.png", error = FALSE)
-```
-
-\linebreak
-
-```{r xrf-kmeans-boxplot, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
+```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for the whole dataset, grouped by K-means cluster.
+Note that cluster 1 feature all those samples with a Si value above 90%, with cluster 2 featuring higher Fe/Ti/Zr values.", echo = FALSE}
 #quietly(eval(parse("./analysis/paper/009-xrf-kmeans-boxplot.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-kmeans-boxplotNOSR.png", error = FALSE)
+knitr::include_graphics("../figures/009-xrf-kmeans-boxplot.png", error = FALSE)
 ```
@@
 Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
+Other metamorphic rocks, such as schists, containing graphite are a common enough occurence in other regions [e.g.
+@ukar_cloos16].
 Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp.
@@
 
+# (APPENDIX) Appendix {-}
+
+# Appendix A
+
+\linebreak
+
+```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components  NIR data.
+", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-loadings.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca-loadings.R", encoding = "UTF-8"))
+knitr::include_graphics("../figures/004-nir-pca-loadings.png", error = FALSE)
+```
+
+\linebreak
+
+```{r raman-pca-loadings, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-loadings.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/007-raman-pca-loadings.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/007-raman-pca-loadings.png", error = FALSE)
+```
+
+\linebreak
+
+```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data.
+a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.
+", echo = FALSE}
+knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
+```
+\newpage
+
 ### Colophon

### 2022-12-12 analysis/paper/011-nir-sample-comparison.R
Updated with new sample and package for reading PNG

@@ -1,7 +1,8 @@
 # Loading package
 library(png)
 library(grid)
-library(jpeg)
+library(loder)
+suppressPackageStartupMessages(library(ChemoSpec))
 suppressPackageStartupMessages(library(cowplot))
 suppressPackageStartupMessages(library(data.table))
 suppressPackageStartupMessages(library(flextable))
@@ -131,8 +132,8 @@ Points.raman <-
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.xrf <-
   xrf.merged %>%
-  dplyr::select(sample_id, `Al`, `Si`, `K`, `Ca`, `Fe`, `Zr`, `Ti`) %>% 
-  dplyr::filter(sample_id == 231 | sample_id == 404| sample_id == 391)
+  dplyr::select(sample_id, `Mg`, `Al`, `Si`, `P`, `S`, `K`, `Ca`, `Ti`, `Fe`, `Sr`, `Zr`,  `Ba`) %>% 
+  dplyr::filter(sample_id == 386 | sample_id == 404| sample_id == 391)
 
 ft <- 
   flextable(Points.xrf) %>% 
@@ -182,8 +183,8 @@ pn.231 <-
   geom_line(aes(y = Absorbance, colour = ""), size = 1, stat = "identity") +
   xlab("Wavelength (nm)") +
   ylab("Absorbance") +
-  labs(title="231") +
-  scale_color_manual(name = "231",
+  labs(title="386") +
+  scale_color_manual(name = "386",
                      values = "black") + 
   scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 5)) +
   theme_classic() +
@@ -264,7 +265,7 @@ pr.231 <-
   geom_line(aes(y = Intensity, colour = "", group = sample_id), size = 1, stat = "identity") +
   xlab("Wavenumber (cm-1)") +
   ylab("Intensity") +
-  labs(title="231") +
+  labs(title="386") +
   scale_color_manual(name = "Dark",
                      values = "black") + 
   theme_classic() +
@@ -309,13 +310,13 @@ pr.391 <-
         axis.title.y = element_blank(),
         legend.title = element_text(size = 12, face = "bold", colour = "black"))
 
-img.231 <- readJPEG("./analysis/figures/231.JPG",  native=TRUE)
[…]

### 2022-12-12 analysis/paper/009-xrf-pca-alt-impute.R
updated xrf script

@@ -1,0 +1,661 @@
+suppressPackageStartupMessages(library(factoextra))
+suppressPackageStartupMessages(library(ggpubr))
+suppressPackageStartupMessages(library(robCompositions))
+suppressPackageStartupMessages(library(tidyverse))
+
+#### Load data ####
+
+#Import xrf data, set empty fields to NA
+xrf.csv <-
+  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+
+#Import nir data, set empty fields to NA
+nir.csv <-
+  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#merge XRF data with metadata
+xrf.merged <- 
+  as.data.frame(merge(metadata.csv, xrf.csv, by='sample_id'))
+
+#### Prepare dataset ####
+
+#Filter xrf data to focus on points and preforms made from quartz/quartzite material 
+Points.xrf <-
+  xrf.merged %>%
+  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
[…]

### 2022-12-12 analysis/paper/paper.Rmd
Updating layout and text

@@
 The loadings can be seen in Appendix \@ref(fig:xrf-pca-loadings-scree).
-The initial PCA with the full dataset included resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition.
+The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.
+K-means cluster analysis was performed on the PCA data in order identify potential groups with similar elemental composition.
 Using the elbow graph method 3 clusters were deemed suitable for the analysis.
@@
 \@ref(fig:xrf-pca):A), the three K-means clusters remain separated.
-As expected, cluster 1 is made up of those samples that feature a high ~14~Si content, and what would traiditionally be considered a quartzite.
-There is no pattern in terms of hue within this group, and interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1.
+As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (fig.
+\@ref(fig:xrf-kmeans-boxplot)).
+There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values.
+Loadings for elements that you would expect to follow changes in ~14~Si content also end up near cluster 1 (e.g.
+Al, Ca).
+Interestingly one of the "quartz" labelled samples ends up in cluster 3 rather than 1, and upon examining the sample visually it does not showcase any typical quartz characteristics.
 
@@
 It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
-Cluster 3 is mainly marked by its, on average, lower elemental content than the other two clusters.
-That said, it also features a more limited range of values between the samples.
-The material tends to feature a lighter hue with all of the brecciated quartz samples ending up in this cluster.
+~13~Al and ~19~K is similarly higher in cluster 2 than compared to cluster 3, which is marked by its, on average, lower elemental content than the other two clusters.
+That said, cluster 3 does features a more limited range of values between the samples.
+The material in cluster 3 also includes all of the *brecciated quartz* samples as well as other light - colourless objects.
 
-In the second PCA model, with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%.
-The elbow graph method still suggests 3 clusters to be adequate.
+In the second PCA model (model B), with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%.
+Again, 3 clusters were deemed suitable for the K-means analysis.
+Cluster 3 stands out by having a smaller range of values in most elements, as well as lower values (fig.
+\@ref(fig:xrf-pca):B), than compared to the other clusters.
+The *brecciated quartz* samples all remain in cluster 3, along with mainly light-to-colourless samples.
+Cluster 2, on the other hand, feature the main bulk of darker samples with higher ~22~Ti, ~26~Fe and ~40~Zr values.
+Cluster 1 likewise feature a large portion of the darker materials, with comparable ~22~Ti, ~26~Fe and ~40~Zr values.
+It also has higher values of ~12~Mg, ~13~Al, ~16~S, ~19~K, and ~20~Ca.
 
-separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
-The remaining elemental variables are all higher on average than the other clusters, as is to be expected considering the higher ~14~Si value.
-
-Cluster 2 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples.
-As in the case of cluster 1, apart from a couple of outliers, most of the samples in cluster 2 have a recorded ~14~Si content below 50%.
-While the clay mineral related elements (Al, Ca, K) are comparable to cluster 1, cluster 2 measured lower in ~22~Ti, ~26~Fe and ~40~Z.
-Where it stands out, however, is in the large amount of samples with ~38~Sr readings, as almost all ~38~Sr records come from samples in cluster 2.
-~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 2.
+~38~Sr was detected at trace amounts.
+It is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays [@ahijado_etal05;@banner95], which might be one explanation for its presence in cluster 1 in model B.
+Interestingly, nearly all samples with ~38~Sr are located in cluster 2 in model A.
 
@@
 
-```{r xrf-pca, fig.cap= "xrf score plot", echo = FALSE}
+```{r xrf-pca, fig.cap= "XRF PCA score plots: a) Score plot featuring the full dataset, cluster 1 includes all samples with Si > 90%, while the remaining samples seem to cluster somewhat based on Fe and Zr content.
+This results in most of the darker materials clustering towards the bottom of the plot.
+b) Score plot with > 90% Si samples excluded.
+Similar to plot A there is a separation based Fe and Zr values, with darker material towards the bottom of the plot.
+Light-to-colourless samples also remain largely together in cluster 3.
+Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg.", echo = FALSE}
 #quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca.png", error = FALSE)
+knitr::include_graphics("../figures/009-xrf-pca-with-photos.png", error = FALSE)
 ```
@@
 
-Visualising the K-means cluster analysis of the XRF PCA in the NIRS model further emphasizes the variation within the general color groups (fig.
+The K-means cluster analysis of model B is visualised in the NIRS model, which further emphasizes the variation within the general color groups (fig.
 \@ref(fig:nir-xrf-kmeans)).
-Most of the dark material belongs to K-means cluster 1, which on average features higher ~22~Ti, ~26~Fe, and ~40~Zr values.
-This is not surprising as Fe may contribute to a darker hue in the material.
-Cluster 2, which features low Si values and clustered alongside clay mineral indicating elements, is fairly evenly distributed in the NIR PCA, including some of the darker objects.
-The same can be said about cluster 3, which indicates no real patterning in the NIR model.
+Cluster 2, with higher ~26~Fe and ~40~Zr values, mainly coincides with the darker material to the right side of the plot.
+This was expected as the trend along PC 1 in the NIR PCA mainly separates the material according to absorbance.
+Cluster 3 in turn coincides with the lighter material in the opposite side of the model.
+Cluster 1, however, is fairly evenly distributed within the model along PC 1 and PC 2.
 
-This variation can be exemplified by looking closer at three samples selected from the dark dark group, that each belong to a different K-means cluster.
+The variation that occurs within the general colour groups can be further exemplified by looking closer at three samples selected from the dark dark group, that each belong to a different K-means cluster.
 They all share a similar hue with D/G peaks detected in the Raman analysis (fig.

### 2023-01-05 analysis/paper/paper.Rmd
Updated version of the manuscript

@@
 The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
-Characterization of the material, and identification of its potential source(s), represents the important first step in the reconstruction of the *chaîne opératoire*, and inferences of strategies for raw material selection and management.
-Flint and obsidian have both been central to the study of trade and exhange networks in European prehistory, due in part to the well-surveyed geological sources for these materials.
+The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources.
+Flint and obsidian have both been central to the study of trade and exchange networks in European prehistory, due in part to the well-surveyed geological sources for these materials.
 The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
@@
 Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
-Spectroscopic instrumentations feature fast measurements that can be collected directly on the surface of an object.
+Spectroscopic instrumentation feature fast measurements that can be collected directly on the surface of an object.
 With newer mobile models of these instruments readily available, this makes it an attractive option for field-based screening purposes.
@@
 The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
-The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, and the Earth sciences [@dubessy_etal12].
+The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].
 
@@
 Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups.
-The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water [@hunt77 p.508].
+The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77 p.508].
 
@@
 By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect.
-The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12].
+The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12].
 The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency.
@@
 In both cases an interesting feature at 1 800 nm have also been amplified, which happens to be a filter shift in the instrumentation.
-After reviewing the dark spectra, out of a total of `r length(spec.d.total)` spectra, `r length(spec.d.filtered)` were kept in the dark group.
+After reviewing the dark spectra, out of a total of `r length(spec.d.total)`, `r length(spec.d.filtered)` were kept in the dark group.
 
@@
 ```{r nir-spectra-summary, fig.cap= "Summary of the raw spectra for each colour group.
-Note the differeces in absorbance between the groups, with the dark group featuring featuring a larger range of absorbance values as well as less distinct peaks.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
+Note the differences in absorbance between the groups, with the dark group featuring a larger range of absorbance values as well as less distinct peaks.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
 quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
@@
 ```{r nir-savitzky-transform, fig.cap= "Comparison of NIR spectra between two samples from the dark group.
-a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform2.R"), echo = FALSE}
-quietly(source("./analysis/paper/003-nir-savitzky-transform2.R", encoding = "UTF-8"))
+a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform.R"), echo = FALSE}
+quietly(source("./analysis/paper/003-nir-savitzky-transform.R", encoding = "UTF-8"))
 knitr::include_graphics("../figures/003-nir-savitzky-transform.png", error = FALSE)
@@
 
-```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
+```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
 quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
@@
 
-```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-alt.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
+```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
+quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
 nir.pc1var <- nir.pc1var
@@
 
-```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-alt.R"), echo = FALSE}
+```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating out a couple of light samples.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
 #quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca-alt-with-photos.png", error = FALSE)
+knitr::include_graphics("../figures/004-nir-pca-with-photos.png", error = FALSE)
 ```
@@
 ```{r raman-spectra-summary, fig.cap= "The non-processed Raman spectra visualised by colour group.
-Note that the dark material in general feature more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/005-raman-spectra-summary2.R"), echo = FALSE}
-quietly(source("./analysis/paper/005-raman-spectra-summary2.R", encoding = "UTF-8"))
+Note that the dark material in general feature more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/005-raman-spectra-summary.R"), echo = FALSE}
+quietly(source("./analysis/paper/005-raman-spectra-summary.R", encoding = "UTF-8"))
 knitr::include_graphics("../figures/005-raman-spectra-summary.png", error = FALSE)
@@
 
-```{r raman-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-alt.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/007-raman-pca-alt.R", encoding = "UTF-8")))
+```{r raman-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
 raman.pc1var <- raman.pc1var
@@
 Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributeted towards the right of the plot.
-In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-alt.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/007-raman-pca-alt.R", encoding = "UTF-8")))
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
 knitr::include_graphics("../figures/raman-pca-with-photos.png", error = FALSE)
@@
 
-```{r xrf-si, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
+```{r xrf-si, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
 quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))
@@
 ```{r xrf-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-alt-impute.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca-alt-impute.R", encoding = "UTF-8")))
+quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
 
@@
 While this could be due to the limitations of the instrumentation, it is perhaps more likely a result of the sampling.
-As quartzite is a metamorph rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on measurement.
+As quartzite is a metamorph rock where the distribution of quartz is not necessarily uniform in the material, the sample point can have a considerable influence on the measurement.
 Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content higher than 90%.
-~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of missing values (below LOD), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
+~12~Mg, ~23~V, ~25~Mn, ~39~Y and ~30~Zn recorded a high number of values below the LOD ("missing values"), although in the case of ~12~Mg and ~25~Mn the recorded values reached up to 0.3% and 3% respectively.
 
@@
 -->
-The PCA was run twice, once with the > 90% ~14~Si samples and a second time without them, as their inclusion would certainly influence the model as a whole.
+The PCA was run twice, once with the > 90% ~14~Si samples included and a second time without them.
 The loadings can be seen in Appendix \@ref(fig:xrf-pca-loadings-scree).
@@
 Using the elbow graph method 3 clusters were deemed suitable for the analysis.
+
 As can be seen in the score plot (fig.
@@
 Again, 3 clusters were deemed suitable for the K-means analysis.
-Cluster 3 stands out by having a smaller range of values in most elements, as well as lower values (fig.
+Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (fig.
 \@ref(fig:xrf-pca):B), than compared to the other clusters.
@@
 
-```{r xrf-violin-plot, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
+```{r xrf-violin-plot, fig.cap= "Violin plot of the XRF values per element.
+Median and interquartile range represented with a boxplot, surrounded by a kernel density estimation.
+Values above LOD is noted by *n*.
+The bimodal distribution of Si is visible in the plot, as well as the low number of values above LOD for V, Y and Zn.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
 quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))
@@
 
-```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for the whole dataset, grouped by K-means cluster.
+```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for model A, grouped by K-means cluster.
 Note that cluster 1 feature all those samples with a Si value above 90%, with cluster 2 featuring higher Fe/Ti/Zr values.", echo = FALSE}
@@
 
-The variation that occurs within the general colour groups can be further exemplified by looking closer at three samples selected from the dark dark group, that each belong to a different K-means cluster.
+The variation that occurs within the general colour groups can be further exemplified by looking closer at three samples selected from the dark group, that each belong to a different K-means cluster.
 They all share a similar hue with D/G peaks detected in the Raman analysis (fig.
@@
 
-```{r nir-xrf-kmeans, fig.cap= "XRF PCA K-means cluster on NIR PCA", echo = FALSE}
-knitr::include_graphics("../figures/010-nir-pca-xrf-kmeans.png", error = FALSE)
+```{r nir-xrf-kmeans, fig.cap= "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points.
+White points are samples that were excluded in the XRF PCA.
+The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot.
+This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD.", echo = FALSE}
+knitr::include_graphics("../figures/009-nir-pca-xrf-kmeans.png", error = FALSE)
 ```
@@
 
-```{r nir-sample-comparison, fig.cap= "sample comparison", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/011-sample-comparison.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/011-nir-sample-comparison.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/011-nir-sample-comparison.png", error = FALSE)
+```{r nir-sample-comparison, fig.cap= "Comparison of three dark samples from separate K-cluster groups (model B).
+Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples.
+The main difference between samples 386 and 404 seems to mainly be in the XRF data.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-sample-comparison.R"), echo = FALSE}
+quietly(eval(parse("./analysis/paper/010-nir-sample-comparison.R", encoding = "UTF-8")))
+knitr::include_graphics("../figures/010-nir-sample-comparison.png", error = FALSE)
 ```
@@
 
-The study represents a good overview of the range of material which exists within the frames of what archaeologists refer to as *quartzite*, at least within Northern Sweden.
+The exploratory evaluation of spectroscopic data generated on quartz/quartzite points and preforms demonstrates that valuable information on the composition and structural properties of the material can be gained via non-destructive analysis.
+As the identified features are related to the diagenetic/paragenetic environment they should be useful in future survey efforts aimed at locating possible geological sources.
+This does not, however, address the issues related to the resource procurement strategies employed by people within the region during the 4 000 - 2 000 BP period, as the material may have been sourced from other glacial and postglacial landforms, into which the material may have been incorporated.
+One way of addressing this issue would be to construct a predictive model based solely on the spectroscopic fingerprints of the large unknown dataset, which could form the basis of spatial analysis and hypothesis testing.
+
+One benefit of non-destructive screening is the relative ease with which large datasets can be generated.
+Although the resolution may be affected to some degree, it lends itself quite well to the exploratory approach and hypothesis generation.
+A non-destructive approach is quite effective in regions such as Northern Sweden, where little development has been made in regards to the chemical characterization of quartzite and there are many unknowns.
+There is now a better understanding of the chemical and structural variation that exists under the umbrella of what archaeologists refer to as *quartzite*, at least within the Västerbotten area.
 Visually the material varies greatly, both in terms of texture as well as color.
-The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, and as was to be expected there seems to be some possible mischaracterization due to this.
-The intention of keeping these original classifications was to further test whether the instrumentation could pick up on these possible discrepancies.
-These instances are not represented as outliers in the different PCA models, however, but rather distinguish themselves by behaving contrary to expectations.
-In a similar manner the *brecciated quartz* exhibited no unique signals in the analysis, and behaved much like the other semi-translucent material in the dataset.
-The two quartz samples, on the other hand, generated somewhat different results.
-While they did not contrast with the dataset as a whole they placed in different cluster within the models, with only one them recording > 90% ~14~Si.
-This, with a visual inspection of the material, suggests that at least one of them have been mislabeled.
+The initial classification of the material is primarily based on the archaeological documentation from the surveys and excavations, but as was highlighted in the results, there seems to be some instances of possible mischaracterization due to this.
+It is an important reminder to be mindful when aggregating data where the analyst is not the primary producer.
 
-The most easily recognizable megascopic feature of petrographic material is its color, which, together with texture, has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g.
-@forsberg85 pp.60–64; @kaeck09 pp.53–55].
-While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite.
-Indeed, destructive techniques have been identified as necessary in order to document the metamorphic features that are necessary for distinguishing between sedimentary and metamorphic quartzite [@prieto_etal19].
-Alhough field-based megascopic methods for petrographic characterization of quartzite have been proposed [e.g.
-@howard05], it would be fair to suggest that these have seen little to no use among archaeologists.
-As such, archaeological classification of quartzite is mainly based on a general understanding of the material.
-Development of field-based spectroscopic methods is therefore desirable, as these not only would increase the accuracy, but also open up for comparative analysis of the material and aid in guiding the survey and excavation efforts.
-
-The three instrumentations (NIRS, Raman, XRF) are capable of characterizing some aspects of quartzite.
-They highlight the complexity of the heterogeneous material, as well as the limitations of restricting oneself to one singular technique.
-Despite the limitations inherent with dark and translucent (colourless) materials in the near-infrared when using the ASD, it does provide a good estimation of differences related to general hue.
-That being said, the results also highlights the complexity inherent with categorizing the color of a heterogeneous material which lacks a uniform surface.
-Although the surface area sampled by the contact probe is large enough to cover any potential variation in color it will ultimately generate an average spectra of these features.
-Including other techniques, such as imaging techniques [@sciuto_etal19], or Vis-spectrophotometers [@ramacciotti_etal22], would open up for a more systematic way of characterizing color.
-Apart from the absorbance of the material, what differences could be observed between the samples in the NIR spectra were limited, and mainly occurred within the metal-OH range.
-
-The presence of the “graphite peaks” is another interesting aspect of the material revealed by the Raman analysis.
-These graphite-like carbon structures likely relate to the diagenetic/paragenetic environment in which the material formed, and should thus be critical to identifying the area of plausible geological origin.
-Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day.
-Other metamorphic rocks, such as schists, containing graphite are a common enough occurence in other regions [e.g.
-@ukar_cloos16].
+As quartzite is a metamorphic rock there are two factors to consider that will affect its chemical and structural properties; the make-up of the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former).
+The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus helping in delimiting possible geological source areas.
+Usage of the term, "graphitic quartzite", can actually be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden.
+Metamorphic rocks containing graphite, including schist and quartzite, are a common enough occurrence in other regions [e.g.
+@gaweda_cebulak99; @ukar_cloos16].
 Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp.
 482-ff], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.
-Naturally, identifying the geological origin for these materials does not necessitate a direct link between that area and the procurement strategies employed by people, but it is nonetheless a key point for understanding the material flow within the region.
+The metal-OH features in the NIRS data is another likely piece of evidence which could be helpful in separating quartzite based on metamorphic grade.
 
-Perhaps the most clear overview of the material comes from the XRF analysis, with reservations to the sampling.
-The results hint at a general division between material with a higher and lower amount of impurities related to Fe/Ti/Zr, as well as higher and lower amounts of Si.
-Perhaps the biggest conundrum in this case are those samples with < 50% ~14~Si, which is accompanied with a high BAL.
-At this point it is difficult to say with any certainty whether this is a result of instrumentation, sampling or merely the reality of the situation.
-While 1 sample was taken per object as a rule, several of these points and preforms came from the same site, and likely stem from the same material.
-Nevertheless, more data is needed in order to form a clearer picture of the situation.
-
-The exploratory approach presented in this paper holds potential for enabling provenance studies within geologically complex regions.
-Screening a large enough dataset makes it possible to look at in-group variations with the potential for isolating spectroscopic distinct fingerprints [e.g.
-@sciuto_etal19].
-The characterization of quartz and quartzite is fundamental if questions of resource procurement, trade and exchange and human mobility are to be developed within Northern Sweden.
-Currently, the resolution at which these differences manifest themselves within the models are limiting.
-However, by applying data fusion techniques there may be the potential to further develop these models.
-Being able to distinguish between different types of quartz and quartzite with these techniques would enable the spatial analysis and modelling of distribution patterns, regardless of whether the geological source is known or not.
+Due to the high representation of quartz and quartzite in the lithic assemblages dating to the Mesolithic - Bronze Age in Northern Sweden [@baudou78], it is vital that methods for characterization of these materials are developed.
+As they represent a primary lithic resource they hold a critical position in the study of resource procurement, trade and exchange and human mobility within the region (as well as many other regions [e.g.
+@prieto_etal19]).
+Previous studies have primarily relied on visual characterization following a system that was setup in order to manage the large quantities of data generated following the large scale development for hydroelectric power during the 1940s - 1980s [@biornstad06; @kaeck09 pp.53-55].
+This was necessary in order to manage the overabundance of data and make sense of it at the time, and provided a lot of new insights [e.g.
+@forsberg85; @holm91; @lundberg97], but there has been little development in regards to questions of provenance of these materials since.
+This study addresses those needs and demonstrates the potential which exists for increasing the understanding of human resource management of locally available lithic material.
 
@@
 Nevertheless, they do provide a more in-depth understanding of the composition and structure of the material, as well as showcase the complexity in the petrographic character of quartz and quartzite materials.
-The study provides an example of an exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
+The study provides an example of a non-destructive and exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
 Not only is there a need for methodological development in the study of quartzite in archaeology, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
@@
 
-I would like to thank my supervisors, Philip I. Buckland and Johan Linderholm, for valuable comments on previous versions of this paper.
+I would like to thank Professor Paul Geladi, as well as my supervisor Dr. Johan Linderholm, for valuable insights and comments on the methodological aspects of this paper.
 
@@
 
-```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components  NIR data.
+```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components for the NIR data.
 ", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-loadings.R"), echo = FALSE}
@@
 
-```{r raman-pca-loadings, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-loadings.R"), echo = FALSE}
+```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman data", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-loadings.R"), echo = FALSE}
 quietly(eval(parse("./analysis/paper/007-raman-pca-loadings.R", encoding = "UTF-8")))
@@
 ```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data.
-a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.
-", echo = FALSE}
+a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.", echo = FALSE}
 knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)

### 2023-01-09 analysis/paper/paper.Rmd
Updated relative paths for subscripts to use "here" package, removed old texts and code chunks that were greyed out

@@
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
@@
 Made with ©SGU, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
-knitr::include_graphics('../figures/site_distribution_soils_annotated.png', error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "site_distribution_soils_annotated.png"), error = FALSE)
 ```
@@
 ```{r sample-hue-pictures, fig.cap= "An assortment of points demonstrating the variation in color and heterogeneity of the material."}
-knitr::include_graphics('../figures/sample_hue.jpg', error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "sample_hue.jpg"), error = FALSE)
 ```
@@
 
-\linebreak
-<!--
-## O2PLS
-
-O2PLS is a multivariate regression method used to integrate data from different sources into a single model [@trygg02].
-It builds on the Orthogonal Partial Least Squares (OPLS) method, which was developed to correct for systematic noise in a single dataset and identify latent structures [@bylesjo_etal07].
-The O2PLS method takes the systematic variation in two datasets and divides it into two parts, joint variation between the two blocks of data, and the orthogonal variation that is block specific.
-This makes it possible to test the potential overlap between separate instrumentation, in this case NIRS and XRF.
-
-O2PLS was implemented in R [@rcoreteam21] using the *OmicsPLS* package [@bouhaddani_etal18].
-The data was pre-processed according to the above descriptions, and then cross-validation was used in order to determine the number of components to fit to the model.
--->
 \pagebreak
@@
 ```{r nir-spectra-summary, fig.cap= "Summary of the raw spectra for each colour group.
-Note the differences in absorbance between the groups, with the dark group featuring a larger range of absorbance values as well as less distinct peaks.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/001-nir-spectra-summary.R"), echo = FALSE}
-quietly(source("./analysis/paper/001-nir-spectra-summary.R", encoding = "UTF-8"))
-knitr::include_graphics('../figures/001-nir-spectra-summary.png', error = FALSE)
+Note the differences in absorbance between the groups, with the dark group featuring a larger range of absorbance values as well as less distinct peaks.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "001-nir-spectra-summary.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "001-nir-spectra-summary.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.png"), error = FALSE)
 ```
@@
 ```{r nir-savitzky-transform, fig.cap= "Comparison of NIR spectra between two samples from the dark group.
-a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform.R"), echo = FALSE}
-quietly(source("./analysis/paper/003-nir-savitzky-transform.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/003-nir-savitzky-transform.png", error = FALSE)
+a) the non-processed spectra for sample 258 and 411, b) sample 258 spectra with 2^nd^ derivative Savitzky-Golay filtering, c) sample 411 spectra with 2^nd^ derivative Savitzky-Golay filtering.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "003-nir-savitzky-transform.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.png"), error = FALSE)
 ```
@@
 
-```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/003-nir-savitzky-transform-alt.R"), echo = FALSE}
-quietly(source("./analysis/paper/003-nir-savitzky-transform-alt.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/003-nir-savitzky-transform-alt.png", error = FALSE)
+```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.png"), error = FALSE)
 ```
@@
 The 2 200 nm and 2 350 nm bands indicate metal-OH combinations, where a more intense 2 200 nm band has typically been related to the presence of AlOH, and a more intense 2 350 nm band MgOH.
-Although difficult to detect in the non-processed spectra, weaker features also appear at 1 065 nm and 1 300 nm, that may be the result of ferrous iron (Fe^2+^)", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/002-nir-peak-summary.R"), echo = FALSE}
-quietly(source("./analysis/paper/002-nir-peak-summary.R", encoding = "UTF-8"))
-knitr::include_graphics('../figures/002-nir-peak-summary.png', error = FALSE)
+Although difficult to detect in the non-processed spectra, weaker features also appear at 1 065 nm and 1 300 nm, that may be the result of ferrous iron (Fe^2+^)", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "002-nir-peak-summary.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "002-nir-peak-summary.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "002-nir-peak-summary.png"), error = FALSE)
 ```
@@
 
-```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
+```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca.R")), encoding = "UTF-8", echo = FALSE}
+quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
 nir.pc1var <- nir.pc1var
@@
 
-```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating out a couple of light samples.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
+```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating out a couple of light samples.", echo = FALSE}
 #quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca-with-photos.png", error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.png"), error = FALSE)
 ```
@@
 ```{r raman-spectra-summary, fig.cap= "The non-processed Raman spectra visualised by colour group.
-Note that the dark material in general feature more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/005-raman-spectra-summary.R"), echo = FALSE}
-quietly(source("./analysis/paper/005-raman-spectra-summary.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/005-raman-spectra-summary.png", error = FALSE)
+Note that the dark material in general feature more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "005-raman-spectra-summary.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "005-raman-spectra-summary.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "005-raman-spectra-summary.png"), error = FALSE)
 ```
@@
 The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
-The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/006-raman-peak-identification.R"), echo = FALSE}
-quietly(source("./analysis/paper/006-raman-peak-identification.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/006-raman-peak-identification.png", error = FALSE)
+The 1 287 cm-^-1^ and 1 598 cm-^-1^ peaks fall within what is referred to as the D and G band respectively, indicating the presence of carbon structures.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "006-raman-peak-identification.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "006-raman-peak-identification.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "006-raman-peak-identification.png"), error = FALSE)
 ```
@@
 
-```{r raman-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
+```{r raman-pca-var, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca.R")), echo = FALSE}
+quietly(eval(parse(here::here("analysis", "paper", "007-raman-pca.R"), encoding = "UTF-8")))
 raman.pc1var <- raman.pc1var
@@
 Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributeted towards the right of the plot.
-In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
-knitr::include_graphics("../figures/raman-pca-with-photos.png", error = FALSE)
+In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca.R")), echo = FALSE}
+knitr::include_graphics(here::here("analysis", "figures", "raman-pca-with-photos.png"), error = FALSE)
 ```
 
-```{r xrf-si, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))
+```{r xrf-si, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "008-xrf-violin-plot.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
 
@@
   filter(`Si` > 90)
-#total number of samples in dataset
 
 #calculate percentage of samples above 80% Si
-#si_perc <- round(((length(unique(Si_80$sample_id))/length(unique(Points.xrf$sample_id)))*100), 0)
 si_90 <- length(unique(Si_90$sample_id))
@@
 
-```{r xrf-pca-var, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-alt-impute.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
+```{r xrf-pca-var, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "009-xrf-pca-alt-impute.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "009-xrf-pca.R"), encoding = "UTF-8"))
 
@@
 Values above LOD is noted by *n*.
-The bimodal distribution of Si is visible in the plot, as well as the low number of values above LOD for V, Y and Zn.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/008-xrf-violin-plot.png", error = FALSE)
+The bimodal distribution of Si is visible in the plot, as well as the low number of values above LOD for V, Y and Zn.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "008-xrf-violin-plot.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "008-xrf-violin-plot.png"), error = FALSE)
 ```
-
-\linebreak
-<!--
-```{r xrf-pca-scree, fig.cap= "xrf loadings", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-scree.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca-scree.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-scree.png", error = FALSE)
-```
-
-\linebreak
- 
-```{r xrf-pca-loadings, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-loadings.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/009-xrf-pca-loadings.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-loadings.png", error = FALSE)
-```
--->
 
@@
 Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg.", echo = FALSE}
-#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-pca-with-photos.png", error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-with-photos.png"), error = FALSE)
 ```
@@
 Note that cluster 1 feature all those samples with a Si value above 90%, with cluster 2 featuring higher Fe/Ti/Zr values.", echo = FALSE}
-#quietly(eval(parse("./analysis/paper/009-xrf-kmeans-boxplot.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/009-xrf-kmeans-boxplot.png", error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.png"), error = FALSE)
 ```
@@
 This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD.", echo = FALSE}
-knitr::include_graphics("../figures/009-nir-pca-xrf-kmeans.png", error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-nir-pca-xrf-kmeans.png"), error = FALSE)
 ```
@@
 Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples.
-The main difference between samples 386 and 404 seems to mainly be in the XRF data.", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-sample-comparison.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/010-nir-sample-comparison.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/010-nir-sample-comparison.png", error = FALSE)
+The main difference between samples 386 and 404 seems to mainly be in the XRF data.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "010-sample-comparison.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "010-nir-sample-comparison.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "010-nir-sample-comparison.png"), error = FALSE)
 ```
 
-<!--# O2PLS
 \linebreak
- \linebreak
-
-The cross-validation was performed using the *crossval_o2m_adjR2()* function from the *OmicsPLS* package, and based on the results it was decided to use 2 joint components, 3 X specific (NIRS block), and 4 Y specific (XRF block).
-The data was then fitted to the O2PLS model, with summary statistics of the joint and orthogonal variation in table \@ref(tab:o2pls-summary).
-
-\linebreak
-
-```{r o2pls-summary, fig.width=8, fig.height=10, cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS-model-fit.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/010-O2PLS-model-fit.R", encoding = "UTF-8")))
-#setup a tempd data.frame with the image of the table, and then us kable to give it a table caption
-temp.df = data.frame(image="![](../figures/010-O2PLS-model-fit.png)")
-temp.mat <- as.matrix(temp.df)
-colnames(temp.mat) <- NULL
-knitr::kable(temp.mat, caption="A summary of the O2PLS estimated variation for the X (NIRS) and Y (XRF) data")
-```
-
-```{r o2pls-plot, fig.cap= "o2pls score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/010-O2PLS.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/010-O2PLS.png", error = FALSE)
-```
--->
 
@@
 ```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components for the NIR data.
-", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca-loadings.R"), echo = FALSE}
-quietly(source("./analysis/paper/004-nir-pca-loadings.R", encoding = "UTF-8"))
-knitr::include_graphics("../figures/004-nir-pca-loadings.png", error = FALSE)
+", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca-loadings.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.png"), error = FALSE)
 ```
@@
 
-```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman data", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca-loadings.R"), echo = FALSE}
-quietly(eval(parse("./analysis/paper/007-raman-pca-loadings.R", encoding = "UTF-8")))
-knitr::include_graphics("../figures/007-raman-pca-loadings.png", error = FALSE)
+```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca-loadings.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.png"), error = FALSE)
 ```
@@
 a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.", echo = FALSE}
-knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
 ```

### 2023-01-09 analysis/paper/001-nir-spectra-summary.R
Updated relative paths in all subscripts to use "here" package, including minor updates to deprecated functions and streamlining

@@ -6,11 +6,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -143,7 +143,7 @@ fig <-
 ggsave("001-nir-spectra-summary.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        width=25, 
        height=20,
        units = "cm",

### 2023-01-09 analysis/paper/002-nir-dark-filtered-summary.R

@@ -5,11 +5,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -145,7 +145,7 @@ fig <-
 ggsave("002-nir-dark-filtered-summary.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        width=20, 
        height=20,
        units = "cm",

### 2023-01-09 analysis/paper/002-nir-peak-summary.R

@@ -5,11 +5,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -156,7 +156,7 @@ fig <-
 ggsave("002-nir-peak-summary.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        width=25, 
        height=25,
        units = "cm",

### 2023-01-09 analysis/paper/003-nir-savitzky-transform-alt.R

@@ -6,11 +6,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 nir.averaged <- 
@@ -76,7 +76,7 @@ fig <-
 ggsave("003-nir-savitzky-transform-alt.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        width=25, 
        height=15,
        units = "cm",

### 2023-01-09 analysis/paper/003-nir-savitzky-transform.R

@@ -6,11 +6,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -119,7 +119,7 @@ fig <-
 ggsave("003-nir-savitzky-transform.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        width=25, 
        height=20,
        units = "cm",

### 2023-01-09 analysis/paper/004-nir-pca-loadings.R

@@ -4,11 +4,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 nir.averaged <- 
@@ -98,7 +98,7 @@ fig <-
 ggsave("004-nir-pca-loadings.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        scale = 1, 
        width=25, 
        height=15,

### 2023-01-09 analysis/paper/004-nir-pca.R

@@ -5,11 +5,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220127.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 #nir.averaged <- 
@@ -139,7 +139,7 @@ fig <-
 ggsave("004-nir-pca.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        scale = 1, 
        width=25, 
        height=25,

### 2023-01-09 analysis/paper/005-raman-spectra-summary.R

@@ -4,11 +4,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 
@@ -138,7 +138,7 @@ fig <-
 ggsave("005-raman-spectra-summary.png",
        fig,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        width=20, 
        height=20,
        units = "cm",

### 2023-01-09 analysis/paper/006-raman-peak-identification.R

@@ -4,11 +4,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 
@@ -24,7 +24,7 @@ raman.transposed <-
 
 #Write csv with transposed raman data for baseline de-trending in ChemoSpec package
 #ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
 
 #prepare vector with sample_id for reading raman data into spectra object
 sample_id <- 
@@ -37,11 +37,11 @@ raman <- suppressWarnings(matrix2SpectraObject(
   freq.unit ="Wavelength cm-1",
   int.unit ="Intensity",
   descrip ="Bifacial points measurements",
-  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
   sep = ";",
   dec = ".",
   chk = TRUE,
-  out.file = "./analysis/data/derived_data/raman_spec_object"))
+  out.file = here::here("analysis", "data", "derived_data", "raman_spec_object")))
 
[…]

### 2023-01-09 analysis/paper/007-raman-pca-loadings.R

@@ -6,10 +6,10 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 
@@ -21,24 +21,24 @@ raman.transposed <-
 
 #Write csv with transposed raman data for baseline de-trending in ChemoSpec package
 #ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
 
 #prepare vector with sample_id for reading raman data into spectra object
 sample_id <- 
   raman.averaged$sample_id
 
 #read transposed raman data into spectra object
-raman <-suppressWarnings(matrix2SpectraObject(
+raman <- suppressWarnings(matrix2SpectraObject(
   gr.crit = sample_id,
   gr.cols = c("auto"),
   freq.unit ="Wavelength cm-1",
   int.unit ="Intensity",
   descrip ="Bifacial points measurements",
-  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
   sep = ";",
   dec = ".",
   chk = TRUE,
-  out.file = "./analysis/data/derived_data/raman_spec_object"))
[…]

### 2023-01-09 analysis/paper/007-raman-pca.R

@@ -6,10 +6,10 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 
@@ -21,24 +21,24 @@ raman.transposed <-
 
 #Write csv with transposed raman data for baseline de-trending in ChemoSpec package
 #ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = "./analysis/data/derived_data/raman_transposed.csv", col.names = FALSE, sep = ";", dec = )
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
 
 #prepare vector with sample_id for reading raman data into spectra object
 sample_id <- 
   raman.averaged$sample_id
 
 #read transposed raman data into spectra object
-raman <-suppressWarnings(matrix2SpectraObject(
+raman <- suppressWarnings(matrix2SpectraObject(
   gr.crit = sample_id,
   gr.cols = c("auto"),
   freq.unit ="Wavelength cm-1",
   int.unit ="Intensity",
   descrip ="Bifacial points measurements",
-  in.file = "./analysis/data/derived_data/raman_transposed.csv",
+  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
   sep = ";",
   dec = ".",
   chk = TRUE,
-  out.file = "./analysis/data/derived_data/raman_spec_object"))
[…]

### 2023-01-09 analysis/paper/008-xrf-violin-plot.R

@@ -3,11 +3,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2(here::here("analysis", "data", "raw_data", "XRF", "xrf_quantitative_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #merge XRF data with metadata
 xrf.merged <- 
@@ -148,24 +148,28 @@ scales <- c(
 )
 
 #Violin plot of elemental content, trimmed, with fill based on elemental groups and number of values under LOD as text annotation
-Points_long %>%
-  ggplot(aes(x=Element, y=Value, fill = Group)) +
-  geom_violin(trim = TRUE) +
-  geom_boxplot(width=.05, outlier.shape = NA) +
-  facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
-  #facetted_pos_scales(y = scales) +
-  theme_bw() +
-  geom_text(data=na_values, aes(x=0.25, y=ypos, label=na_values, fontface = "bold"), nudge_x = 0.4, inherit.aes = FALSE, size = 4) +
-  theme(
-    legend.title = element_text(size = 12, face = "bold", colour = "black"),
-    strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
-    axis.title.x = element_blank(),
-    plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
-    axis.title.y=element_blank(),
-    axis.text.x=element_blank(),
-    axis.ticks.x=element_blank())
+fig <- 
+  Points_long %>%
+    ggplot(aes(x=Element, y=Value, fill = Group)) +
+    geom_violin(trim = TRUE) +
[…]

### 2023-01-09 analysis/paper/009-xrf-pca.R

@@ -7,15 +7,15 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2(here::here("analysis", "data", "raw_data", "XRF", "xrf_quantitative_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #merge XRF data with metadata
 xrf.merged <- 
@@ -389,7 +389,7 @@ fig1 <-
 ggsave("009-xrf-pca-load-scree.png",
        fig1,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        scale = 1, 
        width=25, 
        height=20,
@@ -505,7 +505,7 @@ fig2 <-
 ggsave("009-xrf-pca.png",
        fig2,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        scale = 1, 
        width=20, 
        height=20,
@@ -555,7 +555,7 @@ fig3b <-
 ggsave("009-xrf-kmeans-boxplot.png",
        fig3,
        device = "png",
-       here::here("analysis/figures/"),
+       here::here("analysis", "figures"),
        scale = 1, 
        width=20, 
        height=15,
[…]

### 2023-01-09 analysis/paper/010-nir-sample-comparison.R

@@ -11,19 +11,19 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
 metadata.csv <-
-  read.csv2("./analysis/data/raw_data/metadata.csv", sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2("./analysis/data/raw_data/NIR/asd_raw_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2("./analysis/data/raw_data/RAMAN/raman_samples_data_non_treated_20220407.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
-  read.csv2("./analysis/data/raw_data/XRF/xrf_quantitative_data_20220407.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2(here::here("analysis", "data", "raw_data", "XRF", "xrf_quantitative_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 nir.averaged <- 
@@ -65,7 +65,7 @@ raman.transposed <-
 
 #Write csv with transposed raman data for baseline de-trending in ChemoSpec package
 #ChemoSpec requires data to be read from file
[…]

### 2023-01-24 analysis/paper/paper.Rmd
Updated manuscript

@@
 bibliography: references.bib
-csl: "../templates/journal-of-archaeological-and-anthropological-science.csl"
+csl: "../templates/archaeological-and-anthropological-sciences.csl"
 abstract: |
-  Northern Fennoscandia represents a geologically complex region with both glacial and postglacial processes having an impact on the archaeology.
-The widely distributed moraine makes provenience studies difficult, and there is a severe lack of methodological development in the characterization of local geological materials used for tool making.
-The question is how to approach provenience studies in a region where there is a lack of reference samples as well as documentation of known geological sources.
-Quartz and quartzite are two central material types utilized among hunter-gatherers in Northern Sweden, and thus important to the understanding of raw material procurement and material flow within the region.
-A dataset of 445 quartz and quartzite points were analyzed non-destructively with three separate spectroscopic instrumentations, including Near-Infrared, Raman scattering and X-Ray Fluorescence.
-The aim was to test whether non-destructive application of these spectroscopic techniques can distinguish between different quartz and quartzite types.
-Multivariate modelling of the spectral data was performed using Principal Component Analysis (PCA) in which the features observed in the spectra contribute to the clustering within the model.
-While the results demonstrate that there are a number of structural and compositional differences within the dataset it is difficult to identify any distinct material classes based on any singular technique.
-The features identified do, however, provide good grounds for survey efforts and the identification of potential geological origin of some of the material.
-Future efforts will endeavor to model the joint relations between the separate blocks in order to test whether the combined information can aid in discerning different material classes.
+  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes.
+Provenance of locally available geological material thus becomes complex and fraught with uncertainty.
+This leads to a severe lack of methodological development in the characterization of these same materials in an archaeological context.
+Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region.
+125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden.
+The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence).
+The data was evaluated with exploratory analysis using Principal Component Analysis.
+The results demonstrate that detectable differences exist within the dataset for each instrumentation.
+A number of features identified, such as the presence of graphite, can likely be related to diagenetic/paragenetic differences in the material.
+Not only should these results aid in future surveys of geological sources in the region, but they will also aid in the construction of predictive models for classifying quartzite.
 keywords: |
-  Spectroscopy; Quartzite; Characterization; Lithics
+  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
 highlights: |
@@
 A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
-The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the modern day landscape.
-What interpretations of lithic material sourcing within Northern Sweden there are tends to describe strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
-This does not, however, account for the possibility that the material was sourced locally from exposed veins in the bedrock, or as loose rocks and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
+The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the current landscape.
+Existing theories on lithic material sourcing within the North Swedish region tend to describe people employing strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
 Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.
@@
 It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05].
-Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
-Spectroscopic instrumentation feature fast measurements that can be collected directly on the surface of an object.
-With newer mobile models of these instruments readily available, this makes it an attractive option for field-based screening purposes.
+Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
+Spectroscopic instrumentation features fast measurements that can be collected directly on the surface of an object.
+With newer mobile instrumentations readily available, this makes it an attractive option for field-based screening purposes.
 In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable.
@@
 The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].
+
+# Aims
 
@@
 @holm91 p.24].
-As the material frequently has been characterized based on megascopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.
+As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.
 
@@
 
-```{r sample-hue-pictures, fig.cap= "An assortment of points demonstrating the variation in color and heterogeneity of the material."}
-knitr::include_graphics(here::here("analysis", "figures", "sample_hue.jpg"), error = FALSE)
+```{r sample-hue-pictures, fig.cap= "Not only is the material heterogeneous in terms of colour, but the surface of the artefacts themselves is not uniform.
+Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation."}
+knitr::include_graphics(here::here("analysis", "figures", "figure_2.png"), error = FALSE)
 ```
@@
 All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/].
-The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens.
+The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens.
 Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult.
-For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
+When sampling materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
 An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
-This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind.
+This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind.
 The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.
@@
 XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
-Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e.
+Fluorescence produced when a material is exposed to Moseley's law [@moseley1913] recognizes that
+
+Exposing a material to high energy particles (e.g.
+X-Ray photons) will excite atoms as bound electrons are ejected upon colliding with the previous particles.
+In order for the atom to reach its ground state again one of the electrons in a higher energy level orbital will need to move to a lower energy orbital.
+The energy of the fluorescence produced due to this transition is characteristic for
+
+causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e.
 *fluourescence*).
@@
 \@ref(fig:xrf-pca):A), the three K-means clusters remain separated.
-As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (fig.
-\@ref(fig:xrf-kmeans-boxplot)).
+As expected, cluster 1 is made up of those samples that feature a high ~14~Si content (> 90%).
 There is no real pattern to this cluster aside from higher ~14~Si values with lower ~26~Fe and ~40~Zr values.
@@
 In the second PCA model (model B), with the high ~14~Si samples excluded, four PCs feature an eigenvalue above 1 with a cumulative explained variance of `r xrf_var_low`%.
-Again, 3 clusters were deemed suitable for the K-means analysis.
+Again, 3 clusters were deemed suitable for the K-means analysis (fig.
+\@ref(fig:xrf-kmeans-boxplot)).
 Cluster 3 stands out by having a smaller range of values in most elements, as well as overall lower values (fig.
@@
 
-```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for model A, grouped by K-means cluster.
-Note that cluster 1 feature all those samples with a Si value above 90%, with cluster 2 featuring higher Fe/Ti/Zr values.", echo = FALSE}
+```{r xrf-kmeans-boxplot, fig.cap= "Boxplot of XRF values (%) for model B, grouped by K-means cluster.
+Cluster 1 is made up of samples with lower values of Si, but higher values for those elements that are typically associated with the same (e.g.
+Al, Ca).
+Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content.", echo = FALSE}
 knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.png"), error = FALSE)
@@
 As quartzite is a metamorphic rock there are two factors to consider that will affect its chemical and structural properties; the make-up of the original depositional context, and the grade of metamorphism (in the case of sedimentary orthoquartzite it is mainly the former).
-The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus helping in delimiting possible geological source areas.
-Usage of the term, "graphitic quartzite", can actually be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden.
+The "graphite peaks" are a good indicator of such information, which could point towards the material forming in carbon-rich formations, thus delimiting possible geological source areas.
+Usage of the term "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in modern day Sweden.
 Metamorphic rocks containing graphite, including schist and quartzite, are a common enough occurrence in other regions [e.g.
@@
 482-ff], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.
-The metal-OH features in the NIRS data is another likely piece of evidence which could be helpful in separating quartzite based on metamorphic grade.
+
+Questions of metamorphic grade can also be raised based on the metal-OH features in the NIRS data.
+Should these be related to the Al content of the material and the presence of mica, the band positions may indicate differing metamorphic grades in the material (muscovite - biotite).
+Not only does such information provide grounds for discussions on provenance, but it also aids in the construction of a predictive model.
+The main issue, however, seems to be the struggle with sampling dark material using NIRS (and to an extent Raman), which also feature many of the indicators for biotite in the material.
+Sampling in general is a core issue, as the question becomes how representative the results really are of the material/objects in question?
+This does not simply extend to the measurements, however, but also to the visual characterization.
+The division of a heterogeneous material into larger simplified color groups might give a mistaken impression of homogeneity.
+Should a light grey point with streaks of black mineral running through it be classified as a light or dark quartzite?
+The necessity for adopting such classifications in the past is understandable, a certain level of abstraction is necessary in higher level modelling after all [@sjolander22].
+Spectroscopic screening, however, does provide a more quantitative and empirical approach to such tasks [@ramacciotti_etal22].
 
@@
 
-<!--
-
-Although there are trends that align with the perceived color of the objects there are also frequent exceptions to this rule.
-It is also worth mentioning that the color classification does not take the heterogeneous nature of the material into consideration.
-
-The heterogenous nature of the material also raises the issue of sampling.
-While NIRS provides a somewhat less subjective approximation of the color, the limited integration surface could result in a biased represenation of the material as a whole.
-This is true for all of the instrumentations used in the current study, especially for Raman and XRF where the spot size is far smaller than compared to the ASD contact probe.
-A significant number of points in the dataset feature veins of mineral inclusions running throughout the material, making it difficult to get a representative reading using probes.
-Such issues could be avoided using spectral cameras where the entire surface might be captured, but it would also defeat the purpose of using field-based methods.
-
-NIR 
-- absorbance
-- small metal-OH differences
-- material with translucence can still have dark impurities -> higher absorbance and placed next to black matrix
-- difficulties in sampling lead to issues with normalization methods? Or just material itself?
-
-Raman
-- graphite indicating
-
-XRF
-- clay minerals (K, Ca, Al)
-- Higher iron (dark samples)
-- High/Low Si
-
-Sampling 
--
-
- While no extreme outliers showed up in the analysis there were some instances where objects were located contrary to the observed trend.
-The two examples of points/preforms with a dark hue in the Raman model is a good example of this.
-It highlights the complexity of the material and
-
-Taking the sampling of the XRF data into consideration, some interesting patterns did emerge from the analysis. The first being the bimodal distribution of ~14~Si
-
-The exploratory approach presented in this paper holds potential for enabling provenance studies within geologically complex regions.
-Screening a large enough dataset makes it possible to look at in-group variations and isolate spectroscopic distinct fingerprints [e.g.
-@sciuto_etal19].
-These can then be used as the basis for later classification of the material into structurally different groups.
-The characterization of quartz and quartzite is fundamental if questions of resource procurement, trade and exchange and human mobility are to be developed within Northern Sweden.
-Being able to distinguish between different types of quartz and quartzite would enable the spatial analysis and modelling of distribution patterns, regardless of whether the geological source is known or not.
-In the current case study modelling the results of single instrumentation may have been limited for the purpose of classification, but modelling the joint relations between the separate data blocks may provide further aid in isolating distinct types of quartz and quartzite.
-
--->
 \linebreak
@@
 
+\linebreak
+ \linebreak
+
 The results demonstrate clearly that non-destructive application of all three instrumentations (NIRS, Raman, XRF) are capable of petrographically characterizing some aspect of the quartz and quartzite, however, taken on their own they lack the resolution needed to provide a detailed enough fingerprint for classification.
@@
 The study provides an example of a non-destructive and exploratory approach to the characterization of petrographic material where there also is an absence of reference material.
-Not only is there a need for methodological development in the study of quartzite in archaeology, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
+Not only is there a need for methodological development in the study of quartzite in an archaeological context, but there is also a need for methodologies that can be applied in regions with a complex geological setting and ambiguous distribution of potential geological sources.
 Taking full advantage of the short integration time of these instruments it is possible to build large comprehensive datasets that then can be explored and mined for information, and ultimately used to produce predictive models which can be applied to spatiotemporal questions of raw material management.
 
-\linebreak
 \linebreak
@@
 ```
+
 \newpage

### 2023-01-25 analysis/paper/001-artefact-size.R
Added artefact size script and figure

@@ -1,0 +1,49 @@
+# Loading package
+library(data.table)
+library(ggplot2)
+library(gridBase)
+suppressPackageStartupMessages(library(tidyverse))
+
+#Import descriptive metadata
+metadata.csv <-
+  read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
+
+#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+Points <-
+  metadata.csv %>%
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+                  site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
+                  site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
+                  site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
+                  site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
+                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
+                  site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
+                  site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
[…]

### 2023-01-25 analysis/paper/paper.Rmd
Updated text

@@
 ---
-title: "Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach"
+title: "The Quartzite Conundrum: Non-destructive analysis of Bronze Age quartzite points from Västerbotten, Sweden"
 author:
@@
 The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden.
-The sites are distributed along the Ångermanälven river valley, from the lake Kultsjön in the north, down towards the town of Åsele in the south.
+The sites are distributed along the Ångermanälven river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south.
 A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
-The material features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
-The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
+No distinction is made between arrow- and spearheads.
 
@@
 
-<!--
-A large portion of the collections stored at North Swedish museums were recovered from excavations and surveys performed in conjunction with the river regulations during the 1940s – 1980s [@biornstad06; @janson_hvarfner60], as well as subsequent follow-up survey projects [e.g.
-@andersson15].
-Vilhelmina parish in particular has been the target for numerous excavations and survey projects during this period.
--->
 \linebreak
@@
 The inclusion of quartz and brecciated quartz in the study serves two purposes.
-The first being to test the potential for the instrumentation to separate quartz from quartzite based on their chemical and structural characteristic, and the second related purpose is to account for potential mistakes in the original classification.
+The first being to test the potential for separation of quartz from quartzite based on non-destructive analysis of their chemical and structural characteristics, and the second related purpose is to account for potential mistakes in the original classification.
 
@@
 The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens.
+The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape.
+As the study is non-destructive sampling was performed on the surface of a material that is not uniform.
+The material ranges from fairly small and flat point fragments, to large and rough preforms.
+Therefore the integration could at times be difficult, which in turn could lead to scattering effects and saturation.
+
 Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult.
@@
 
-Data analysis and visualisation was performed in R [@Rteam21] using a suite of packages, including Chemospec [@hanson16] and zCompositions [@palarea-albaladejo_martin-fernandez15] for data processing, and ggplot2 [@wickham16] for visualisation.
-Pre-processing and analysis of the data is different for each instrumentation and based on relevant literature.
-Data exploration and analysis was performed using Principal Component Analysis (PCA).
+The artefacts features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent.
+While the colour of the material has been classified according to the Munsell system, for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent).
+In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.
+
+(ref:artefact) Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts. Note that the term ""point"" does not differentiate between spearheads or arrowheads.
+
+\linebreak
+
+```{r artefact-size, fig.cap= "(ref:artefact)", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "001-artefact-size.R")), echo = FALSE}
+quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
+knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.png"), error = FALSE)
+```
 
@@
 XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
-Fluorescence produced when a material is exposed to Moseley's law [@moseley1913] recognizes that
-
-Exposing a material to high energy particles (e.g.
-X-Ray photons) will excite atoms as bound electrons are ejected upon colliding with the previous particles.
-In order for the atom to reach its ground state again one of the electrons in a higher energy level orbital will need to move to a lower energy orbital.
-The energy of the fluorescence produced due to this transition is characteristic for
-
-causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e.
-*fluourescence*).
-As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can be tested statistically [@tykot03].
+Moseley's law [@moseley1913] recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms.
+This is what enables the spectral features in XRF spectrometry to be related to certain elements.
+Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12 pp.
+143-159].
+In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample.
+One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.
 
 XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand.
-The reference calibration used for element quantification was set to *mining* mode [@knight_etal21 p.2].
-The integration time was set to 130 seconds in order for the instrument to cycle through all of the filters (main, low, high, light).
-Due to time constraints, and the number of objects included in the project, it was only possible to sample each object once, apart from some exceptions.
+The reference calibration used for element quantification was *mining* mode [@knight_etal21 p.2].
+The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light).
+Due to time constraints, and the number of artefacts included in the project, it was only possible to sample each object once, apart from some exceptions.
 
-Before engagaging in any analysis it is important to address the *closed sum* effect of compositional data [@pawlowsky-glahn_egozcue06].
+It is also important to address the *closed sum* effect of compositional data [@pawlowsky-glahn_egozcue06].
 As the variables are given as proportions of a constant sum (e.g.
@@
 They all have their own nuances and as such the choice of method is dependent on the type of analysis and material [@pawlowsky-glahn_etal15].
-In the PCA the Principal Components are orthonormal, and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
+In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
 @reimann_etal12].
@@
 A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement.
-Since missing values are not permitted during PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15].
+Since missing values are not permitted in the PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15].
 *lrEM* stands for "Log-ratio Expectation-Maximisation", which is an algorithm used to impute "left-censored data" (e.g.
@@
 This effect can be seen in figure \@ref(fig:nir-spectra-summary) which shows a summary of the centered NIR spectra by color group.
-Apart from the dark material featuring a wider range of absorbance, likely owing some to the range of hues within the group, they generally features less distinct peaks than the material with a lighter hue.
+Apart from the dark material exhibiting a wider range of absorbance, likely owing some to the range of hues within the group, they generally show less distinct peaks than the material with a lighter hue.
 This is not to say that all spectra collected on dark quartzite features less information, but any model including them will likely affect the clustering of the dataset as a whole.
@@
 The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81].
-2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the cost of potentially amplifying noise [@mark_workman21 pp.
+2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21 pp.
 351-397; @schmid_etal22].
@@
 Weak water features also have been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water, however, these overlap with the Fe^2+^ bands in, for instance, biotite [@adams75 pp.104-106].
-It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this point.
+It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.
 
@@
 
-```{r nir-savitzky-transform-alt, fig.cap= "SG transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R")), echo = FALSE}
+```{r nir-savitzky-transform-alt, fig.cap= "Savitzky-Golay transformed spectra contrasting the results between the spectra in the dark (black) and light (orange) groups.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R")), echo = FALSE}
 quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R"), encoding = "UTF-8"))

### 2023-01-26 analysis/paper/009-xrf-pca.R
Updated script to produce two figures

@@ -514,7 +514,7 @@ ggsave("009-xrf-pca.png",
 
 
 #Boxplot of XRF raw data by element and K-means cluster for the whole dataset
-fig3 <- 
+fig3a <- 
   ggplot(xrf_long, aes(x=variable, y=value, fill = cluster)) +
   geom_boxplot() +  
   facet_wrap( ~ variable, scales = "free", labeller = labeller(Element = Elements)) +
@@ -553,7 +553,7 @@ fig3b <-
 
 #Save K-means boxplot
 ggsave("009-xrf-kmeans-boxplot.png",
-       fig3,
+       fig3b,
        device = "png",
        here::here("analysis", "figures"),
        scale = 1, 

### 2023-01-26 analysis/paper/008-xrf-violin-plot.R
Updated violin plot to use violin-halfs and boxplots

@@ -34,7 +34,8 @@ Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
 Points_long <-
   drop_na(Points.xrf %>%
             filter(max_length_mm >= 10 | max_width_mm >= 10) %>% 
-            dplyr::select(`Mg`,`Al`, `Si`, `P`, `S`, `K`, `Ca`, `V`, `Mn`, `Fe`, `Zn`, `Sr`, `Y`, `Zr`, `Ba`, `Ti`) %>%
+            #leave out Zn and Y as these have less than 30 values above LOD
+            dplyr::select(`Mg`,`Al`, `Si`, `P`, `S`, `K`, `Ca`, `V`, `Mn`, `Fe`, `Sr`, `Zr`, `Ba`, `Ti`) %>%
             gather(key = "Element", value = "Value"))
 
 # Add column with elemental groups for visualisation
@@ -51,9 +52,7 @@ Points_long <-
     endsWith(Element, "V") ~ "Lithophile",
     endsWith(Element, "Mn") ~ "Siderophile",
     endsWith(Element, "Fe") ~ "Siderophile",
-    endsWith(Element, "Zn") ~ "Chalcophile",
     endsWith(Element, "Sr") ~ "Lithophile",
-    endsWith(Element, "Y") ~ "Lithophile",
     endsWith(Element, "Zr") ~ "Lithophile",
     endsWith(Element, "Ba") ~ "Lithophile",
     endsWith(Element, "Ti") ~ "Chalcophile",
@@ -82,9 +81,7 @@ Elements <- c(
   'V' = "V",
   'Mn' = "Mn",
   'Fe' = "Fe",
-  'Zn' = "Zn",
   'Sr' = "Sr",
-  'Y' = "Y",
   'Zr' = "Zr",
   'Ba' = "Ba",
   'Ti' = "Ti"
@@ -108,9 +105,7 @@ na_values <-
     paste("n = ", length(which(!is.na(Points.xrf$`V`)))),
     paste("n = ", length(which(!is.na(Points.xrf$`Mn`)))),
     paste("n = ", length(which(!is.na(Points.xrf$`Fe`)))),
-    paste("n = ", length(which(!is.na(Points.xrf$`Zn`)))),
     paste("n = ", length(which(!is.na(Points.xrf$`Sr`)))),
-    paste("n = ", length(which(!is.na(Points.xrf$`Y`)))),
[…]

### 2023-01-26 analysis/paper/paper.Rmd
Updated manuscript

@@
       correspondence: true
+  - Johan Linderholm:
+      email: johan.linderholm@umu.se
+      institute: [UmU]
+      correspondence: false
+  - Paul Geladi:
+      email: paul.geladi@slu.se
+      institute: [SLU]
+      correspondence: false
 institute:
-  - UmU: Umeå University
+  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Sweden
+  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Sweden 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
     bookdown::word_document2:
+      number_sections: false
       fig_caption: yes
-      reference_docx: "../templates/template.docx" 
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
       pandoc_args:
@@
       - --lua-filter=../templates/pagebreak.lua
-mainfont: Times New Roman
+header-includes:
+    \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
 bibliography: references.bib
@@
   warning = FALSE,
-  error = TRUE,
   message = FALSE,
@@
 
-# Introduction
+# Introduction {-}
 
@@
 
-# Aims
+# Aims {-}
 
@@
 
-# Material
+# Material {-}
 
@@
 
-# Methods
+# Methods {-}
 
@@
 
-## Instrumentaion and pre-processing
+## Instrumentaion and pre-processing {-}
 
@@
 
-### Near Infrared
+### Near Infrared {-}
 NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
@@
 
-### Raman
+### Raman {-}
 Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01].
@@
 
-### X-Ray Fluourescence
+### X-Ray Fluourescence {-}
 XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
@@
 
-## PCA
+## PCA {-}
 
@@
 
-# Results
+# Results {-}
 
@@
 
-## NIR - near infrared range
+## NIR - near infrared range {-}
 ```{r nir-sample-count, results='hide'}
@@
 
-The loading lines for the PCA can be seen in Appendix \@ref(fig:nir-pca-loadings).
+The loading lines for the PCA can be seen in [Appendix A](#appendix-a).
 The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig.
@@
 
-## Raman
+## Raman {-}
 There are two distinguishing features present in the Raman data (fig.
@@
 
-The loading lines for the PCA can be seen in Appendix \@ref(fig:raman-pca-loadings).
+The loading lines for the PCA can be seen in [Appendix A](#appendix-a).
 The PCA model (fig.
@@
 
-## XRF
+## XRF {-}
 
@@
 The PCA was run twice, once with the > 90% ~14~Si samples included and a second time without them.
-The loadings can be seen in Appendix \@ref(fig:xrf-pca-loadings-scree).
+The loadings can be seen in [Appendix A](#appendix-a).
 The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.
@@
 
-# Discussion
+# Discussion {-}
 
@@
 
-# Conclusion
+# Conclusion {-}
 
@@
 
-# Acknowledgements
+# CRediT authorship contribution statement {-}
+Mattias Sjölander: Conceptualization, Data curation, Methodology, Software, Validation, Formal analysis, Resources, Writing - original draft, Visualization, Project administration.
 
-I would like to thank Professor Paul Geladi, as well as my supervisor Dr. Johan Linderholm, for valuable insights and comments on the methodological aspects of this paper.
+Johan Linderholm: Conceptualization, Methodology
+
+Paul Geladi: Methodology
+
+\linebreak
+
+# Statements and Declarations
+
+# Funding source {-}
+This work was supported by
+
+\linebreak
+
+# Acknowledgements {-}
+
+We would like to thank Västerbottens museum (www.vbm.se) for giving us access to their collections, and Susanne Sundström for her assistance with putting the artefacts together.
+We would also like to acknowledge the XXX consortium for providing access to the instrumentation used in this study.
 
@@
 
-# References
+# References {-}
 
@@
 
-# Appendix A
+\linebreak
+ \linebreak
+
+## Appendix A {#appendix-a}
+
+Collection of loadings for each single instrumentation PCA.
+
+\beginsupplement
 
@@
 
-```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components for the NIR data.
-", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca-loadings.R")), echo = FALSE}
+```{r nir-pca-loadings, fig.width=20, fig.height=20, fig.cap= "PCA loadings for the first 3 principal components for the NIR model.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca-loadings.R")), echo = FALSE}
 quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
@@
 
-```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman data", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca-loadings.R")), echo = FALSE}
+```{r raman-pca-loadings, fig.cap= "PCA loadings for the first 3 principal components for the Raman model", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca-loadings.R")), echo = FALSE}
 quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
@@
 ```{r xrf-pca-loadings-scree, fig.cap= "PCA Loadings and eigenvalues for the XRF data.
-a) eigenvalues and loadings for the full dataset, b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded.", echo = FALSE}
+a) eigenvalues and loadings for the full dataset (model A), b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded (model B).", echo = FALSE}
 knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
@@
 
-### Colophon
+### Colophon {-}
 

### 2023-03-29 analysis/paper/001-artefact-size.R
theme update

@@ -1,7 +1,5 @@
 # Loading package
-library(data.table)
 library(ggplot2)
-library(gridBase)
 suppressPackageStartupMessages(library(tidyverse))
 
 #Import descriptive metadata
@@ -34,7 +32,7 @@ fig <-
                     values=ggpubfigs::friendly_pal("contrast_three")) +
   theme_minimal() +
   theme(legend.position = c(.9,.1),
-        axis.title.x = element_blank(),
+        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
         axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
         legend.title = element_text(size = 12, face = "bold", colour = "black"))
 

### 2023-03-29 analysis/paper/001-nir-spectra-summary.R
Updated due to changes in data file, layout changes

@@ -10,7 +10,7 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -30,44 +30,44 @@ Points.nir <-
          type == "Point" | type == "Point fragment" | type == "Preform", 
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
   replace_na(list(munsell_hue = "Colourless")) %>% 
-  group_by(across(sample_id:river)) %>% 
+  group_by(across(sample_id:weight_g)) %>% 
   dplyr::summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")
 
 #Filter NIR data to focus on colourless 
-#and select the NIR range 1 000 - 2 500 nm
+#and select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 n.colourless <- Points.nir %>%
   dplyr::filter(hue == "Colourless") %>% 
-  dplyr::select(1, c(681:2180))
+  dplyr::select(1, c(679:2178))
 
 #Melt into long format
 n.colourless.long <- 
   suppressWarnings(data.table::melt(setDT(n.colourless), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with dark hues 
-#and select the NIR range 1 000 - 2 500 nm
+#and select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 n.dark <- Points.nir %>%
   dplyr::filter(hue == "Dark") %>% 
-  dplyr::select(1, c(681:2180))
+  dplyr::select(1, c(679:2178))
 
 #Melt into long format
 n.dark.long <- 
   suppressWarnings(data.table::melt(setDT(n.dark), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with light hues 
-#and select the NIR range 1 000 - 2 500 nm
[…]

### 2023-03-29 analysis/paper/002-nir-dark-filtered-summary.R
Updated due to data file change, include layout changes

@@ -9,7 +9,7 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -32,44 +32,44 @@ Points.nir <-
                            "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                            "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
   replace_na(list(munsell_hue = "Colourless")) %>% 
-  group_by(across(sample_id:river)) %>% 
+  group_by(across(sample_id:weight_g)) %>% 
   dplyr::summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")
 
 #Filter NIR data to focus on colourless 
-#and select the NIR range 1 000 - 2 500 nm
+#and select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 n.colourless <- Points.nir %>%
   dplyr::filter(hue == "Colourless") %>% 
-  dplyr::select(1, c(681:2180))
+  dplyr::select(1, c(679:2178))
 
 #Melt into long format
 n.colourless.long <- 
   suppressWarnings(data.table::melt(setDT(n.colourless), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with dark hues 
-#and select the NIR range 1 000 - 2 500 nm
+#and select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 n.dark <- Points.nir %>%
   dplyr::filter(hue == "Dark") %>% 
-  dplyr::select(1, c(681:2180))
+  dplyr::select(1, c(679:2178))
 
 #Melt into long format
 n.dark.long <- 
   suppressWarnings(data.table::melt(setDT(n.dark), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with light hues 
[…]

### 2023-03-29 analysis/paper/002-nir-peak-summary.R
Updated due to data file changes, include layout changes

@@ -9,7 +9,7 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -32,14 +32,14 @@ Points.nir <-
                                   "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
                                   "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
   replace_na(list(munsell_hue = "Colourless")) %>% 
-  group_by(across(sample_id:river)) %>% 
+  group_by(across(sample_id:weight_g)) %>% 
   dplyr::summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")
 
 #Filter NIR data to focus on material with dark hues 
-#and select the NIR range 1 000 - 2 500 nm
+#and select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 Points.d <- Points.nir %>%
   dplyr::filter(hue == "Dark") %>% 
-  dplyr::select(681:2180) %>% 
+  dplyr::select(679:2178) %>% 
   summarise_if(is.numeric, mean)
 
 #Melt into long format
@@ -47,10 +47,10 @@ Points.d <-
   suppressWarnings(data.table::melt(setDT(Points.d), variable.name = "Wavelength", variable.factor = FALSE, value.name = "Absorbance"))
 
 #Filter NIR data to focus on material with light hues 
-#and select the NIR range 1 000 - 2 500 nm
+#and select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 Points.l <- Points.nir %>%
   dplyr::filter(hue == "Light") %>% 
-  dplyr::select(681:2180) %>% 
+  dplyr::select(679:2178) %>% 
   summarise_if(is.numeric, mean)
 
 #Melt into long format
@@ -58,10 +58,10 @@ Points.l <-
[…]

### 2023-03-29 analysis/paper/003-nir-savitzky-transform-alt.R
Updated due to data file changes, include layout changes

@@ -10,7 +10,7 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 nir.averaged <- 
@@ -19,9 +19,6 @@ nir.averaged <-
 #merge NIR data with metadata
 nir.merged <- 
   as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
-
-#Fill the NA fields in the munsell_hue column and mark them as colourless samples
-nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
 
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
@@ -38,20 +35,21 @@ Points.nir <-
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
   filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                            "215","216","229","234","235","237","238","251","262","265","268","272","278","281","282","359","377","392","393","397","405",
-                           "406","410","411","413","414","415","416","417","424","425","426","428","430","55","56"))
+                           "406","410","411","413","414","415","416","417","424","425","426","428","430","55","56")) %>% 
+  replace_na(list(munsell_hue = "Colourless"))
 
 
 #Filter NIR data to focus on light and dark hue material
 #and select the NIR range 1 000 - 2 500 nm, as well as the hue variable
 Points <- Points.nir %>%
   filter(hue == "Light" | hue == "Dark") %>% 
-  dplyr::select(7, c(681:2180))
+  dplyr::select(7, c(678:2178))
 
 #Reorder the hue factor level so light spectra will be drawn behind dark spectra
 Points$hue <- factor(Points$hue, c("Light", "Dark"))
 
[…]

### 2023-03-29 analysis/paper/003-nir-savitzky-transform.R
Updated due to data file changes

@@ -10,7 +10,7 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -30,13 +30,13 @@ Points.nir <-
          type == "Point" | type == "Point fragment" | type == "Preform", 
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
   replace_na(list(munsell_hue = "Colourless")) %>% 
-  group_by(across(sample_id:river)) %>% 
+  group_by(across(sample_id:weight_g)) %>% 
   dplyr::summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")
 
-#Filter NIR data to focus on sample 258 and 411, then select the NIR range 1 000 - 2 500 nm
+#Filter NIR data to focus on sample 258 and 411, then select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 raw.spec <- Points.nir %>%
   filter(sample_id == "258" | sample_id == "411") %>% 
-  select(1, c(681:2180))
+  select(1, c(679:2178))
 
 #Melt into long format, data.table needs to be specified to avoid r using reshape2 version that sets variable to factor
 raw.spec.long <- 

### 2023-03-29 analysis/paper/004-nir-pca-loadings.R
Updated due to data file changes

@@ -8,7 +8,7 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 nir.averaged <- 
@@ -17,9 +17,6 @@ nir.averaged <-
 #merge NIR data with metadata
 nir.merged <- 
   as.data.frame(merge(metadata.csv, nir.averaged, by='sample_id'))
-
-#Fill the NA fields in the munsell_hue column and mark them as colourless samples
-nir.merged[8][is.na(nir.merged[8])] <- "Colourless"
 
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
@@ -36,7 +33,8 @@ Points.nir <-
          material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
   filter(!sample_id %in% c("153","167","168","169","172","174","175","177","182","183","190","191","193","194","196","198","200","204","207","210","213","214",
                            "215","216","229","234","235","237","238","251","262","265","268","269","272","278","281","282","359","377","385","392","393","397","405",
-                           "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56"))
+                           "406","410","411","413","414","415","416","417","424","425","426","428","430","432","55","56")) %>% 
+  replace_na(list(munsell_hue = "Colourless"))
 
 #perform PCA with SNV normalization and mean-center
 nir.pca <-

### 2023-03-29 analysis/paper/004-nir-pca.R
Layout changes

@@ -53,7 +53,7 @@ nir.pc1lab <- as.data.frame(paste0("PC1 (",as.character(round(summary(nir.pca)$i
 nir.pc2lab <- as.data.frame(paste0("PC2 (",as.character(round(summary(nir.pca)$importance[2,2]*100, digits=2)),"%)"))
 nir.pc3lab <- as.data.frame(paste0("PC3 (",as.character(round(summary(nir.pca)$importance[2,3]*100, digits=2)),"%)"))
 
-pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
+pca.colors <- c("#56B4E9", "#36454F", "#E69F00", "#F9F6EE")
 pca.hue <- c("Colourless", "Dark", "Light", "White")
 
 #prepare a basic score plot with fviz_pca_ind using PC1 and PC2
@@ -66,28 +66,29 @@ basic_plot2 <-
 fig.1 <- 
   ggplot(cbind(basic_plot1$data, Points.nir[, c(7,10)]),
          aes(x=x, y=y, shape = material, fill = hue)) +
+  #to add sample id as text
   #geom_text(aes(label=Points.nir$sample_id, hjust=0.5,vjust=-1.0)) +
-  geom_point(size=3) +
+  geom_point(size=3.5) +
   theme_bw() +
   ggtitle("Near infrared 1 000 - 2 500 nm") +
   labs(x = nir.pc1lab,
        y = nir.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=c(24,22,21)) +
+                     values=c(25,24,21)) +
   scale_fill_manual(name = "Hue",
                     values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol = 3,
+  guides(fill = guide_legend(override.aes = list(shape = 22,
+                                                 ncol = 1,
+                                                 color = "black",
+                                                 size=3),
+                             title.position="top", 
+                             title.hjust = 0.5,
+                             order = 1),
+         shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol = 1,
                               title.position="top", 
                               title.hjust = 0.5,
-                              order = 2),
-         fill = guide_legend(override.aes = list(shape = 22,
-                                                 fill = pca.colors,
-                                                 ncol = 4,
-                                                 color = "black",
-                                                 size=3,
-                                                 order = 1),
-                             title.position="top", title.hjust = 0.5))+
+                              order = 2)) +
[…]

### 2023-03-29 analysis/paper/005-raman-spectra-summary.R
Layout changes

@@ -76,7 +76,7 @@ p.c <-
     xlab("Wavenumber (cm-1)") +
     ylab("Intensity") +
     scale_color_manual(name = "Colourless",
-                       values = "#9467BDFF") + 
+                       values = "#56B4E9") + 
     theme_classic() +
     theme(legend.position = "none",
           axis.title.x = element_blank(),
@@ -91,7 +91,7 @@ p.d <-
     xlab("Wavenumber (cm-1)") +
     ylab("Intensity") +
     scale_color_manual(name = "Dark",
-                      values = "black") + 
+                       values = "#36454F") + 
     theme_classic() +
     theme(legend.position = "none",
           axis.title.x = element_blank(),
@@ -106,7 +106,7 @@ p.l <-
     xlab("Wavenumber (cm-1)") +
     ylab("Intensity") +
     scale_color_manual(name = "Light",
-                       values = "#FF7F0EFF") + 
+                       values = "#E69F00") + 
     theme_classic() +
     theme(legend.position = "none",
           axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
@@ -121,7 +121,7 @@ p.w <-
     xlab("Wavenumber (cm-1)") +
     ylab("Intensity") +
     scale_color_manual(name = "White",
-                       values = "#1F77B4FF") + 
+                       values = "#CC79A7") + 
     theme_classic() +
     theme(legend.position = "none",
           axis.title.x = element_text(size = 12, face = "bold", colour = "black"),

### 2023-03-29 analysis/paper/007-raman-pca.R
Layout changes

@@ -101,7 +101,7 @@ raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
 raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
 raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")
 
-pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
+pca.colors <- c("#56B4E9", "#36454F", "#E69F00", "#F9F6EE")
 pca.hue <- c("Colourless", "Dark", "Light", "White")
 
 #prepare a basic score plot with fviz_pca_ind using PC1 and PC2
@@ -113,28 +113,29 @@ basic_plot1 <-
 fig <- 
   ggplot(cbind(basic_plot1$data, Points.raman[, c(7,10)]),
          aes(x=x, y=y, shape = material, fill = hue)) +
+  #to add sample id as text
   #geom_text(aes(label=Points.raman$sample_id, hjust=0.5,vjust=-1.0)) +
-  geom_point(size=3) +
+  geom_point(size=3.5) +
   theme_bw() +
   ggtitle(bquote(bold("Raman 100 - 1 800 " ~cm^-1))) +
   labs(x = raman.pc1lab,
        y = raman.pc2lab) +
   scale_shape_manual(name = "Material",
-                     values=c(24,22,21)) +
+                     values=c(25,24,21)) +
   scale_fill_manual(name = "Hue",
                     values=pca.colors) +
-  guides(shape = guide_legend(override.aes = list(fill = "black"), 
-                              ncol = 3,
-                              title.position="top", 
-                              title.hjust = 0.5,
-                              order = 2),
-         fill = guide_legend(override.aes = list(shape = 22,
+  guides(fill = guide_legend(override.aes = list(shape = 22,
                                                  fill = pca.colors,
                                                  ncol = 4,
                                                  color = "black",
                                                  size=3,
                                                  order = 1),
-                             title.position="top", title.hjust = 0.5))+
+                             title.position="top", title.hjust = 0.5),
+         shape = guide_legend(override.aes = list(fill = "black"), 
+                              ncol = 3,
+                              title.position="top", 
+                              title.hjust = 0.5,
+                              order = 2)) +
   theme(plot.title = element_text(size = 12, colour = "black", vjust = 1, hjust = 0.02),
[…]

### 2023-03-29 analysis/paper/008-xrf-violin-plot.R
Color changes

@@ -65,11 +65,7 @@ Points_long <-
   group_by(Element) %>% 
   dplyr::mutate(ypos = max(Value) * 0.95)  
 
-#Prepare colors
-loading.colors <- c("lightgoldenrod1", "darkorange", "lightpink1")
-
 #Lookup for element labeller
-#Ändra ordning så att de mer avvikande (Si, Mn etc) hamnar uppe till vänster
 Elements <- c(
   'Mg' = "Mg",
   'Al' = "Al",
@@ -147,7 +143,8 @@ fig <-
     see::geom_violinhalf(trim = TRUE) +
     geom_boxplot(width=.1, outlier.shape = NA, position= position_nudge(x=-.1)) +
     facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
-    #facetted_pos_scales(y = scales) +
+    scale_fill_manual(name = "Group",
+                      values = c("#E69F00", "#56B4E9", "#CC79A7")) +
     theme_bw() +
     geom_text(data=na_values, aes(x=0.25, y=ypos, label=na_values, fontface = "bold"), nudge_x = 0.4, inherit.aes = FALSE, size = 4) +
     theme(

### 2023-03-29 analysis/paper/009-xrf-pca.R
Layout changes and additional comments

@@ -73,12 +73,12 @@ xrf <-
 #### PCA full ####
 
 #log-transform data in order to address issues with closure
-#clr can also be used, but gives similar results
+#clr can also be used, in that case comment this log10 code, but gives similar results in this instance
 xrf.log <- 
   log10(xrf[,5:16])
 
 
-#if you want to transform data using centered log-ratio
+#remove comment from code below if you want to transform data using centered log-ratio
 #potential issues due to using geometric mean
 #xrf.clr <- 
 #  compositions::clr(xrf[,5:16], ifclose = TRUE)
@@ -98,7 +98,7 @@ xrf.pc3lab <- paste0("PC3 (",as.character(xrf.pc3var),"%)")
 
 
 #prepare color/fill and symbols for score plots
-pca.colors <- c("#9467BDFF", "#7F7F7FFF", "#FF7F0EFF", "#1F77B4FF")
+pca.colors <- c("#56B4E9", "#36454F", "#E69F00", "#F9F6EE")
 pca.hue <- c("Colourless", "Dark", "Light", "White")
 
 #Extract loadings from xrf.pca
@@ -192,10 +192,10 @@ xrf.low <-
   column_to_rownames(var = "reading_no") 
 
 #log-transform data in order to address issues with closure
-#clr gives similar results in this case, at least in terms of exploration of the dataset
 xrf.log.low <- 
   log10(xrf.low[,5:16])
 
+#clr gives similar results in this case, at least in terms of exploration of the dataset
 #xrf.clr <- 
 #  compositions::clr(xrf.low[,5:16], ifclose = TRUE)
 
@@ -396,7 +396,6 @@ ggsave("009-xrf-pca-load-scree.png",
        units = "cm",
        dpi = 300)
 
-
 #Biplot for the full dataset, with ellipses showing the results of the k-means cluster analysis 
 fig2a <- 
   fviz_pca_biplot(xrf.pca,
@@ -415,12 +414,13 @@ fig2a <-
   labs(title = "XRF PCA Biplot", 
        x = xrf.pc1lab,
        y = xrf.pc2lab) +
+  #to add sample id as text
   #geom_text(aes(label=xrf$sample_id, hjust=0.5,vjust=-1.0)) +
   geom_point(aes(fill = xrf$hue, 
                  shape = factor(xrf$material)), 
              size = 2) +
   scale_shape_manual(name = "Material", 
-                     values=c(24,22,21),
+                     values=c(25,24,21),
                      guide = guide_legend(override.aes = list(fill = "black"),
[…]

### 2023-03-29 analysis/paper/010-nir-sample-comparison.R
Layout changes

@@ -186,7 +186,7 @@ pn.231 <-
   labs(title="386") +
   scale_color_manual(name = "386",
                      values = "black") + 
-  scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 5)) +
+  scale_x_continuous(limits=c(1000,2500)) +
   theme_classic() +
   theme(legend.position = "none",
         plot.title = element_text(size=16, 
@@ -203,8 +203,8 @@ pn.404 <-
   #ylab("Absorbance") +
   labs(title="404") +
   scale_color_manual(name = "404",
-                     values = "black") + 
-  scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 5)) +
+                     values = "black") +
+  scale_x_continuous(limits=c(1000,2500)) +
   theme_classic() +
   theme(legend.position = "none",
         plot.title = element_text(size=16, 
@@ -223,7 +223,7 @@ pn.391 <-
   labs(title="391") +
   scale_color_manual(name = "391",
                      values = "black") + 
-  scale_x_continuous(limits = c(1000, 2500), breaks = scales::pretty_breaks(n = 5)) +
+  scale_x_continuous(limits=c(1000,2500)) +
   theme_classic() +
   theme(legend.position = "none",
         plot.title = element_text(size=16, 

### 2023-03-29 analysis/paper/paper.Rmd
Updated manuscript

@@
 ---
-title: "The Quartzite Conundrum: Non-destructive analysis of Bronze Age quartzite points from Västerbotten, Sweden"
+title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
 author:
@@
 header-includes:
-    \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
 bibliography: references.bib
@@
 Provenance of locally available geological material thus becomes complex and fraught with uncertainty.
-This leads to a severe lack of methodological development in the characterization of these same materials in an archaeological context.
 Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region.
+However, there is a severe lack of methodological development in the characterization of the same these materials.
 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden.
 The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence).
-The data was evaluated with exploratory analysis using Principal Component Analysis.
-The results demonstrate that detectable differences exist within the dataset for each instrumentation.
-A number of features identified, such as the presence of graphite, can likely be related to diagenetic/paragenetic differences in the material.
-Not only should these results aid in future surveys of geological sources in the region, but they will also aid in the construction of predictive models for classifying quartzite.
+Evaluation of the spectra and exploratory analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin and potentially the metamorphic grade.
+These features indicate the presence of graphite, muscovite, and biotite.
+Apart from reaffirming the potential of field-based screening instrumentation, these results should aid in future surveys of geological sources in the region.
+There also seems to be a good ground for the construction of a predictive model which that could classify the quartzite based on its chemical characteristic.
+Such a model could prove useful in spatial analysis and testing of models of raw material management.
 keywords: |
@@
 
-Petrographic characterization and provenance studies are instrumental in the modelling and inference of trade and exchange networks among prehistoric societies (REF).
+Petrographic characterization and provenance studies are instrumental for inferences of trade and exchange among prehistoric societies [@tykot03].
 The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery.
 The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources.
-Flint and obsidian have both been central to the study of trade and exchange networks in European prehistory, due in part to the well-surveyed geological sources for these materials.
+Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17].
 The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].
-The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent^[This is not to say that flint was not available or used in these regions] has been highlighted by Prieto et al.
-[-@prieto_etal19 p.
-15].
-In Northern Sweden both quartz and quartzite are instrumental in the inference of settlement and mobility within the region as they represent the most common material groups at archaeological sites.
-While quartz and quartzite technology have been the target of a number of studies over the years, there have been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
-While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden has been put forth [@halen94 pp.
-115-122; @lundberg97 pp.
-161-171], in contrast, there has been little development on the topic of quartz and quartzite.
+The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent has been highlighted by Prieto et al.
+[-@prieto_etal19: 15].
+In Northern Sweden both quartz and quartzite dominate the lithic assemblages and have been used in studies of settlements, exchange and mobility [@baudou78].
+While quartz and quartzite technology have been the target of several studies over the years, there has been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10].
+M While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden haves been put forth [@halen94: 115-122; @lundberg97: 161-171].
+On the other hand there haves been little development on the topic of quartz and quartzite [@holm91: 23-24].
 A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16].
-The abundance of moraine formations and the degree to which geological materials have been mixed and transported necessitates considerable effort in order to reconstruct the events that have formed the current landscape.
-Existing theories on lithic material sourcing within the North Swedish region tend to describe people employing strategies leaning towards direct procurement from the mountain zone [@forsberg85; -@forsberg12].
-This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06 pp.249-250].
-Due to this ambiguous nature of availability of raw materials provenience studies are complex and fraught with uncertainty.
+The abundance of moraine deposits and the degree to which geological materials have been mixed and transported necessitates considerable effort to reconstruct the events that have formed the current landscape.
+Existing theories on lithic material sourcing within the North Swedish region tend to describe strategies regarding direct procurement from the mountain zone [@forsberg85; -@forsberg12].
+This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06: 249-250].
+Due to this ambiguous nature of availability of raw materials, provenance studies are complex.
 
 Prieto et al.
-[-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
-It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05].
-Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies.
-Spectroscopic instrumentation features fast measurements that can be collected directly on the surface of an object.
-With newer mobile instrumentations readily available, this makes it an attractive option for field-based screening purposes.
-In the case of archaeological research it is also worth considering situations where destructive techniques, such as thin-section analysis, are not suitable or applicable.
+[-@prieto_etal19: 15; -@prieto_etal20: 32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite.
+It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature.
+Because of this , and different researchers have thus emphasized different different features as being characteristics of the material [@howard05].
+Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05: 708] also emphasises the need for field-based methodologies.
+Spectroscopic surface measurements are fast, and can be adapted for use in the field.
+In the case of archaeological research there are situations where destructive techniques, such as thin-section analysis, are not suitable or applicable.
 Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.
-While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information.
-While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other complementary instrumentations that may provide further insight into the structural properties of a material.
-The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
-The same can be said for Raman spectroscopy, which has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].
+
+X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03]), due to its capability of providing element compositional data.
+The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF, as described by Norris [-@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
+Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].
 
@@
 The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (fig.
-\@ref(fig:nir-spectra-summary)).
+\@ref(fig:distribution-map)).
 The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources.
@@
 
-The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden.
-The sites are distributed along the Ångermanälven river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south.
-A dataset of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
+The current study includes spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden (fig.
+\@ref(fig:distribution-map)).The sites are distributed along the Ångerman river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south.
+The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF.
 No distinction is made between arrow- and spearheads.
@@
 A number of these points have, according to prior classifications, been worked using bifacial technology.
-While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, @mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca.
-3 800 - 2 000 BP [@kaeck09 p.60].
+The geographically closest chronology for bifacial points can be found in southeast Norway [@mjaerum12], but so far few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89], ).
+These attempts suggests a period of use between ca.
+3 800 - 2 000 BP [@kaeck09: 60].
 Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].
@@
 ```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine.
-Made with ©SGU, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
+Tectonic symbology after [@gee_stephens20].
+Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
 knitr::include_graphics(here::here("analysis", "figures", "site_distribution_soils_annotated.png"), error = FALSE)
@@
 
-When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67 p.8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
+When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden.
 Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis.
 *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g.
-@holm91 p.24].
+@holm91: 24].
 As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.
@@
 
-```{r sample-hue-pictures, fig.cap= "Not only is the material heterogeneous in terms of colour, but the surface of the artefacts themselves is not uniform.
-Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation."}
+```{r artefacts, fig.cap= "The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform.
+Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation.
+Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 2-76 mm.
+"}
 knitr::include_graphics(here::here("analysis", "figures", "figure_2.png"), error = FALSE)
@@
 
-All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/en/research/infrastructure/mal/].
+All measurements were performed at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/).
 The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens.
-The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape.
-As the study is non-destructive sampling was performed on the surface of a material that is not uniform.
+The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape (fig.
+\@ref(fig:artefacts)-\@ref(fig:artefact-size)) [see Appendix A](#appendix-a).
+The study is non-destructive sampling was performed on the surface of a material that is not uniform.
 The material ranges from fairly small and flat point fragments, to large and rough preforms.
-Therefore the integration could at times be difficult, which in turn could lead to scattering effects and saturation.
+Therefore, the signal integration could at times be difficult, which in turn could lead to scattering effects and saturation.
 
 Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult.
-When sampling materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.
+Sampling materials that featured high heterogeneity the sample point was focused on the surrounding matrix, away from deviating inclusions.
 An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions.
@@
 
-(ref:artefact) Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts. Note that the term ""point"" does not differentiate between spearheads or arrowheads.
+(ref:artefact) Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts.
+8 objects have no recorded thickness, which ranges between 2-76 mm.
+Note that the term *point* does not differentiate between spearheads or arrowheads.
 
@@
 
-## Instrumentaion and pre-processing {-}
+## Instrumentation and data pre-processing {-}
 
@@
 NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum.
+Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near infrared region.
+The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify the bond.
+Overtones occur at frequencies that are a multiple of the infrared vibration.
+Most of them end up in the near infrared region [@osborne_etal93].
+
 Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups.
-The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77 p.508].
+The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508].
 
@@
 The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser.
-The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^.
-Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå^[www.slu.se/en/departments/forest-biomaterials-technology/].
-
+The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^.
 After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data.
-This was achieved using modified polynomial fitting [@lieber_mahadevan-jansen03] from the ChemoSpec package [@hanson16].
-The spectra were then mean-centred <!--and normalized using Standard Normal Variate [@shaver01 pp.284-285] --> before PCA.
+This was achieved using Lieber and Mahadevan-Jansens' [-@lieber_mahadevan-jansen03] algorithm for polynomial fitting, with the ChemoSpec package [@hanson16].
+The spectra were then mean-centred before PCA.
 
@@
 XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a].
-Moseley's law [@moseley1913] recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms.
-This is what enables the spectral features in XRF spectrometry to be related to certain elements.
-Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12 pp.
-143-159].
+Spectral features in XRF spectrometry can be related to certain elements according to Moseley's law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24].
+Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159].
 In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample.
@@
 XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand.
-The reference calibration used for element quantification was *mining* mode [@knight_etal21 p.2].
+The reference calibration used for element quantification was *mining* mode [@knight_etal21: 2].
 The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light).
@@
 
-It is also important to address the *closed sum* effect of compositional data [@pawlowsky-glahn_egozcue06].
-As the variables are given as proportions of a constant sum (e.g.
-percentage, parts per million) this means that if one element was to change in value, others would follow [@grunsky_caritat20].
-This is managed by transforming the data using log-ratio transformation [@aitchison99], and currently there are three main methods for achieving this; additive log-ratio (alr), centered log-ratio (clr) or isometric log-ratio.
-They all have their own nuances and as such the choice of method is dependent on the type of analysis and material [@pawlowsky-glahn_etal15].
+As the variables in XRF analysis is given as proportions of a constant sum (e.g.
+percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06].
+This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20].
+This has implications for classical statistical methods, and is managed by transforming the data using log-ratio transformation [@aitchison99].
+There are currently three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15].
 In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g.
@@
 The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20].
-Multivariate approaches, such as PCA,  has proven useful for exploring complex spectroscopic datasets [@sciuto18].
+Multivariate approaches, such as PCA,  were proven useful for exploring complex spectroscopic datasets [@sciuto18].
 Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material.
-These groupings can then be followed up with supervised methods, and aid in the construction of predicitive models.
+These groupings can then be followed up with supervised methods, and aid in the construction of predictive models.
 
@@
 points.d.filt <-
-  metadata.csv %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
-           site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
-           site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" | site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | 
-           site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" | site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | 
-           site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" | site_id == "Vilhelmina 458" | 
-           site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | 
-           site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
-           site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
-           site_id == "Åsele 99", 
-         type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  points.d.tot %>% 
   filter(!sample_id %in% c("153","167","168","169","172","174","175","176","177","182","190","191","193","194","196","198","200","204","214",
                            "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
-                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56")) %>% 
-  filter(hue == "Dark")
+                           "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56"))
 
-#pass to vector so that length() can be used in text to quote nr. of dark samples
+#pass to vector so that length() can be used in text to quote total nr. of samples
 spec.d.total <- 
@@
 
-#pass to vector so that length() can be used in text to quote nr. of dark samples
+#pass to vector so that length() can be used in text to quote nr. of dark samples after SG analysis
 spec.d.filtered <- 
@@
 
-The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81].
-2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21 pp.
-351-397; @schmid_etal22].
+The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17: 78-81].
+2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the risk of amplifying noise [@mark_workman21: 351-397; @schmid_etal22].
 An example of this can be seen in fig.
@@
 These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig.
-\@ref(fig:nir-savitzky-transform-alt) & \@ref(fig:nir-peak-summary)) [@hunt77 p.
-508-510; @sciuto_etal19].
+\@ref(fig:nir-savitzky-transform-alt) & \@ref(fig:nir-peak-summary)) [@hunt77: 508-510; @sciuto_etal19].
 OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region.
-As a general rule, the 2 200 nm and 2 350 nm positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664].
+As a general rule, the 2 200 nm and 2 350 nm positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77: 509; @clark_etal90: 12664].
 While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH.
@@
 
-The above absorption bands can be found in mica, such as muscovite and biotite, which would not be strange to find in quartzite [@clark_etal90 p.12; @hunt77 p.510; @ramanaidou_etal15 p.207].
+The above absorption bands can be found in mica, such as muscovite and biotite, which would not be strange to find in quartzite [@clark_etal90: 12; @hunt77: 510; @ramanaidou_etal15: 207].
 Although muscovite and biotite share some band positions, the altered Al content as a response to metamorphism has been documented to shift some positions and increase others, such as the 2250 nm band for biotite [@duke94].
@@
 \@ref(fig:nir-savitzky-transform-alt)), and are often vague and difficult to make out in the non-processed spectra.
-These could be caused by ferrous iron (Fe^2+^) [@hunt77 pp.503-504], which has been demonstrated to produce a response in the NIR in a number of minerals.
+These could be caused by ferrous iron (Fe^2+^) [@hunt77: 503-504], which has been demonstrated to produce a response in the NIR in a number of minerals.
 In studies of olivine the absorption region has been determined to be broad and centered around 1 000 nm, but with three features at 900 nm, 1 100 nm and 1 300 nm [@trang_etal13].
-Weak water features also have been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water, however, these overlap with the Fe^2+^ bands in, for instance, biotite [@adams75 pp.104-106].
+Weak water features also have been noted in the 950 nm and 1 150 nm bands, particularly among clays with adsorped water, however, these overlap with the Fe^2+^ bands in, for instance, biotite [@adams75: 104-106].
 It is possible that the features at 1 065 nm and 1 300 nm can be linked to the presence of ferrous iron, but it is difficult to say what the contributing mineral is at this stage.
@@
 
-```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca.R")), encoding = "UTF-8", echo = FALSE}
+```{r nir-pca, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "004-nir-pca.R")), encoding = "UTF-8", echo = FALSE}
 quietly(source(here::here("analysis", "paper","004-nir-pca.R"), encoding = "UTF-8"))
@@
 
-The loading lines for the PCA can be seen in [Appendix A](#appendix-a).
+The loading lines for the PCA can be seen in [Appendix B](#appendix-B).
 The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig.
@@
 
-While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) (Appendix \@ref(fig:nir-pca-loadings)).
+While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1) ([Appendix A](#appendix-a)).
 Adding PC 3 to a score plot (fig.
@@
 
-```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separating out a couple of light samples.", echo = FALSE}
-#quietly(source("./analysis/paper/004-nir-pca-alt.R", encoding = "UTF-8"))
+```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separate three of the light samples.", echo = FALSE}
 knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.png"), error = FALSE)
@@
 Whereas the 2^nd^ peak falls within the expected position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region.
-It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280].
+It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07: 1280].
 This could thus explain the discrepancy between the observed peak in the dataset and the literature.
@@
 ```{r raman-spectra-summary, fig.cap= "The non-processed Raman spectra visualised by colour group.
-Note that the dark material in general feature more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "005-raman-spectra-summary.R")), echo = FALSE}
+Note that the dark material in general features more intense peaks at wavenumber ~ 1 300 - 1 600 (cm^-1^).", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "005-raman-spectra-summary.R")), echo = FALSE}
 quietly(source(here::here("analysis", "paper", "005-raman-spectra-summary.R"), encoding = "UTF-8"))
@@
 
-The loading lines for the PCA can be seen in [Appendix A](#appendix-a).
+The loading lines for the PCA can be seen in [Appendix B](#appendix-b).
 The PCA model (fig.
@@
 ```{r raman-pca, fig.cap= "PCA score plot for Raman data.
-Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributeted towards the right of the plot.
+Objects that feature only quartz peaks are clustered towards the top left of the plot, whilst the darker material that feature dominant D/G peaks are distributed towards the right of the plot.
 In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca.R")), echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "raman-pca-with-photos.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.png"), error = FALSE)
 ```
@@
 -->
-The PCA was run twice, once with the > 90% ~14~Si samples included and a second time without them.
-The loadings can be seen in [Appendix A](#appendix-a).
+The PCA was run twice, firstly with the > 90% 14Si samples included and secondly excluded.
+The loadings can be seen in [Appendix B](#appendix-b).
 The initial PCA (model A), with the full dataset included, resulted in 3 PCs with an eigenvalue above 1, and a cumulative explained variance of `r xrf_var`%.
@@
 ~13~Al and ~19~K is similarly higher in cluster 2 than compared to cluster 3, which is marked by its, on average, lower elemental content than the other two clusters.
+As ~13~Al and ~19~K signifies a sedimentary depositional context that has been subjected to metamorphism, this may be a possible explanation.
 That said, cluster 3 does features a more limited range of values between the samples.
@@
 ```{r xrf-violin-plot, fig.cap= "Violin plot of the XRF values per element.
-Median and interquartile range represented with a boxplot, surrounded by a kernel density estimation.
+Median and interquartile range represented with a boxplot, with the half-violin showing kernel density estimation.
 Values above LOD is noted by *n*.
-The bimodal distribution of Si is visible in the plot, as well as the low number of values above LOD for V, Y and Zn.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "008-xrf-violin-plot.R")), echo = FALSE}
+Of particular note is the bimodal distribution of Si values.
+Y and Zn were left out as they had less than 30 values above LOD.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "008-xrf-violin-plot.R")), echo = FALSE}
 quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
@@
 ```{r nir-sample-comparison, fig.cap= "Comparison of three dark samples from separate K-cluster groups (model B).
-Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples.
-The main difference between samples 386 and 404 seems to mainly be in the XRF data.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "010-sample-comparison.R")), echo = FALSE}
+Sample 391 from cluster 3 stands out the most with visible differences in both the NIR and Raman spectra, when compared to the other samples.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "010-sample-comparison.R")), echo = FALSE}
 quietly(source(here::here("analysis", "paper", "010-nir-sample-comparison.R"), encoding = "UTF-8"))
@@
 @gaweda_cebulak99; @ukar_cloos16].
-Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20 pp.
-482-ff], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.
+Considering the presence of Alum shale and similar carbon-rich tectonic strata within the Scandian mountain chain [@stephens_bergmanweihed20: 482-599], it would not be strange to encounter carbon structures in quartzite or quartzite-like material.
 
@@
 @prieto_etal19]).
-Previous studies have primarily relied on visual characterization following a system that was setup in order to manage the large quantities of data generated following the large scale development for hydroelectric power during the 1940s - 1980s [@biornstad06; @kaeck09 pp.53-55].
+Previous studies have primarily relied on visual characterization following a system that was setup in order to manage the large quantities of data generated following the large scale development for hydroelectric power during the 1940s - 1980s [@biornstad06; @kaeck09: 53-55].
 This was necessary in order to manage the overabundance of data and make sense of it at the time, and provided a lot of new insights [e.g.
@@
 # CRediT authorship contribution statement {-}
-Mattias Sjölander: Conceptualization, Data curation, Methodology, Software, Validation, Formal analysis, Resources, Writing - original draft, Visualization, Project administration.
+Mattias Sjölander: Conceptualization, Data curation, Formal analysis, Methodology, Project administration, Resources, Software, Visualization, Validation, Writing - original draft.
 
-Johan Linderholm: Conceptualization, Methodology
+Johan Linderholm: Conceptualization, Methodology, Supervision
 
-Paul Geladi: Methodology
+Paul Geladi: Formal analysis, Methodology
 
@@
 # Funding source {-}
-This work was supported by
+This work was supported by PhD fundings and facilities at the Environmental Archaeology Laboratory at Umeå University.
 
@@
 
-We would like to thank Västerbottens museum (www.vbm.se) for giving us access to their collections, and Susanne Sundström for her assistance with putting the artefacts together.
-We would also like to acknowledge the XXX consortium for providing access to the instrumentation used in this study.
+We would like to thank Västerbottens museum (www.vbm.se) for access to their collections, and Susanne Sundström in particular for her assistance with preparing the artefacts.
+We would also like to acknowledge good cooperation with the division of Biomass Technology and Chemistry at the Swedish University of Agricultural Sciences.
 
@@
 ## Appendix A {#appendix-a}
+
+\beginsupplement
+
+\linebreak
+
+(ref:points) Metadata for the `r length(tot_uniq_samples)` samples included in the study.
+
+```{r point-data, echo = FALSE, results = 'asis'}
+knitr::kable(tot_samples, caption = "(ref:points)")
+```
+
+\endsupplement
+
+## Appendix B {#appendix-b}
 
@@
 
+\endsupplement
+
 \newpage

### 2023-04-03 analysis/paper/paper.Rmd
Updated manuscript

@@
 institute:
-  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Sweden
-  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Sweden 
+  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
+  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 Apart from reaffirming the potential of field-based screening instrumentation, these results should aid in future surveys of geological sources in the region.
-There also seems to be a good ground for the construction of a predictive model which that could classify the quartzite based on its chemical characteristic.
+There also seems to be a good ground for the construction of a predictive model that could classify the quartzite based on its chemical characteristic.
 Such a model could prove useful in spatial analysis and testing of models of raw material management.
@@
   metadata.csv %>% 
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
                 site_id == "Vilhelmina 114" | site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | 
@@
                 site_id == "Åsele 101" | site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | 
-           site_id == "Åsele 182" |  site_id == "Åsele 188" |site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
+                site_id == "Åsele 182" | site_id == "Åsele 188" | site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | 
+                site_id == "Åsele 99", 
          type == "Point" | type == "Point fragment" | type == "Preform", 
@@
 
-X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03]), due to its capability of providing element compositional data.
+X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data.
 The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF, as described by Norris [-@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18].
@@
 Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
-knitr::include_graphics(here::here("analysis", "figures", "site_distribution_soils_annotated.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "site_distribution_v1.jpg"), error = FALSE)
 ```
@@
 "}
-knitr::include_graphics(here::here("analysis", "figures", "figure_2.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "figure_2.jpg"), error = FALSE)
 ```
@@
 (ref:artefact) Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts.
-8 objects have no recorded thickness, which ranges between 2-76 mm.
+8 objects have no recorded thickness, which ranges between 3-76 mm.
 Note that the term *point* does not differentiate between spearheads or arrowheads.
@@
 quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.jpeg"), error = FALSE)
 ```
@@
 The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^, with a resolution of 9.5 cm^-1^.
+Integration time was set to 20 seconds with 30% power.
 After averaging the spectra for each artefact the results were baseline corrected in order to detrend the data.
@@
 Following log-transformation the data was then mean-centered and scaled to unit variance before PCA.
+Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].
 
@@
 
-PCA analysis was performed using the base R *prcomp()* function [@Rteam21; @rcoreteam21].
+PCA analysis was performed using the base R *prcomp()* function [@Rteam21a; @Rteam21b].
 
@@
 quietly(source(here::here("analysis", "paper", "001-nir-spectra-summary.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "001-nir-spectra-summary.jpeg"), error = FALSE)
 ```
@@
 quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform.jpeg"), error = FALSE)
 ```
@@
 quietly(source(here::here("analysis", "paper", "003-nir-savitzky-transform-alt.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "003-nir-savitzky-transform-alt.jpeg"), error = FALSE)
 ```
@@
 quietly(source(here::here("analysis", "paper", "002-nir-peak-summary.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "002-nir-peak-summary.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "002-nir-peak-summary.jpeg"), error = FALSE)
 ```
@@
 ```{r nir-pca-fig, fig.width=20, fig.height=20, fig.cap= "NIR PCA score plot for PC 1-3: a) dark and light material is separated out along PC 1, due to absorbance, with white and colourless material separating along PC 2, b) PC 3 adds little new information, mainly separate three of the light samples.", echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-with-photos.jpg"), error = FALSE)
 ```
@@
 quietly(source(here::here("analysis", "paper", "005-raman-spectra-summary.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "005-raman-spectra-summary.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "005-raman-spectra-summary.jpeg"), error = FALSE)
 ```
@@
 
-```{r raman-peak-ident, fig.cap= "The two features that show up in the Raman spectra can be related to quartz and carbon structures in the material.
+```{r raman-peak-ident, fig.cap= "The features in the Raman spectra can be related to quartz and carbon structures in the material.
 The initial three peaks in the 100 - 500 cm^-1^ range correspond to some of the more well-known vibration modes of quartz.
@@
 quietly(source(here::here("analysis", "paper", "006-raman-peak-identification.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "006-raman-peak-identification.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "006-raman-peak-identification.jpeg"), error = FALSE)
 ```
@@
 In the middle cluster the samples feature both quartz and D/G peaks, with those that have higher intensity D/G peaks in the bottom of the cluster.", cache=TRUE, cache.extra=tools::md5sum(here::here("analysis", "paper", "007-raman-pca.R")), echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-with-photos.jpg"), error = FALSE)
 ```
@@
 
-<!--The initial PCA model of the full dataset produced 5 outliers, which significantly impacted the interpretation of the data (sample 183, 193, 259, 359 and 400).
-These samples were therefore excluded and the PCA run again, which resulted in 4 PCs with an eigenvalue above 1 (fig.
-\@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%.
-K-means cluster analysis was then performed on the PCA data in order to help identify potential groups exhibiting similar chemofacies.
-Using the elbow graph method 3 clusters were deemed suitable for the analysis.
-
-As can be seen in the score plot of PC 1-2 (fig.
-\@ref(fig:xrf-pca-a)), the three K-means clusters remain separated with next to no overlap.
-Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%.
-It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue.
-Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD.
-Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples.
-As in the case of cluster 3, apart from 1 outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%.
-Except for ~22~Ti, ~26~Fe and ~40~Zr, the cluster features comparable values to cluster 3.
-Where it stands out is in the large amount of samples with ~38~Sr readings.
-~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays, and has been the focus of study of carbonate diagenesis/paragenesis [@ahijado_etal05;@banner95].
--->
+Initial analysis used clr transformed data, but encountered issues related to the structure of the ~14~Si values.
+log10-transformation was used instead, producing a similar model, but that managed the loading for ~14~Si better.
 The PCA was run twice, firstly with the > 90% 14Si samples included and secondly excluded.
@@
 quietly(source(here::here("analysis", "paper", "008-xrf-violin-plot.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "008-xrf-violin-plot.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "008-xrf-violin-plot.jpeg"), error = FALSE)
 ```
@@
 Cluster 1 stands out by featuring slightly lower Si, Fe, Ti, and Zr values, but higher values in Al, Ca, K, and Mg.", echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-with-photos.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-with-photos.jpg"), error = FALSE)
 ```
@@
 Cluster 2 includes the samples with higher Fe/Ti/Zr content, whilst cluster 3 is mainly distinguished by a higher Si content and low Fe/Ti/Zr content.", echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-xrf-kmeans-boxplot.jpeg"), error = FALSE)
 ```
@@
 ```{r nir-xrf-kmeans, fig.cap= "NIR PCA score plot for PC 1-2, with the K-means cluster from XRF PCA model B visualised in the colour of the points.
-White points are samples that were excluded in the XRF PCA.
+Red points are samples that were excluded in the XRF PCA.
 The trend separating dark-light samples in the NIR PCA mostly keeps cluster 2-3 separated, showing some overlap in the results, but cluster 1 is fairly evenly distributed in the score plot.
 This means that whatever differences can be discerned in the compositional data does not seem to be detectable using the ASD.", echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "009-nir-pca-xrf-kmeans.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-nir-pca-xrf-kmeans.jpeg"), error = FALSE)
 ```
@@
 quietly(source(here::here("analysis", "paper", "010-nir-sample-comparison.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "010-nir-sample-comparison.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "010-nir-sample-comparison.jpeg"), error = FALSE)
 ```
@@
 
+**Competing interests**
+The authors declare no competing interests.
+
 # Funding source {-}
-This work was supported by PhD fundings and facilities at the Environmental Archaeology Laboratory at Umeå University.
+This work was supported by PhD funding and facilities at the Environmental Archaeology Laboratory at Umeå University.
 
@@
 
-# (APPENDIX) Appendix {-}
+# Appendices{-}
 
@@
 
-(ref:points) Metadata for the `r length(tot_uniq_samples)` samples included in the study.
-
 ```{r point-data, echo = FALSE, results = 'asis'}
-knitr::kable(tot_samples, caption = "(ref:points)")
+suppressPackageStartupMessages(library(flextable))
+#show basic point metadata in a flextable
+tot_samples %>% 
+  select(sample_id, inv_nr, sub_nr, site_id, material, hue, type, max_length_mm, max_width_mm, max_height_mm) %>% 
+  flextable() %>% 
+  set_header_labels(sample_id = "Sample ID", inv_nr = "Inventory Nr.", sub_nr = "Sub Nr.", site_id = "National Site ID", material = "Material", hue = "Hue", type = "Type", 
+                    max_length_mm = "Length(mm)", max_width_mm = "Width(mm)", max_height_mm = "Thickness(mm)") %>% 
+  theme_vanilla() %>% 
+  set_caption(caption = "Sample metadata")
 ```
@@
 \endsupplement
+
+\newpage
 
@@
 quietly(source(here::here("analysis", "paper", "004-nir-pca-loadings.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "004-nir-pca-loadings.jpeg"), error = FALSE)
 ```
@@
 quietly(source(here::here("analysis", "paper", "007-raman-pca-loadings.R"), encoding = "UTF-8"))
-knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "007-raman-pca-loadings.jpeg"), error = FALSE)
 ```
@@
 a) eigenvalues and loadings for the full dataset (model A), b) eigenvalues and loadings with samples measuring > 90% ~14~Si excluded (model B).", echo = FALSE}
-knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.png"), error = FALSE)
+knitr::include_graphics(here::here("analysis", "figures", "009-xrf-pca-load-scree.jpeg"), error = FALSE)
 ```

### 2023-04-03 analysis/paper/001-artefact-size.R
updated to save as jpg

@@ -6,26 +6,28 @@ suppressPackageStartupMessages(library(tidyverse))
 metadata.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
-#Filter nir data to focus on points and preforms made from quartz/quartzite material 
+#Filter data to focus on points and preforms made from quartz/quartzite material in Vilhelmina and Åsele parish
 Points <-
-  metadata.csv %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-                  site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-                  site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-                  site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-                  site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
-                  site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
-                  site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
[…]

### 2023-04-03 analysis/paper/001-nir-spectra-summary.R

@@ -19,16 +19,18 @@ nir.merged <-
 #Filter nir data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
-           site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
-           site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
-         type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
[…]

### 2023-04-03 analysis/paper/002-nir-dark-filtered-summary.R

@@ -18,16 +18,18 @@ nir.merged <-
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
-           site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
-           site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
-         type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
[…]

### 2023-04-03 analysis/paper/002-nir-peak-summary.R

@@ -18,14 +18,16 @@ nir.merged <-
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
-  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-                  site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-                  site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-                  site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-                  site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-                  site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
-                  site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
-                  site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
[…]

### 2023-04-03 analysis/paper/003-nir-savitzky-transform-alt.R

@@ -23,16 +23,18 @@ nir.merged <-
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
-           site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
-           site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
-           site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
-           site_id == "Vilhelmina 458" | site_id == "Vilhelmina 539" | site_id == "Vilhelmina 542" | site_id == "Vilhelmina 611" | site_id == "Vilhelmina 619" | site_id == "Vilhelmina 636" | 
-           site_id == "Vilhelmina 637" | site_id == "Vilhelmina 643" | site_id == "Vilhelmina 769" | site_id == "Vilhelmina 949" | site_id == "Vilhelmina 95" | site_id == "Åsele 101" | 
-           site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
-           site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
-         type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | 
[…]

### 2023-04-03 analysis/paper/003-nir-savitzky-transform.R

@@ -19,7 +19,7 @@ nir.merged <-
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
            site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
            site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
            site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
@@ -36,7 +36,7 @@ Points.nir <-
 #Filter NIR data to focus on sample 258 and 411, then select the NIR range 1 001 - 2 500 nm, exclude 1000 nm due to filter shift
 raw.spec <- Points.nir %>%
   filter(sample_id == "258" | sample_id == "411") %>% 
-  select(1, c(679:2178))
+  dplyr::select(sample_id, `1001.0`:`2500.0`) 
 
 #Melt into long format, data.table needs to be specified to avoid r using reshape2 version that sets variable to factor
 raw.spec.long <- 
@@ -73,7 +73,7 @@ sg.411.long <- suppressWarnings(data.table::melt(setDT(sg.411), variable.name = 
 p.sg.411 <-
   sg.411.long %>% 
   ggplot() + 
-  geom_line(aes(x = as.numeric(Wavelength), y = as.numeric(Absorbance), colour = ""), size = 1, stat = "identity") +
[…]

### 2023-04-03 analysis/paper/004-nir-pca-loadings.R

@@ -21,7 +21,7 @@ nir.merged <-
 #Filter xrf data to focus on points and preforms made from quartz/quartzite material 
 Points.nir <-
   nir.merged %>%
-  filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
+  dplyr::filter(site_id == "Vilhelmina 1069" | site_id == "Vilhelmina 109" | site_id == "Vilhelmina 112" | site_id == "Vilhelmina 1124" | site_id == "Vilhelmina 1127" | site_id == "Vilhelmina 114" |
            site_id == "Vilhelmina 115" | site_id == "Vilhelmina 117" | site_id == "Vilhelmina 118" | site_id == "Vilhelmina 1254" | site_id == "Vilhelmina 216" | site_id == "Vilhelmina 235" |
            site_id == "Vilhelmina 240" | site_id == "Vilhelmina 245" | site_id == "Vilhelmina 252" | site_id == "Vilhelmina 263" | site_id == "Vilhelmina 335" | site_id == "Vilhelmina 356" |
            site_id == "Vilhelmina 399" | site_id == "Vilhelmina 411" | site_id == "Vilhelmina 419" | site_id == "Vilhelmina 439" | site_id == "Vilhelmina 444" | site_id == "Vilhelmina 450" |
@@ -93,9 +93,9 @@ fig <-
                     nrow = 3,
                     hjust = -1.5)
 
-ggsave("004-nir-pca-loadings.png",
+ggsave("004-nir-pca-loadings.jpeg",
        fig,
-       device = "png",
+       device = "jpeg",
        here::here("analysis", "figures"),
        scale = 1, 
        width=25, 

### 2023-04-03 analysis/paper/004-nir-pca.R

@@ -2,18 +2,13 @@
 suppressPackageStartupMessages(library(factoextra))
 suppressPackageStartupMessages(library(tidyverse))
 
-
 #Import descriptive metadata
 metadata.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220127.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
-
-#aggregate observations by group(sample) and calculate average of wavelength measurements
-#nir.averaged <- 
-#  aggregate(nir.csv[, 4:2154], list(sample_id = nir.csv$sample_id), mean)
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #merge NIR data with metadata
 nir.merged <- 
@@ -36,7 +31,7 @@ Points.nir <-
                            "215","216","229","238","262","265","272","282","359","391","392","393","397","405",
                            "408","410","411","413","414","415","416","417","424","425","428","430","432","55","56")) %>% 
   replace_na(list(munsell_hue = "Colourless")) %>% 
-  group_by(across(sample_id:river)) %>% 
+  group_by(across(sample_id:weight_g)) %>% 
   summarise(across(`350.0`:`2500.0`, mean), .groups = "drop")
 
 #perform PCA with SNV normalization and mean-center
@@ -64,7 +59,7 @@ basic_plot2 <-
 
 #bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
 fig.1 <- 
-  ggplot(cbind(basic_plot1$data, Points.nir[, c(7,10)]),
+  ggplot(cbind(basic_plot1$data, Points.nir[, c(9,12)]),
          aes(x=x, y=y, shape = material, fill = hue)) +
   #to add sample id as text
   #geom_text(aes(label=Points.nir$sample_id, hjust=0.5,vjust=-1.0)) +
@@ -97,7 +92,7 @@ fig.1 <-
         legend.position = "bottom")
 
[…]

### 2023-04-03 analysis/paper/005-raman-spectra-summary.R

@@ -8,7 +8,7 @@ metadata.csv <-
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 
@@ -35,7 +35,7 @@ Points.raman <-
 #Filter raman data to focus on dark samples 
 r.dark <- Points.raman %>%
   filter(hue == "Dark") %>% 
-  select(1, c(30:494))
+  select(sample_id, c(`92.88`:`2503.59`))
 
 #Melt into long format
 r.dark.long <- 
@@ -44,7 +44,7 @@ r.dark.long <-
 #Filter raman data to focus on light samples
 r.light <- Points.raman %>%
   filter(hue == "Light") %>% 
-  select(1, c(30:494))
+  select(sample_id, c(`92.88`:`2503.59`))
 
 #Melt into long format
 r.light.long <- 
@@ -53,7 +53,7 @@ r.light.long <-
 #Filter raman data to focus on colourless samples
 r.colourless <- Points.raman %>%
   filter(hue == "Colourless") %>% 
-  select(1, c(30:494))
+  select(sample_id, c(`92.88`:`2503.59`))
 
 #Melt into long format
 r.colourless.long <- 
@@ -62,7 +62,7 @@ r.colourless.long <-
 #Filter raman data to focus on white samples 
 r.white <- Points.raman %>%
   filter(hue == "White") %>% 
-  select(1, c(30:494))
+  select(sample_id, c(`92.88`:`2503.59`))
 
 #Melt into long format
 r.white.long <- 
@@ -71,7 +71,7 @@ r.white.long <-
 #plot the colourless spectra
 p.c <-
   ggplot(r.colourless.long,aes(x = as.numeric(as.character(Wavenumber)))) + 
-    geom_line(aes(y = Intensity, colour = "", group = sample_id), size = 1, stat = "identity") +
[…]

### 2023-04-03 analysis/paper/006-raman-peak-identification.R

@@ -8,7 +8,7 @@ metadata.csv <-
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 raman.averaged <- 
@@ -87,15 +87,13 @@ Points.raman <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
          type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  replace_na(list(munsell_hue = "Colourless"))
 
 #Filter raman data to focus on dark samples 
 r.dark <- Points.raman %>%
   filter(hue == "Dark") %>% 
-  select(30:494)
+  select(`92.88`:`2503.59`)
 
 #Calculate the mean of all dark spectra
 r.dark.mean <-
@@ -104,7 +102,7 @@ r.dark.mean <-
 #Filter raman data to focus on light samples 
 r.light <- Points.raman %>%
   filter(hue == "Light") %>% 
-  select(30:494)
+  select(`92.88`:`2503.59`)
 
 #Calculate the mean of all dark spectra
 r.light.mean <-
@@ -150,9 +148,9 @@ fig <-
                     nrow = 2)
 
 #Save figure
-ggsave("006-raman-peak-identification.png",
+ggsave("006-raman-peak-identification.jpeg",
        fig,
[…]

### 2023-04-03 analysis/paper/007-raman-pca-loadings.R

@@ -6,7 +6,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 metadata.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
@@ -84,14 +84,12 @@ Points.raman <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
          type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite")
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.raman[8][is.na(Points.raman[8])] <- "Colourless"
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  replace_na(list(munsell_hue = "Colourless"))
 
 #perform PCA with SNV normalization and mean-center
 raman.pca <-
-  prcomp(Points.raman[,c(30:321)], center = TRUE, scale = FALSE)
+  prcomp(Points.raman[,c(30:322)], center = TRUE, scale = FALSE)
 
 #prepare labels for PCs, doing it in 2 steps allows variance percentage to be called on in the Rmd file
 raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
@@ -149,9 +147,9 @@ fig <-
                     nrow = 3,
                     align = "v")
 
-ggsave("007-raman-pca-loadings.png",
+ggsave("007-raman-pca-loadings.jpeg",
        fig,
[…]

### 2023-04-03 analysis/paper/007-raman-pca.R

@@ -6,7 +6,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 metadata.csv <-
   read.csv2(here::here("analysis", "data", "raw_data", "metadata.csv"), sep = ";", header = TRUE, na = c("", "NA", "NULL"), encoding = "UTF-8")
@@ -21,7 +21,7 @@ raman.transposed <-
 
 #Write csv with transposed raman data for baseline de-trending in ChemoSpec package
 #ChemoSpec requires data to be read from file
-write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"), col.names = FALSE, sep = ";", dec = )
+write.table(raman.transposed, file = here::here("analysis", "data", "derived_data", "raman_averaged_transposed.csv"), col.names = FALSE, sep = ";", dec = )
 
 #prepare vector with sample_id for reading raman data into spectra object
 sample_id <- 
@@ -34,7 +34,7 @@ raman <- suppressWarnings(matrix2SpectraObject(
   freq.unit ="Wavelength cm-1",
   int.unit ="Intensity",
   descrip ="Bifacial points measurements",
-  in.file = here::here("analysis", "data", "derived_data", "raman_transposed.csv"),
+  in.file = here::here("analysis", "data", "derived_data", "raman_averaged_transposed.csv"),
   sep = ";",
   dec = ".",
   chk = TRUE,
@@ -84,14 +84,12 @@ Points.raman <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
          type == "Point" | type == "Point fragment" | type == "Preform", 
[…]

### 2023-04-03 analysis/paper/008-xrf-violin-plot.R

@@ -3,7 +3,7 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "XRF", "xrf_quantitative_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #Import descriptive metadata
 metadata.csv <-
@@ -25,10 +25,8 @@ Points.xrf <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
          type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") 
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  replace_na(list(munsell_hue = "Colourless"))
 
 #Prepare XRF data in long format for violin plots, exclude samples with a dimension smaller than 10 mm 
 Points_long <-
@@ -156,9 +154,9 @@ fig <-
       axis.text.x=element_blank(),
       axis.ticks.x=element_blank())
 
-ggsave("008-xrf-violin-plot.png",
+ggsave("008-xrf-violin-plot.jpeg",
        fig,
-       device = "png",
+       device = "jpeg",
        here::here("analysis", "figures"),
        width=25, 
        height=20,

### 2023-04-03 analysis/paper/009-xrf-pca.R

@@ -7,11 +7,11 @@ suppressPackageStartupMessages(library(tidyverse))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "XRF", "xrf_quantitative_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import descriptive metadata
 metadata.csv <-
@@ -35,10 +35,8 @@ Points.xrf <-
            site_id == "Åsele 107" | site_id == "Åsele 115" | site_id == "Åsele 117" | site_id == "Åsele 119" | site_id == "Åsele 129" | site_id == "Åsele 182" | site_id == "Åsele 188" |
            site_id == "Åsele 393" | site_id == "Åsele 56" | site_id == "Åsele 91" | site_id == "Åsele 92" | site_id == "Åsele 99", 
          type == "Point" | type == "Point fragment" | type == "Preform", 
-         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") 
-
-#fill the NA fields in the munsell hue column to mark them as colourless/translucent material
-Points.xrf[8][is.na(Points.xrf[8])] <- "Colourless"
+         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
+  replace_na(list(munsell_hue = "Colourless"))
 
 #filter XRF data on sample size (only include samples with smallest dimension >= 10mm), 
 #select elements to include in PCA, as well as columns to group by later on
@@ -386,9 +384,9 @@ fig1 <-
                     nrow = 2)
[…]

### 2023-04-03 analysis/paper/010-nir-sample-comparison.R

@@ -1,6 +1,5 @@
 # Loading package
 library(png)
-library(grid)
 library(loder)
 suppressPackageStartupMessages(library(ChemoSpec))
 suppressPackageStartupMessages(library(cowplot))
@@ -15,15 +14,15 @@ metadata.csv <-
 
 #Import nir data, set empty fields to NA
 nir.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "NIR", "asd_raw_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "asd_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import raman data, set empty fields to NA
 raman.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "RAMAN", "raman_samples_data_non_treated_20220407.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
+  read.csv2(here::here("analysis", "data", "raw_data", "raman_raw_data.csv"), sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))
 
 #Import xrf data, set empty fields to NA
 xrf.csv <-
-  read.csv2(here::here("analysis", "data", "raw_data", "XRF", "xrf_quantitative_data_20220407.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
+  read.csv2(here::here("analysis", "data", "raw_data", "xrf_quantitative_raw_data.csv"), sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))
 
 #aggregate observations by group(sample) and calculate average of wavelength measurements
 nir.averaged <- 
@@ -133,7 +132,9 @@ Points.raman <-
 Points.xrf <-
   xrf.merged %>%
   dplyr::select(sample_id, `Mg`, `Al`, `Si`, `P`, `S`, `K`, `Ca`, `Ti`, `Fe`, `Sr`, `Zr`,  `Ba`) %>% 
-  dplyr::filter(sample_id == 386 | sample_id == 404| sample_id == 391)
+  dplyr::filter(sample_id == 386 | sample_id == 404| sample_id == 391) %>% 
[…]

### 2023-04-03 README.Rmd
Grammar correction
name update
updated readme

@@
 # Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
+Title <- "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
 ```
 
-# MethodPaper
+# Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden
 
@@
 
-> Sjoelander Mattias, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P (2023). _`r Title`_. Journal of Archaeological and Anthropological Sciences <https://doi.org/xxx/xxx>
 
-Our pre-print is online here:
+The pre-print is online here:
 
-> Sjoelander Mattias, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P (2023). _`r Title`_. Journal of Archaeological and Anthropological Sciences `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
 
@@
 
-> Sjoelander Mattias, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+> Sjölander M., Linderholm, J., Geladi P (2023).
+(`r format(Sys.Date(), "%Y")`).
+_Compendium of R code and data for `r Title`_.
+Accessed `r format(Sys.Date(), "%d %b %Y")`.
+Online at <https://doi.org/xxx/xxx>
 
@@
 
-## How to run in your broswer or download and run locally
+## How to run in your browser or download and run locally
 
